---
title: "About MIDFIELD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About MIDFIELD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
	echo = FALSE, # varies from one Rmd to another 
  message = TRUE,
  warning = TRUE, 
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE, 
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/about-"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
options(tibble.print_min = 8L, tibble.print_max = 8L)

# normally, purl is FALSE. But in some code chunks, I want purl = TRUE so that # the code in the vignette is executed to pass R CMD check. These next few 
# lines set that up. In the code chunk header, add opts.label = 'dopurl'
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_template$set(dopurl = list(purl = TRUE, error = FALSE))
```

<a href="https://engineering.purdue.edu/MIDFIELD" target="blank">MIDFIELD</a> is the acronym for the Multiple-Institution Database for Investigating Engineering Longitudinal Development. 

Longitudinal studies using MIDFIELD are not limited to Engineering---the database includes *all* student records from member institutions.


## Who we are

```{r}
library(tibble)
df <- tribble(
  ~name, ~position, ~email,
  "Matthew Ohland", "Principal Investigator/Director", "ohland@purdue.edu",
  "Marisa Orr", "Associate Director", "marisakorr@gmail.com",
  "Catherine Brawner", "Director Policy Analysis", "brawnerc@bellsouth.net",
  "Nichole Ramirez", "Associate Director Policy Analysis", "nramire@purdue.edu",
  "Russell Long", "Managing Director/Data Steward", "ralong@purdue.edu",
  "Susan Lord", "Director MIDFIELD Institute", "slord@sandiego.edu",
  "Richard Layton", "Director Data Display", "layton@rose-hulman.edu"
)
knitr::kable(df, col.names = NULL)
```
 
 
 
## Funding

Current funding is provided by the US National Science Foundation, [Grant 1545667](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1545667) Expanding Access to and Participation in the Multiple-Institution Database for Investigating Engineering Longitudinal Development.    













