---
title: "About MIDFIELD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{About MIDFIELD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, message = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(collapse = TRUE, comment = "#>")
opts_chunk$set(warning = FALSE, message = FALSE)
opts_chunk$set(error = TRUE, purl = FALSE)
opts_chunk$set(fig.width = 6)
options(tibble.print_min = 6L, tibble.print_max = 6L)
knit_hooks$set(inline = function(x) {
	if (!is.numeric(x)) { 
		x 
		} else if (x >= 10000) { 
		prettyNum(round(x, 2), big.mark = ",")
		} else {
		prettyNum(round(x, 2))
		}
	})
```

<a href="https://engineering.purdue.edu/MIDFIELD" target="blank">MIDFIELD</a> is the acronym for the Multiple-Institution Database for Investigating Engineering Longitudinal Development. 

That our work focuses on Engineering (the "E" in MIDFIELD) is a consequence of our specific research interests. However, longitudinal studies using the MIDFIELD data are not limited to Engineering---the database includes *all* student records from member institutions. Therefore, one may study any of the instructional programs in the data sample. 
  


## Who we are

```{r echo = FALSE}
library(tibble)
df <- tribble(~name, ~position, ~email, 
  "Matthew Ohland",    "Principal Investigator/Director",    "ohland@purdue.edu", 
  "Marisa Orr",        "Associate Director",                 "marisakorr@gmail.com", 
  "Catherine Brawner", "Director Policy Analysis",           "brawnerc@bellsouth.net", 
  "Nichole Ramirez",   "Associate Director Policy Analysis", "nramire@purdue.edu", 
  "Russell Long",      "Managing Director/Data Steward",     "ralong@purdue.edu", 
  "Susan Lord",        "Director MIDFIELD Institute",        "slord@sandiego.edu", 
  "Richard Layton",     "Director Data Display",             "layton@rose-hulman.edu"
  )
kable(df, col.names = NULL)
```
 
 
 
## Funding

Current funding is provided by the US National Science Foundation, [Grant 1545667](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1545667) Expanding Access to and Participation in the Multiple-Institution Database for Investigating Engineering Longitudinal Development.    













