---
title: "Introduction to midfieldr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to midfieldr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/body-and-society.csl
link-citations: yes
nocite: | 
  @Dowle+Srinivasan:2020:data.table, @Harrell:2021:Hmisc. @Waring+Quinn:2021:skimr
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here("man/figures", 
                                            "art-00-introduction-"))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
```

## Introduction 

The *Multiple-Institution Database for Investigating Engineering Longitudinal Development* (MIDFIELD) is a partnership of US higher education institutions with engineering programs. MIDFIELD contains registrar's data for 1.7M undergraduates in all majors at 19 institutions from 1987--2019 [@Ohland+Long:2016].

The midfieldr package provides functions specialized for manipulating MIDFIELD data. The midfielddata package provides practice data for users to learn about student record analysis using R. In this introduction, we inspect the practice data and briefly summarize the midfieldr functions. Subsequent vignettes illustrate the application of midfieldr functions in detail. 

The midfielddata package organizes the student records in four tables keyed by student ID. 

- `student`  
- `course`  
- `term`   
- `degree`  

To maintain confidentiality of human subject data, the identifiers for students and institutions have been anonymized.

In this vignette, we introduce the four data tables and briefly examine their contents before moving on in subsequent vignettes to operate on the data using the functions in midfieldr. 

We suggest installing and loading the data.table package. While we use base R functions in the vignettes to reduce the cognitive load on new users, there are times when the data.table syntax offers advantages in clarity and speed.


### Outline

- Organize for practice 
- Loading data
- Inspect the `student` data 
- Inspect the `course` data 
- Inspect the `term` data 
- Inspect the `degree` data 
- Overview of midfieldr functions 


## Organize for practice 

We recommend that you always work within an RStudio Project and that you organize the file structure at the start of the project.  For practice working with midfieldr you might create an RStudio project `practice-midfieldr` that has the minimal structure:

```
/practice-midfieldr
        /data       # raw data from any external source 
        /results    # reproducible results:  graphs and tabulated data
        /scripts    # R files, one per vignette 
        practice-midfieldr.Rproj 
```

Some of the code snippets in the midfieldr vignettes are written with relative file paths that assume the existence of this minimal structure. Use any organization that suits your needs. 

Reading:

- Wilson, Bryan, et al. [-@Wilson+Bryan+others:2017] offer detailed advice for organizing computational projects. Highly recommended.  
- [Using RStudio Projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects) provides instructions for creating RStudio projects.


## Loading data

These code snippets can be run in the Console, though for repeatability and exploration, some are more useful if typed into an R script. 

```{r}
# packages used in this vignette
library("midfieldr")
library("midfielddata")
library("data.table")
```


Use `data()` to load the midfielddata data tables into memory for use. For example, after loading the package with `library("midfielddata")` as we did above, you can run `ls()` to see that the environment contains zero objects, 

```{r}
# the names of objects in the environment
ls()
```

To load the `student` data from midfielddata into memory, we run

```{r echo = FALSE, cache = FALSE}
# cache = TRUE speeds up repeats
if(!exists("student")) data(student)
```

```r
# loading a data set 
data(student)
```

Checking the environment again, 

```{r}
# the names of objects in the environment
ls()
```

If you look in the RStudio *Environment* pane, you should see `student` listed. We will typically load the data sets we need at the top of a script just after loading the midfielddata package. You need to load them only once in a session.  

```{r echo = FALSE, cache = FALSE}
# cache = TRUE speeds up repeats 
if(!exists("course"))  data(course)
if(!exists("term"))    data(term)
if(!exists("degree"))  data(degree)
```

```r
# load remaining data used in this vignette
data(course)
data(term)
data(degree)
```

Checking the environment again, 

```{r}
# the names of objects in the environment
ls()
```


```{r echo = FALSE}
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}

# cip
n_cip <- nrow(unique(cip))
  
# student
obs_students <- nrow(student)
var_students <- ncol(student)
size_students <- "19 MB"

# degree
n_id_degrees <- nrow(unique(degree[, .(mcid)]))
n_graduates <- nrow(unique(degree[!is.na(degree), .(mcid)]))
n_term <- nrow(unique(degree[, .(term)]))
obs_degrees <- nrow(degree)
var_degrees <- ncol(degree)
size_degrees <- "10 MB"

# term
n_id_terms <- nrow(unique(term[, .(mcid)]))
obs_terms <- nrow(term)
var_terms <- ncol(term)
n_institutions <- nrow(unique(term[, .(institution)]))
n_programs_terms <- nrow(unique(term[, .(institution, cip6)]))
n_terms <- nrow(unique(term[, .(term)]))
year_span <- c(substr(min(term[, term]), 1, 4), substr(max(term[, term]), 1, 4))
size_terms <- "82 MB"

# course
n_id_courses <- nrow(unique(course[, .(mcid)]))
n_courses <- nrow(unique(course[, .(institution, abbrev, number)]))
n_term_course <- nrow(unique(course[, .(term)]))
obs_courses <- nrow(course)
var_courses <- ncol(course)
size_courses <- "349 MB"
```





## Inpect the `student` data

`student` is a table of `r var_students` student-related variables for the students in the sample, collected at the time they were admitted to their institutions as  degree-seeking undergraduates. For the data dictionary, you can navigate to the midfielddata website  [`student`](https://midfieldr.github.io/midfielddata/reference/student.html) page or view the same information in the help page by running 

```r
? student
``` 

Each observation is a unique student keyed by student ID. Student IDs and institution names are anonymized but are consistent across the four data sets. To see the data structure, run

```{r}
# view the structure of the data set 
str(student)
```

`student` has `r obs_students` observations and occupies approximately `r size_students` memory. We can confirm that the student IDs are unique by running, 

```{r}
# number of unique IDs
length(unique(student$mcid))
```

Summarize some of the quantitative variables. 

```{r}
summary(student$sat_math)

summary(student$age)
```

Examine a few of the categorical variables and their values. 

```{r}
sort(unique(student$sex))

sort(unique(student$race))
```

### Exercises
 
1. Determine the possible values in the `institution` column. 
2. Determine the possible values in the `transfer` column. 
3. Determine the median `sat_verbal` score.

## Inpect the `course` data

`course` is a table of `r var_courses` course-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`course`](https://midfieldr.github.io/midfielddata/reference/course.html) page or view the same information in the help page by running 

```r
? course
```

Each observation is one course in one term for one student keyed by student ID and a combination of course, abbreviation, number, and section. To see the data structure, run

```{r}
# view the structure of the data set 
str(course)
```

Of the four practice data sets in midfielddata, `course` is by far the largest, with `r obs_courses` observations and occupying `r size_courses` of memory. Depending on your machine, loading `course` can cause a noticeable delay in execution time (13 seconds on my machine), so you probably want to load it only when needed. 

The number of unique students.  

```{r}
# number of unique IDs
length(unique(course$mcid))
```
 
Summarize a quantitative variable. 

```{r}
summary(course$hours_course)
```

Examine a categorical variable and its values. 

```{r}
# type of courses
sort(unique(course$type))
```

### Exercises

1. Determine the possible values in the `faculty_rank` column. 
2. Determine the range of terms in `term` column. 
3. Determine the possible values in the `grade` column. 

## Inpect the `term` data

`term` is a table of `r var_terms` term-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`term`](https://midfieldr.github.io/midfielddata/reference/term.html) page or view the same information in the help page by running

```r
? term
```

Each observation is one term for one student keyed by student ID and term. To see the data structure, run

```{r}
str(term)
```

`term` has `r obs_terms` observations and occupies approximately `r size_terms` of memory. The number of unique students is found by, 

```{r}
# number of unique IDs
length(unique(term$mcid))
```

The number of unique students in `student` and `term` are not the same. To be recorded in `student`, students must be degree-seeking. To be recorded in `term`, students must complete a term. The data can include degree-seekers who fail to complete a term as well as term-completers not seeking a degree---hence the difference in numbers

Summarize some of the quantitative variables. 

```{r}
summary(term$hours_term)

summary(term$gpa_term)
```

Examine some categorical variables and their values.  

```{r}
sort(unique(term$level))

sort(unique(term$standing))
```

### Exercises

1. Determine the possible values in the `coop` column. 
2. Determine the range of terms in `term` column. 
3. Determine the median value of the `gpa_cumul` column. 

## Inpect the `degree` data

`degree` is a table of `r var_degrees` degree-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`degree`](https://midfieldr.github.io/midfielddata/reference/degree.html) page or view the same information in the help page by running

```r
? degree
```

Each observation is a unique student keyed by student ID. To see the data structure, run

```{r}
str(degree)
```

`degree` has `r obs_degrees` observations and occupies approximately `r size_degrees` memory. The number of unique students is found by

```{r}
# number of unique IDs
length(unique(degree$mcid))
```

Thus we have `r obs_degrees` graduates (in `degree`) from the `r obs_students` students admitted (in `student`). 

Examine some categorical variables and their values. 

```{r}
sort(unique(degree$institution))

sort(unique(degree$degree))
```

Checking if any of the degrees are recorded as NA results in an empty table, thus all the students in `degree` are graduates. 

```{r}
degree[is.na(degree)]
```

### Exercises

Some packages have functions specialized for data frame summaries. Try these two and compare results. 

1. Install the `Hmisc` package, load it with `library("Hmisc")`, and run `describe(x)`, where `x` is one of the data tables, for example, `describe(student)`, `describe(degree)`, etc. 
2. Install the `skimr` package, load it with `library("skimr")`, and run `skim(x)`, where `x` is one of the data tables, for example, `skim(student)`, `skim(degree)`, etc.  



## Overview of midfieldr functions

midfieldr functions are introduced in detail in subsequent vignettes, but we can introduce their general organization here. 

**`add_*`**    
This family of functions takes working data frames as input and adds specific columns of interest without subsetting rows. 

```{r echo = FALSE}
df <- wrapr::build_frame(
  "Function", "Action" |
  "add_completion_timely()", "Add a column to evaluate program completion" |
  "add_data_sufficiency()", "Add a column to evaluate data sufficiency" |
  "add_institution()", "Add a column of institution names" |
  "add_race_sex()", "Add two columns for student race/ethnicity and sex" |
  "add_timely_term()", "Add a column of terms delimiting timely completion" 
)
kable2html(df, font_size = 12)
```

**`filter_*`**    
This family of functions takes working data frames as input and subsets rows that match conditions; selected columns are subset if specified.  

```{r echo = FALSE}
df <- wrapr::build_frame(
  "Function", "Action" |
  "filter_match()", "Subset rows by matching values in shared key columns" |
  "filter_search()", "Subset rows that include matches to search strings" 
)
kable2html(df, font_size = 12)
```

**`condition_*`**    
This family of functions takes highly-structured data frames as input and produces output data frames conditioned for particular tasks. 

```{r echo = FALSE}
df <- wrapr::build_frame(
  "Function", "Action" |
  "condition_fye()", "Condition FYE data for multiple imputation" |
  "condition_multiway()", "Condition multiway data for graphing" 
)
kable2html(df, font_size = 12)
```


### Exercises

View the help pages for the functions `add_timely_term()` and `add_completion_timely()`. 

1. Summarize the difference(s) between the two functions. 
2. Describe the manner in which one depends on the other.  

View the help pages for the functions `add_institution()` and `add_race_sex()`.

3. List similarities in what the two functions do. 
3. List differences in what the two functions do. 

## References

<div id="refs"></div>


## Appendix

### Complete script

We conclude each vignette by collecting its code chunks in a single, condensed script for the convenience of those who wish to copy it into their own R file. 

```{r eval = FALSE}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# loading data
data(student)
data(course)
data(term)
data(degree)

# student
str(student)
length(unique(student$mcid))
summary(student$sat_math)
summary(student$age)
sort(unique(student$sex))
sort(unique(student$race))

# course
str(course)
length(unique(course$mcid))
summary(course$hours_course)
sort(unique(course$type))

# term
str(term)
length(unique(term$mcid))
summary(term$hours_term)
summary(term$gpa_term)
sort(unique(term$level))
sort(unique(term$standing))

# degree
str(degree)
length(unique(degree$mcid))
sort(unique(degree$institution))
sort(unique(degree$degree))
degree[is.na(degree)]
```


