---
title: "Case study: Goals"
vignette: >
  %\VignetteIndexEntry{Case study: Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-001-")
```


Part 1 of a case study in three parts, illustrating how we work with longitudinal student-level records.  

1. Goals (this page). Introducing the study.   

2. [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.

3. [Results.](art-003-case-results.html) Calculating summary statistics and metrics and displaying the results. 






## Definitions

```{r child = "../man/rmd/define-student-level-data.Rmd"}
```

```{r child = "../man/rmd/define-stickiness.Rmd"}
``` 

: Stickiness is a more-inclusive alternative to graduation rate as a measure of a program's success in attracting, keeping, and graduating their undergraduates. Stickiness includes many students excluded by graduation rate such as part-time students, transfers, students admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

 




## Goals

The technical goal of the case study is to compare the program stickiness of Civil, Electrical, Industrial, and Mechanical Engineering programs with students grouped by race/ethnicity and sex. 

The meta-goal of the case study is to illustrate a typical procedure for working with student-level data. 






## What the case study does

- We focus on our procedures and underlying rationale for working with student-level data. 

- For reproducibility, we provide all the necessary code. 

- We use the [data.table](https://rdatatable.gitlab.io/data.table/)  package instead of base R or the dplyr package for speed, memory efficiency, and concise transforms via the []-indexing operator.

- We use the [ggplot2](https://ggplot2.tidyverse.org/) package instead of base R or lattice (or other R packages) because: ggplot2's layered approach allows  plots to evolve (or devolve) with minimal changes to code; easily-invoked facets and legends simplify multivariate exploration; and its popularity yields extensive online help. 











## What the case study does not do

To render the case study complete but brief, we omit the details of package syntax or functions, leaving such discussions to later articles. One can always use the R help system to read more about a data set or function.









## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
