---
title: "Case study: Data"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study: Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: |
  @Dowle+Srinivasan:2021:data.table
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = FALSE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
  fig.path = "../man/figures/art-002-case-data-",
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```


Part 2 of a case study in three parts:

- [Goals.](art-001-case-goals.html) Describing the study parameters.  
- [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.
- [Results.](art-003-case-results.html) Calculating summary statistics and metrics and displaying the results. 

In this part, we use the midfieldr functions:  

- `add_completion_status()` 
- `add_data_sufficiency()`
- `add_timely_term()`
- `filter_search()` 

## Preface

Our data processing goal is to reduce the source data tables to the specific observations needed to compute our metrics. The data processing tasks include a combination of filtering observations (rows), creating, renaming, and recoding variables (columns), and joining data frames. 

The analysis is organized to produce two data frames---students ever enrolled in the programs and students graduating from the programs---that are joined and written to file as a starting point for the next section of the case study. 

If you are writing your own script to follow along, we start with these packages: 

```{r}
# Packages
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# Printing options for data.table
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```






## Import data

All MIDFIELD data are in block-record form, that is, every row is an observation (keyed by student ID) and every column is a variable. Data structure is explored in greater detail in  the [Getting started](art-000-getting-started.html) vignette. 

**Importing.** Load three of the midfielddata data tables. 

```{r}
# Load three data sets from midfielddata
data(student, term, degree)
```


## Gather ever enrolled 

**Copying.** Copy the `term` data set to create the working data frame (`DT`), leaving `term` unaltered for later use. 

```{r}
# Create a working data frame
DT <- copy(term)
str(DT)
```

The result has `r nrow(DT)` observations. We will usually note the number of observations as they change. 

**Filtering.** In this study, only two variables from the `term` data are needed to start: student ID `mcid` and program code `cip6`. We drop unnecessary columns then remove duplicate rows such that we have unique combinations of student ID and program CIP. 

```{r}
# Minimize the dimensions of the data
DT <- DT[, .(mcid, cip6)]
DT <- unique(DT)
DT[]

# Count unique IDs
length(unique(DT$mcid))
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students. The number of unique students is smaller than the number of observations because some students change programs over their course of study.  Here we have one row per unique combination of student (`mcid`) and program code (`cip6`).  



## Filter for data sufficiency

Some student records near the lower and upper terms that bound the available data must be excluded to prevent false summaries involving timely degree  completion. To apply this filter, we determine the timely completion term then data sufficiency. 

**Creating variables.** Using information in `term`, we add the  `timely_term` variable as well as supporting variables used in its construction.

```{r}
# Calculate a timely completion term for every student
DT <- add_timely_term(DT, term)
DT[]
```

You can view the help page by running

```{r}
#| eval: false
# Run in Console
? add_timely_term
```

**Creating variables.** Using information in `term`, we add the  `data_sufficiency` variable as well as supporting variables used in its construction.

```{r}
# Determine data sufficiency for every student
DT <- add_data_sufficiency(DT, term)
DT[]
```

You can view the help page by running

```{r}
#| eval: false
# Run in Console
? add_data_sufficiency
```


**Filtering.** We filter to retain observations for which the data are sufficient then drop unnecessary columns.   

```{r}
# Retain observations having sufficient data
DT <- DT[data_sufficiency == "include"]
DT <- DT[, .(mcid, cip6)]
DT[]
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students.







## Filter for degree-seeking

In studying stickiness, we exclude non-degree-seeking students. We filter such students using the `student` data which contains only degree seeking students. 

**Filtering.** An inner join with `student` that returns rows with matching IDs from both data frames, then selected columns.

```{r}
# Inner join that filters DT
DT <- DT[student, .(mcid, cip6), on = c("mcid"), nomatch = NULL]
DT[]
```

The result has `r nrow(DT)` observations. (No change is expected in this example because all students in the midfielddata practice data are degree-seeking.) 






## Filter by program

To filter by program, we have to construct a data frame of labels for our four engineering programs keyed by the US *Classification of Instructional Programs* (CIP) codes. These codes are bundled with midfieldr in the `cip` data set. 

The `cip6` variable in our data is the 6-digit code for programs in which students are enrolled in a given term. 

Having already searched the `cip` data set for our programs, we report the results: Civil Engineering CIPs start with 1408, Electrical Engineering 1410, Mechanical Engineering 1419, and Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 

**Filtering** Gather the 6-digit CIP codes for the programs in the case study. Drop unnecessary variables.  We keep the `cip4name` because, in engineering at least, the 4-digit name tends to be a disciplinary or departmental name. 

```{r}
# Gather program CIP codes
study_program <- filter_search(cip, keep_text = c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437"))
study_program <- study_program[, .(cip6, cip4name)]
study_program[]
```

You can view the help page by running

```{r}
#| eval: false
# Run in Console
? filter_search
```

**Creating variables.** Add a variable to label each of these `r nrow(study_program)` programs with one of the four conventional program abbreviations we will use in comparing metrics, i.e., Civil (CE), Electrical (EE), Mechanical (ME), and Industrial/Systems Engineering (ISE). 

```{r}
# Assign four program names by CIP code
study_program[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %chin% c("142701", "143501", "143601", "143701"), "ISE"
)]

# Confirm that abbreviations match the longer program names
print(study_program[, .(cip4name, program)])
```

Having checked that the new abbreviations correctly represent the programs, we can finalize the data frame of program CIPs and names. 

```{r}
# Case study CIPs and program names
study_program[, cip4name := NULL]
study_program[]
```

This data frame is bundled with midfieldr as the `study_program_labels` data set. You can view the help page by running

```{r}
#| eval: false
# Run in Console
? study_program_labels
```

**Joining.** We join the program names to the working data frame using a left-outer join and return selected rows.    

```{r}
# Left-outer join to DT
DT <- study_program[DT, .(mcid, cip6, program), on = c("cip6")]
DT[]
```

**Filtering.**  All instances of NA programs are dropped.

```{r}
# Retain observations in our four programs
DT <- DT[!is.na(program)]
DT[]
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students.

**Filtering.** A program label can be associated with multiple CIP codes. We drop the code and filter to remove duplicates to ensure unique combinations of ID and program. 

```{r}
# Drop duplicate rows
DT[, cip6 := NULL]
DT <- unique(DT)
DT[]
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students.

**Creating variables.** Add a variable to label this group as "ever" enrolled to distinguish them from the group of graduates to be created next. Copy the data frame to a new name to save for later. 

```{r}
# Conclude development of the ever-enrolled observations
ever <- copy(DT)
ever[, group := "ever"]
ever[]
```









## Gather graduate subset

**Filtering.** Students potentially counted as graduates must be a subset of those in the ever enrolled population. We re-initialize the working data frame, retaining unique IDs only. 

```{r}
# Unique IDs of students ever enrolled in the study programs
DT <- DT[, .(mcid)]
DT <- unique(DT)
DT[]
```

These IDs are bundled with midfieldr as the `study_mcid` data set. You can view the help page by running

```{r}
#| eval: false
# Run in Console
? study_mcid
```

**Filtering.** An inner join with `degree` that returns rows with matching IDs from both data frames, then selected columns. 

```{r}
# Inner join that filters DT
DT <- DT[degree, .(mcid, cip6), on = c("mcid"), nomatch = NULL]
DT[]
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students. These IDs have already been filtered for data sufficiency and  degree-seeking.




## Filter by completion status

To be counted as a graduate in the stickiness metric, a student must complete their program in a timely manner (default 6 years after admission, with reduced spans for some transfer students). 

**Add variables.** Completion status requires the timely completion term variable. 

```{r}
# Calculate a timely completion term for every student
DT <- add_timely_term(DT, term)
DT[]
```


**Creating variables.** Using information in `degree`, we add the `completion_status` variable as well as supporting variables used in its construction. 

```{r}
# Determine completion status for every student
DT <- add_completion_status(DT, degree)
DT[]
```

You can view the help page by running

```{r}
#| eval: false
# Run in Console
? add_completion_status
```

**Filtering.** Retain records with "positive"  completion status (i.e. completion in 6 years or less) and drop unnecessary variables.

```{r}
# Retain timely completers
DT <- DT[completion_status == "timely"]
DT <- DT[, .(mcid, cip6)]
DT[]
```

The result has `r nrow(DT)` unique observations.


## Filter by program

We join the program names to the working data frame using a left-outer join and return selected rows. We then filter to retain program names that are not NA. 

```{r}
# Left-outer join to DT
DT <- study_program[DT, .(mcid, program), on = c("cip6")]
DT <- DT[!is.na(program)]

# Unique observations
DT <- unique(DT)
DT[]
```

The result has `r nrow(DT)` unique observations.

**Creating variables.** Add a variable to label this group as "grad" to distinguish them from the group of ever enrolled students created earlier.  

```{r}
# Conclude development of the graduate observations
grad <- copy(DT)
grad[, group := "grad"]
grad[]
```

## Combine ever and grad

**Binding.** Combine two data frames by matching column names. 

```{r}
# Combine two data frames
DT <- rbindlist(list(ever, grad), use.names = TRUE)
DT[]
```







## Add demographics

**Creating variables.** Using columns from `student`, we add variables for race/ethnicity and sex using a left-outer join. All rows and columns of DT are returned with two columns added from `student`. 

```{r}
# Selecting columns from student
student_cols <- student[, .(mcid, race, sex)]

# Add them to DT with a left outer join
DT <- student_cols[DT, on = c("mcid")]
DT[]
```

The result has `r nrow(DT)` observations of `r length(unique(DT$mcid))` unique students. IDs labeled "grad" are a subset of those labeled "ever". 





## Review 

To better grasp the structure of these data, we can review selected sets of observations by ID. In the first example, the ID produces one observation only. The student was enrolled in Civil Engineering but completion, if any, was not timely. 

```{r collapse = TRUE}
DT[mcid == "MID26694225"]
```

In the next example, the ID produces two observations indicating that the student was enrolled in Civil Engineering and a timely graduate of that program. 

```{r collapse = TRUE}
DT[mcid == "MID25786154"]
```

In the next example, the ID produces two observations indicating that the student was enrolled in Electrical Engineering and in Mechanical Engineering but a timely graduate of neither program. 

```{r collapse = TRUE}
DT[mcid == "MID25868925"]
```

In the next example, the ID produces three observations indicating that the student was enrolled in Electrical Engineering and in Civil Engineering and a timely graduate of Civil.  

```{r collapse = TRUE}
DT[mcid == "MID26526757"]
```








## Closing

Starting with the `student`, `term`, and `degree` data tables, we filtered the data for data sufficiency, degree seeking, program, and timely completion in four programs (Civil, Electrical, Industrial/Systems, and Mechanical Engineering) and processed the data to obtain the observations needed to determine stickiness.  

**Write to file.**  Results like these, used as inputs for calculating summary statistics and metrics, are often written to file as an intermediate result. In this case, the data frame above is bundled with midfieldr as `study_observations`. You can view the help page (and data dictionary) by running

```{r}
#| eval: false
# Run in Console
? study_observations
```




## References

<div id="refs"></div>







## Appendix

We conclude each vignette by collecting its code chunks in a single, condensed script for the convenience of those who wish to copy it into their own R file. 

```{r}
#| eval: false
# Packages
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# Import data
data(student, term, degree)

# Gather ever enrolled
DT <- copy(term)
DT <- DT[, .(mcid, cip6)]
DT <- unique(DT)

# Filter for data sufficiency
DT <- add_timely_term(DT, term)
DT <- add_data_sufficiency(DT, term)
DT <- DT[data_sufficiency == "include"]
DT <- DT[, .(mcid, cip6)]

# Filter for degree-seeking
DT <- DT[student, .(mcid, cip6), on = c("mcid"), nomatch = NULL]

# Filter by program
study_program <- filter_search(cip, keep_text = c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437"))
study_program <- study_program[, .(cip6, cip6name)]
study_program[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %chin% c("142701", "143501", "143601", "143701"), "ISE"
)]
DT <- study_program[DT, .(mcid, cip6, program), on = c("cip6")]
DT <- DT[!is.na(program)]
DT[, cip6 := NULL]
DT <- unique(DT)
ever <- copy(DT)
ever[, group := "ever"]

# Gather graduate subset
DT <- DT[, .(mcid)]
DT <- unique(DT)
DT <- DT[degree, .(mcid, cip6), on = c("mcid"), nomatch = NULL]

# Filter by completion status
DT <- add_timely_term(DT, midfield_term = term)
DT <- add_completion_status(DT, degree)
DT <- DT[completion_status == "positive"]
DT <- DT[, .(mcid, cip6)]

# Filter by program
DT <- study_program[DT, .(mcid, program), on = c("cip6")]
DT <- DT[!is.na(program)]
DT <- unique(DT)
grad <- copy(DT)
grad[, group := "grad"]

# Combine two data frames and match the column names
DT <- rbindlist(list(ever, grad), use.names = TRUE)

# Add demographics
student_cols <- student[, .(mcid, race, sex)]
DT <- student_cols[DT, on = c("mcid")]
```

```{r echo = FALSE}
# to change the CSS file for block quotes
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 0px solid
}
```

