---
title: "Getting started"
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    css: extra.css
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = ".../man/figures/art-000-")
```

Familiarity with the MIDFIELD data structure is a prerequisite for working with midfieldr functions so we start by inspecting the practice data in [midfielddata](https://midfieldr.github.io/midfielddata/), a companion R package that provides anonymized student-level records for 98,000 undergraduates at three US institutions from 1988 through 2018. (If you haven't yet installed midfielddata, see the [Installation](https://midfieldr.github.io/midfieldr/index.html#installation) instructions.)

The practice data are organized in four datasets keyed by student ID.  

```{r}
#| echo: false
library(gt)
wrapr::build_frame(
  "Dataset", "Each row is", "N students", "N rows", "N columns", "Memory" |
    "course", "one student per course & term", "97,555", "3,289,532", 12L, "324.3 MB" |
    "term", "one student per term", "97,555", "639,915", 13L, "72.8 MB" |
    "student", "one student", "97,555", "97,555", 13L, "17.3 MB" |
    "degree", "one student per degree", "49,543", "49,665", 5L, "5.2 MB"
) |>
  gt() |>
  tab_caption("Table 1. Practice datasets in midfielddata") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  )
```



## Definitions

```{r child = "../man/rmd/define-student-level-data.Rmd"}
```

observation

: Row of a data frame (`student`, `course`, `term`, `degree`) keyed by student ID. 

variable

: Column of a data frame







## Method

In this article: 

- Overview of each dataset 
- Summary of variables typically encountered when using midfieldr functions
- Closer look: For one student, all records in all datasets 
- Introduce helper function `select_required()`  and wrapr `check_equiv_frames()`

```{r child = "../man/rmd/note-for-practice-only-1.Rmd"}
```






## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```


*Load data tables.*  &nbsp; Data tables can be loaded individually or collectively as needed.  View data dictionaries via `?student`, `?course`, `?term`, or `?degree`. 

```{r}
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```






## `student`

Contains one observation per student. Data are assumed to be current at the time the student was admitted to their institution.

```{r}
student
```

Student IDs and institution names have been anonymized to remove identifiable information. 

```{r}
# Anonymized IDs
sample(student$mcid, 8)

# Anonymized institutions
sort(unique(student$institution))
```

Race/ethnicity and sex are often used as grouping variables.

```{r}
# Possible values
sort(unique(student$race))

# Possible values
sort(unique(student$sex))
```

Counts in each category. 

```{r}
# N by institution
student[order(institution), .N, by = "institution"]

# N by race
student[order(race), .N, by = "race"]

# N by sex
student[order(sex), .N, by = "sex"]
```







## `course`

Contains one observation per student per course.

```{r}
course
```

The `abbrev`, `number`, and `discipline_midfield` columns have no NA values, so they might be useful if one is filtering for specific course types. The `course` column, on the other hand, has a high number of NA values. 

```{r}
# Many NA values in this column
sum(is.na(course$course))

# No NA values in these columns.
sum(is.na(course$abbrev))
sum(is.na(course$number))
sum(is.na(course$discipline_midfield))
```










## `term`

Contains one observation per student per term. 

```{r}
term
```

Terms are encoded `YYYYT`, where 

- `YYYY` is the year at the start of the academic year, and  
- `T` encodes the semester or quarter---Fall (`1`), Winter (`2`), Spring (`3`), and Summer (`4`, `5`, and `6`)---within an academic year

For example, for academic year 1995--96, 

- `19951` encodes Fall 95--96
- `19953` encodes Spring 95--96
- `19954` encodes Summer 95--96 (first session)

Different institutions supply data over different time spans. 

```{r}
# Range of data by institution
term[, .(min_term = min(term), max_term = max(term)), by = "institution"]
```

Programs are encoded in the `cip6` variable, a 6-digit character based on the 2010 Classification of Instructional Programs (CIP) [@NCES:2010].

```{r}
# A sample of cip6 values
sort(unique(sample(term$cip6, 8)))
```

Student level is used when determining timely completion terms of transfer students.

```{r}
# Possible values
sort(unique(term$level))
```






## `degree`

Contains one observation per student per degree. 

This dataset contains records for graduates only, thus the number of observations in `degree` (`r nrow(degree)`) is less than the number of observations in `student` (`r nrow(student)`). The `term_degree` and `cip6` variables indicate when and from which program a student graduates.  

```{r}
degree
```

Number of degrees earned per student. 

```{r}
# Count students by number of degrees
by_id <- degree[, .(degree_count = .N), by = "mcid"]
by_id[, .(N_students = .N), by = "degree_count"]
```






## Closer look

We display the records for one specific student, using their ID to subset each dataset.  

```{r}
# One student ID
id_we_want <- "MCID3112192438"
```

*Student.* &nbsp; As expected, `student` yields one row per student.

```{r}
# Observations for a selected ID
student[mcid == id_we_want]
```

*Course.* &nbsp; For this student, the records span `r nrow(course[mcid == id_we_want])` rows, one row per course. 

```{r}
# Observations for a selected ID
course[mcid == id_we_want]
```

*Term.* &nbsp; Here, the records span `r nrow(term[mcid == id_we_want])` rows, one row per term. 

```{r}
# Observations for a selected ID
term[mcid == id_we_want]
```

*Degree.* &nbsp; In this example, the records span `r nrow(degree[mcid == id_we_want])` rows, one row per degree. The degrees were earned in the same term, Spring 2009. 

```{r}
# Observations for a selected ID
degree[mcid == id_we_want]
```

Not all students with more than one degree earn them in the same term. For example, the next student earned a degree in 1996 and a second degree in 1999. In most analyses, only the first baccalaureate degree would be used.

```{r}
# Observations for a different ID
degree[mcid == "MCID3111315508"]
```















## `select_required()`

A midfieldr convenience function to reduce the number of columns of a MIDFIELD data table after loading. Using this function is optional. 

`select_required()` selects only those columns typically required by other midfieldr functions. Operates on a data frame to retain columns having names that match or partially match search terms. Rows are unaffected. 

The primary benefit is reducing screen clutter when viewing data frames during an interactive session. The disadvantage is that the deleted columns are unavailable unless you first set aside a copy of the source file or reload it using `data()` when you need it. 

*Arguments.*

- **`midfield_x`** &nbsp; MIDFIELD data frame, typically `student`, `term`, or `degree`. 

- **`select_add`** &nbsp; Optional character vector of search terms to add to the default vector given by `c("mcid", "institution", "race", "sex", "^term", "cip6", "level")`. Argument, if used, must be used by name.

For example, term records are significantly more compact if we select this minimum set of columns.  

```{r}
# Select variables required by midfieldr functions
select_required(term)
```

We can add columns if we need them. 

```{r}
# Select additional columns
select_required(term, select_add = c("gpa_term"))
```








## `check_equiv_frames()` 

A function imported from the wrapr package that confirms two data frames are equivalent after reordering columns and rows. Accessible by loading midfieldr. 

*Example.* &nbsp; Demonstrate that the following implementations of `select_required()` yield identical results. 

```{r}
# Required argument explicitly named
x <- select_required(midfield_x = term)

# Required argument not named
y <- select_required(term)

# Optional argument, if used, must be named. NULL yields the default columns.
z <- select_required(term, select_add = NULL)

# Demonstrate equivalence
check_equiv_frames(x, y)
check_equiv_frames(x, z)
```

Demonstrate that row and column order are ignored. 

```{r}
# Two columns from student, use key to order rows
x <- student[, .(mcid, institution)]
setkey(x, mcid)
x

# Same information with different row order, column order, and key
y <- student[, .(institution, mcid)]
setkey(y, institution)
y

# Demonstrate equivalence
check_equiv_frames(x, y)
```

If the two data tables do not have the same content, the return is FALSE. 

```{r}
# Demonstrate non-equivalence
check_equiv_frames(student, degree)
```

To explore the differences between non-equivalent data frames, `janitor::compare_df_cols()` returns a comparison of column names and class. 

```{r}
library(janitor)
compare_df_cols(student, degree)
```











## Reusable code

*Preparation.* &nbsp; The immediate prerequisites or “intake” required by the reusable code chunk are the source data tables.  

```{r}
#| eval: false
# Load source data
data(student, term, degree)
```

*Initial data processing.* &nbsp; A summary code chunk for ready reference. 

```{r}
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)
source_degree <- copy(degree)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
term <- select_required(source_term)
degree <- select_required(source_degree)
```

The `copy()` function ensures that "by-reference" changes to `student`, for example, have no effect on `source_student` [@data.table-reference-semantics]. Thus the `source_*` objects retain all the original columns, if needed later.



## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
