---
title: "Getting started"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: |
  @Dowle+Srinivasan:2021:data.table,  
  @Waring+Quinn:2022:skimr
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  cache = FALSE # speeds up repeats
)

# figures
knitr::opts_chunk$set(
  fig.path = "../man/figures/art-000-getting-started-",
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```




## Student unit records 

The MIDFIELD Student Unit Records (SUR) are registrar's student-level data compiled by an institution and anonymized and curated by the MIDFIELD data steward. SURs include personal information (e.g., age, sex, and race/ethnicity), enrollment information (e.g., institution, program, and grade level), and academic information (e.g., courses, grades, and degrees). MIDFIELD data are organized into four tables---student, course, term, and degree---keyed by student ID. 

From the MIDFIELD *research* data, we've taken a proportionate stratified sample and made it available as a *practice* data set in  [midfielddata](https://midfieldr.github.io/midfielddata/)---an R data package containing anonymized SURs for nearly 98,000 undergraduates at 12 institutions from 1987--2016. Table 1 briefly introduces the four data tables.

```{r}
#| echo: false

wrapr::build_frame(
    "Practice data table", "Each row is", "No. of rows", "No. of columns", "Memory" |
        "course" , "a student in a course"   , "3,439,936", "12", "340 Mb" |
        "term"   , "a student in a term"     ,   "710,841", "13",  "80 Mb" |
        "student", "a degree-seeking student",    "97,640", "13",  "19 Mb" |
        "degree" , "a student who graduates" ,    "47,499",  "5",  "10.2 Mb"
) |>
    kableExtra::kbl(align = "llrrr", 
                    caption = "Table 1: Attributes of the practice data tables in midfielddata") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::row_spec(0, background = "#c7eae5") |>
    kableExtra::column_spec(1, monospace = TRUE) |>
    kableExtra::column_spec(1:4, color = "black", background = "white")
```



## Method

Prerequisites: 

- [Requirements.](../index.html#requirements) R and data.table.
- [midfieldr.](../index.html#install-midfieldr) Install and check. 
- [midfielddata.](../index.html#install-midfielddata) Install and check. 

In this vignette, we

- Load the practice data from midfielddata
- Inspect the four data tables 
- Preview the midfieldr functions 

*Caveat:* The Student Unit Records included with midfieldr and midfielddata are practice data, not research data, suitable for practice working with SURs but not for drawing inferences about program attributes or student experiences.




## Load practice data

*Start a script*. If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
# Getting started
# midfieldr vignette

# Packages used
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# Printing options for data.table
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

*Load data* from the midfielddata package. 

Use `data()` to load the practice data tables into memory for use. For example, after loading the package with `library("midfielddata")` as we did above, you can run `ls()` to see that the environment contains zero objects, 

```{r}
# Names of objects in the environment
ls()
```

To load the `student` data from midfielddata into memory, we run

```{r}
# Load practice data
data(student, package = "midfielddata")
```

Checking the environment again, 

```{r}
# Names of objects in the environment
ls()
```

If you look in the RStudio *Environment* pane, you should see `student` listed. We will typically load the data sets we need at the top of a script just after loading the midfielddata package. You need to load them only once in a script.  

```{r}
#| echo: false

if (!exists("course")) data(course, package = "midfielddata")
if (!exists("term"))   data(term, package = "midfielddata")
if (!exists("degree")) data(degree, package = "midfielddata")
```

```{r}
#| eval: false

# Load remaining data used in this vignette
data(course, term, degree, package = "midfielddata")
```

Checking the environment again, 

```{r}
# Names of objects in the environment
ls()
```


```{r}
#| echo: false

# cip
n_cip <- nrow(unique(cip))

# student
obs_students <- nrow(student)
var_students <- ncol(student)
size_students <- "19 MB"

# degree
n_id_degrees <- nrow(unique(degree[, .(mcid)]))
n_graduates <- nrow(unique(degree[!is.na(degree), .(mcid)]))
n_term <- nrow(unique(degree[, .(term)]))
obs_degrees <- nrow(degree)
var_degrees <- ncol(degree)
size_degrees <- "10 MB"

# term
n_id_terms <- nrow(unique(term[, .(mcid)]))
obs_terms <- nrow(term)
var_terms <- ncol(term)
n_institutions <- nrow(unique(term[, .(institution)]))
n_programs_terms <- nrow(unique(term[, .(institution, cip6)]))
n_terms <- nrow(unique(term[, .(term)]))
year_span <- c(substr(min(term[, term]), 1, 4), substr(max(term[, term]), 1, 4))
size_terms <- "82 MB"

# course
n_id_courses <- nrow(unique(course[, .(mcid)]))
n_courses <- nrow(unique(course[, .(institution, abbrev, number)]))
n_term_course <- nrow(unique(course[, .(term)]))
obs_courses <- nrow(course)
var_courses <- ncol(course)
size_courses <- "349 MB"
```





## Inspect the `student` data

`student` is a table of `r var_students` student-related variables for  degree-seeking students at the time they were admitted to their institutions. You can view the help page (and data dictionary) by running

```{r}
#| eval: false

# Run in Console
? student
```

Each observation is a unique student keyed by student ID. Student IDs and institution names are de-identified but are consistent across the four data sets. To see the data structure, run

```{r}
# View the structure of the data set
str(student)
```

`student` has `r obs_students` observations and occupies approximately `r size_students` memory. We can confirm that the student IDs are unique by running, 

```{r}
nrow(student)

length(unique(student$mcid))
```

Summarize some of the quantitative variables. 

```{r}
summary(student$sat_math)

summary(student$age)
```

Examine a few of the categorical variables and their values. 

```{r}
sort(unique(student$sex))

sort(unique(student$race))
```

### Exercises
 
1. Determine the possible values in the `institution` column. 
2. Determine the possible values in the `transfer` column. 
3. Determine the median `sat_verbal` score.






## Inspect the `course` data

`course` is a data frame of `r var_courses` course-related variables with observations of students who complete a course or remain in a course beyond their institution's course-withdrawal deadline. You can view the help page (and data dictionary) by running

```{r}
#| eval: false

# Run in Console
? course
```

Each observation is one course in one term for one student keyed by student ID. To see the data structure, run

```{r}
# View the structure of the data set
str(course)
```

Of the four practice data sets in midfielddata, `course` is by far the largest, with `r obs_courses` observations and occupying `r size_courses` of memory. Depending on your machine, loading `course` can cause a noticeable delay in execution time, so you probably want to load it only when needed. 

To be recorded in `course` (unlike `student`), students can be non-degree-seeking but they must complete a course. The number of unique students in `course` is given by, 

```{r}
length(unique(course$mcid))
```
 
Summarize a quantitative variable. 

```{r}
summary(course$hours_course)
```

Examine a categorical variable and its values. 

```{r}
sort(unique(course$type))
```

### Exercises

1. Determine the possible values in the `faculty_rank` column. 
2. Determine the range of terms in `term` column. 
3. Determine the possible values in the `grade` column. 






## Inspect the `term` data

`term` is a data frame of `r var_terms` term-related variables with  observations of students who complete a term or remain in a program beyond their institution's term-withdrawal deadline. You can view the help page (and data dictionary) by running

```{r}
#| eval: false

# Run in Console
? term
```

Each observation is one term for one student keyed by student ID. To see the data structure, run

```{r}
str(term)
```

`term` has `r obs_terms` observations and occupies approximately `r size_terms` of memory. 

To be recorded in `term` (unlike `student`), students can be non-degree-seeking but they must complete a term. The number of unique students in `term` is given by, 

```{r}
length(unique(term$mcid))
```

Summarize some of the quantitative variables. 

```{r}
summary(term$hours_term)

summary(term$gpa_term)
```

Examine some categorical variables and their values.  

```{r}
sort(unique(term$level))

sort(unique(term$standing))
```




### Exercises

1. Determine the possible values in the `coop` column. 
2. Determine the range of terms in `term` column. 
3. Determine the median value of the `gpa_cumul` column. 





## Inspect the `degree` data

`degree` is a table of `r var_degrees` degree-related variables with observations of students who complete their program(s). You can view the help page (and data dictionary) by running

```{r}
#| eval: false

# Run in Console
? degree
```

Each observation is a unique combination of student and degree keyed by student ID. To see the data structure, run

```{r}
str(degree)
```

`degree` has `r obs_degrees` observations and occupies approximately `r size_degrees` memory. We can confirm that the student IDs are unique by running, 

```{r}
nrow(degree)

length(unique(degree$mcid))
```

Had the practice data included students earning a double degree (or more) in the same term, then the number of rows would exceed the number of unique IDs. 

By student ID, the students in `degree` are a subset of the students in `student`. We can confirm by running the following, where a sum of `r length(unique(degree$mcid))`  indicates that all IDs in `degree` are present in `student`. 

```{r}
# Isolate IDs
degree_mcid  <- unique(degree$mcid)
student_mcid <- unique(student$mcid)

# Sum of logical results, 1 for every TRUE, 0 for every FALSE
sum(degree_mcid %chin% student_mcid)
```


  

Examine some categorical variables and their values. 

```{r}
sort(unique(degree$institution))

sort(unique(degree$degree))
```

Checking if any of the degrees are recorded as NA results in an empty table, thus all the students in `degree` are graduates. 

```{r}
degree[is.na(degree)]
```

### Exercises

Some packages have functions specialized for data frame summaries. Try this one---you may find it useful in future data explorations. 

1. Install the `skimr` package, load it with `library("skimr")`, and run `skim(x)`, where `x` is one of the data tables, for example, 

```{r}
#| eval: false

library(skimr)
skim(student)
skim(degree)
```


## Review the midfieldr functions

Table 2 lists the functions in midfieldr and briefly describes what they do. The vignettes give examples of usage.     

```{r}
#| echo: false

wrapr::build_frame(
    "Function", "Action" |
        "add_completion_status()", "Determine degree completion status for every student" |
        "add_data_sufficiency()", "Determine data sufficiency for every student" |
        "add_timely_term()", "Calculate a timely completion term for every student" |
        "filter_search()", "Subset rows that include matches to search strings" |
        "order_multiway()", "Condition multiway data for graphing" |
        "prep_fye_mice()", "Construct and format FYE data for multiple imputation" |
        "select_required()", "Select variables required by other midfieldr functions"
) |>
    kableExtra::kbl(align = "ll", 
                    caption = "Table 2: Summary of midfieldr functions") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::column_spec(1, monospace = TRUE) |>
    kableExtra::column_spec(1:2, color = "black", background = "white") |>
    kableExtra::row_spec(c(0), background = "#c7eae5")
```


### Exercises

View the help pages for the functions `add_timely_term()` and `add_completion_status()`. 

1. Summarize the difference(s) between the two functions. 
2. Describe the manner in which one depends on the other.  





## R help

This section is primarily for users fairly new to R. 

R's built-in help system is an essential part of solving problems you might encounter. Here we provide a short list of R help features applied to midfieldr and midfielddata content. 

**Help function.** From the R Console you can use `?` or `help()`. For example, the following commands yield the same result:

```{r}
#| eval: false

# Type in the Console
help("add_timely_term")
? add_timely_term
```

The help function can also be used to learn more about data sets included with midfieldr and midfielddata. 

```{r}
#| eval: false

# Type in the Console
? cip
```

**Data.** Use the `data()` function to view the names of the data sets included with a package. 

```{r}
#| eval: false

# Type in the Console
data(package = "midfieldr")
#> Data sets in package 'midfieldr':
#> 
#> cip                      
#> fye_proxy                
#> study_mcid               
#> ...                
```

**List functions.** Use `lsf.str()` to list all functions in a package and their arguments, 

```{r}
# Type in the Console
lsf.str("package:midfieldr")
```

**Run examples.** Use the `example()` function to run the function examples that are included with a package.

```{r}
#| eval: false

# Type in the Console
example("add_timely_term")
```







## References

<div id="refs"></div>







## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r}
#| eval: false

# Packages used
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# Loading data
data(student, package = "midfielddata")
data(course, package = "midfielddata")
data(term, package = "midfielddata")
data(degree, package = "midfielddata")

# student
str(student)
length(unique(student$mcid))
summary(student$sat_math)
summary(student$age)
sort(unique(student$sex))
sort(unique(student$race))

# course
str(course)
length(unique(course$mcid))
summary(course$hours_course)
sort(unique(course$type))

# term
str(term)
length(unique(term$mcid))
summary(term$hours_term)
summary(term$gpa_term)
sort(unique(term$level))
sort(unique(term$standing))

# degree
str(degree)
length(unique(degree$mcid))
sort(unique(degree$institution))
sort(unique(degree$degree))
degree[is.na(degree)]
```

```{r}
#| echo: false

# to change the CSS file
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding:     10px 20px;
    margin:      0 0 20px;
    border-left: 0px
}
caption {
    color:       #525252;
    text-align:  left;
    font-weight: normal;
    font-size:   medium;
    line-height: 1.5;
}
```
