---
title: "Starters"
vignette: >
  %\VignetteIndexEntry{Starters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette:
    css: extra.css
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-070-")
```

A degree-seeking student enrolled in their first degree-granting program is a *starter* in that program. Identifying starters is typically performed as part of a graduation rate calculation, though it can also be a useful measure in its own right. 

This vignette in the MIDFIELD workflow.   

1. Planning  
1. Initial processing  
1. [Blocs]{.accent}  
    - Ever-enrolled  
    - FYE proxies  
    - [Starters]{.accent}  
    - Graduates    
1. Groupings  
1. Metrics  
1. Displays 

## Special cases

In two special cases, an entering student's CIP code does not correspond to a degree-granting program. Our procedure for identifying starters accommodate both special cases.

The first case includes records for which a CIP is unspecified or reported as "undecided". In MIDFIELD data, both conditions are encoded as CIP 999999. Students may *enter* with this CIP but we do not consider them *starters* until and if they enroll in a degree-granting  program. (The midfielddata practice datasets contain no undecided CIP codes.)

The second case is more nuanced. At some US institutions, engineering students are required to complete a First-Year Engineering (FYE) program as a prerequisite for declaring an engineering major. These students are admitted as Engineering majors but we don't know to which degree-granting program they intended to transition. At the 2-digit CIP level, FYE students are starters in Engineering (CIP 14). If we do not restrict a study to 2-digit CIPs, however, we use [FYE proxies](art-060-fye-proxies.html)---our estimates of the degree-granting engineering programs (6-digit CIP level) that FYE students would have declared had they not been required to enroll in FYE.










## Definitions

```{r child = "../man/rmd/define-bloc.Rmd"}
``` 

```{r child = "../man/rmd/define-degree-seeking.Rmd"}
``` 

```{r child = "../man/rmd/define-starters.Rmd"}
``` 

entry term

: A student's first term in the database. 

start term

: The first term in which a student can be considered a starter. Identical to the entry term unless the student enters as undecided/unspecified. 

```{r child = "../man/rmd/define-undecided.Rmd"}
```

```{r child = "../man/rmd/define-fye-and-proxy.Rmd"}
``` 




## Method

We use `student` and `term` to identify starters. 

1. Filter the source student-level records for data sufficiency and degree-seeking. 

2. Filter for a student's first term not assigned an undecided/unknown CIP code.  

3. Identify the program(s) of which a student can be considered a starter. Substitute an FYE proxy when a starting program is FYE. 

4. Filter by program. 

```{r child = "../man/rmd/note-for-practice-only-2.Rmd"}
```





## Load data


*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

*Load.* &nbsp; Practice datasets. View data dictionaries via `?student`, `?term`.  

```{r}
# Load practice data
data(student, term)
```

*Loads with midfieldr.* &nbsp; Prepared data. View data dictionaries via `?study_programs`, `?baseline_mcid`, `?fye_proxy`. 

- `study_programs` (derived in [Programs](art-040-programs.html#assigning-program-names)). 

- `baseline_mcid`  (derived in [Blocs](art-050-blocs.html#initial-processing)).

- `fye_proxy` (derived in [FYE proxies](art-060-fye-proxies.html)). 












## Initial processing

*Select (optional).* &nbsp; Reduce the number of columns. Code reproduced from [Getting started](art-000-getting-started.html#reusable-code). 

```{r}
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
term <- select_required(source_term)
```

```{r child = "../man/rmd/note-baseline-mcid.Rmd"}
```

```{r}
# Working data frame
DT <- copy(baseline_mcid)
```





## Isolate the start term

The *start term* is the first term in which a student can be considered a starter, that is, they are degree-seeking and not recorded as undecided/unspecified. 

*Add variables*. &nbsp;  Left join to add terms and CIPs for these students. 

```{r}
# Term into DT left join
DT <- term[DT, .(mcid, term, cip6), on = c("mcid")]
DT
```

*Filter.*  &nbsp; We remove observations of undecided/unspecified (CIP 999999). Any rows remaining for the same IDs will have CIPs of degree-granting program (or FYE), allowing us to infer their preferred starting programs. (A required step for completeness, but unnecessary when using the practice data.)

```{r}
# Remove undecided/unspecified
DT <- DT[!cip6 %like% "999999"]
DT
```

*Filter.* &nbsp; Order rows by ID and term, then filter to retain the start term observation. If your data contain students enrolled in more than one major in their first term, replace `.SD[1]` with the (slower) `.SD[which.min(term)]`.

```{r}
# Retain observations of the earliest remaining terms by ID
setorderv(DT, cols = c("mcid", "term"))
DT <- DT[, .SD[1], by = "mcid"]
DT
```

  

*Filter.* &nbsp; Remove unnecessary variables and filter for unique observations. 

```{r}
# Unique combinations of ID and CIP
DT <- DT[, .(mcid, cip6)]
DT <- unique(DT)
DT
```







## Starters without FYE

If and only if our study excluded Engineering programs that require FYE, the 
data frame just derived would be the desired bloc of starters. 

In such a case, we would rename `cip6` to `start` to make explicit that the CIP codes in this column represent programs of which the students can be considered starters. The code would have the form  

```{r}
#| eval: false
# Not run
DT <- DT[, .(mcid, start = cip6)]
```

which retains the ID variable and changes the name of the CIP variable. 





## Starters with FYE

*Add a variable.*  &nbsp; Merge `fye_proxy` with the working data frame. The left join introduces NA in the proxy column for students not assigned an FYE proxy.  

```{r}
# Join the proxies to the working data frame
DT <- fye_proxy[DT, on = c("mcid")]
DT
```

*Create a variable.*  &nbsp;  Estimated starting programs for FYE students are in the `proxy` column. Actual, recorded starting programs for non-FYE students are in the `cip6` column. Create the `start` column to combine the two. 

```{r}
# Combine all starting CIPs
DT[, start := fcase(
  cip6 == "140102", proxy,
  cip6 != "140102", cip6
)]
DT
```

```{r}
#| eval: false
#| echo: false
# Searching through the proxies for closer look cases
x <- DT[!is.na(proxy)]

for (jj in x$mcid) {
  y <- term[mcid == jj]
  y <- y[cip6 != "140102"]

  # y <- y[!cip6 %like% "^14"]
  #
  # if (nrow(y) != 0 ){
  #     print(y)
  #     profvis::pause(1)
  # }

  if (nrow(y) == 0) {
    print(term[mcid == jj])
    profvis::pause(1)
  }
}
DT[mcid == "MCID3111303095"]
term[mcid == "MCID3111303095"]
```

*Select.* &nbsp;  Omit unnecessary columns. 

```{r}
# Omit unnecessary columns.
DT <- DT[, .(mcid, start)]
DT
```







### Closer look

Examining the records of selected students in detail.

*Example 1.* &nbsp; In our results, this student is a starter in CIP 143501 (Industrial Engineering).  

```{r}
# Analysis result
DT[mcid == "MCID3111150194"]
```

An excerpt from their record in `term` shows them enrolled in CIP 140102 (FYE) for three terms followed by CIP 143501 They transitioned post-FYE to Industrial Engineering and we consider them a starter in that program. 

```{r}
# Sequence of term records
term[mcid == "MCID3111150194"]
```

*Example 2.* &nbsp;  In our results, this student is a starter in CIP 141801 (Materials Engineering). 

```{r}
# Analysis result
DT[mcid == "MCID3111161837"]
```

An excerpt from their record in `term` shows them enrolled in CIP 140102 (FYE) for three terms followed by CIP 270101 (Mathematics)---they transitioned from FYE to a non-engineering major. Thus we consider them a starter in their proxy program, Materials Engineering. 

```{r, echo=c(4, 5)}
op <- options()
options(datatable.print.nrows = 11)

# Sequence of term records
term[mcid == "MCID3111161837"]

options(op)
```

*Example 3.* &nbsp;  In our results, this student is a starter in CIP 140701 (Chemical Engineering). 

```{r}
# Analysis result
DT[mcid == "MCID3111303095"]
```

An excerpt from their record in `term` shows them enrolled in CIP 140102 (FYE) for two terms and then leaving the database. Again, we consider them a starter in their proxy program, Chemical Engineering. 

```{r}
term[mcid == "MCID3111303095"]
```







## Filter by program

*Filter.*  &nbsp; Because "starter" usually means "starter in specific programs," this bloc concludes with a filter by program.

```{r}
# Rename cip6 as start
join_labels <- copy(study_programs)
join_labels <- join_labels[, .(program, start = cip6)]

# Filter by program
DT <- join_labels[DT, on = c("start"), nomatch = NULL]
DT
```

*Select.*  &nbsp; Omit unnecessary variables. 

```{r}
DT <- DT[, .(mcid, program)]
DT <- unique(DT)
DT
```








## Reusable code

*Preparation.* &nbsp; The data frame of baseline IDs is the intake for this section. 

```{r}
DT <- copy(baseline_mcid)
```

*Starters.* &nbsp; Summary code chunks for ready reference. 

```{r}
# Isolate starting term
DT <- term[DT, .(mcid, term, cip6), on = c("mcid")]
DT <- DT[!cip6 %like% "999999"]
setorderv(DT, cols = c("mcid", "term"))
DT <- DT[, .SD[1], by = "mcid"]
# Alternatively
# DT <- DT[, .SD[which.min(term)], by = "mcid"]
DT <- DT[, .(mcid, cip6)]
DT <- unique(DT)
```

For starters without FYE, finish by renaming `cip6` 

```{r}
#| eval: false
# Not run
DT <- DT[, .(mcid, start = cip6)]
```

For starters with FYE, continue with FYE proxies. 

```{r}
DT <- fye_proxy[DT, .(mcid, cip6, proxy), on = c("mcid")]
DT[, start := fcase(
  cip6 == "140102", proxy,
  cip6 != "140102", cip6
)]
DT <- DT[, .(mcid, start)]

# Filter by program on start
join_labels <- copy(study_programs)
join_labels <- join_labels[, .(program, start = cip6)]
DT <- join_labels[DT, on = c("start"), nomatch = NULL]
DT <- DT[, .(mcid, program)]
DT <- unique(DT)
```




## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```

