---
title: "Multiway graphs"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Multiway graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files:
  - ../man/figures/art-050-multiway-graphs-fig1-1.png
  - ../man/figures/art-050-multiway-graphs-fig2-1.png
  - ../man/figures/art-050-multiway-graphs-fig3-1.png
  - ../man/figures/art-050-multiway-graphs-fig4-1.png
  - ../man/figures/art-050-multiway-graphs-fig5-1.png
  - ../man/figures/art-050-multiway-graphs-fig6-1.png
---

```{r setup, include = FALSE}
source("article-preamble.R")
set_fig_path("art-050-multiway-graphs-")
```


## Introduction 

We introduce graphs of *multiway data* in which there is is one quantitative variable and two categorical variables. As explained by  Cleveland [-@Cleveland:1993], 

> ... the quantitative variable is a response, and the goal is to study how it depends on the categorical variables, which are factors. What distinguishes multiway data is the cross-classification of the categorical variables; there is a value of the response for each combination of levels of the two categorical variables. 

Multiway data have a specific structure: a categorical variable with $m$ levels; an independent categorical variable with $n$ levels; and a single  quantitative value associated with every combinations of levels of the two categories. Multiway data are typically organized as three columns with $m \times n$ rows. 

Note that "multiway" in our context refers to the data structure and graph design defined by Cleveland, not to the methods of analysis described by Kroonenberg [-@Kroonenberg:2008]. 

### Outline

- Start with the case study stickiness results (`study_stickiness`) 
- Practice conditioning multiway data for display 
- Practice creating multiway graphs and their duals 
- Practice different options for ordering rows and panels

### In this vignette

midfieldr functions 
 
- [`condition_multiway()`](../reference/condition_multiway.html)

packages

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
library("ggplot2")
```

We can control how a data.table is printed to the screen with the following options.  

```{r}
# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```


## Prepare data

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@Ohland+Orr+others:2012]. The stickiness metric is developed in a separate vignette---here we focus on graphing a result. 

We computed stickiness for a case study in Civil, Electrical, Industrial, and Mechanical Engineering and saved the results as the data set, `study_stickiness`. You may view its help page by running

```
? study_stickiness
```

The data is loaded with midfieldr (we don't have to use `data()` to access it.)

```{r}
# examine the built-in data
study_stickiness
```

Categorical variables are often provided as character variables. Without ordering the levels, the rows and panels of the multiway graph are ordered alphabetically---not the best design for gaining insights from the graph. 

To condition the categorical variables, we use `condition_multiway()` to convert them to factors with levels ordered by the quantitative response variable in some fashion, To see its help page, run

```
? condition_multiway()
```

The function requires two categorical variables and one quantitative variable. In this data frame, the two categorical columns are `program` and `race_sex`. the quantitative variable is `stick` (for longitudinal stickiness). Before conditioning, the categorical variables are characters, 

```{r}
# create a new memory location
DT <- copy(study_stickiness)

# first category is a character
class(DT$program)

# second category is a character
class(DT$race_sex)
```

We do some data conditioning before presenting the results. Even with students and institutions de-identified, privacy is at risk for small populations, so we omit observations with fewer than 10 students ever enrolled. 

```{r}
# Omit before graphing
DT <- DT[ever >= 10]
```

We also omit rows with ambiguous values for race/ethnicity (International and Other/Unknown). Having graphed the results in advance, we find that the small population subset removed Native American students from all but one of the disciplines---so we add them to the list of observations to omit. 

```{r}
# data table printing option
options(datatable.print.topn = 3)

# Omit before graphing
DT <- DT[!race_sex %ilike% c("International|Other|Native")]

# Examine the result
DT

temp <- copy(DT)
```
 
We process the data frame

```{r}
# condition as multiway data
DT <- condition_multiway(
  dframe = DT, # input data frame
  categ_col = c("program", "race_sex"), # multiway categorical variables
  quant_col = "stick", # multiway quantitative variable
  detail = FALSE, # do not return detail columns 
  order_by = "percent", # row and panel ordering method
  param_col = c("grad", "ever") # parameters used for ordering
)
```

No change to the row and column structure, but as you can see, the previous character variables have been converted to factors, 

```{r}
DT
```

The levels of the factors are ordered non-alphabetically. `program` is now a factor with `r nlevels(DT[, program])` levels; `race_sex` is a factor with `r nlevels(DT[, race_sex])` levels. 

```{r}
# first category is now a factor
levels(DT$program)

# second category is now a factor
levels(DT$race_sex)
```

The values that determine the order of the levels can be obtained by using the `detail` argument. 

```{r}
# return stickiness by category
DT <- condition_multiway(
  dframe = DT, # input data frame
  categ_col = c("program", "race_sex"), # multiway categorical variables
  quant_col = "stick", # multiway quantitative variable
  detail = TRUE, # return detail columns
  order_by = "percent", # row and panel ordering method
  param_col = c("grad", "ever") # parameters used for ordering
)
```

By ordering the data by `program`, you can see that a given program has a single value of `program_stick` copied as many times as needed. 

```{r}
# optional code to control data.table printing
options(datatable.print.topn = 10)

# programs have a median stickiness across race-sex groups
DT[order(program)]
```

Similarly, by ordering the data by `race_sex`, you can see that a given demographic group has a single value of `race_sex_stick` copied as many times as needed. 

```{r}
# race_sex groupings have a median stickiness across programs
DT[order(race_sex)]
```


### Exercises 

1. Determine the class of the categorical variables in the built-in data set `study_grad_rate`. 
2. Apply `condition_multiway()`. Confirm that the categorical variables are factors. 
3. Determine the order of the factor levels. 


## Create a multiway graph

```{r include = FALSE}
asp_ratio <- asp_ratio_mw(DT, categories = c("program", "race_sex"))
```

We use conventional ggplot2 functions to create the multiway graphs. 

By previously ordering the levels of the factors, we have structured the data so that the rows and panels of the multiway graph are ordered by the method we selected. here we used the "percent" method. Program stickiness is drawn as a vertical reference line in each panel.

*Reminder:* These are practice data, not suitable for drawing inferences about student performance, i.e., not for research.

```{r fig1, fig.asp = asp_ratio[1]}
# create one multiway graph
ggplot(data = DT, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = program_stick),
    linetype = 2,
    color = "gray70"
  ) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Stickiness (%)",
    y = "",
    title = "Practice data (not for research)",
    caption = "Source: midfielddata"
  )
```
<br>

This graph permits a direct visual comparison of how stickiness varies by race and sex within a particular major. 

Reading a multiway graph

- We can more effectively compare values within a panel than between panels.
- Because rows are ordered, one expects a generally increasing trend within a panel. A response greater or smaller than expected creates a visual asymmetry. The interesting stories are often in these visual anomalies.
- To explore the comparison between panels, we swap the row and panel categories, creating the dual multiway.

### Exercise

1. Create a multiway graph using the conditioned `study_grad_rate` data 



## Create the dual multiway graph

The previous graph does not facilitate visual comparisons of members of the same race-sex group. As Cleveland says,

> Because of this asymmetry, it is often important to explore multiway data by as many multiway dot plots as there are categorical variables, with each variable assigned once to the levels.


To create the *dual multiway* graph, we swap the roles of the rows and panels. 

- `y = race_sex` becomes `y = program`
- `facet_wrap(vars(program) ...`  becomes `facet_wrap(vars(race_sex) ...` 

```{r fig2, fig.asp = asp_ratio[2]}
# create the dual multiway graph
ggplot(data = DT, aes(x = stick, y = program)) +
  facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = race_sex_stick),
    linetype = 2,
    color = "gray70"
  ) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Stickiness (%)",
    y = "",
    title = "Practice data (not for research)",
    caption = "Source: midfielddata"
  )
```
<br>

This graph permits a direct visual comparison of how stickiness varies by major for one group of students. Race/ethnicity/sex group stickiness is drawn as a vertical reference line in each panel. 

Both multiways should be created for discussion. Usually one or the other will convey a more compelling visual story that gets included in the published findings. 


### Exercise

1. Create the dual multiway graph using the conditioned `study_grad_rate` data 






## Panel and row ordering options 

The `order_by` argument in `condition_multiway()` provides options for ordering the two categorical variables (panel order and row order) in a multiway graph. 

```{r}
# create a common graph framework to use in the example below
p <- ggplot(data = temp, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Longitudinal stickiness (%)",
    y = "",
    caption = "Source: midfielddata"
  )
```

The following values of `order_by` are possible:

percent
: Orders by ratios computed by category. Used when the quantitative response variable is the ratio (in percent) of two columns of integer counts (the two columns are the numerator and denominator of the ratio), such as the ratio of graduates to ever-enrolled (stickiness) or the ratio of graduates to starters (graduation rate). The counts in the two columns are summed by category and ratios for that category are computed. These grouped ratios are used to order the levels of the categories. 

: The graphs shown above are representative of this type of ordering. 

: Requires the optional `param_col` argument with the names of the columns of the two terms of the ratio.

median
: (default) Orders by the median of values in the quantitative column grouped by category. Here, the only quantity manipulated is the 

```{r fig3, fig.asp = asp_ratio[1]}
dtf <- condition_multiway(
  dframe = temp,
  categ_col = c("program", "race_sex"),
  quant_col = "stick",
  detail = TRUE
)
p %+% dtf +
  labs(title = "Levels ordered by median values") +
  geom_vline(aes(xintercept = program_stick),
    linetype = 2,
    col = "gray50"
  )
```

mean
: Orders by the mean of values in the quantitative column grouped by category.

```{r fig4, fig.asp = asp_ratio[1]}
dtf <- condition_multiway(
  dframe = temp,
  categ_col = c("program", "race_sex"),
  quant_col = "stick",
  order_by = "mean",
  detail = TRUE
)
p %+% dtf +
  labs(title = "Levels ordered by mean values") +
  geom_vline(aes(xintercept = program_stick),
    linetype = 2,
    col = "gray50"
  )
```

sum
: Orders by the sum of values in the quantitative column grouped by category. Useful when the response variable is a count or frequency.

```{r fig5, fig.asp = asp_ratio[1]}
dtf <- condition_multiway(
  dframe = temp,
  categ_col = c("program", "race_sex"),
  quant_col = "ever",
  order_by = "sum",
  detail = TRUE
)
ggplot(data = dtf, aes(x = ever, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  scale_x_continuous(trans = "log2") +
  geom_vline(aes(xintercept = program_ever),
    linetype = 2,
    col = "gray50"
  ) +
  labs(
    x = "Number of students ever enrolled (log-2 scale)",
    y = "",
    title = "Levels ordered by count totals",
    caption = "Source: midfielddata"
  )
```

alphabet
: Orders the levels of the categorical variables alphabetically. Rarely useful for perceiving effects, but can be useful for value look up.

```{r fig6, fig.asp = asp_ratio[1]}
dtf <- condition_multiway(
  dframe = temp,
  categ_col = c("program", "race_sex"),
  quant_col = "stick",
  order_by = "alphabet"
)
p %+% dtf +
  labs(title = "Levels ordered alphbetically")
```


## References

<div id="refs"></div>






## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# packages used
library("midfieldr")
suppressMessages(library("data.table"))
library("ggplot2")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)


```

```{r echo = FALSE}
# to change the CSS file for block quotes
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 0px solid
}
```



