---
title: "Getting started"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: |
  @Dowle+Srinivasan:2020:data.table,  
  @Waring+Quinn:2021:skimr
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
    fig.path = "../man/figures/art-000-get-started-", 
    fig.width = 6,
    fig.asp = 1 / 1.6,
    out.width = "70%",
    fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```


## Introduction 

Before beginning our survey of student unit record (SUR) data, check that the necessary prerequisites have been met: 

- [Requirements](../index.html#requirements)
- [Install midfieldr](../index.html#install-midfieldr) 
- [Install midfielddata](../index.html#install-midfielddata)

In this vignette, we

- Load the practice data from midfielddata
- Inspect the four data tables 
- Preview the midfieldr functions 


## Student unit record data 

The MIDFIELD Student Unit Record (SUR) data are de-identified undergraduate student-level information covering admission characteristics, demographics, programs, courses, instructors, grades, and degrees. The data are organized into four tables---student, course, term, and degree---keyed by student ID. 

In this vignette, we examine the contents of practice data from [midfielddata](https://midfieldr.github.io/midfielddata/), an R data package with longitudinal SURs for nearly 98,000 undergraduates at 12 institutions from  1987--2016: 


```{r echo = FALSE}
wrapr::build_frame(
    "Practice data table", "Each row is", "No. of rows", "No. of columns" |
        "[`student`](https://midfieldr.github.io/midfielddata/reference/student.html)", "a degree-seeking student", "97,640", "13" |
        "[`course`](https://midfieldr.github.io/midfielddata/reference/course.html)", "a student in a course", "3.5M", "12" |
        "[`term`](https://midfieldr.github.io/midfielddata/reference/term.html)", "a student in a term", "728,000", "13" |
        "[`degree`](https://midfieldr.github.io/midfielddata/reference/degree.html)", "a student who graduates", "48,000", "5"
) |>
    kableExtra::kbl(align = "llrr") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::column_spec(1:4, color = "black", background = "white")
```














## Load practice data

If you are writing your own script to follow along, we start with these packages: 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# optional code to control data.table printing
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```




Use `data()` to load the midfielddata data tables into memory for use. For example, after loading the package with `library("midfielddata")` as we did above, you can run `ls()` to see that the environment contains zero objects, 

```{r}
# the names of objects in the environment
ls()
```

To load the `student` data from midfielddata into memory, we run

```{r echo = FALSE, cache = FALSE}
# cache = TRUE speeds up repeats
if (!exists("student")) data(student)
```

```r
# loading a data set 
data(student)
```

Checking the environment again, 

```{r}
# the names of objects in the environment
ls()
```

If you look in the RStudio *Environment* pane, you should see `student` listed. We will typically load the data sets we need at the top of a script just after loading the midfielddata package. You need to load them only once in a session.  

```{r echo = FALSE, cache = FALSE}
# cache = TRUE speeds up repeats
if (!exists("course")) data(course)
if (!exists("term")) data(term)
if (!exists("degree")) data(degree)
```

```r
# load remaining data used in this vignette
data(course)
data(term)
data(degree)
```

Checking the environment again, 

```{r}
# the names of objects in the environment
ls()
```


```{r echo = FALSE}
# cip
n_cip <- nrow(unique(cip))

# student
obs_students <- nrow(student)
var_students <- ncol(student)
size_students <- "19 MB"

# degree
n_id_degrees <- nrow(unique(degree[, .(mcid)]))
n_graduates <- nrow(unique(degree[!is.na(degree), .(mcid)]))
n_term <- nrow(unique(degree[, .(term)]))
obs_degrees <- nrow(degree)
var_degrees <- ncol(degree)
size_degrees <- "10 MB"

# term
n_id_terms <- nrow(unique(term[, .(mcid)]))
obs_terms <- nrow(term)
var_terms <- ncol(term)
n_institutions <- nrow(unique(term[, .(institution)]))
n_programs_terms <- nrow(unique(term[, .(institution, cip6)]))
n_terms <- nrow(unique(term[, .(term)]))
year_span <- c(substr(min(term[, term]), 1, 4), substr(max(term[, term]), 1, 4))
size_terms <- "82 MB"

# course
n_id_courses <- nrow(unique(course[, .(mcid)]))
n_courses <- nrow(unique(course[, .(institution, abbrev, number)]))
n_term_course <- nrow(unique(course[, .(term)]))
obs_courses <- nrow(course)
var_courses <- ncol(course)
size_courses <- "349 MB"
```





## Inspect the `student` data

`student` is a table of `r var_students` student-related variables for the students in the sample, collected at the time they were admitted to their institutions as  degree-seeking undergraduates. For the data dictionary, you can navigate to the midfielddata website  [`student`](https://midfieldr.github.io/midfielddata/reference/student.html) page or view the same information in the help page by running 

```r
? student
``` 

Each observation is a unique student keyed by student ID. Student IDs and institution names are de-identified but are consistent across the four data sets. To see the data structure, run

```{r}
# view the structure of the data set
str(student)
```

`student` has `r obs_students` observations and occupies approximately `r size_students` memory. We can confirm that the student IDs are unique by running, 

```{r}
length(unique(student$mcid))
```

Summarize some of the quantitative variables. 

```{r}
summary(student$sat_math)

summary(student$age)
```

Examine a few of the categorical variables and their values. 

```{r}
sort(unique(student$sex))

sort(unique(student$race))
```

### Exercises
 
1. Determine the possible values in the `institution` column. 
2. Determine the possible values in the `transfer` column. 
3. Determine the median `sat_verbal` score.

## Inspect the `course` data

`course` is a table of `r var_courses` course-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`course`](https://midfieldr.github.io/midfielddata/reference/course.html) page or view the same information in the help page by running 

```r
? course
```

Each observation is one course in one term for one student keyed by student ID and a combination of course, abbreviation, number, and section. To see the data structure, run

```{r}
# view the structure of the data set
str(course)
```

Of the four practice data sets in midfielddata, `course` is by far the largest, with `r obs_courses` observations and occupying `r size_courses` of memory. Depending on your machine, loading `course` can cause a noticeable delay in execution time (13 seconds on my machine), so you probably want to load it only when needed. 

The number of unique students.  

```{r}
# number of unique IDs
length(unique(course$mcid))
```
 
Summarize a quantitative variable. 

```{r}
summary(course$hours_course)
```

Examine a categorical variable and its values. 

```{r}
# type of courses
sort(unique(course$type))
```

### Exercises

1. Determine the possible values in the `faculty_rank` column. 
2. Determine the range of terms in `term` column. 
3. Determine the possible values in the `grade` column. 

## Inspect the `term` data

`term` is a table of `r var_terms` term-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`term`](https://midfieldr.github.io/midfielddata/reference/term.html) page or view the same information in the help page by running

```r
? term
```

Each observation is one term for one student keyed by student ID and term. To see the data structure, run

```{r}
str(term)
```

`term` has `r obs_terms` observations and occupies approximately `r size_terms` of memory. The number of unique students is found by, 

```{r}
# number of unique IDs
length(unique(term$mcid))
```

The number of unique students in `student` and `term` are not the same. To be recorded in `student`, students must be degree-seeking. To be recorded in `term`, students must complete a term. The data can include degree-seekers who fail to complete a term as well as term-completers not seeking a degree---hence the difference in numbers

Summarize some of the quantitative variables. 

```{r}
summary(term$hours_term)

summary(term$gpa_term)
```

Examine some categorical variables and their values.  

```{r}
sort(unique(term$level))

sort(unique(term$standing))
```

### Exercises

1. Determine the possible values in the `coop` column. 
2. Determine the range of terms in `term` column. 
3. Determine the median value of the `gpa_cumul` column. 

## Inspect the `degree` data

`degree` is a table of `r var_degrees` degree-related variables for the students in the sample. For the data dictionary, you can navigate to the midfielddata website  [`degree`](https://midfieldr.github.io/midfielddata/reference/degree.html) page or view the same information in the help page by running

```r
? degree
```

Each observation is a unique student keyed by student ID. To see the data structure, run

```{r}
str(degree)
```

`degree` has `r obs_degrees` observations and occupies approximately `r size_degrees` memory. The number of unique students is found by

```{r}
# number of unique IDs
length(unique(degree$mcid))
```

Thus we have `r obs_degrees` graduates (in `degree`) from the `r obs_students` students admitted (in `student`). 

Examine some categorical variables and their values. 

```{r}
sort(unique(degree$institution))

sort(unique(degree$degree))
```

Checking if any of the degrees are recorded as NA results in an empty table, thus all the students in `degree` are graduates. 

```{r}
degree[is.na(degree)]
```

### Exercises

Some packages have functions specialized for data frame summaries. Try this one---you may find it useful in future data explorations. 

1. Install the `skimr` package, load it with `library("skimr")`, and run `skim(x)`, where `x` is one of the data tables, for example, `skim(student)`, `skim(degree)`, etc.  



## Review the midfieldr functions

midfieldr functions are introduced in detail in subsequent vignettes, but we can introduce their general organization here. 

<br>

**`add_*`**    
This family of functions takes working data frames as input and adds specific columns of interest without subsetting rows. 

```{r echo = FALSE}
wrapr::build_frame(
  "Function", "Action" |
    "add_completion_timely()", "Add a column to evaluate program completion" |
    "add_data_sufficiency()", "Add a column to evaluate data sufficiency" |
    "add_institution()", "Add a column of institution names" |
    "add_race_sex()", "Add columns for student race/ethnicity and sex" |
    "add_timely_term()", "Add a column of terms delimiting timely completion"
) |>
    kableExtra::kbl(align = "ll") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::column_spec(1:2, color = "black", background = "white")
```

<br>

**`filter_*`**    
This family of functions takes working data frames as input and subsets rows that match conditions; selected columns are subset if specified.  

```{r echo = FALSE}
wrapr::build_frame(
  "Function", "Action" |
    "filter_match()", "Subset rows by matching values in shared key columns" |
    "filter_search()", "Subset rows that include matches to search strings"
) |>
    kableExtra::kbl(align = "ll") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::column_spec(1:2, color = "black", background = "white")
```

<br>

**`condition_*`**    
This family of functions takes highly-structured data frames as input and produces output data frames conditioned for particular tasks. 

```{r echo = FALSE}
wrapr::build_frame(
  "Function", "Action" |
    "condition_fye()", "Condition FYE data for multiple imputation" |
    "condition_multiway()", "Condition multiway data for graphing"
) |>
    kableExtra::kbl(align = "ll") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::column_spec(1:2, color = "black", background = "white")
```


### Exercises

View the help pages for the functions `add_timely_term()` and `add_completion_timely()`. 

1. Summarize the difference(s) between the two functions. 
2. Describe the manner in which one depends on the other.  

View the help pages for the functions `add_institution()` and `add_race_sex()`.

3. List similarities in what the two functions do. 
4. List differences in what the two functions do. 

## References

<div id="refs"></div>







## Appendix

### Complete script

We conclude each vignette by collecting its code chunks in a single, condensed script for the convenience of those who wish to copy it into their own R file. 

```{r eval = FALSE}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))

# loading data
data(student)
data(course)
data(term)
data(degree)

# student
str(student)
length(unique(student$mcid))
summary(student$sat_math)
summary(student$age)
sort(unique(student$sex))
sort(unique(student$race))

# course
str(course)
length(unique(course$mcid))
summary(course$hours_course)
sort(unique(course$type))

# term
str(term)
length(unique(term$mcid))
summary(term$hours_term)
summary(term$gpa_term)
sort(unique(term$level))
sort(unique(term$standing))

# degree
str(degree)
length(unique(degree$mcid))
sort(unique(degree$institution))
sort(unique(degree$degree))
degree[is.na(degree)]
```


