---
title: "Data selection"
vignette: >
  %\VignetteIndexEntry{Data selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
    rmarkdown::html_vignette:
        highlight: "tango"
        css: "style.css"
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-141-", 
                      message = FALSE)
```

## Getting started

As shown in the [How to](art-140-data-manipulation.html) article.

```{r}
# packages
library(midfieldr)    # same_content()
library(midfielddata) # datasets
library(data.table)   # data manipulation system
library(dplyr)        # data manipulation system
library(tidyr)        # drop_na()

# Initialize data frames 
data(student, term)
student_df  <- data.frame(student)
student_dt  <- copy(student)
student_tbl <- tibble(student)
term_df  <- data.frame(term)
term_dt  <- copy(term)
term_tbl <- tibble(term)
```






## Subsetting rows and columns

A common task when working with data is selecting a subset of rows,  columns, or both. 




### Index using intermediate variables

*Scenario.* &nbsp; Summarize domestic students by race, sex, and institution.

Our first step might be to subset columns (institution, race, sex, and us_citizen) and subset rows such that US citizenship is true. We often construct intermediate variables `columns_we_want` and `rows_we_want` as indices for column and row subsetting. The extra clarity is often worth the extra lines of code.  

```{r}
# Column names in a character vector 
columns_we_want <- c("institution", "race", "sex", "us_citizen")
```

```{r class.source = "base"}
# base R
rows_we_want <- student_df$us_citizen == "Yes"
result_df <- student_df[rows_we_want, columns_we_want, drop = FALSE]
```

```{r class.source = "dtbl"}
# data.table
rows_we_want <- student_dt$us_citizen == "Yes"
result_dt <- student_dt[rows_we_want, ..columns_we_want]
```

```{r class.source = "dplyr"}
# dplyr
rows_we_want <- student_tbl$us_citizen == "Yes"
result_tbl <- student_tbl %>%
    select(all_of(columns_we_want)) %>%
    filter(rows_we_want)
```

```{r}
result_dt

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```


*Comments*

- *base R.* When working with `[,]` always add a third argument `drop = FALSE` to ensure a data frame is returned. Base R's default behavior when selecting a single column is to return a vector. 

- *data.table.* The `..` notation indicates that `columns_we_want` refers to a character vector of column names and is not itself a column name. 

- *dplyr.* The added notation `all_of()` in effect serves the same purpose in dplyr as `..` does in data.table. 







### Direct indexing

*Scenario.* &nbsp; Summarize domestic students by race, sex, and institution.

Column and row selection can be typed directly into the subsetting operators `[,]` or `filter()` and `select()`. 

```{r class.source = "base"}
# base R
result_df <- student_df[student_df$us_citizen == "Yes", 
                        c("institution", "race", "sex", "us_citizen"), 
                        drop = FALSE]
```

```{r class.source = "dtbl"}
# data.table
result_dt <- student_dt[us_citizen == "Yes", .(institution, race, sex, us_citizen)]
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- student_tbl %>%
    select(institution, race, sex, us_citizen) %>%
    filter(us_citizen == "Yes")
```

```{r}
result_dt

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* For direct indexing, intermediate variable names in `[,]` are replaced verbatim by their values. 

- *data.table.* The notation `.()` subsets columns by name without quotation marks;  rows are subset without the `$` operator.

- *dplyr.* Similarly, `select()` operates on unquoted column names; rows are subset without the `$` operator. 






















### Subset rows with multiple conditions

*Scenario.* &nbsp; Find IDs of transfer students with 15 or more hours transferred. 

```{r class.source="base"}
# base R
columns_we_want <- c("mcid", "transfer", "hours_transfer")
rows_we_want <- student_df$transfer == "First-Time Transfer" & student_df$hours_transfer >= 15
result_df <- student_df[rows_we_want, columns_we_want, drop = FALSE]

# remove rows with NA in every column
rows_we_want <- rowSums(is.na(result_df)) != ncol(result_df)
result_df <- result_df[rows_we_want, , drop = FALSE]
```

```{r class.source="dtbl"}
# data.table
rows_we_want <- student_dt$transfer == "First-Time Transfer" & student_dt$hours_transfer >= 15
result_dt <- student_dt[rows_we_want, ]
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- student_tbl %>%
    filter(transfer == "First-Time Transfer" & hours_transfer >= 15)
```

```{r}
str(result_dt)

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* Two steps required. Subset rows by the original logical conditions; then subset to remove rows with NA across all columns. 

- *data.table.* Multiple conditions separated by logical AND `&`. 

- *dplyr.* Similar. 





### Subset one column as a vector

*Scenario.* &nbsp; Suppose we want a vector of unique student IDs from the `term` dataset.

```{r class.source="base"}
# base R
result_df <- term_df[, "mcid"]
result_df <- unique(result_df)
```

```{r class.source="dtbl"}
# data.table
result_dt <- term_dt[, mcid]
result_dt <- unique(result_dt)
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- term_tbl %>%
    pull(.data$mcid) %>%
    unique()
```

```{r}
str(result_dt)

all.equal(result_dt, result_df)
all.equal(result_dt, result_tbl)
```


*Comments*

- *base R.* 

- *data.table.* 

- *dplyr.* The conventional syntax would be `pull(mcid)` except for a subtle flaw in the current version of `pull()`. For details, see [@Layton:2024]. 









## Removing records with incomplete data

An important variation of subsetting data is removing rows of data that have missing values. If appropriate, limit the dropped rows to those with missing values in specific columns. 

*Scenario.* &nbsp; We are asked to compare SAT scores by institution. Before doing any analysis, we first remove rows with missing SAT scores. 




### Using complete.cases

We select three columns, then obtain a logical vector with one entry per row of the data frame: TRUE if a row has no missing entries, FALSE otherwise. 

```{r class.source = "base"}
# base R
result_df <- student_df[, c("institution", "sat_math", "sat_verbal"), drop = FALSE]
result_df <- result_df[complete.cases(result_df), , drop = FALSE]
```

```{r class.source = "dtbl"}
# data.table
result_dt <- student_dt[, .(institution, sat_math, sat_verbal)]
result_dt <- result_dt[complete.cases(result_dt), ]
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- student_tbl %>%
    select(institution, sat_math, sat_verbal) %>%
    filter(complete.cases(.))
```

```{r}
result_dt

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* Where before we used the name `columns_we_want` to reference a character vector of column names, here we use the character vector itself. 

- *data.table.* 

- *dplyr.* In magrittr pipe notation a dot `.` is taken to mean the item being piped. Here for example, the output of `select()` is piped to `filter()` and `complete.cases(.)`.  









### Using na.omit or equivalent

We can perform the same task without explicit Boolean row-selection in any of the systems. 

```{r class.source = "base"}
# base R
result_df <- student_df[, c("institution", "sat_math", "sat_verbal"), drop = FALSE]
result_df <- na.omit(result_df)
```

```{r class.source = "dtbl"}
# data.table
result_dt <- student_dt[, .(institution, sat_math, sat_verbal)]
result_dt <- na.omit(result_dt)
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- student_tbl %>%
    select(institution, sat_math, sat_verbal) %>%
    drop_na()
```

```{r}
result_dt

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* 

- *data.table.*  

- *dplyr.* 













### Missing values in specified columns

Retain all columns. Delete rows with missing values in specified columns. 

```{r class.source = "base"}
# base R
row_index <- with(student_df, !(is.na(institution) | is.na(sat_math) | is.na(sat_verbal)))
result_df <- student_df[row_index, , drop = FALSE]
```

```{r class.source = "dtbl"}
# data.table
result_dt <- na.omit(student_dt, cols = c("institution", "sat_math", "sat_verbal"))
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- student_tbl %>%
     drop_na(institution, sat_math, sat_verbal)
```

```{r}
dim(result_dt)
same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* Using logical operators `!` NOT and `|` OR to create a logical vector used as a row index. 

- *data.table.* The same function name `na.omit()` as base R but with an argument to drops rows with missing values in specified columns. Retains all columns. 

- *dplyr.* Similarly, `drop_na()` drops rows with missing values in  specified columns, retaining all columns. 









## Ordering rows

In some instances, it is important for us to control the order of rows in our data. 

*Scenario.* &nbsp; Suppose we want to know a student's level and declared major in their first term at their institution. We use the `term` dataset.  

We decide to order the rows first by institution, then by ID, then by term. In each case, we select columns in a separate operation so we can highlight the row-ordering operation.  

```{r}
columns_we_want <- c("institution", "mcid", "term", "level", "cip6")
```

```{r class.source = "base"}
# base R
result_df <- term_df[, columns_we_want, drop = FALSE]

order_index <- with(result_df, order(institution, mcid, term))
result_df <- result_df[order_index, , drop = FALSE]
```

```{r class.source = "dtbl"}
# data.table
result_dt <- term_dt[, ..columns_we_want]

order_cols <- c("institution", "mcid", "term")
setorderv(result_dt, order_cols)
```

```{r class.source = "dplyr"}
# dplyr
result_tbl <- term_tbl %>%
    select(institution, mcid, term, level, cip6) %>%
    arrange(institution, mcid, term)
```

```{r}
result_dt

same_content(result_dt, data.table(result_df))
same_content(result_dt, data.table(result_tbl))
```

*Comments*

- *base R.* Using `with()`, we can list the variables to order by without using the `$` operator. 

- *data.table.* `setorderv()` reorders data in place---hence no assignment operator `<-` is required---and it operates on a character vector of column names.  

- *dplyr.* `arrange()` orders the rows of a data frame by the values of selected columns. 






## References

<div id="refs"></div>






### Appendix

For future reference, the R and package version numbers (as of this writing) are: 

```{r}
R.version$version.string
packageVersion("midfielddata")
packageVersion("data.table")
packageVersion("dplyr")
packageVersion("tidyr")
```






```{r child = "../man/rmd/common-closing.Rmd"}
```
