---
title: "Case study: Goals"
vignette: >
  %\VignetteIndexEntry{Case study: Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-001-")
```


Part 1 of a case study in three parts, illustrating how we work with longitudinal student-level records.  

1. *Goals.* &nbsp; Introducing the study.   

2. *Data.* &nbsp; Transforming the data to yield the observations of interest.

3. *Results.* &nbsp; Summary statistics, metric, chart, and table.  






## Definitions

```{r child = "../man/rmd/define-student-level-data.Rmd"}
```

```{r child = "../man/rmd/define-stickiness.Rmd"}
``` 

: Stickiness is a more-inclusive alternative to graduation rate as a measure of a program's success in attracting, keeping, and graduating their undergraduates. Stickiness includes many students excluded by graduation rate such as part-time students, transfers, students admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

 




## Goals

The technical goal of the case study is to compare the program stickiness of Civil, Electrical, Industrial, and Mechanical Engineering programs with students grouped by program, race/ethnicity, and sex. 

The meta-goal of the case study is to illustrate a typical procedure for working with student-level data. 






*What the case study does*. &nbsp; We focus on our procedures and underlying rationale for working with student-level data. For reproducibility, we provide all the necessary code. 

*What the case study does not do*. &nbsp; To render the case study complete but brief, we omit the details of package syntax or functions, leaving such discussions to later articles. One can always use the R help system to read more about a data set or function.









## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
