---
title: "Subsetting MIDFIELD data"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Subsetting MIDFIELD data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-015-subsetting-data-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
```

## Introduction

In the previous vignette, we practiced gathering 6-digit CIP codes of programs we want to study and assigning a custom `program` label to each. 

In this vignette, we use the programs codes to subset the MIDFIELD data tables. The examples and practice exercises  illustrate a number of data manipulation methods that we regularly use in working with student unit record data. 





### Outline

- Subset by program code
- Subset by ID
- Merge program names

### In this vignette

midfieldr functions 

- `filter_match()` 

packages

```{r}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

data

```{r}
data(student, term, degree)
```




## Subset by program code

In these examples, we use the `study_program` data frame as our starting point to subset the MIDFIELD data tables. You can view its help page by running 

``` r 
? study_program
```

Review its structure, 

```{r}
study_program
```

To extract all the students in these programs who graduated, we subset rows of `degree` to keep values of `cip6` that match values of `cip6` in `study_program`.

```{r}
# subset degree rows
rows_we_want <- degree$cip6 %chin% study_program$cip6
dframe <- degree[rows_we_want]

# examine the result
dframe
```

We often want to select specific rows as well, e.g., 

```{r}
# subset degree columns
cols_we_want <- c("mcid", "institution", "cip6")
dframe <- dframe[, ..cols_we_want]

# examine the result
dframe
```

These two operations could have been combined, as in 

```{r}
# subset rows and columns 
rows_we_want <- degree$cip6 %chin% study_program$cip6
cols_we_want <- c("mcid", "institution", "cip6")
dframe <- degree[rows_we_want, ..cols_we_want]

# examine the result
dframe
```

This code chunk above represents a routine we encounter fairly often: subset rows of one data frame to keep values of a key variable that match values of the same key variable in a second data frame.  We use `filter_match()` to do the entire task:  subset rows of `degree` to keep values of `cip6` that match values of `cip6` in `study_program` and return three columns of the `degree` subset. 

```{r}
# subset degree table
case_degree <- filter_match(degree,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "institution", "cip6")
)

# compare to the DT we obtained above after ordering rows the same way
all.equal(case_degree[order(mcid)], dframe[order(mcid)])
```


If duplicate rows were a possibility and we want to delete them, `unique()` will do  the job. 

```{r}
# omit duplicate rows if any
case_degree <- unique(case_degree)

# examine the results
case_degree
```








## Subset by ID

Another common data access task is to retrieve information from `student` by ID. Suppose we wanted to examine the transfer status at the time of admission of the  students with degrees we subset above. We use `filter_match()` again to subset rows of `student` to keep values of `mcid` that match values of `mcid` in `case_degree` and return three columns of the `student` subset. 

```{r}
# subset student table
case_student <- filter_match(student,
  match_to = case_degree,
  by_col = "mcid",
  select = c("mcid", "transfer", "hours_transfer")
)
# examine the result
case_student
```

If the next step was to compare the numbers of students by their transfer status, we use `.N` from data.table. For example, 

```{r}
# omit any duplicate rows before counting
case_student <- unique(case_student)
case_student[, .N, by = "transfer"]
```








## Merge program names

Another common use of the `study_program` data is to merge its program names with another data frame by matching `cip6` values. The program names can then be used for grouping and summarizing. 

For example, let's join the program names to the case degree data frame we created above. The `all.x = TRUE` argument performs the merge as a "left outer join", retaining all rows of `case_degree` (the "left" data frame).  

```{r}
# join program names
case_degree <- merge(case_degree, study_program, by = "cip6", all.x = TRUE)

# examine the result
case_degree
```

If the next step was to group and summarize by program, we use `.N` again, for example,

```{r}
case_degree[, .N, by = "program"]
```


### Exercises 

Start by reading the  `exercise_program`  file you saved in the exercises in the previous vignette. Try `fread()`. You can view its help page by running, 

``` r 
? fread
```

To read the saved data from your `results` directory, 

```{r eval = FALSE}
exercise_program <- fread(
  "results/case_program.csv",
  colClasses = list(character = c("cip6"))
)
```

1. Use `filter_match()` to find all terms of all students ever enrolled in these programs by matching the `term` data table to `exercise_program` by the `cip6` column. Select the ID and CIP variables. Assign the result to the object `exercise_student`.
2. Eliminate duplicate rows using `unique()`. 
3. Merge the program names in `exercise_program` to the students in `exercise_student` by the CIP code. 
4. Summarize the number of students in these history programs by the `program` column. 

```{r echo = FALSE}
# copied from previous vignette
exercise_program <- wrapr::build_frame(
   "cip6"    , "program"           |
     "540101", "General History"   |
     "540102", "US History"        |
     "540103", "European History"  |
     "540104", "Science and Techn" |
     "540105", "Resource Admin"    |
     "540106", "Asian History"     |
     "540107", "Canadian History"  |
     "540108", "Military History"  |
     "540199", "General History"   )

# exercise solution
exercise_student <- filter_match(term,
  match_to = exercise_program,
  by_col = "cip6",
  select = c("mcid", "cip6")
)
exercise_student <- unique(exercise_student)
exercise_student <- merge(exercise_student,
  exercise_program,
  by = "cip6",
  all.x = TRUE
)
exercise_summary <- exercise_student[, .N, by = "program"]
```

```{r}
# answer
exercise_summary
```





## References

<div id="refs"></div>










## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval=FALSE}
# packages used
library("midfieldr")
library("data.table")


```
