---
title: "Case study---metrics"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study---metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
    fig.path = "../man/figures/art-003-case-metrics-", 
    fig.width = 6,
    fig.asp = 1 / 1.6,
    out.width = "70%",
    fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 3 of a case study in four parts:

- [Goals.](art-001-case-goals.html) Describing the study parameters.  
- [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.
- [Metrics.](art-003-case-metrics.html) Calculating summary statistics and metrics.
- [Displays.](art-004-case-displays.html) Creating tables and charts to display results.


## Preface

The next step in our case study is to ...  

If you are writing your own script(s) to follow along, you can continue the previous script or start a new script for this section. 

If starting a new script, we start with these packages and data. 

```{r}
# Packages
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# data.table printing options
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

The `study_reduced_data` data frame captures the state of the working data frame at the close of the "Data" part of the case study. It is bundled with midfieldr for the convenience of the tutorial user. 
