---
title: "Stickiness"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stickiness}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/body-and-society.csl
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here("man/figures", 
                                            "art-05-stickiness-"))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
```

## Introduction

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@Ohland+Orr+others:2012]. 

The stickiness metric accounts for all students in a program. Unlike the graduation rate metric, stickiness includes students who begin college part-time, who enroll mid-year, and who switch majors or transfer between institutions. 

Every student counted in a program's stickiness metric was enrolled in that program for at least one term. The program succeeded in attracting these students---stickiness tells us how well the program succeeded in retaining them.

### This vignette uses

midfieldr functions 
 
- `f()` 
- `f()` 
- `f()` 

packages

```{r}
 # packages used
library("midfieldr")
library("midfielddata")
library("data.table")
library("ggplot2")

# print max 20 rows, otherwise 5 rows each head/tail
options(datatable.print.nrows = 20, datatable.print.topn = 5)
```

## Under construction


## References

<div id="refs"></div>





## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script.

```{r eval=FALSE}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")
library("ggplot2")


```




