---
title: "Case study stickiness"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Case study stickiness}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
nocite: | 
resource_files: 
  - ../man/figures/art-130-case-study-stickiness-fig1-1.png
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-130-case-study-stickiness-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```

## Introduction

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program after accounting for the data sufficiency and timely completion  criteria  [@Ohland+Orr+others:2012]. 

The stickiness metric accounts for all students in a program, including  students who begin college part-time, enroll mid-year, switch majors, or transfer. Every student counted in a program's stickiness metric was enrolled in that program for at least one term. The program succeeded in attracting these students---stickiness tells us how well the program succeeded in retaining them.

### Outline

- Gather eligible students (start with `study_program`)
- Group, summarize, and join enrollees and graduates  
- Condition the findings
- Compute and graph stickiness (conclude with `study_stickiness`)

### This vignette uses

midfieldr functions 
 
- `filter_match()` 
- `add_timely_term()` 
- `add_data_sufficiency()` 
- `add_race_sex()` 
- `add_completion_timely()` 
- `condition_multiway()` 

packages

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))
library("ggplot2")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

data

```{r}
# load data tables from midfielddata
data(student, term, degree)
```








## Gather eligible students

We start with the case study program codes and names saved in the data set `study_program`, with which you are familiar from previous vignettes. You may review its help page by running

``` r
? study_program
```

Review the contents of the saved data:

```{r}
# Case study programs codes and names 
study_program
```

Use `filter_match()` to subset rows of `term` by values of `cip6` that match values of `cip6` in `study_program`. 

```{r}
# Get all students in case programs 
DT <- filter_match(term,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "institution", "cip6")
)

# Retain unique observations
DT <- unique(DT)

# Limit the study to degree-seeking students 
DT <- filter_match(DT,
  match_to = student,
  by_col = "mcid"
)
DT

# Apply the data sufficiency criterion 
DT <- add_timely_term(DT, midfield_term = term)
DT <- add_data_sufficiency(DT, midfield_term = term)
DT <- DT[data_sufficiency == TRUE]
DT
```



```{r}
# Add the grouping variables: program, race, and sex
DT <- merge(DT, study_program, by = "cip6", all.x = TRUE)
DT <- add_race_sex(DT, midfield_student = student)
DT
```


## Group, summarize, and join

Ever enrolled

```{r}
# Count ever enrolled
ever_enrolled <- DT[, .(ever = .N), by = .(program, race, sex)]
ever_enrolled
```

Graduates

```{r}
DT <- add_completion_timely(DT, midfield_degree = degree)

graduates <- DT[completion_timely == TRUE]
graduates



graduates <- graduates[, .(grad = .N), by = .(program, race, sex)]
graduates

```

Join 

```{r}
DT <- merge(ever_enrolled, 
            graduates, 
            by = c("program", "race", "sex"),
            all = TRUE)
DT
```



## Condition the results

```{r}
# Condition
DT[is.na(ever)]
DT[is.na(grad)]

DT <- DT[is.na(grad), grad := 0]

sort(unique(DT$ever))
sort(unique(DT$grad))
sort(unique(DT$race))
sort(unique(DT$sex))



DT <- DT[ever > 10]
DT <- DT[race %chin% c("Asian", "Black", "Hispanic/Latinx", "Native American", "White" )]



sort(unique(DT$ever))
sort(unique(DT$grad))
sort(unique(DT$race))
sort(unique(DT$sex))
```

## Compute stickiness

Expect 40 rows

```{r}
# stickiness
options(datatable.print.topn = 40)
DT[, stick := round(100 * grad / ever, 1)]
DT <- DT[order(program, sex, race)]



DT[, race_sex := paste(race, sex)]
DT <- DT[, .(program, race_sex, ever, grad, stick)]
DT
```

Save result for graph

```{r}
case_stickiness <- DT[, .(program, race_sex, stick)]
case_stickiness
```

Same as built-in data `study_stickiness`. 

```{r}
all.equal(case_stickiness, study_stickiness)
```



```{r}
# data preparation
DT <- condition_multiway(case_stickiness)
DT_med <- condition_multiway(DT, details = TRUE)
```

```{r echo = FALSE}
asp_ratio <- asp_ratio_mw(DT_med, 
                                categories = c("program", "race_sex"))
```


## Graph the results

```{r fig1, fig.asp = asp_ratio[1]}
# creating a multiway graph
ggplot(data = DT_med, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = med_program),
    linetype = 2
  ) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Stickiness",
    y = "",
    title = "Practice data (not for research)",
    caption = "Source: midfielddata"
  )
```


## References

<div id="refs"></div>










## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script.

```{r eval=FALSE}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")
library("ggplot2")
```




