---
title: "Case study programs"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Case study programs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
source("vignettes/article-preamble.R")
set_fig_path("art-002-case-study-programs-")
```

## Introduction

The case study is a condensed tour of a typical workflow with student unit record data. We provide the R code with minimum explanation and focus instead on the overall logic of the process. Other vignettes explore the details of the midfieldr functions.  

Assuming you completed the introduction to the four data tables (in [Getting started](art-000-getting-started.html)), the case study has four parts:

- *Case study programs* &#9664; You are here.
- [Case study students](art-004-case-study-students.html) 
- [Case study stickiness](art-006-case-study-stickiness.html)
- [Case study graduation rate](art-006-case-study-grad-rate.html)

Longer vignettes with detailed development of methods and functions related to this part of the study include:   

- [Program codes and names](art-010-program-codes.html){target="_blank"} 
- [Subsetting MIDFIELD data](art-015-subsetting-midfield-data.html){target="_blank"} 


### In this vignette

In studying student records, a common first step is to select a specific set of programs to explore. In this vignette, we start a case study of four engineering programs in which we: 

- Identify their CIP codes   
- Assign custom program names
- Save the results

We use the following midfieldr functions
 
- [`filter_search()`](../reference/filter_search.html)

The vignette output is the data frame `study_program`. 






## Identify program codes

If you are writing your own script to follow along, we start with these packages: 

```{r}
# packages used
library("midfieldr")
suppressMessages(library("data.table"))

# optional code to control data.table printing
options(datatable.print.nrows = 10,
        datatable.print.topn = 5,
        datatable.print.class = TRUE)
```

midfieldr includes `cip`,  a data table of academic program codes and names. View its help page by running

```r
# type in the Console
? cip
```

We know the conventional names of the desired programs (Civil, Electrical, Industrial, and Mechanical Engineering) but these are not necessarily the exact program names used in the CIP. Thus we have to search CIP thoroughly to make sure we find all 6-digit CIP codes associated with these disciplines.

We start with a text search of `cip` for `engineering` using `filter_search()`. 

```{r}
# subset rows of the CIP data matching conditions
pass01 <- filter_search(cip, "engineering")

# examine the result
pass01
```

We have a number of 2-digit codes reported. Let's examine those more closely using data.table syntax for selecting columns. 

```{r}
cols_we_want <- c("cip2", "cip2name")
unique(pass01[, ..cols_we_want])
```

Only one of these (14) will have the programs we are looking for. So in our second pass, we apply the filter to the results of the first pass and restrict our search to codes starting with 14. 

```{r}
# subset the previous pass
pass02 <- filter_search(pass01, "^14")

# examine the result
pass02
```

In the next pass, applying the filter to the results of the second pass, let's try restricting our search to the four conventional terms: civil, electrical, industrial, and mechanical. The search is not case-sensitive. 

```{r}
# optional code to control data.table printing
options(datatable.print.topn = 13)

# subset the previous pass
pass03 <- filter_search(pass02, c("civil", "electrical", "industrial", "mechanical"))

# examine the result
cols_we_want <- c("cip6", "cip6name")
pass03[, ..cols_we_want]
```

In this result,  only Electromechanical Engineering is superfluous, so we can drop it in the next pass. 

```{r}
# subset the previous pass
pass04 <- filter_search(pass03, drop_text = "electromechanical")

# examine the result
pass04
```

We see that 

- the 4-digit level describes these programs with the customary names  
- Civil Engineering encompasses six programs at the 6-digit level
- Electrical Engineering encompasses four programs at the 6-digit level
- Mechanical and Industrial have one program each at the 6-digit level

All MIDFIELD data tables that have a column of CIP codes use 6-digit codes. Therefore, for a case study, we only need the `cip6` column from the work above. The names we want (in this case), however, are at the 4-digit level. We can select the columns we need. 

```{r}
# select columns from the previous pass
cols_we_want <- c("cip6", "cip4name")
case_cip <- pass04[, ..cols_we_want]

# examine the result
case_cip
```


## Assign program names

As the study proceeds, we will want to compare and contrast the programs by name. The names in `cip4name`  are close to what we want, but all can be shortened for later reporting in graphs and tables. 

We use the data.table `fcase()` function and `%ilike%` to add a `program` column with values that depend on the values in the `cip4name` column. 

View their help pages by running

``` r
# type in the Console
? `%ilike%`

? fcase
```

```{r}
# create a new memory location to avoid updating by reference
case_program <- copy(case_cip)

# add a new column for custom program labels
case_program[, program := fcase(
  cip4name %ilike% "civil", "Civil",
  cip4name %ilike% "electrical", "Electrical",
  cip4name %ilike% "mechanical", "Mechanical",
  cip4name %ilike% "industrial", "Industrial"
)]

# examine the result
case_program
```

In subsequent work we need only the `cip6` and `program` values, so we can drop the `cip4name` column. 

```{r}
# drop an unnecessary column
case_program[, cip4name := NULL]

# examine the result
case_program
```


## Save results

We save the `case_program` data frame as a CSV file to the `results` directory for easy access as we work through the study. Here we use fast write `fwrite()`  from data.table. You may prefer other read/write functions such as those in base R or packages readr or rio. Note that relative file path shown assumes you are working within an RStudio project and that you have a `results` subdirectory (as described in [Getting started](art-00-midfield-data.html)).  

``` r
# Assumes you have a results folder in the project directory
fwrite(case_program, file = "results/case_program.csv")
```

We suggest saving intermediate results like these at each significant milestone of a study, especially after steps requiring a lot of computing time.

These results have been included with midfieldr as the built-in data set `study_program`. View its help page by running 

```r
# type in the Console
? study_program
```
View the data frame, 

```{r}
study_program
```

Check that the two data frames are identical, 

```{r}
all.equal(case_program, study_program)
```

In subsequent vignettes, you can read in the `case_program` data from your `results` directory or you can use the `study_program` data provided with midfieldr. 


## Related articles

- [Program codes and names](art-010-program-codes.html) for practice working with CIP codes and assigning program names. 








## References

<div id="refs"></div>






## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# packages used
library("midfieldr")
suppressMessages(library("data.table"))

# identify program codes
pass01 <- filter_search(cip, "engineering")
pass02 <- filter_search(pass01, "^14")
pass03 <- filter_search(pass02, c("civil", "electrical", "industrial", "mechanical"))
pass04 <- filter_search(pass03, drop_text = "electromechanical")
cols_we_want <- c("cip6", "cip4name")
case_cip <- pass04[, ..cols_we_want]

# assign program names
case_program <- copy(case_cip)
dframe <- copy(case_program)
case_program[, program := fcase(
  cip4name %ilike% "civil", "Civil",
  cip4name %ilike% "electrical", "Electrical",
  cip4name %ilike% "mechanical", "Mechanical",
  cip4name %ilike% "industrial", "Industrial"
)]
case_program[, cip4name := NULL]

# save results
fwrite(case_program, file = "results/case_program.csv")
```
