---
title: "Data sufficiency"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Data sufficiency}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: 
  - ../man/figures/art-020-data-sufficiency-fig1-1.png
  - ../man/figures/art-020-data-sufficiency-fig2-1.png
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-020-data-sufficiency-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```






## Introduction

The data provided by institutions to MIDFIELD cover a finite span of years, with different starting and ending terms for each institution.

For students admitted too near the upper limit of their institution's data range, the  available data cover an insufficient number of years to fairly assess student  performance. Only students for whom the data range is sufficient should be included in any analysis. 

In this vignette, we introduce two concepts and their related midfieldr functions:

timely completion (TC) term 
: The last term in which program completion would be considered timely for a given student. For example, the TC term in many cases is defined as the admission term plus 6 years.     

data sufficiency criterion
: A study must be limited to students whose timely completion term is within the range of data provided by their institution. 

In the figure we compare two first-time-in-college students admitted in different terms, both with 6-year spans during which program completion would be considered timely. 

- Student A enters in Fall 1990 with a timely completion (TC) term of  Spring 1995. 
- Student B enters in Fall 1992 with a TC term of Spring 1997. 
- Institution data is available from 1986 to 1996.




<br>
```{r fig1, echo = FALSE, fig.asp = 5/10}
library("data.table")
library("ggplot2")

# parameters
callout_color <- "gray60"
callout_line_size <- 0.3
anno_size <- 3.5 # 3.5 approx 10 point
vert_baseline <- 1.07

# construct coordinates
y <- c(1.4, 1.75, 2.0)
y_lim <- c(min(y) - 0.3, max(y) + 0.1)
x <- c(1986, 1990, 1992, 1995, 1996, 1997)
x_lim <- c(min(x) - 0.5, max(x) + 1.5)

# arrows
df_arrow <- wrapr::build_frame(
  "x", "x_end", "y" |
    x[1], x[5], y[3] |
    x[2], x[4], y[2] |
    x[3], x[6], y[1]
)
setDT(df_arrow)
df_arrow[, x_mid := (x + x_end) / 2]

# dimension lines
del <- 0.03
dim_lines <- wrapr::build_frame(
  "x", "y", "y_end" |
    x[1], y[3], y[3] |
    x[5], y[1], y[3] |
    x[2], y[2], y[2] |
    x[4], y[2], y[2] |
    x[3], y[1], y[1] |
    x[6], y[1], y[1]
)
setDT(dim_lines)
dim_lines[, `:=`(y = y - 1.5 * del, y_end = y_end + del)]
dim_lines[x == 1996, y := y - 4 * del]

# graph
ggplot(data = df_arrow) +
  scale_x_continuous(
    limits = x_lim,
    breaks = x
  ) +
  scale_y_continuous(limits = y_lim) +
  labs(
    y = "",
    x = "Year",
    title = "Illustrating data sufficiency"
  ) +
  theme_light() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  # horizontal arrows
  geom_segment(aes(x = x, xend = x_end, y = y, yend = y),
    color = callout_color,
    size = callout_line_size,
    arrow = arrow(
      type = "closed",
      ends = c("both", "both", "both"), # both, first, last
      length = unit(c(2, 2, 2), "mm") # "0" length = no arrow head
    ),
    arrow.fill = callout_color
  ) +
  # dimension lines
  geom_segment(
    data = dim_lines,
    aes(x = x, xend = x, y = y, yend = y_end),
    color = callout_color,
    size = callout_line_size,
    linetype = 1
  ) +
  # student labels
  annotate("text",
    x = df_arrow$x[2] - 0.7,
    y = df_arrow$y[c(2, 3)],
    size = anno_size,
    label = c("student A:", "student B:"),
    vjust = 0.5,
    hjust = 1
  ) +
  # label arrows
  annotate("text",
    x = df_arrow$x_mid,
    y = df_arrow$y,
    size = anno_size,
    label = c("institution data range", "timely span", "timely span"),
    vjust = -0.5
  ) +
  # label dimension lines
  annotate("text",
    x = dim_lines$x,
    y = dim_lines$y,
    size = anno_size,
    label = c(
      "",
      "data limit",
      "enter",
      "TC term",
      "enter",
      "TC term"
    ),
    vjust = 1,
    hjust = c(-0.1, 0.5, -0.1, 1.1, -0.1, -0.1)
  )
```
<br>

The TC term for Student B is not within the range of data provided by their institution. By the data sufficiency criterion, we exclude student B from the study.

The criterion applies even if Student B were known to have graduated within the data limit. Other students like Student B will graduate after the data limit, but we cannot know if their program completion was timely or not. Because such students must be excluded from the study, so must every student whose TC term is not encompassed by the available data.  

### Outline

- Start with the case study programs (`study_program`) 
- Obtain all students ever enrolled in these programs 
- Determine their timely completion terms 
- Determine data sufficiency by institution and student 
- Filter observations for data sufficiency 
- Result is set of students for further analysis (`study_student`)

### This vignette uses

midfieldr functions 
 
- `add_data_sufficiency()`
- `add_institution()`
- `add_timely_term()` 
- `filter_match()` 


packages

```{r}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# optional code to format data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

data

```{r}
# load data tables from midfielddata
data(student, term, degree)
```
   





## Case study students

Recall that the case study program codes and names are saved in the midfieldr package as the data set `study_program`. View its help page by running

```r
? study_program
```

View the data frame, 

```{r}
study_program
```

To obtain the students enrolled in the case study programs, we use `filter_match()` to access the `term` table. 

```{r}
case_students <- filter_match(term,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "cip6")
)
case_students
```

Delete duplicate rows, if any, 

```{r}
case_students <- unique(case_students)
case_students
```






## Estimate the timely completion term 

A necessary prerequisite for applying the data sufficiency criterion is to determine a timely completion term for every prospective student in a study. To confirm that the pool of students are all degree-seeking, we use `filter_match()` to match to the IDs in the `student` table. 

```{r}
# limit population to degree-seeking students
DT <- filter_match(case_students,
  match_to = student,
  by_col = "mcid"
)

# examine the result
DT <- unique(DT)
DT
```

We use `add_timely_term()` to estimate the timely completion term. View its help page by running

```r
? add_timely_term
```

As the function name suggests, `add_timely_term()` adds the `timely_term` column to the input data frame. The new variable contains the estimated timely completion term for each student. The first argument is your working data frame to which the column is to be added; the second argument is `term` (or your equivalent term-variables data frame). Both are required. 

```{r}
# estimate the timely completion term
DT <- add_timely_term(DT, midfield_term = term)
DT
```

The basic heuristic starts with span number of years for each student (default 6 years) and adjusts the span by subtracting a whole number of years based on the level at which the student is admitted.

For example, a student admitted at the second-year level is assumed to have completed one year of a program, so their span is reduced by one year. Similarly, spans are reduced by two years for students admitted at the 3rd-year level and by three years for students admitted at the fourth-year level. The adjusted span of years added to their starting term produces timely completion term.

Optional arguments are `details`, `span`, and `sched_span`. When used, these arguments must be named. 

```r
add_timely_term(dframe,
                midfield_term,
                ...,
                details = NULL,     # default FALSE
                span = NULL,        # default 6 years
                sched_span = NULL)  # default 4 years
```

- `span` must be no less than `sched_span`. 
- `sched_span` is the number of years the institution schedules for program completion, typically 4 years. 
- `details`, when TRUE, provides additional columns (described next) 

Setting `details` to TRUE yields additional columns that were used to determine the timely-term results: 

- `term_i` admission term 
- `level_i` level upon being admitted 
- `adj_span` adjusted span 

The potential use of these data are to implement one's own heuristic for determining timely completion terms. 

```{r}
# show details
DT <- add_timely_term(DT,
  midfield_term = term,
  details = TRUE
)
DT
```
 
If the input data frame has an existing column name matching any of the added column names, the existing columns are deleted. For example, having examined the details, you can delete them (especially useful in an interactive session) by running the same code with details set to FALSE. 

```{r}
# remove details
DT <- add_timely_term(DT,
  midfield_term = term,
  details = FALSE
)
DT
```


### Exercises


If we start with a data frame with details shown 

```{r}
x <- add_timely_term(DT, term, details = TRUE)
```

then the following code chunk, 

```r
x[adj_span == 6, .SD[1], by = cip6]
```
subsets `x` as follows: 

- `adj_span == 6` subsets rows for a 6 year span
- `by = cip6` groups the data by the four CIP codes
- `.SD[1]` retains the first row of each group

with the resulting 4 rows (one for each CIP): 

```{r echo = FALSE}
x[adj_span == 6, .SD[1], by = cip6]
```

1. Recall (from the `term` help page) that all terms of an academic year are encoded with the same YYYY value. Choose any row. Confirm that a student starting in the initial term shown and graduating a full 6 academic years later would graduate in the term reported in the timely-term column. 
2. Reuse the code chunk but change the adjusted span to 4 or 5 years. Select one row and repeat the assessment. 





## Determine data sufficiency 

Data sufficiency depends on the range of data from an institution, so our first step is to add a column for institution keyed by the student ID. We use the `add_institution()`. View its help page by running 

```r
? add_institution
```
The function has two arguments, the data frame to which the new variable is added and the `term` MIDFIELD data table (or equivalent). In the event that a student were admitted to more than one institution over their history in the data, the function reports the institution at which the student was enrolled for the greatest number of terms.  

```{r}
# add institutions
DT <- add_institution(DT,
  midfield_term = term
)

# examine the result
DT
```

Like  the other midfieldr `add_` functions, existing columns are deleted or overwritten if they have the same name as the variables to be added (here, `institution`). 

Now that our working data frame has `timely_term` and `institution` columns, we can use `add_data_sufficiency()` to add a logical variable (TRUE/FALSE) that answers the question, "Is the data sufficiency constraint satisfied?" View its help page by running, 

```r
? add_data_sufficiency

```

Similar to `add_timely_term()`, this function has two required arguments and one optional `details` argument. The details, in this case, add one extra column to show the institution's upper data limit. 

```{r}
# add column with details
DT <- add_data_sufficiency(DT,
  midfield_term = term,
  details = TRUE
)
# examine the result
DT
```

With details FALSE, only the `data_sufficieny` column is added.

```{r}
# without details
DT <- add_data_sufficiency(DT,
  midfield_term = term,
  details = FALSE
)
# examine the result
DT
```



### Exercises

If we start with the `term` data table, select two columns, and order its rows by institution and term, 

```{r}
# select columns from term
cols_we_want <- c("institution", "term")
x <- term[order(institution, term), ..cols_we_want]
```

then the following code chunk, 

```r
# optional code to format data.table printing
options(datatable.print.nrows = 12)

# filter for last term grouped by instititution
x[, .SD[.N], by = institution]
```
subsets `x` as follows: 

- `by = institution` groups the data by the 12 institutions
- `.SD[.N]` retains the last row of each group (having previously ordered the rows by term within each institution)

yielding a data frame (not shown) giving the last term in the data for each institution. Obtain the data frame for yourself by running the code chunks above. Then run:

```{r include = FALSE}
options(datatable.print.nrows = 12)
x[, .SD[.N], by = institution]
```

```{r}
y <- add_data_sufficiency(DT,
  midfield_term = term,
  details = TRUE
)
```

Filter for one institution using

```{r results = 'hide'}
y[institution == "Institution A", .SD[1:5]]
```

where 

- `institution == "Institution A"` filters for Institution A only 
- `.SD[1:5]` pulls out the first 5 rows, 

1. Confirm that the last term for Institution A found earlier is a match to the `inst_limit` reported by `add_data_sufficiency()`. 
2. Repeat the last code chunk for another institution and confirm the result. 




## Subset data for data sufficiency

Observations (rows) satisfying the data sufficiency criterion are labeled TRUE in the `data_sufficiency` column; those that do not are labeled FALSE. Limiting the population to students satisfying the criterion is straightforward---we retain the TRUE data sufficiency rows. 

```{r}
# limit population to data sufficient
DT <- DT[data_sufficiency == TRUE]

# examine the result
DT
```

Having performed the subset, the working columns we used to get to this point can be deleted. We retain the two columns with which we started this vignette, `mcid` and `cip6`. 

```{r}
DT <- DT[, .(mcid, cip6)]
```

Again, we suggest saving intermediate results like these. This data frame of IDs and CIP codes is saved in midfieldr as the data set `study_student`. View its help page by running

```r
? study_student
```
View the data frame, 

```{r}
study_student
```

Confirm the two are the same.

```{r}
DT <- DT[order(mcid)]
study_student <- study_student[order(mcid)]
all.equal(DT, study_student)
```


## Conclusion

The data sufficiency criterion must be applied whenever a study incorporates program completion (graduation) as part of a metric. Different heuristics might be used to determine the term of timely completion (discussed in a separate vignette), but once the timely term is determined, the sufficiency of the data available immediately determines whether the student remains in the study.








## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# load data tables from midfielddata
data(student, term, degree)

# Case study students
case_students <- filter_match(term,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "cip6")
)
case_students <- unique(case_students)

# Estimate the timely completion term
DT <- filter_match(case_students,
  match_to = student,
  by_col = "mcid"
)
DT <- unique(DT)
DT <- add_timely_term(DT, midfield_term = term)

# Determine data sufficiency
DT <- add_institution(DT, midfield_term = term)
DT <- add_data_sufficiency(DT,
  midfield_term = term,
  details = TRUE
)

# Subset data for data sufficiency
DT <- DT[data_sufficiency == TRUE]
DT <- DT[, .(mcid, cip6)]
```


