---
title: "Program codes and names"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Program codes and names}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-010-program-codes-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
```

## Introduction

In the US, instructional programs are encoded by "CIP" codes. CIP is the acronym for *Classification of Instructional Programs*, a taxonomy of academic programs curated by the US Department of Education [@NCES:2010]. 
The 2010 CIP codes are included with midfieldr in the data set `cip`. 

The IPEDS standard encoding format is a two-digit number followed by a period, followed by a four-digit number, for example, 14.0102 [@NCES:2020]. MIDFIELD uses the same numerals, but omits the period, i.e., 140102, and stores the variable as a  character string. 


In this vignette, we practice strategies of searching `cip` for programs we want to study. 

### Outline

- Explore the taxonomy of academic programs 
- Inspect the CIP data 
- Practice searching  the CIP data using different approaches 
- Assigning program names
- Save results

### This vignette uses

midfieldr functions 

- `filter_search()` 

packages

```{r}
# packages used
library("midfieldr")
library("data.table")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```



## Taxonomy of programs

Academic programs have three levels of codes and names:

- 6-digit code, a specific program
- 4-digit code, a group of 6-digit programs of comparable content 
- 2-digit code, a grouping of 4-digit groups of related content 

Loosely, the 6-digit level can be thought of as a specialty within a discipline, the 4-digit level is often equivalent to an academic department, and the 2-digit level is roughly equivalent to an academic college. For example, Geotechnical Engineering (140802) is a specialty in  Civil Engineering (1408) which is a department in the college of Engineering (14). 

```{r echo = FALSE}
df <- filter_search(cip, "^41")
n41 <- nrow(df)
n4102 <- nrow(filter_search(df, "^4102"))
n4103 <- nrow(filter_search(df, "^4103"))
name41 <- unique(df[, cip2name])

df24 <- filter_search(cip, "^24")
n24 <- nrow(df24)
name24 <- unique(df24[, cip2name])

df51 <- filter_search(cip, "^51")
n51 <- nrow(df51)
name51 <- unique(df51[, cip2name])

df1313 <- filter_search(cip, "^1313")
n1313 <- nrow(df1313)
name1313 <- unique(df1313[, cip2name])

cip <- filter_search(cip) # removes data.frame class if present
```

In the table below, we show all programs in the broad area of "`r name41`" (code 41) to illustrate the relationships among 2-, 4-, and 6-digit program codes and names in more detail. This 2-digit grouping is subdivided into `r length(unique(df[, cip4]))` groups at the 4-digit level (codes 4100--4199) which are further subdivided into `r n41` programs at the 6-digit level (codes 410000--419999). 

```{r echo=FALSE}
sub_cip <- filter_search(cip, "^41")
kable2html(sub_cip)
```
<br>

The number of 6-digit programs in a 4-digit group varies by program. Four-digit groupings include anywhere from 1 program (code 4100 above for example) to `r n1313` programs (code 1313 `r name1313`). 

Similarly, 2-digit groupings include anywhere from four 4-digit programs  (code 24 `r name24`) to `r n51` programs (code 51 `r name51`).

## Inspect the `cip` data  

Data set that loads with midfieldr. For definitions of variables, view the help page by running 

```r
? cip
```

To see the data structure, run the following. Note that all variables are character strings, protect the leading zeros on some of the codes.  

```{r}
str(cip)
```

To see a few of the observations, run 

```{r}
cip
```

The number of unique programs.  

```{r}
# at the 2-digit level
sort(unique(cip$cip2))

# at the 4-digit level
sort(unique(cip$cip4))

# at the 6-digit level
length(unique(cip$cip6))
```

A sample of program names (`sample()` uses a random number generator, so your result will differ from that shown)

```{r echo = FALSE}
set.seed(20210613)
```

```{r}
some_programs <- cip[, cip4name, drop = FALSE]
sample(some_programs, 20)
```

```{r echo = FALSE}
set.seed(NULL)
```

## Filter using a keyword search

`filter_search()` is the midfieldr function for searching the `cip` data set for program codes and names. View the help page by running

```r
? filter_search()
```

The variables in `cip` are all characters and `filter_search()` subsets a data frame using character search terms. The arguments are: 

- `dframe` data frame to be subset
- `keep_text` character vector of search text for retaining rows, not case-sensitive 
- `drop_text` (optional) character vector of search text for dropping rows
- `select` (optional) character vector of column names to search and return, default all columns  

For example, filtering the CIP data for all programs containing the word "engineering" yields `r nrow(filter_search(cip, "engineering"))` observations. 

```{r}
# filter basics
filter_search(dframe = cip, keep_text = "engineering")
```

The first two arguments in `filter_search()` do not have to be named explicitly, but the  optional `drop_text` and `select` arguments do.

```{r}
# applying optional argument drop_text and select
filter_search(cip,
  "engineering",
  drop_text = c(
    "related",
    "technology",
    "technologies"
  ),
  select = c("cip6", "cip6name")
)
```

Columns in `select`  are subset after filtering for `keep_text` and `drop_text`. Thus a repeat search with a different column selection should return the same rows but different columns. 

```{r}
# repeat search with different columns selected
filter_search(cip,
  "engineering",
  drop_text = c(
    "related",
    "technology",
    "technologies"
  ),
  select = c("cip4", "cip4name")
)
```


Suppose we want to find the CIP codes and names for all programs in Civil Engineering. The search is insensitive to case, so we start with the following code chunk. 

```{r}
# example 1 filter using keywords
filter_search(cip, "civil")
```

The same information is tabulated below to help you see the structure.  In the examples that follow, we show a table like this one instead of the R output. 

```{r echo = FALSE}
sub_cip <- filter_search(cip, "civil")
# using kable_styling() for output but conceal from novice user
kable2html(sub_cip)
```
<br>

The search returns some programs with Civilization in their names as well as Engineering Technology. If we wanted Civil Engineering only, we can use a sequence of function calls, where the outcome of the one operation is assigned to the first argument of the next operation.  

The following code chunk could be read as, "Start with the `cip` data frame, then keep any rows in which 'civil' is detected, then keep any rows in which 'engineering' is detected, then drop any rows in which 'technology' is detected."

```{r results = "hide"}
# first search
first_pass <- filter_search(cip, "civil")

# refine the search
second_pass <- filter_search(first_pass, "engineering")

# refine further
third_pass <- filter_search(second_pass, drop_text = "technology")
```

```{r echo = FALSE}
kable2html(third_pass)
```
<br>

Seeing that all Civil Engineering programs have the same `cip4name`, we could have used 
`keep_text = c("civil engineering")` to narrow the search to rows that match the full phrase. For example, 

```{r results = 'hide'}
filter_search(cip, "civil engineering", drop_text = "technology")
```

produces the same result reported in the table above. 

### Exercises

1. Determine the set of 6-digit CIP codes you would expect to be used by a College of Business. 
2. The social sciences 2-digit CIP does not necessarily include all possible social science programs. Let anthropology, economics, geography, linguistics, political science, jurisprudence, psychology, and sociology be defined a social sciences. Determine the set of 6-digit codes we would collect to study these programs. 

## Filter using a numerical code search

Suppose we want to study programs relating to German culture, language, and literature. Using  "german" for the `keep_text` value yields 

```{r results = "hide"}
# example 2 filter using numerical codes
filter_search(cip, "german")
```

```{r echo = FALSE}
sub_cip <- filter_search(cip, "german")
kable2html(sub_cip)
```
<br>

From the 6-digit program names we find only two that are of interest, German Studies (050125) and German Language and Literature (160501). We use a character vector to assign these two codes to the `keep_text` argument. 
 
```r
# character vector of search terms 
filter_search(cip, c("050125", "160501"))
```

```{r echo = FALSE}
sub_cip <- filter_search(cip, c("050125", "160501"))
kable2html(sub_cip)
```
<br>

If the 6-digit codes are entered as integers, they produce an error. 
 
```{r error = TRUE, purl = FALSE}
filter_search(cip, c(050125, 160501))
```

### Exercise

1. In the Social sciences exercise in the previous section, searching for  "jurisprudence" produced two codes: "220203" and "220204." Search `cip` for these two codes and determine the 2-digit program to which they belong. 
2. Do an initial search on the term, "history." From the results, 
    - Identify the 2-digit CIP for history.   
    - Use a numerical search for programs that start with that 2-digit CIP. 
    - Subset for the 6-digit CIPs and 6-digit CIP names.  
    - Assign the result to the object name `exercise_cip`.
    - Answer given below, check your work. 

```{r include = FALSE}
# partial solution to exercise
pass01 <- filter_search(cip, "history")
pass02 <- filter_search(pass01, "^54")
cols_we_want <- c("cip6", "cip6name")
exercise_cip <- pass02[, ..cols_we_want]
```


```{r}
# Answer to exercise 2.
exercise_cip
```


## Filter using a regular expression search

Specifying 4-digit codes yields a data frame all 6-digit programs containing the 4-digit string. We use the regular expression notation `^` to match the start of the strings.

```r
# example 3 filter using regular expressions
filter_search(cip, c("^1407", "^1408"))
```

```{r echo = FALSE}
sub_cip <- filter_search(cip, c("^1407", "^1408"))
kable2html(sub_cip)
```
<br>

The 2-digit series represent the most general groupings of related programs. Here, the result includes all History programs. 

```r
# 2-digit example
filter_search(cip, "^54")
```

```{r echo = FALSE}
sub_cip <- filter_search(cip, "^54")
kable2html(sub_cip)
```
<br>

The series argument can include any combination of 2, 4, and 6-digit codes. It can also be passed to the function as a character vector. 

```{r results = "hide"}
# character vector of search terms
codes_we_want <- c("^24", "^4102", "^450202")

# a series with 2, 4, and 6-digits specified
filter_search(cip, codes_we_want)
```

```{r echo = FALSE}
sub_cip <- filter_search(cip, codes_we_want)
kable2html(sub_cip)
```
<br>

## When search terms cannot be found

If the `keep_text` argument includes terms that cannot be found in the CIP data frame, the unsuccessful terms are identified in a message and the successful terms produce the usual output. 

For example, the following `keep_text` argument includes three search terms that are not present in the CIP data ("111111", "^55", and "Bogus") and two that are ("050125" and "160501").  

```{r message = TRUE}
# unsuccessful terms produce a message
sub_cip <- filter_search(cip, c("050125", "111111", "160501", "Bogus", "^55"))

# but the successful terms are returned
sub_cip
```

However, as seen earlier, if none of the search terms are found, an error occurs. 

```{r error = TRUE, purl = FALSE}
filter_search(cip, c("111111", "Bogus", "^55"))
```


### Exercise

1. Determine the set of programs having the same 4-digit codes as "220203" and "220204". 

## CIP data from another source

If you use a CIP data set from another source, it must have the same structure as `cip`: six character columns named as follows,  

```{r}
# name and class of variables (columns) in cip
unlist(lapply(cip, FUN = class))
```






## Assigning program names

You may have noticed that the program names in CIP tend to be verbose. In addition, some groups of programs at a particular institution may be grouped differently than they are in the CIP structure, e.g., Liberal Arts, Humanities, or Arts and Sciences. 

We have always found it useful, therefore, to add a `program` column to our data frame of CIP codes with a customized label for a group of programs. 

We use our case study group of four engineering programs to illustrate several methods for adding a `program` column to the CIP data frame. 

First, assume we've found the 4-digit codes for the four engineering programs: Civil (1408), Electrical (1410), Mechanical (1419), and Industrial (1435). We obtain their CIP codes and names, 

```{r}
options(datatable.print.nrows = 12)

# Four engineering programs
DT <- filter_search(cip, c("^1408", "^1410", "^1419", "^1435"))

# Retain the needed columns
cols_we_want <- c("cip6", "cip4name")
DT <- DT[, ..cols_we_want]

# Examine the result
DT
```

Our approach here will be to assign a new `program` column with values from the `cip4name` column, then edit the new column values


```{r}
# Assign a new column 
DT[, program := cip4name]

# Examine the result
DT
```

### 1. Use `program %like%` to edit one value

The `%like%` function is essentially a wrapper function around the base R `grepl()` function. See its help page by running (the back-ticks facilitate a help search for terms starting with a symbol):

``` r
? `%like%`
```

In this approach, we search for one distinctive term only.

```{r}
# return matches in the cip4name column
temp <- copy(DT)
rows_to_edit <- temp$program %ilike% "electrical"
temp[rows_to_edit, program := "Electrical Engineering"]
temp
```


### 2. Use `cip6 %like%` to edit one value

In our second approach, we use the `%like%` function again, but apply it to the CIP codes. Here we use the regular expression `^1410` meaning "starts with 1410." 

```{r}
# return matches that start with 1410
temp <- copy(DT)
rows_to_edit <- temp$cip6 %like% "^1410"
temp[rows_to_edit, program := "Electrical Engineering"]
temp
```

### 3. Use `program := fcase()` to edit all values 

In this approach, we use the data.table `fcase()` function to change all of the program names to one word labels. The `%ilike%` variant of `%like%` sets `ignore.case = TRUE`.

```{r}
DT[, program := fcase(
  program %ilike% "civil", "Civil",
  program %ilike% "electrical", "Electrical",
  program %ilike% "mechanical", "Mechanical",
  program %ilike% "industrial", "Industrial"
)]
DT
```

Having completed labeling the programs, we can remove the original verbose `cip4nme` column. If needed, we can always recover these names using the values in the `cip6` column. 

```{r}
DT[, cip4name := NULL]

DT
```



### Exercises

1. Assign a copy of `exercise_cip` (from the exercise in the previous section) to the name `exercise_program`. 
2. Add a `program` column to `exercise_program` with the initial values the same as `cip6name`. 
3. Edit some of the values in the `program` column as follows:
    - Rename both "General" and "Other" to "General History" 
    - Shorten "History and Philosophy..." to "Science and Techn" 
    - Shorten "American History ..." to "US History"
    - Shorten "Public Applied ..." to "Resource Admin"
3. Delete the `cip6name` column. 
4. Answer given below to check your work. 

```{r echo = FALSE}
# exercise solution
exercise_program <- copy(exercise_cip)
exercise_program[, program := cip6name]
exercise_program[
  cip6name %ilike% "General",
  program := "General History"
]
exercise_program[
  cip6name %ilike% "Other",
  program := "General History"
]
exercise_program[
  cip6name %ilike% "Philosophy",
  program := "Science and Techn"
]
exercise_program[
  cip6name %ilike% "Public",
  program := "Resource Admin"
]
exercise_program[
  cip6name %ilike% "United States",
  program := "US History"
]
exercise_program[, cip6name := NULL]

# cat(wrapr::draw_frame(exercise_program))
```

```{r}
# Answer
exercise_program
```





```{r include = FALSE}
# work out the exercises
library("midfieldr")
library("midfielddata")
library("data.table")
data(student, term)

# Identify program codes
pass01 <- filter_search(cip, "history")
pass02 <- filter_search(pass01, "^54")
cols_we_want <- c("cip6", "cip6name")
exercise_cip <- pass02[, ..cols_we_want]
exercise_cip

# Assign program names
exercise_program <- copy(exercise_cip)
exercise_program[, program := cip6name]
exercise_program[
  cip6name %ilike% "General",
  program := "General History"
]
exercise_program[
  cip6name %ilike% "Other",
  program := "General History"
]
exercise_program[
  cip6name %ilike% "Philosophy",
  program := "Science and Technology"
]
exercise_program[
  cip6name %ilike% "Public",
  program := "Historical Resources"
]
exercise_program[
  cip6name %ilike% "United States",
  program := "US History"
]
exercise_program[, cip6name := NULL]
exercise_program

# Subset by CIP
exercise_student <- filter_match(term,
  match_to = exercise_program,
  by_col = "cip6",
  select = c("mcid", "cip6")
)
exercise_student <- unique(exercise_student)
exercise_student

# Merge program names
exercise_student <- merge(exercise_student,
  exercise_program,
  by = "cip6",
  all.x = TRUE
)
exercise_student

exercise_student[, .N, by = "program"]
```




## Save results

Having identified a set of codes and names for the programs we want to study, we should save it to use later without rerunning the code that created it. For example, we could save the program codes and names in the `DT` data frame above to the project `results` directory for easy access as we work through the study. 

The file path names shown in the read/write commands are relative file paths incorporating two assumptions---first described in [Getting started](art-000-getting-started.html)

1. You are working in an RStudio project. 
2. The project main directory contains a `results` folder. 


Here we use fast write `fwrite()` from data.table. You can view its help page by running 

``` r
? fwrite

```

You may prefer other read/write functions such as those in base R or packages readr or rio, but we'll use the data.table functions. 

```r
# write to the results directory
fwrite(DT, file = "results/case_program.csv")
```

We could read it using `fread()`. We use the `colClasses` argument to ensure that the `cip6` variable type is read as a character, not an integer. 

```r
# read from the results directory
case_program <- fread(
  "results/case_program.csv",
  colClasses = list(character = c("cip6"))
)
```


We suggest saving intermediate results like these at each significant milestone of a study, especially after steps requiring a lot of computing time. 


### Exercises

1. Write a copy of the exercise results above (`exercise_program`) to file in your local `results` directory. 


## References

<div id="refs"></div>










## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval=FALSE}
# packages used
library("midfieldr")
library("data.table")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
# cip
str(cip)
sort(unique(cip$cip2))
sort(unique(cip$cip4))
length(unique(cip$cip6))
some_programs <- cip[, cip4name, drop = FALSE]
sample(some_programs, 20)

# filter using keywords
filter_search(dframe = cip, keep_text = "engineering")
filter_search(cip,
  "engineering",
  drop_text = c(
    "related",
    "technology",
    "technologies"
  ),
  select = c("cip6", "cip6name")
)
filter_search(cip,
  "engineering",
  drop_text = c(
    "related",
    "technology",
    "technologies"
  ),
  select = c("cip4", "cip4name")
)
filter_search(cip, "civil")
first_pass <- filter_search(cip, "civil")
second_pass <- filter_search(first_pass, "engineering")
third_pass <- filter_search(second_pass, drop_text = "technology")
filter_search(cip, "civil engineering", drop_text = "technology")

# filter using numerical codes
filter_search(cip, "german")
filter_search(cip, c("050125", "160501"))
filter_search(cip, c("^1407", "^1408"))
filter_search(cip, "^54")
codes_we_want <- c("^24", "^4102", "^450202")
filter_search(cip, codes_we_want)

# when search terms cannot be found
sub_cip <- filter_search(cip, c("050125", "111111", "160501", "Bogus", "^55"))
sub_cip

# Assigning program names
DT <- filter_search(cip, c("^1408", "^1410", "^1419", "^1435"))
cols_we_want <- c("cip6", "cip4name")
DT <- DT[, ..cols_we_want]
DT[, program := cip4name]
temp <- copy(DT)
rows_to_edit <- temp$program %ilike% "electrical"
temp[rows_to_edit, program := "Electrical Engineering"]
temp <- copy(DT)
rows_to_edit <- temp$cip6 %like% "^1410"
temp[rows_to_edit, program := "Electrical Engineering"]
DT[, program := fcase(
  program %ilike% "civil", "Civil",
  program %ilike% "electrical", "Electrical",
  program %ilike% "mechanical", "Mechanical",
  program %ilike% "industrial", "Industrial"
)]
DT[, cip4name := NULL]

# results directory read/write
fwrite(DT, file = "results/case_program.csv")
case_program <- fread(
  "results/case_program.csv",
  colClasses = list(character = c("cip6"))
)
```
