---
title: "Case study students"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Case study students}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-004-case-study-students-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
```

## Introduction

We continue the case study, a tour of a typical workflow with student unit record data. Again, for the sake of brevity we focus on the logic of the analysis and leave the detailed discussion of functions and arguments to later vignettes. 

Assuming you completed the introduction to the four data tables (in [Getting started](art-000-getting-started.html)), the case study has four parts:

- [Case study programs](art-002-case-study-programs.html)
- *Case study students* &#9665; You are here.
- [Case study stickiness](art-006-case-study-stickiness.html)
- [Case study graduation rate](art-006-case-study-grad-rate.html)

Longer vignettes with detailed development of methods and functions related to this part of the study include:   

- [Subsetting MIDFIELD data](art-015-subsetting-midfield-data.html){target="_blank"} 
- [Data sufficiency](art-020-data-sufficiency.html){target="_blank"} 

### In this vignette

Starting with the `study_program` data frame from the previous vignette ([Case study programs](art-002-case-study-programs.html){target="_blank"}), we: 

- Gather all degree-seeking students ever enrolled  in the programs
- Join columns for program names and student race/ethnicity and sex 
- Subset rows for data sufficiency 
- Save the results 

We use the following midfieldr functions
 
- [`filter_match()`](../reference/filter_match.html)
- [`add_race_sex()`](../reference/add_race_sex.html)
- [`add_timely_term()`](../reference/add_timely_term.html)
- [`add_data_sufficiency()`](../reference/add_data_sufficiency.html)

The vignette output is the data frame `study_student`. 

## Gather students

If you are writing your own script to follow along, we start with these packages: 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# optional code to control data.table printing
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)

# load data tables from midfielddata
data(student, term)
```

If you saved your results from the previous vignette in the file  `results/case_program.csv`, you can read them back in to your workspace using the following where we specify that the CIP column is a string variable to prevent it being read as an integer. 

``` r
# Optional 
case_program <- fread("results/case_program.csv", 
                      colClasses = c(cip6 = "character"))
```

If not, the data are included with midfieldr as the `study_program` data set. View its help page by running

``` r
? study_program
```

In the remainder of the vignette, we assume we're using the `study_program` data frame. A reminder of its contents:

```{r}
# Loaded with midfieldr
study_program
```

We start with `filter_match()` to subset rows of `term` with `cip6` values that match the `cip6` values in `study_program` and returning three columns for student ID, institution, and CIP code.  

```{r}
# All terms for all students in the study programs
DT <- filter_match(term,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "institution", "cip6")
)

# Examine the result
DT
```

This result has duplicate rows because the data include all terms for each student. We remove duplicates with `unique()`.   

```{r}
DT <- unique(DT)

# Examine the result
DT
```

In general, the `term` data can include non-degree-seeking students. To limit the study to degree-seeking students only, we subset rows of `DT` with `mcid` values that match the `mcid` values in the `student` table (all students in th `student` data table are degree-seeking). As it happens, in this example all students are degree-seeking, so the results don't change. 

```{r}
# Limit the study to degree-seeking students
DT <- filter_match(DT,
  match_to = student,
  by_col = "mcid"
)

# Examine the result
DT
```




## Join grouping variables 

In support of examining student characteristics (race/ethnicity and sex) and curricular pathways (program), we join these characteristics to the student data. These are the "group by" variables used in our persistence metrics. 

Program names are joined by merging `study_program` to our working data frame. 

```{r}
# Program name is a commonly used grouping variable
DT <- merge(DT, study_program, by = "cip6", all.x = TRUE)

# Examine the result
DT
```

Student race/ethnicity and sex are joined using the midfieldr `add_race_sex()` function. 

```{r}
# Race/ethnicity and sex are commonly used grouping variables
DT <- add_race_sex(DT, midfield_student = student)

# Examine the result
DT
```

## Filter for data sufficiency 

For students admitted too near the upper limit of their institution's data range, the available data cover an insufficient number of years to fairly assess student performance. Only students for whom the data range is sufficient should be included in any analysis.

Data sufficiency is defined in terms of a timely completion term: the last term in which program completion would be considered timely for a given student. The timely completion term for many students is their admission term plus 6 years. 

The midfieldr function `add_timely_term()` estimates a timely completion term for each student. 

```{r}
# Timely completion term is required for determining data sufficiency
DT <- add_timely_term(DT, midfield_term = term)

# Examine the result
DT
```

The data sufficiency criterion is that a study must be limited to students whose timely completion term is within the range of data provided by their institution.

The midfieldr function `add_data_sufficiency` accesses finds the upper limit of their institution's data range, compares it to the timely completion term, and reports in a TRUE/FALSE column whether the data are sufficient. 

```{r}
# Establish the criterion
DT <- add_data_sufficiency(DT, midfield_term = term)

# Examine the result
DT
```

We limit the study to students for whom the data sufficiency criterion is met.

```{r}
# Apply the criterion
DT <- DT[data_sufficiency == TRUE]

# Examine the result
DT
```







## Save results

We can order the columns of the result using the data.table function  `setcolorder()`---not strictly necessary, but it does facilitate  reproducibility.  

```{r}
# Optional step to order the rows
setcolorder(DT, c("mcid", "institution"))
```

We save the `DT` data frame as a CSV file to the `results` directory for easy access as we work through the study.  

``` r
# Assumes you have a results folder in the project directory
fwrite(DT, file = "results/case_student.csv")
```
These results have been included with midfieldr as the built-in data set `study_student`. View its help page by running 

```r
? study_student
```
View the data frame, 

```{r}
study_student
```

Check that the two data frames are identical, 

```{r}
all.equal(DT, study_student)
```

In subsequent vignettes, you can read in the `case_student` data from your `results` directory or you can use the `study_student` data provided with midfieldr. 

















## References

<div id="refs"></div>






## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# packages used
library("midfieldr")
library("midfielddata")
library("data.table")

# load data tables from midfielddata
data(student, term)

# Gather students
DT <- filter_match(term,
  match_to = study_program,
  by_col = "cip6",
  select = c("mcid", "institution", "cip6")
)
DT <- unique(DT)
DT <- filter_match(DT,
  match_to = student,
  by_col = "mcid"
)

# Join grouping variables
DT <- merge(DT, study_program, by = "cip6", all.x = TRUE)
DT <- add_race_sex(DT, midfield_student = student)

# Filter for data sufficiency
DT <- add_timely_term(DT, midfield_term = term)
DT <- add_data_sufficiency(DT, midfield_term = term)
DT <- DT[data_sufficiency == TRUE]

# Save results
setcolorder(DT, c("mcid", "institution"))
fwrite(DT, file = "results/case_student.csv")
```

