---
title: "Case study---displays"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study---displays}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
    fig.path = "../man/figures/art-004-case-displays-", 
    fig.width = 6,
    fig.asp = 1 / 1.6,
    out.width = "70%",
    fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 4 of a case study in four parts:

- [Goals.](art-001-case-goals.html) Describing the study parameters.  
- [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.
- [Metrics.](art-003-case-metrics.html) Calculating summary statistics and metrics.
- [Displays.](art-004-case-displays.html) Creating tables and charts to display results.


## Preface

The next step in our case study is to ...  

If you are writing your own script(s) to follow along, you can continue the previous script or start a new script for this section. 

If starting a new script, we start with these packages and data. 

```{r}
# Packages
library("midfieldr")
library("midfielddata")
suppressPackageStartupMessages(library("data.table"))

# data.table printing options
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

