---
title: "Case study graduation rate"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Case study graduation rate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
source("article-preamble.R")
set_fig_path("art-008-case-study-grad-rate-")
```





## Introduction

We continue the case study, a tour of a typical workflow with student unit record data. Again, for the sake of brevity we focus on the logic of the analysis and leave the detailed discussion of functions and arguments to later vignettes. 

Assuming you completed the introduction to the four data tables (in [Getting started](art-000-getting-started.html)), the case study has four parts:

- [Case study programs](art-002-case-study-programs.html)
- [Case study students](art-004-case-study-students.html)
- [Case study stickiness](art-006-case-study-stickiness.html)
- *Case study graduation rate* &#9665; You are here.

Longer vignettes with detailed development of methods and functions related to this part of the study include:  

- [Timely completion](art-030-timely-completion.html){target="_blank"} 
- [FYE programs](art-040-fye-programs.html){target="_blank"} 
- [Multiway graphs](art-050-multiway-graphs.html){target="_blank"}  
- [Tabulating data](art-060-tabulating-data.html){target="_blank"} 

### Graduation rate

**As defined by IPEDS:** In the US, the Integrated Postsecondary Education Data System (IPEDS) defines "graduation rate" as the fraction of a cohort of full-time, first-time, degree-seeking undergraduates who complete their program within a percentage (100%, 150%, or 200%) of the "normal" time (typically 4 years) as defined by the institution [@IPEDS:2020]. The MIDFIELD team typically uses the 150% (6-year) definition. 

The IPEDS definition limits the cohort being studied by excluding students who attend college part-time, who transfer from one institution to another, and who start college in any term other than the Fall term. Moreover, IPEDS policy does not extend the time limit for students who change majors (switchers) even if the student and the institution are satisfied that the student graduated in the new major in a timely fashion. 

The American Council on Education (ACE) estimates that this definition of graduation rate may exclude up to 60% of students at 4-year institutions [@Cook+Hartle:2011]. So why use the metric at all? As Cook and Hartle explain, 

> ... in the eyes of the public, policy makers, and the media, graduation
> rate is a clear, simple, and logical---if often misleading---number.


**As defined by MIDFIELD:** Compared to IPEDS, a more inclusive cohort is used in MIDFIELD calculations of graduation rate. The differences are summarized in the table.  


<br>
```{r echo = FALSE}
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
Constraints <- c(
  "timely completion in:",
  "transfer students are:",
  "extended span for switchers is:",
  "part-time students are:",
  "cohort admitted in:"
)
IPEDS <- c(
  "6 years",
  "excluded",
  "excluded",
  "excluded",
  "Fall only"
)
MIDFIELD <- c(
  "6 years",
  "included",
  "planned",
  "included",
  "any term"
)
Notes <- c(
  "can be changed by user",
  "timely completion span depends on entry level",
  "heuristic in development for future release",
  "",
  ""
)

DT <- data.table(Constraints, IPEDS, MIDFIELD, Notes)
setnames(DT, old = c("Notes"), new = c("MIDFIELD notes"))
kable2html(DT, caption = "Comparing graduation rate criteria")
```
<br>



### In this vignette

Starting with the `study_student` data frame from a previous vignette ([Case study students](art-004-case-study-students.html){target="_blank"}), we:

- A
- T
- T
- Compute and graph the graduation rate 
- Save the results

We use the following midfieldr functions

- [`add_completion_timely()`](../reference/add_completion_timely.html)
- [`condition_multiway()`](../reference/condition_multiway.html)

The vignette output is the data frame `study_grad_rate`. 





## Apply the timely completion criterion 

If you are writing your own script to follow along, we start with these packages: 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
library("ggplot2")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)

# load data tables from midfielddata
data(degree)
```

We gathered the eligible students for the case study programs in a previous case study vignettes.  The data are saved in the `study_student` data set that loads with midfieldr. View its help page by running

```r  
? study_student
```

Review the data. The data sufficiency criterion has been applied. As usual, we use assign our working data frame to `DT`. 

```{r}
# Create a new memory location to avoid updating by reference
DT <- copy(study_student)

# Examine the result
DT
```

Before we can count the number of graduates, we have to assess whether graduates complete their program in a timely manner. A student graduating later than their timely-completion term are grouped with non-graduates for purposes of computing a persistence metric.

The `add_completion_timely()` function requires that the input data frame has a `timely_term` column---which ours does. The function accesses the `degree` data to determine the term  of a student's first degree and reports whether the degree term (if any) satisfies the timely completion criterion. 

```{r}
DT <- add_completion_timely(DT, midfield_degree = degree)
DT
```

## Under construction 

## References

<div id="refs"></div>





## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
library("ggplot2")
```



