---
title: "Case study stickiness"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Case study stickiness}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
nocite: | 
resource_files: 
  - ../man/figures/art-006-case-study-stickiness-fig1-1.png
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
source("vignettes/article-preamble.R")
set_fig_path("art-006-case-study-stickiness-")
```





## Introduction

We continue the case study, a condensed tour of a typical workflow with student unit record data. We provide the R code with minimum explanation and focus instead on the overall logic of the process. Other vignettes explore the details of the midfieldr functions. 

Assuming you completed the introduction to the four data tables (in [Getting started](art-000-getting-started.html)), the case study has four parts:

- [Case study programs](art-002-case-study-programs.html)
- [Case study students](art-004-case-study-students.html)
- *Case study stickiness* &#9664; You are here.
- [Case study graduation rate](art-006-case-study-grad-rate.html)

Longer vignettes with detailed development of methods and functions related to this part of the study include:  

- [Data sufficiency](art-020-data-sufficiency.html)
- [Timely completion](art-030-timely-completion.html){target="_blank"} 
- [Multiway graphs](art-050-multiway-graphs.html){target="_blank"} 

### Longitudinal stickiness
 
*Longitudinal stickiness* is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program after accounting for the data sufficiency and timely completion  criteria  [@Ohland+Orr+others:2012]. 

Stickiness measures the extent to which a program succeeds in its basic goal of graduating the students it admits, without regard to how or when a student is admitted to a program---the metric includes students who begin college part-time, enroll mid-year, switch majors, or transfer, in addition to first-time-in-college students. 

### In this vignette

Starting with the `study_student` data frame from the previous vignette ([Case study students](art-004-case-study-students.html){target="_blank"}), we:

- Apply the timely completion criterion 
- Tally students ever enrolled in a program by race/ethnicity, and sex
- Tally students completing a program by race/ethnicity, and sex
- Compute and graph the stickiness 
- Save the results

We use the following midfieldr functions

- [`add_completion_timely()`](../reference/add_completion_timely.html)
- [`condition_multiway()`](../reference/condition_multiway.html)

The vignette output is the data frame `study_stickiness`. 







## Apply the timely completion criterion 

If you are writing your own script to follow along, we start with these packages: 

```{r}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
library("ggplot2")

# optional code to control data.table printing
options(datatable.print.nrows = 10,
        datatable.print.topn = 5,
        datatable.print.class = TRUE)

# load data tables from midfielddata
data(degree)
```


If you saved your results from the previous vignette in the file  `results/case_student.csv`, you can read them back in to your workspace using the following where we specify that the CIP and timely term columns are character variables to prevent them being read as integers. 

``` r
# Optional 
case_program <- fread("results/case_student.csv", 
                      colClasses = c(
                        cip6 = "character",
                        timely_term = "character"
                      ))
```

If not, the data are included with midfieldr as the `study_student` data set. View its help page by running

``` r
? study_student
```

In the remainder of the vignette, we assume we're using the `study_student` data frame. A reminder of its contents:

```{r}
# Loaded with midfieldr
study_student
```

The data sufficiency criterion has been applied. As usual, we assign our working data frame to `DT`. 

```{r}
# Create a new memory location to avoid updating by reference
DT <- copy(study_student)
```

Before we can count the number of graduates, we have to assess whether graduates complete their program in a timely manner. A student graduating later than their timely-completion term are grouped with non-graduates for purposes of computing a persistence metric.

The `add_completion_timely()` function accesses the `degree` data to determine the term  of a student's first degree (if any), compares it to the student's timely completion term, and reports whether the degree term satisfies the timely completion criterion. 

```{r}
# Determine whether the timely completion criterion is satisfied
DT <- add_completion_timely(DT, midfield_degree = degree)

# Examine the result
DT
```



## Summarize by group 

The variables we group by are program, race/ethnicity, and sex. It has proven convenient over the years to combine race/ethnicity and sex into its own column.  

```{r}
# Construct one of the grouping variables.
DT[, race_sex := paste(race, sex)]

# Examine the result
DT
```

We assign the grouping variables to a vector. 

```{r}
# A reusable vector for grouping, summarizing, and joining.
grouping_variables <- c("program", "race_sex")
```


```{r echo = FALSE}
temp <- copy(DT)
temp <- temp[, .(ever = .N), by = grouping_variables]
n_temp <- nrow(temp)
```

Starting with our working data frame `DT`, we count the enrollees by groups and assign the result to the new `ever` column. The result is a new data frame with only `r n_temp`  rows since we have aggregated the counts. 

```{r}
# Create a new memory location to avoid updating by reference
enrolled <- copy(DT)

# Count
enrolled <- enrolled[, .(ever = .N), by = grouping_variables]

# Examine the result
enrolled
```

Returning to our working data frame `DT`, we subset rows satisfying the timely-completion criterion. 

```{r}
# Create a new memory location to avoid updating by reference
completed <- copy(DT)

# Limit the graduate count to those with timely completion
completed <- completed[completion_timely == TRUE]

# Examine the result
completed
```

We count the graduates and assign the result to the new `grad` column. 

```{r}
# Obtain count of graduates
completed <- completed[, .(grad = .N), by = grouping_variables]

# Examine the result
completed
```

Now we join the two data frames, setting `all.x = TRUE` to retain all rows of  `enrolled`. 

```{r}
# Join
DT <- merge(enrolled, completed, by = grouping_variables, all.x = TRUE)

# Examine the result
DT

# Merging created a data.table "key" that we delete
setkey(DT, NULL)
```

In this case, the `completed` data has fewer rows than the `enrolled` data so the `merge()` introduced some NA values in the `grad` column. 

```{r}
# Checking if any NAs were introduced in the grad column
sum(is.na(DT$grad))
```

Those NA values are better encoded as zeros.

```{r}
# NA in this case represents a count of zero
DT[is.na(grad), grad := 0][]

# Examine the result
DT
```





## Compute stickiness

Stickiness is the ratio of graduates to enrollees by group. 

```{r}
# data table printing option
options(datatable.print.topn = 40)

# Reached our goal of computing longitudinal stickiness
DT[, stick := round(100 * grad / ever, 1)]

# Examine the result
DT
```

You could save the `DT` data frame as a CSV file to your `results` directory for easy access as we work through the study.  

``` r
# Assumes you have a results folder in the project directory
fwrite(DT, file = "results/case_stickiness.csv")
```

These results have been included with midfieldr as the built-in data set `study_stickiness`. View its help page by running 

```r
? study_stickiness
```
View the data frame, 

```{r}
study_stickiness
```

Check that the two data frames are identical, 

```{r}
all.equal(DT, study_stickiness)
```

In subsequent vignettes, you can read in the `case_stickiness` data from your `results` directory or you can use the `study_stickiness` data provided with midfieldr. 







## Graph the results

We do some data conditioning before presenting the results. Even with students and institutions de-identified, privacy is at risk for small populations, so we omit observations with fewer than 10 students ever enrolled. 

```{r}
# Omit before graphing
DT <- DT[ever >= 10]
```

We also omit rows with ambiguous values for race/ethnicity (International and Other/Unknown). Having graphed the results in advance, we find that the small population subset removed Native American students from all but one of the disciplines---so we add them to the list of observations to omit. 

```{r}
# data table printing option
options(datatable.print.topn = 3)

# Omit before graphing
DT <- DT[!race_sex %ilike% c("International|Other|Native")]

# Examine the result
DT
```

Condition the data for graphing using `condition_multiway()`.  The comments give a brief clue to the purpose of each argument. We'll use it here without lengthy explanation (covered in detail in a separate vignette). 

```{r}
DT <- condition_multiway(
  dframe = DT, # input data frame
  categ_col = c("program", "race_sex"), # multiway categorical variables
  quant_col = "stick", # multiway quantitative variable
  detail = FALSE, # do not return detail columns
  order_by = "percent", # stickiness (a percentage) used to order rows and panels
  param_col = c("grad", "ever") # parameters for computing group stickiness
)

# Examine the result
DT
```

```{r echo = FALSE}
mw <- copy(DT[, .(program, race_sex, stick)])
asp_ratio <- asp_ratio_mw(mw, categories = c("program", "race_sex"))
```

Because stickiness is a percentage, we can use the "percent" method for ordering: panels are ordered by program stickiness (ignoring race/ethnicity and sex) and rows are ordered by race/sex stickiness (ignoring program). 

```{r fig1, fig.asp = asp_ratio[1]}
ggplot(data = DT, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Stickiness (%)",
    y = "",
    title = "Ordered by computed group metric",
    caption = "Source: midfielddata."
  )
```

**Observations:** The graph illustrates how the ordering of rows and panels are crucial to what Cleveland calls the "asymmetry in the perception of effects." For example, consider the low stickiness of Hispanic/Latinx women in Civil Engineering (given their overall high stickiness across all programs) or the high stickiness of Hispanic/Latinx men in Industrial Engineering (given their overall low stickiness across all programs).

Recall, however, that these are practice data, not research data---suitable for learning about student-record analysis, but not for drawing inferences about student performance. 




## References

<div id="refs"></div>










## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script.

```{r eval = FALSE}
# packages used
library("midfieldr")
library("midfielddata")
suppressMessages(library("data.table"))
library("ggplot2")

# load data tables from midfielddata
data(degree)

# Apply the timely completion criterion
DT <- copy(study_student)
DT <- add_completion_timely(DT, midfield_degree = degree)

# Summarize by group
DT[, race_sex := paste(race, sex)]
grouping_variables <- c("program", "race_sex")
enrolled <- copy(DT)
enrolled <- enrolled[, .(ever = .N), by = grouping_variables]
completed <- copy(DT)
completed <- completed[completion_timely == TRUE]
completed <- completed[, .(grad = .N), by = grouping_variables]
DT <- merge(enrolled, completed, by = grouping_variables, all.x = TRUE)
setkey(DT, NULL)
DT[is.na(grad), grad := 0][]

# Compute longitudinal stickiness
DT[, stick := round(100 * grad / ever, 1)]

# Graph the results
DT <- DT[ever >= 10]
DT <- DT[!race_sex %ilike% c("International|Other|Native")]
DT <- condition_multiway(
  dframe = DT, # input data frame
  categ_col = c("program", "race_sex"), # multiway categorical variables
  quant_col = "stick", # multiway quantitative variable
  detail = FALSE, # do not return detail columns
  order_by = "percent", # stickiness (a percentage) used to order rows and panels
  param_col = c("grad", "ever") # parameters for computing group stickiness
)
ggplot(data = DT, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(
    x = "Stickiness (%)",
    y = "",
    title = "Ordered by computed group metric",
    caption = "Source: midfielddata."
  )
```




