---
title: "Case study---classifying"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study---classifying}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
    fig.path = "../man/figures/art-002-case-classify-", 
    fig.width = 6,
    fig.asp = 1 / 1.6,
    out.width = "70%",
    fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 3 of a case study in five parts:

- [Introduction.](art-001-case-introduction.html) Goals.  
- [Subsetting.](art-002-case-filter.html) Filter observations and select variables.  
- [Classifying.](art-003-case-classify.html) Construct variables that characterize or classify the record. 
- [Summarizing.](art-004-case-summarize.html) Group variables and count observations.
- [Displaying.](art-005-case-display.html) Create charts and tables for discussion.

Overall, we use R packages midfieldr, midfielddata, data.table, and ggplot2 without detailed explanation.  We provide those details in separate vignettes.  


## Load 


```{r}
# do this after classifying by custom program name
#>  1408 Civil 
#>  1410 Electrical 
#>  1419 Mechanical 
#>  1427 Systems  
#>  1435 Industrial Engineering 143501
#>  1436 Manufacturing Engineering 
#>  1437 Operations Research 

# program_cip <- c("", "", "", "", "", "", "")
# # Subset observations for programs
# x <- copy(DT)
# x <- x[cip6 %like% "^1408|^1410|^1419|^1427|^1435|^1436|^1437"]
# sort(unique(x$cip6))
```

