---
title: "Getting started"
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  rmarkdown::html_vignette
date: "`r Sys.Date()`"
csl: ../inst/information-science-and-technology.csl
bibliography: ../inst/REFERENCES.bib
link-citations: yes
---

```{r setup, child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-000-")
```

Familiarity with the MIDFIELD data structure is a prerequisite for working with `midfieldr` functions so we start by inspecting the practice data in [`midfielddata`](https://midfieldr.github.io/midfielddata/), a companion R package that provides anonymized student-level records for 98,000 undergraduates at three US institutions from 1988 through 2018. 

The practice data are organized in four datasets keyed by student ID.  

```{r}
#| echo: false
wrapr::build_frame(
    "Dataset", "Each row is"           , "Students", "Rows"     , "Columns", "Memory"   |
    "course" , "one student per course", "97,555"  , "3,289,532", 12L      , "324.3 MB" |
    "term"   , "one student per term"  , "97,555"  , "639,915"  , 13L      , "72.8 MB"  |
    "student", "one student"           , "97,555"  , "97,555"   , 13L      , "17.3 MB"  |
    "degree" , "one student per degree", "49,543"  , "49,665"   ,  5L      , "5.2 MB"   
) |>
  kableExtra::kbl(align = "llrrrr", caption = "Table 1. Practice datasets in `midfielddata`.") |>
  kableExtra::kable_paper(lightable_options = "basic", full_width = FALSE) |>
  kableExtra::row_spec(0, background = "#c7eae5") |>
  kableExtra::column_spec(1, monospace = TRUE) |>
  kableExtra::column_spec(1:6, color = "black", background = "white")
```



## Definitions

```{r child = "../man/rmd/define-student-level-data.Rmd"}
```








## Method

In this vignette: 

- Overview of each dataset 
- Summary of variables typically encountered when using `midfieldr` functions
- Closer look: For one student, all records in all datasets 
- Introduce helper functions `select_required()` and `same_content()` 

*Notes on syntax.* &nbsp; We use `data.table` for data manipulation. Some users may prefer base R or `dplyr`. Each system has its strengths---users are welcome to translate our examples to their preferred syntax.  

```{r child = "../man/rmd/note-practice-not-research.Rmd"}
```






## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

*Load data tables.*  &nbsp; Data tables can be loaded individually or collectively as needed.  

```{r}
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```






## `student`

Run `?student` to view its help page and data dictionary. Contains one observation per student. Data are assumed to be current at the time the student was admitted to their institution. 

```{r}
student
```

Student IDs and institution names have been anonymized to remove identifiable information. 

```{r}
# Anonymized IDs
sample(student$mcid, 8)

# Anonymized institutions
sort(unique(student$institution))
```

Race/ethnicity and sex are often used as grouping variables.

```{r}
# Possible values
sort(unique(student$race))

# Possible values
sort(unique(student$sex))
```

Counts in each category. 

```{r}
# N by institution
student[order(institution), .N, by = "institution"]

# N by race
student[order(race), .N, by = "race"]

# N by sex
student[order(sex), .N, by = "sex"]
```







## `course`

Run `?course` to view its help page and data dictionary. Contains one observation per student per course.  

```{r}
course
```

The `abbrev`, `number`, and `discipline_midfield` columns have no NA values, so they might be useful if one is filtering for specific course types. The `course` column, on the other hand, has a high number of NA values. 

```{r}
# Many NA values in this column
sum(is.na(course$course))

# No NA values in these columns.
sum(is.na(course$abbrev))
sum(is.na(course$number))
sum(is.na(course$discipline_midfield))
```










## `term`

Run `?term` to view its help page and data dictionary. Contains one observation per student per term.  

```{r}
term
```

Terms are encoded `YYYYT`, where 

- `YYYY` is the year at the start of the academic year, and  
- `T` encodes the semester or quarter---Fall (`1`), Winter (`2`), Spring (`3`), and Summer (`4`, `5`, and `6`)---within an academic year

For example, for academic year 1995--96, 

- `19951` encodes Fall 95--96
- `19953` encodes Spring 95--96
- `19954` encodes Summer 95--96 (first session)

Different institutions supply data over different time spans. 

```{r}
# Range of data by institution
term[, .(min_term = min(term), max_term = max(term)), by = "institution"]
```

Programs are encoded in the `cip6` variable, a 6-digit character based on the 2010 Classification of Instructional Programs (CIP) [@NCES:2010]. CIP codes are explored in detail in [Programs](art-040-programs.html). 

```{r}
# A sample of cip6 values
sort(unique(sample(term$cip6, 8)))
```

Student level is used when determining timely completion terms of transfer students.

```{r}
# Possible values
sort(unique(term$level))
```






## `degree`

Run `?degree` to view its help page and data dictionary. This dataset contains records for graduates only, thus the number of unique students in `degree` (`r length(unique(degree$mcid))`) is less than the number of students in the other three datasets (`r length(unique(student$mcid))`). 

Contains one observation per student per degree. The `term_degree` and `cip6` variables indicate when and from which program a student graduates. 

```{r}
degree
```

Number of degrees earned per student. 

```{r}
# Count students by number of degrees
by_id <- degree[, .(degree_count = .N), by = "mcid"]
by_id[, .(N_students = .N), by = "degree_count"]
```






## Closer look

We display the records for one specific student, using their ID to subset each dataset.  

```{r}
# One student ID
id_we_want <- "MCID3112192438"
```

*Student.* &nbsp; As expected, `student` yields one row per student.

```{r}
# Observations for a selected ID
student[mcid == id_we_want]
```

*Course.* &nbsp; For this student, the records span `r nrow(course[mcid == id_we_want])` rows, one row per course. 

```{r}
# Observations for a selected ID
course[mcid == id_we_want]
```

*Term.* &nbsp; Here, the records span `r nrow(term[mcid == id_we_want])` rows, one row per term. 

```{r}
# Observations for a selected ID
term[mcid == id_we_want]
```

*Degree.* &nbsp; In this example, the records span `r nrow(degree[mcid == id_we_want])` rows, one row per degree. The degrees were earned in the same term, Spring 2009. 

```{r}
# Observations for a selected ID
degree[mcid == id_we_want]
```

Not all students with more than one degree earn them in the same term. For example, the next student earned a degree in 1996 and a second degree in 1999. In most analyses, only the first baccalaureate degree would be used.

```{r}
# Observations for a different ID
degree[mcid == "MCID3111315508"]
```















## `select_required()`

A midfieldr convenience function to reduce the number of columns of a MIDFIELD data table after loading. Using this function is optional. 

`select_required()` selects only those columns typically required by other midfieldr functions. Operates on a data frame to retain columns having names that match or partially match search terms. Rows are unaffected. 

The primary benefit is reducing screen clutter when viewing data frames during an interactive session. The disadvantage is that the deleted columns are unavailable unless you first set aside a copy of the source file or reload it using `data()` when you need it. 


*Arguments.*

- **`midfield_x`** &nbsp; MIDFIELD data frame, typically `student`, `term`, or `degree`. 

- **`select_add`** &nbsp; Optional character vector of search terms to add to the default vector given by `c("mcid", "institution", "race", "sex", "^term", "cip6", "level")`. Argument, if used, must be used by name.

*Before.* &nbsp; The original source file. 

```{r}
# Create a copy of the source file
source_term <- copy(term)
source_term[]
```

*After* &nbsp; The output retains only the variables required by other midfieldr functions.  

```{r}
# Select variables required by midfieldr functions
term <- select_required(source_term)
term[]
```

*Equivalent usage.* &nbsp; The following implementations yield identical results

```{r}
# Required argument explicitly named
x <- select_required(midfield_x = term)

# Required argument not named
y <- select_required(term)

# Optional argument, if used, must be named. NULL yields the default columns.
z <- select_required(term, select_add = NULL)
```


## `same_content()` 

In the code chunk just above, we claim the three results are equivalent. To verify that the three data tables have the same content, we use `same_content()`, a midfieldr function that wraps `all.equal()` for data.table objects that ignores row order, column order, and keys (if any). 

*Arguments.*

- **`x`, `y`** &nbsp; Two data frames or data.tables to compare. 

*Example.* &nbsp; Demonstrate the equivalence of the data tables from the previous section. 

```{r}
# Demonstrate equivalence
same_content(x, y)
same_content(x, z)
```

Demonstrate that row and column order are ignored. 

```{r}
# Two columns from student, use key to order rows
x <- student[, .(mcid, institution)]
setkey(x, mcid)
x[]

# Same information with different row order, column order, and key
y <- student[, .(institution, mcid)]
setkey(y, institution)
y[]

# Demonstrate equivalence
same_content(x, y)
```

If the two data tables do not have the same content, the printout indicates the types of differences. 

```{r}
# Demonstrate non-equivalence
same_content(student, degree)
```











## Reusable code

*Preparation.* &nbsp; The immediate prerequisites or “intake” required by the reusable code chunk are the source data tables.  

```{r}
#| eval: false
# Load source data
data(student, term, degree)
```

*Initial data processing.* &nbsp; A summary code chunk for ready reference.  

```{r}
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_term    <- copy(term)
source_degree  <- copy(degree)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
term    <- select_required(source_term)
degree  <- select_required(source_degree)
```








```{r child = "../man/rmd/common-closing.Rmd"}
```

