---
title: "Blocs"
vignette: >
  %\VignetteIndexEntry{Blocs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-050-")
```

A *bloc* is a grouping of Student Unit Records (SURs) dealt with as a unit, for example, a grouping of starters in a program, graduates of a program, or ever enrolled in a program. We often use a *left join* merging operation to add one or more variables to a working data frame and filter on those variables to construct the desired bloc. 

Different metrics require different blocs. Graduation rate, for example, requires starters and their graduating subset; stickiness requires ever-enrolled and their graduating subset. Subsequent articles describe [FYE proxies](art-060-fye-proxies.html) (special case of starters), [Starters](art-070-starters.html), and [Graduates](art-080-graduates.html). This article treats the *ever-enrolled* bloc and left joins generally. 

Because a bloc is usually defined for specific programs, the final filter applied in gathering a bloc is often an *inner join* to filter by program labels, developed in [Programs](art-040-programs.html#reusable-code). 

This article in the MIDFIELD workflow. 

1. Planning  
1. Initial processing  
1. `r accent("Blocs")`  
    - `r accent("Ever-enrolled")`  
    - FYE proxies  
    - Starters  
    - Graduates   
1. Groupings  
1. Metrics  
1. Displays 
  






## Definitions

```{r child = "../man/rmd/define-bloc.Rmd"}
```

```{r child = "../man/rmd/define-ever-enrolled.Rmd"}
``` 

```{r child = "../man/rmd/define-migrators.Rmd"}
```





## Method

We use left joins to add variables to a working data frame and filter for students ever-enrolled in the case study programs. Migrators (if any) yield more than one observation (program) for the same ID. 

1. Filter source SURs for data sufficiency and degree-seeking. 

1. Gather ever-enrolled

1. Filter by program. 

```{r child = "../man/rmd/note-practice-not-research.Rmd"}
```






## Load data

*Open.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:


```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

*Load.* &nbsp; MIDFIELD practice data tables. Described in [Getting started](art-000-getting-started.html).

```{r}
# Load practice data
data(student, term, degree)
```

*Loads with midfieldr.* 

- `study_programs` (derived in [Programs](art-040-programs.html#reusable-code)). 








## Initial processing

*(Optional) Select.* &nbsp; Columns required by midfieldr functions. Code reproduced from [Getting started](art-000-getting-started.html#reusable-code). 

```{r}
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)
source_degree <- copy(degree)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
term <- select_required(source_term)
degree <- select_required(source_degree)
```

*Work.* &nbsp; Using `copy()` avoids *by-reference* effects  [@data.table-reference-semantics].

```{r}
# Working data frame
DT <- copy(term)
```

*Data sufficiency.*  &nbsp; Filter to satisfy the data sufficiency criterion. Code reproduced from [Data sufficiency](art-020-data-sufficiency.html#reusable-code).

```{r}
# Filter for data sufficiency, output unique IDs
DT <- add_timely_term(DT, term)
DT <- add_data_sufficiency(DT, term)
DT <- DT[data_sufficiency == "include", .(mcid)]
DT <- unique(DT)
```

*Degree seeking.* &nbsp; Filter to retain degree seeking students via an inner join with `student`. Code reproduced from [Degree seeking](art-030-degree-seeking.html#reusable-code).  

```{r}
# Filter for degree seeking, output unique IDs
DT <- student[DT, .(mcid), on = c("mcid"), nomatch = NULL]
DT <- unique(DT)
```

We preserve this data frame as a baseline for examples in the article.  

```{r}
baseline <- copy(DT)
```






## Left joins

An left join is a merge operation between two data frames which returns all observations (rows) of the "left" data frame `X` and all the matching rows in the "right" data frame `Y`.  Using data.table syntax, we have two approaches: `merge(X, Y, by)` (similar to base R) and `Y[X, j, on]` (native to data.table).

### Using `merge(X, Y, by)` 

The general form for a left join is 

        merge(X, Y, by, all.x = TRUE)

where 

- `X` is the "left" data frame, all rows returned  
- `Y` is the "right" data frame, matching rows returned 
- `by` is the vector of shared column names to merge by  
- `all.x = TRUE` ensures the left join

In this example, the `Y` data frame is `term`, from which we extract the ID and CIP columns before the join. 


```{r}
# Subset of term data frame to join
cols_we_want <- term[, .(mcid, cip6)]

# merge(X, Y, by) left join
merge(DT, cols_we_want, by = c("mcid"), all.x = TRUE)
```

Alternatively, one can select `Y` columns within the merge operation.

```{r}
# merge(X, Y, by) left join
merge(DT, term[, .(mcid, cip6)], by = c("mcid"), all.x = TRUE)
```



### Using `Y[X, j, on]` 

The second approach---native to data.table and computationally more efficient---has the form 

        Y[X, j, on]

where 

- `X` is the "left" data frame, all rows returned  
- `Y` is the "right" data frame, matching rows returned 
- `j` selects columns from the joined data frame to retain
- `on` is the vector of shared column names to merge on  

```{r}
# Y[X, j, on] left join (data.table native syntax)
term[DT, .(mcid, cip6), on = c("mcid")]
```




*Demonstrate equivalence.* Showing that the two approaches produce the same result. 

```{r}
#| collapse: true

# merge(X, Y, by) left join
x <- merge(DT, term[, .(mcid, cip6)], by = c("mcid"), all.x = TRUE)
setkey(x, NULL)

# Y[X, j, on] left join
y <- term[DT, .(mcid, cip6), on = c("mcid")]

# Demonstrate equivalence
same_content(x, y)
```


### Left join matching rules 

*Rows in X with no match in Y* &nbsp; will have NA values in the columns normally filled with `Y` values.  

For example, not all students in `DT` will earn a degree. After a left join (`degree` into `DT`), all rows of `DT` are returned. IDs in `DT` with no match in `degree` have an NA in the `term_degree` column (a variable from  the `degree` source table). 

```{r}
#| echo: false
options(datatable.print.topn = 5)
```

```{r}
x <- degree[DT, .(mcid, term_degree), on = c("mcid")]
setkeyv(x, c("mcid"))
x[]
```


The result has `r length(unique(x$mcid))` unique IDs with `r sum(!is.na(x$term_degree))` degrees. 

*Rows in X with multiple matches in Y* &nbsp; yields a new row in `X` for every matching row in `Y`. 

For example, most students in `DT` will be enrolled in multiple terms. After a left join (`term` into `DT`), all rows in `DT` are returned. IDs in `DT` with multiple matches in `term` have multiple rows in the result, differentiated by the values in the `term` column (a variable from the `term` source table). 

```{r}
x <- term[DT, .(mcid, term), on = c("mcid")]
setkeyv(x, c("mcid", "term"))
x[]
```

The result has `r length(unique(x$mcid))` unique IDs distributed over `r nrow(x)` observations.

```{r}
#| echo: false
options(datatable.print.topn = 3)
```

*"Left" and "right" matter.*  &nbsp; In left joins (unlike inner joins), `X[Y]` and `Y[X]` return different results: 

- `Y[X, j, on]` returns all rows of `X`

- `X[Y, j, on]` returns all rows of `Y`

```{r}
#| collapse: true

# What we want
x <- degree[DT, .(mcid, term_degree), on = c("mcid")]

# Not what we want
y <- DT[degree, .(mcid, term_degree), on = c("mcid")]

# Same content?
same_content(x, y)

# Compare N rows
nrow(x)
nrow(y)
```







## Ever-enrolled

*Work.* &nbsp; The baseline data frame we preserved earlier is the intake for this section. 

```{r}
# Reusable starting state
DT <- copy(baseline)
DT[]
```

*Add a variable.* &nbsp; Use a left join from `term` to `DT` to add the CIP variable.

```{r}
# Left-outer join from term to DT
DT <- term[DT, .(mcid, cip6), on = c("mcid")]
```

*Filter.* &nbsp; Filter to retain unique combinations if ID and CIP code. 

```{r}
# One observation per ID-CIP combination
DT <- unique(DT)
DT[]
```




## Filter by program

*Filter.*  &nbsp; Because "ever-enrolled" usually means "ever-enrolled in specific programs," this bloc concludes with a filter by program. Code reproduced from [Groupings](art-090-groupings.html#program-labels). 

```{r}
# Filter by program
DT <- study_programs[DT, on = c("cip6"), nomatch = NULL]
DT[, cip6 := NULL]
DT <- unique(DT)
DT[]
```



```{r}
#| eval: false
#| echo: false

# Run manually
# Writing external files
setkey(DT, mcid)
setkey(DT, NULL)
study_ever <- copy(DT)
usethis::use_data(study_ever, overwrite = TRUE)
```








## Reusable code: Blocs

*Preparation.* &nbsp; The baseline data frame we preserved earlier is the intake for this section.

```{r}
DT <- copy(baseline)
```

*Ever-enrolled.* &nbsp; A summary code chunk for ready reference.  Requires editing of `study_programs` before reuse with different programs.

```{r}
# Ever-enrolled bloc
DT <- term[DT, .(mcid, cip6), on = c("mcid")]
DT <- unique(DT)

# Filter by program
DT <- study_programs[DT, on = c("cip6"), nomatch = NULL]
DT[, cip6 := NULL]
DT <- unique(DT)
```

*Verify built-in data.* &nbsp; `study_ever`, included with midfieldr, contains the case study information developed above. Here we verify that the two data frames have the same content.  

```{r}
#| collapse: true

# Demonstrate equivalence
same_content(DT, study_ever)
```






```{r child = "../man/rmd/common-closing.Rmd"}
```
