---
title: "Case study: Goals"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study: Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
  @Dowle+Srinivasan:2021:data.table, 
  @Wickham:2016:ggplot2
resource_files: |
---

```{r setup}
#| include: false

# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
  fig.path = "../man/figures/art-001-case-goals-",
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 1 of a case study in three parts, illustrating how we work with longitudinal Student Unit Records (SUR). 

1. Goals (this page). Introducing the study.   

2. [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.

3. [Results.](art-003-case-results.html) Calculating summary statistics and metrics and displaying the results. 






## Definitions

student unit records (SUR)

: Registrar’s student-level data compiled by an institution and anonymized and curated by the MIDFIELD data steward. SURs include personal information (e.g., age, sex, and race/ethnicity), enrollment information (e.g., institution, program, and grade level), and academic information (e.g., courses, grades, and degrees). MIDFIELD data are organized into four tables---student, course, term, and degree---keyed by student ID.

: *Caveat:* The Student Unit Records included with midfieldr and midfielddata are practice data, not research data, suitable for practice working with SURs but not for drawing inferences about program attributes or student experiences.

program stickiness

: The ratio of the number of students graduating from a program to the number of students ever enrolled in the program. 

: Stickiness is a more-inclusive alternative to graduation rate as a measure of of a program's success in attracting, keeping, and graduating their undergraduates. All students excluded by the conventional definition of graduation rate are included in the stickiness metric: part-time, transfers, admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

 




## Goals

The technical goal of the case study is to compare the program stickiness of Civil, Electrical, Industrial, and Mechanical Engineering programs with students grouped by race/ethnicity and sex. 

The meta-goal of the case study is to illustrate a typical procedure for working with SUR data. 






## What the case study does

- We focus on our procedures and underlying rationale for working with SURs. 

- For reproducibility, we provide all the necessary code. 

- We use the [data.table](https://rdatatable.gitlab.io/data.table/)  package instead of base R or the dplyr package for speed, memory efficiency, and concise transforms via the []-indexing operator.

- We use the [ggplot2](https://ggplot2.tidyverse.org/) package instead of base R or lattice (or other R packages) because: ggplot2's layered approach allows  plots to evolve (or devolve) with minimal changes to code; easily-invoked facets and legends simplify multivariate exploration; and its popularity yields extensive online help. 











## What the case study does not do

To keep the case study as brief as possible, we do not explain the details of package syntax or functions except in selected instances. One can always use the R help system to read more about a data set or function. 

Additional online resources are available, such as: 

- [midfieldr vignettes](https://midfieldr.github.io/midfieldr/articles/) for details of syntax, functions, and arguments. 

- [midfielddata](https://midfieldr.github.io/midfielddata/) home page 

- [data.table](https://rdatatable.gitlab.io/data.table/) home page

- [ggplot2](https://ggplot2.tidyverse.org/) home page 






## References

<div id="refs"></div>






```{r}
#| echo: false

# to change the CSS file
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding:     10px 20px;
    margin:      0 0 20px;
    border-left: 0px
}
caption {
    color:       #525252;
    text-align:  left;
    font-weight: normal;
    font-size:   medium;
    line-height: 1.5;
}
```
