---
title: "Case study: Goals"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study: Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
  @Dowle+Srinivasan:2021:data.table, @Wickham:2016:ggplot2
resource_files: |
---

```{r setup}
#| include: false

# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
  fig.path = "../man/figures/art-001-case-goals-",
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 1 of a case study in three parts, illustrating how we work with longitudinal Student Unit Records (SUR). 

1. Goals (this page). Introducing the study.   

2. [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.

3. [Results.](art-003-case-results.html) Calculating summary statistics and metrics and displaying the results. 






## Definitions

student unit records (SUR)

: Registrar’s student-level data compiled by an institution and anonymized and curated by the MIDFIELD data steward. SURs include personal information (e.g., age, sex, and race/ethnicity), enrollment information (e.g., institution, program, and grade level), and academic information (e.g., courses, grades, and degrees). MIDFIELD data are organized into four tables---student, course, term, and degree---keyed by student ID.

program stickiness

: The ratio $(S)$ of the number of students graduating from a program $(N_g)$ to the number of students ever enrolled in the program $(N_e)$.
$$
S=\frac{N_g}{N_e}
$$
: Stickiness is a more-inclusive alternative to graduation rate as a measure of of a program's success in attracting, keeping, and graduating their undergraduates. All students excluded by the conventional definition of graduation rate are included in the stickiness metric: part-time, transfers, admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

 




## Goals

The technical goal of the case study is to compare the program stickiness of Civil, Electrical, Industrial, and Mechanical Engineering programs with students grouped by race/ethnicity and sex. 

The meta-goal of the case study is to illustrate a typical procedure for working with SUR data. 






## What the case study does

- We focus on our procedures and underlying rationale for working with SURs. 

- For reproducibility, we provide all the necessary code. 

- We use the [data.table](https://rdatatable.gitlab.io/data.table/)  package instead of base R or the dplyr package for speed, memory efficiency, and concise transforms via the []-indexing operator.

- We use the [ggplot2](https://ggplot2.tidyverse.org/) package instead of base R or lattice (or other R packages) because: ggplot2's layered approach allows  plots to evolve (or devolve) with minimal changes to code; easily-invoked facets and legends simplify multivariate exploration; and its popularity yields extensive online help. 









## What the case study does not do


- For brevity in the case study, we do not explain the details of midfieldr syntax, functions, or arguments. Those details are covered in other  [vignettes](https://midfieldr.github.io/midfieldr/articles/). 

- We do not explain data.table or ggplot2 syntax except for selected functions. Such details are available in numerous online resources including the [data.table](https://rdatatable.gitlab.io/data.table/) and [ggplot2](https://ggplot2.tidyverse.org/) home pages. 







## References

<div id="refs"></div>

```{r}
#| echo: false

# to change the CSS file for block quotes
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    border-left: 0px
}
```

