---
title: "Tabulating data"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Tabulating data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-060-tabulating-data-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```

## Introduction 

 
### Outline

- Practice tabulating multiway data  

### In this vignette

midfieldr functions 
 
- `condition_multiway()` 

packages

```{r}
# packages used
library("midfieldr")
library("midfielddata")

library("ggplot2")
library("janitor")
library("data.table")
```

We can control how a data.table is printed to the screen with the following options.  

```{r}
# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

```{r echo = FALSE}
DT <- wrapr::build_frame(
  "program", "race_sex", "stick", "med_program", "med_race_sex" |
    "Civil", "Asian Female", 70.6, 58.9, 70.85 |
    "Civil", "Black Female", 51.9, 58.9, 56.95 |
    "Civil", "White Female", 70.5, 58.9, 65.1 |
    "Civil", "Asian Male", 64.9, 58.9, 63.9 |
    "Civil", "Black Male", 43.9, 58.9, 45.9 |
    "Civil", "Hispanic/Latinx Male", 58.9, 58.9, 60.35 |
    "Civil", "International Male", 42.9, 58.9, 52.95 |
    "Civil", "Other/Unknown Male", 31.2, 58.9, 50 |
    "Civil", "White Male", 68.3, 58.9, 65.1 |
    "Electrical", "Asian Female", 71.1, 62.45, 70.85 |
    "Electrical", "Black Female", 61.7, 62.45, 56.95 |
    "Electrical", "Hispanic/Latinx Female", 71.4, 62.45, 72.05 |
    "Electrical", "White Female", 63.2, 62.45, 65.1 |
    "Electrical", "Asian Male", 64.5, 62.45, 63.9 |
    "Electrical", "Black Male", 45.1, 62.45, 45.9 |
    "Electrical", "Hispanic/Latinx Male", 55.2, 62.45, 60.35 |
    "Electrical", "International Male", 59.2, 62.45, 52.95 |
    "Electrical", "Other/Unknown Male", 50, 62.45, 50 |
    "Electrical", "White Male", 63.7, 62.45, 65.1 |
    "Industrial", "Asian Female", 50, 59.6, 70.85 |
    "Industrial", "Black Female", 58.5, 59.6, 56.95 |
    "Industrial", "White Female", 60.7, 59.6, 65.1 |
    "Industrial", "Asian Male", 63.3, 59.6, 63.9 |
    "Industrial", "Black Male", 53.8, 59.6, 45.9 |
    "Industrial", "Hispanic/Latinx Male", 88, 59.6, 60.35 |
    "Industrial", "International Male", 54.5, 59.6, 52.95 |
    "Industrial", "White Male", 65, 59.6, 65.1 |
    "Mechanical", "Asian Female", 82.6, 61.8, 70.85 |
    "Mechanical", "Black Female", 55.4, 61.8, 56.95 |
    "Mechanical", "Hispanic/Latinx Female", 72.7, 61.8, 72.05 |
    "Mechanical", "White Female", 67, 61.8, 65.1 |
    "Mechanical", "Asian Male", 63.1, 61.8, 63.9 |
    "Mechanical", "Black Male", 46.7, 61.8, 45.9 |
    "Mechanical", "Hispanic/Latinx Male", 61.8, 61.8, 60.35 |
    "Mechanical", "International Male", 51.4, 61.8, 52.95 |
    "Mechanical", "Native American Male", 57.1, 61.8, 57.1 |
    "Mechanical", "Other/Unknown Male", 58.3, 61.8, 50 |
    "Mechanical", "White Male", 65.2, 61.8, 65.1
)
setDT(DT)
```


## Tabulating data

Data tables are useful in discussions with one's research team and are often included in publications for readers who want to see the values underlying the data graphics. 

Our example data are currently laid out in "block record" form [@Mount+Zumel:2019:fluid-data], where the `race_sex` key and the `program` key are both in columns and the response, stickiness, is in the last column. This layout is also called "tidy data"  [@Wickham+Grolemund:2017]. 

The graphing package we use, ggplot2, is designed to work most effectively with block-record data, thus the block-record ("long") form is the default layout in midfieldr. 

```{r echo = FALSE}
temp <- copy(DT)
kable2html(temp, caption = "Table 1: Stickiness (block records)")
```
<br>

However, the visual convention preferred by publishers and readers is a row-record ("wide") form, as illustrated by Table 2. The `race_sex` key is in the left column and the `program` key is in the column names. The stickiness findings lie at the row-column intersections. Note that in this form, the four NA observations we mentioned earlier are explicitly shown. 

```{r echo = FALSE}
temp[, race_sex := as.character(race_sex)]
temp[, program := as.character(program)]
temp <- dcast(temp, race_sex ~ program, value.var = "stick")
kable2html(temp, caption = "Table 2: Stickiness (row records)")
```

Depending on your software background, you may have encountered a block-record/row-record transformation using any of these function pairs---a list adapted from [@Mount+Zumel:2019:fluid-data].  

- pivot / anti-pivot or crosstab / shred (databases)
- pivot_to_rowrecs / unpivot_to_blocks (cdata)
- reshape "wide" / reshape "long" (R)
- pivot_wider / pivot_longer (tidyr)
- pivot / unpivot (Microsoft Excel)
- dcast / melt (data.table)
- cast / melt (reshape2)

In our example, we use `dcast()` from the data.table package to reshape the data for publication. We start by limiting the significant figures of floating-point numbers. 

```{r}
# create a new memory location
block_form <- copy(DT)

# limit significant digits
block_form[, stick := round(stick, 2)]
```

If in the final result we want rows and columns ordered alphabetically, we convert factors to characters. 

```{r}
# create a new memory location
row_form <- copy(block_form)

# convert factors to characters
row_form[, race_sex := as.character(race_sex)]
row_form[, program := as.character(program)]
```

`dcast()` is used to reshape the data frame from block-record form to row-record form. 

```{r}
# reshape
row_form <- dcast(row_form, race_sex ~ program, value.var = "stick")

# examine the result
row_form
```

This result is the source for Table 2, repeated below. 

```{r echo = FALSE}
kable2html(temp, caption = "Table 2: Stickiness (row records)")
```


Other good tools for reshaping are provided in packages such as `reshape()` in base R, cdata  [@Mount+Zumel:2020:cdata], and tidyr [@Wickham+Henry:2020:tidyr]. 


### Exercise

1. Create a table of `study_grad_rate` data similar to the table above. 







## Tables with totals

When data have not been summarized, as in the `study_student` data set for example, 

```{r}
study_student
```

and what we want is a count of students by group, the result is a frequency table. The janitor package has convenient functions for creating frequency tables and adding row and column totals.

```{r}
# To access the tabyl() and adorn_total() functions
library("janitor")

# Create a new memory location
DT <- copy(study_student)
```

We join the race/ethnicity and sex variables to construct our usual categories, 

```{r}
# Merge the groups as usual
DT[, race_sex := paste(race, sex)]

# Examine the result
DT
```

Use `tabyl()` to create the two-way frequency table of counts,  

```{r}
# Create the two-way frequency table
DT <- tabyl(DT, race_sex, program)

# Examine the result
DT
```

Use `adorn_totals()` to add row totals, column totals, or (as shown here) both, 

```{r}
# Add row and column totals
DT <- adorn_totals(DT, where = c("row", "col"))

# Examine the result
DT
```









## References

<div id="refs"></div>






## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# packages used
library("midfieldr")
library("data.table")
library("ggplot2")

# optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

```{r echo = FALSE}
# to change the CSS file for block quotes
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 12px;
    border-left: 0px solid
}
```



