---
title: "Case study: Results"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/information-science-and-technology.csl
vignette: >
  %\VignetteIndexEntry{Case study: Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
nocite: | 
resource_files: |
---

```{r setup, include = FALSE}
# code chunks
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = FALSE,
  comment = "#>",
  error = FALSE
)

# figures
knitr::opts_chunk$set(
    fig.path = "../man/figures/art-003-case-results-", 
    fig.width = 6,
    fig.asp = 1 / 1.6,
    out.width = "70%",
    fig.align = "center"
)

# inline numbers
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

Part 3 of a case study in three parts:

- [Goals.](art-001-case-goals.html) Describing the study parameters.  
- [Data.](art-002-case-data.html) Transforming the data to yield the observations of interest.
- [Results.](art-003-case-results.html) Calculating summary statistics and metrics and displaying the results. 

## Preface

Our goal in this segment is to group and summarize the observations we saved previously, calculate the stickiness metric, and display the results. 

If you are writing your own script to follow along, we start with these packages: 

```{r}
# Packages
library("midfieldr")
suppressPackageStartupMessages(library("data.table"))
suppressPackageStartupMessages(library("ggplot2"))

# Printing options for data.table
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

## Import data

**Copying.** Copy the data file `study_observations` containing the final set of case study "ever" and "grad" observations as developed in the previous segment. These data are bundled with midfieldr.

```{r}
# Copy the case study observations
DT <- copy(study_observations)
DT[]
```

View the data set help page by running

    ? study_observations
    
## Group and summarize

To calculate stickiness, we want to count the number of students ever enrolled in each of our programs and the number of students graduating from the programs, grouped by race/ethnicity and sex.

These grouping variables are columns in the data set (`race`, `sex`, `program`, and `group`)---a structure we designed to support this specific  analytical goal. We can now group and summarize in one line of code. 

```{r}
# Group and summarize
DT <- DT[, .N, by = c("group", "program", "race", "sex")]
DT[]
```

## Compute metric

Stickiness is the ratio of the number of graduates to the number of students ever enrolled in a program. To be able to divide `grad / ever` we would like these two values to be in the same row. The `dcast()` function accomplishes that for us. 

This operation is essentially a transformation from block records to row records---a process known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data].  This step  leaves the graphing variables (program, race/ethnicity, and sex) in place. 

```{r}
# Transform to row-record form
DT <- dcast(DT, program + sex + race ~ group, value.var = "N")
DT[]
```

In the `grad` column, a value of NA actually represents a count of zero graduates, so we convert NA to zero. 

```{r}
# Translate NA to zero graduates 
DT <- DT[is.na(grad), grad := 0]
DT[]
```

We calculate percent stickiness. 

```{r}
DT[, stick := round(100 * grad / ever, 1)]
DT[]
```

In this form, the data re bundled with midfieldr as the `study_stickiness` data set. You can view its help page by running 

    ? study_stickiness

## Prepare for dissemination

We take several additional steps to prepare the data for dissemination in  tables or charts.  

**Filtering.** To preserve the anonymity of the people involved, we remove observations with 10 or fewer graduates.   

```{r}
# Preserve anonymity 
DT <- DT[grad >= 10]
DT[]
```

**Filtering.** Let us assume that our study focuses on "domestic" students of known race/ethnicity. In that case, we omit observations labeled "International" and Other/Unknown". 

```{r}
# Filter by study design
DT <- DT[!race %chin% c("International", "Other/Unknown")]
DT[]
```

**Creating variables.** We have found it useful to report such data with a variable that combines race/ethnicity and sex. 

```{r}
DT[, race_sex := paste(race, sex)]
setcolorder(DT, c("program", "race_sex"))
DT[]
```

**Recoding values.** Readers can more readily interpret our ta

```{r}
DT[, program := fcase(
    program %like% "CE", "Civil Engineering", 
    program %like% "EE", "Electrical Engineering", 
    program %like% "ME", "Mechanical Engineering", 
    program %like% "ISE", "Industrial/Systems Engineering"
)]
DT[]
```

With one quantitative variable (stickiness) for every combination of the levels of two categorical variables (program and race/ethnicity/sex), these data are *multiway data* [@Cleveland:1993]. In a faceted chart of multiway data, the order of the rows and panels is critical for visualizing effects. 

**Conditioning.** Convert the two categorical variables to ordered factors to support the ordering of rows and panels in the chart. 

```{r}
# Convert categorical variables to factors
DT <- condition_multiway(DT, 
                        categ_col = c("program", "race_sex"), 
                        quant_col = "stick", 
                        detail = TRUE, 
                        order_by  = "percent", 
                        param_col = c("grad", "ever"))
DT[]
```

The column `program_stick` determines the order of the programs in the chart; `race_sex_stick` determines the order of the race/ethnicity/sex groupings.  

## Charts

In the first multiway chart, the rows are programs and panels are people, facilitating comparisons of different program for a single group. Rows and panels are both ordered from bottom to top in order of increasing stickiness. 

```{r}
#| label: fig-stickiness-01
#| fig-asp: 1.2
ggplot(DT, aes(x = stick, y = program)) +
    facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
    geom_point()  +
    labs(x = "Stickiness (%)", y = "")
```

Alternatively, we can consider the dual chart, swapping the roles of the panels and rows.  Here the rows are people and panels are programs, facilitating comparisons of different people within a program.  Over many years of publishing research using MIDFIELD data, placing people on the rows of the multiway chart has been perhaps our most frequently used design. 

```{r}
#| label: fig-stickiness-02
#| fig-asp: 1
ggplot(DT, aes(x = stick, y = race_sex)) +
    facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
    geom_point()  +
    labs(x = "Stickiness (%)", y = "")
```


## Tables

Data tables are often need for publication. In this example, I place the ratio of grad to ever in parentheses followed by the stickiness result in percent.  

```{r}
# Select the columns I want for the table
tbl <- DT[, .(program, race_sex, grad, ever, stick)]

# Shorten the names that will appear in the header row
tbl[, program := fcase(
    program %like% "Civil Engineering", "Civil", 
    program %like% "Electrical Engineering", "Electrical", 
    program %like% "Mechanical Engineering", "Mechanical", 
    program %like% "Industrial/Systems Engineering", "Industrial/Sys"
)]

# Combine three columns into one character column 
tbl[, results := paste0("(", grad, "/", ever, ") ", stick)]

# Transform from block records to row records 
tbl <- dcast(tbl, race_sex ~ program, value.var = "results")

# Change factor to character so the rows can b alphabetized 
tbl[, race_sex := as.character(race_sex)]
tbl <- tbl[order(race_sex)]

# Edit one column header
setnames(tbl, old = "race_sex", new= "Group", skip_absent = TRUE)
```

Program stickiness in percent with the ratio (`grad/ever`) in parentheses: the number of program graduates and the number of students ever enrolled in the program. Groups with numbers below our reporting threshold are omitted or denoted NA. 

```{r}
#| echo: false
tbl |> 
    kableExtra::kbl(align = "lrrrr") |>
    kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
    kableExtra::row_spec(0, background = "#c7eae5") |>
    # kableExtra::column_spec(1, monospace = TRUE) |>
    kableExtra::column_spec(1:5, color = "black", background = "white")
```




## Closing

Starting with the 

**Write to file.**   

## References

<div id="refs"></div>







## Appendix

### Complete script

We conclude each vignette by collecting its code chunks in a single, condensed script for the convenience of those who wish to copy it into their own R file. 
