---
title: "Case study: Results"
vignette: >
  %\VignetteIndexEntry{Case study: Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| include: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-003-")
```


Part 3 of a case study in three parts, illustrating how we work with longitudinal student-level records. 

1. *Goals.* &nbsp; Introducing the study.   

2. *Data.* &nbsp; Transforming the data to yield the observations of interest.

3. *Results.* &nbsp; Summary statistics, metric, chart, and table.  


## Method 

Our goal in this segment is to group and summarize the observations we saved previously, calculate the stickiness metric, and display the results. 

```{r child = "../man/rmd/note-for-practice-only-2.Rmd"}
```





## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(data.table)
library(ggplot2)
```

*Loads with midfieldr.* &nbsp; Prepared data. View data dictionary via `?study_observations`. 

- `study_observations`  (derived in [Case study: Data](art-002-case-data.html)). 





## Group and summarize

*Initialize.* &nbsp; Assign a working data frame. 

```{r}
# Working data frame
DT <- copy(study_observations)
DT
```

The study observations data frame is designed with one column for each grouping variable: `race`, `sex`, `program`, and `bloc`. 

```{r child = "../man/rmd/define-grouping-variables.Rmd"}
```

*Summarize.* &nbsp; Count the numbers of observations for each combination of the grouping variables. 

```{r}
# Group and summarize
DT <- DT[, .N, by = c("bloc", "program", "race", "sex")]
DT
```


## Reshape

*Reshape.* &nbsp; Transform from block-record form to row-record form to set up the stickiness metric calculation. Transforms the *N* column into two columns, one for ever-enrolled and one for graduates.  This operation is essentially a transformation from block records to row records---a process known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data].  This step  leaves the graphing variables (program, race/ethnicity, and sex) in place. 

```{r}
# Prepare to compute metric
DT <- dcast(DT, program + sex + race ~ bloc, value.var = "N", fill = 0)
DT
```




## Compute the metric

```{r child = "../man/rmd/define-metric.Rmd"}
```

```{r child = "../man/rmd/define-stickiness.Rmd"}
```
\[
S = \frac{N_g}{N_e}
\]


*Create a variable.*  &nbsp; Compute stickiness.  

```{r}
# Compute the metric
DT[, stickiness := round(100 * graduates / ever_enrolled, 1)]
setkey(DT, NULL)
DT
```

*Verify built-in data.* To avoid deriving this data frame each time it is needed in other articles, the same information is provided in the `study_results` data frame included with midfieldr. Here we verify that the two data frames are identical. 

```{r}
# Demonstrate equivalence
same_content(DT, study_results)
```










## Prepare for dissemination

We take several additional steps to prepare the data for dissemination in  tables or charts.  

*Filtering.* To preserve the anonymity of the people involved, we remove observations with fewer than 10 graduates.   

```{r}
# Preserve anonymity
DT <- DT[graduates >= 10]

# Display the result
DT
```

```{r child = "../man/rmd/note-grouping-NAs.Rmd"}
```

*Filtering.* Let us assume that our study focuses on "domestic" students of known race/ethnicity. In that case, we omit observations labeled "International" and Other/Unknown". 

```{r}
# Filter by study design
DT <- DT[!race %chin% c("International", "Other/Unknown")]

# Display the result
DT
```

*Creating variables.* We have found it useful to report such data with a variable that combines race/ethnicity and sex. 

```{r}
# Create a variable
DT[, people := paste(race, sex)]
DT[, c("race", "sex") := NULL]
setcolorder(DT, c("program", "people"))

# Display the result
DT
```

*Recoding values.* Readers can more readily interpret our charts and tables if the programs are unabbreviated. 

```{r}
# Recode values for charts and tables
DT[, program := fcase(
  program %like% "CE", "Civil",
  program %like% "EE", "Electrical",
  program %like% "ME", "Mechanical",
  program %like% "ISE", "Industrial/Systems"
)]

# Display the result
DT
```

With one quantitative variable (stickiness) for every combination of the levels of two categorical variables (program and race/ethnicity/sex), these data are *multiway data* [@Cleveland:1993]. How one orders the categorical variables is critical for visualizing effects. 

```{r child = "../man/rmd/define-multiway-data.Rmd"}
``` 

*Conditioning.* Convert the two categorical variables to ordered factors to support the ordering of rows and panels in the chart.

```{r}
# Convert categorical variables to factors
DT <- order_multiway(DT,
  quantity = "stickiness",
  categories = c("program", "people"),
  method = "percent",
  ratio_of = c("graduates", "ever_enrolled")
)

# Display the result
DT
```

The column `program_stickiness` determines the order of the programs in the chart; `people_stickiness` determines the order of the race/ethnicity/sex groupings; the values in `stickiness` are the quantitative values to be graphed. 




## Charts

In the first multiway chart, the rows are programs and panels are people, facilitating comparisons of different program for a single group. Rows and panels are both ordered from bottom to top in order of increasing stickiness. 

```{r child = "../man/rmd/define-multiway-chart.Rmd"}
``` 

```{r}
#| label: fig01
#| fig-asp: 0.8
#| fig-cap: "Figure 1: Stickiness with programs on rows."

ggplot(DT, aes(x = stickiness, y = program)) +
  facet_wrap(vars(people), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = people_stickiness), linetype = 2, color = "gray60") +
  geom_point() +
  labs(x = "Stickiness (%)", y = "")
```

The vertical reference line is the overall stickiness of the people in a panel. 

Alternatively, we can consider the dual chart, swapping the roles of the panels and rows.  Here the rows are people and panels are programs, facilitating comparisons of different people within a program.  Over many years of publishing research using MIDFIELD data, placing people on the rows of the multiway chart has been perhaps our most frequently used design. 

```{r}
#| label: fig02
#| fig-asp: 0.8
#| fig-cap: "Figure 1: Stickiness with programs in panels."

ggplot(DT, aes(x = stickiness, y = people)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = program_stickiness), linetype = 2, color = "gray60") +
  geom_point() +
  labs(x = "Stickiness (%)", y = "")
```

The vertical reference line is the overall stickiness of the program in a panel.

The chart illustrates the importance of ordering the rows and panels. We would conclude that Industrial/Systems Engineering is the stickiest program of the four, followed by Civil, Mechanical, and Electrical in descending order. 

Because rows are ordered, one expects a generally increasing trend within a panel. A response greater or smaller than expected creates a visual asymmetry. For example, Asian students are asymmetrically lower in Industrial/Systems Engineering. 




## Tables

Data tables are often needed for publication. In this example, we format the data in a conventional row-record form with the groups of people in the first column labeling the rows and the program names labeling the remaining columns. 

```{r}
# Select the columns I want for the table
tbl <- DT[, .(program, people, stickiness)]

# Change factors to characters so rows/columns can be alphabetized
tbl[, people := as.character(people)]
tbl[, program := as.character(program)]

# Transform from block records to row records
tbl <- dcast(tbl, people ~ program, value.var = "stickiness")

# Edit one column header
setnames(tbl, old = "people", new = "People", skip_absent = TRUE)
```

Groups with numbers below our reporting threshold are denoted NA or omitted. 

```{r}
#| echo: false
tbl |>
  kableExtra::kbl(
    align = "lrrrr",
    caption = "Table 1: Program stickiness (%)"
  ) |>
  kableExtra::kable_paper(lightable_options = "basic", full_width = TRUE) |>
  kableExtra::row_spec(0, background = "#c7eae5") |>
  kableExtra::column_spec(1:5, color = "black", background = "white")
```








## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
