---
title: "Multiway graphs and tables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiway graphs and tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/body-and-society.csl
link-citations: yes
resource_files:
  - ../man/figures/art-01-multiway-graphs-fig1-1.png
  - ../man/figures/art-01-multiway-graphs-fig2-1.png
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = here::here("man/figures", 
                                            "art-01-multiway-graphs-"))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```

## Introduction 

We take some time in this early tutorial to create some graphs because, as Wickham and Grolemund state [-@Wickham+Grolemund:2017], 

>  Visualization is a great place to start with R programming, because the payoff is so clear; you get to make elegant and informative plots that help you understand the data. 

In particular, we introduce graphs of *multiway data* in which there is is one quantitative variable and two categorical variables. As explained by  Cleveland [-@Cleveland:1993], 

> ... the quantitative variable is a response, and the goal is to study how it depends on the categorical variables, which are factors. What distinguishes multiway data is the cross-classification of the categorical variables; there is a value of the response for each combination of levels of the two categorical variables. 

The structure of multiway data is specific: a categorical variable with $m$ levels; an independent categorical variable with $n$ levels; and a single  quantitative value associated with every combinations of levels of the two categories. Multiway data are typically organized as three columns with $m \times n$ rows. 

Note that "multiway" in our context refers to the data structure and graph design defined by Cleveland, not to the methods of analysis described by Kroonenberg [-@Kroonenberg:2008]. 

### This vignette uses

midfieldr functions 
 
- `condition_multiway()` 

packages

```{r}
# packages used
library("midfieldr")
library("data.table")
library("ggplot2")

# optional code to control data.table printing
options(datatable.print.nrows = 10, datatable.print.topn = 5)
```

## Prepare data

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@Ohland+Orr+others:2012]. The stickiness metric is developed in a separate vignette---here we focus on graphing a result. 

We computed stickiness for a case study in Civil, Electrical, Industrial, and Mechanical Engineering and saved the results as the data set, `study_stickiness`. You may view its help page by running

```
? study_stickiness
```

The data is loaded with midfieldr. 


```{r}
# examine the built-in data 
study_stickiness
```

Categorical variables are often provided as character variables. Without ordering the levels, the rows and panels of the multiway graph are ordered alphabetically---not the best design for gaining insights from the graph. 

To condition the categorical variables, we use `condition_multiway()` to convert them to factors with levels ordered by the median response values. To see its help page, run

```
? condition_multiway()
```

```{r}
# create a new memory location 
DT <- copy(study_stickiness)
```

The quantitative variable is `stick` (stickiness); the first categorical variable is `program`; the second categorical variable is race/ethnicity/sex. Before conditioning, the categorical variables are characters, 

```{r}
# first category is a character
class(DT[, program])

# second category is a character
class(DT[, race_sex])
```

We process the data frame

```{r}
# condition as multiway data
DT <- condition_multiway(DT)
```

No change to the row and column structure, 

```{r}
DT
```

But the categorical variables are now factors, with levels ordered non-alphabetically. `program` is now a factor with `r nlevels(DT[, program])` levels; `race_sex` is a factor with `r nlevels(DT[, race_sex])` levels. 

```{r}
# first category is now a factor
class(DT[, program])
levels(DT[, program])

# second category is now a factor
class(DT[, race_sex])
levels(DT[, race_sex])
```

We would expect 4 $\times$ 9 = 36 observations. However, some rows are omitted to preserve confidentiality, leaving us with `r dim(DT)[1]` observations.

The median values that determine the order of the levels can be obtained by using the `details` argument. 

```{r}
# return median stickiness by category
DT_med <- condition_multiway(DT, details = TRUE)

# programs have a median stickiness across race-sex groups  
DT_med[order(program, race_sex)]

# race_sex groupings have a median stickiness across programs 
DT_med[order(race_sex, program)]
```

## Create a multiway graph

```{r include = FALSE}
asp_ratio <- asp_ratio_mw(DT, categories = c("program", "race_sex"))
```

We use conventional ggplot2 functions to create the multiway graphs. 

By previously ordering the levels of the factors, we have structured the data so that the rows and panels of the multiway graph are ordered by the appropriate medians. The panel medians are drawn as vertical dashed lines. 

*Reminder:* These are practice data, not suitable for drawing inferences about student performance, i.e., not for research.

```{r fig1, fig.asp = asp_ratio[1]}
# create one multiway graph
ggplot(data = DT_med, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = med_program), 
             linetype = 2, 
             color = "gray70") +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", 
       y = "", 
       title = "Practice data (not for research)", 
       caption = "Source: midfielddata")
```
<br>

This graph permits a direct visual comparison of how stickiness varies by race and sex within a particular major. 

Reading a multiway graph

- We can more effectively compare values within a panel than between panels.
- Because rows are ordered, one expects a generally increasing trend within a panel. A response greater or smaller than expected creates a visual asymmetry. The interesting stories are often in these visual anomalies.
- To explore the comparison between panels, we swap the row and panel categories, creating the dual multiway.



## Create the dual multiway graph

The previous graph does not facilitate visual comparisons of members of the same race-sex group. As Cleveland says,

> Because of this asymmetry, it is often important to explore multiway data by as many multiway dot plots as there are categorical variables, with each variable assigned once to the levels.

To create the *dual multiway* graph, we swap the roles of the rows and panels. 

- `y = race_sex` becomes `y = program`
- `facet_wrap(vars(program) ...`  becomes `facet_wrap(vars(race_sex) ...` 

```{r fig2, fig.asp = asp_ratio[2]}
# create the dual multiway graph
ggplot(data = DT_med, aes(x = stick, y = program)) +
  facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = med_race_sex), 
             linetype = 2, 
             color = "gray70") +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", 
       y = "", 
       title = "Practice data (not for research)", 
       caption = "Source: midfielddata")
```
<br>

This graph permits a direct visual comparison of how stickiness varies by major for one group of students. 

Both multiways should be created for discussion. Usually one or the other will convey a more compelling visual story that gets included in the published findings. 

## Tabulate multiway data

Data tables are useful in discussions with one's research team and are often included in publications for readers who want to see the values underlying the data graphics. 

Our example data are currently laid out in "block record" form [@Mount+Zumel:2019:fluid-data], where the `race_sex` key and the `program` key are both in columns and the response, stickiness, is in the last column. This layout is also called "tidy data"  [@Wickham+Grolemund:2017]. 

The graphing package we use, ggplot2, is designed to work most effectively with block-record data, thus the block-record ("long") form is the default layout in midfieldr. 

```{r echo = FALSE}
temp <- copy(DT)
temp[, stick := round(stick, 2)]
kable2html(temp, caption = "Table 1: Stickiness (block records)")
```
<br>

However, the visual convention preferred by publishers and readers is a row-record ("wide") form, as illustrated by Table 2. The `race_sex` key is in the left column and the `program` key is in the column names. The stickiness findings lie at the row-column intersections. Note that in this form, the four NA observations we mentioned earlier are explicitly shown. 

```{r echo = FALSE}
temp[, race_sex := as.character(race_sex)]
temp[, program := as.character(program)]
temp <- dcast(temp, race_sex ~ program, value.var = "stick")
setnames(temp, 
         old = c("race_sex", 
                 "Civil Engineering", 
                 "Electrical Engineering", 
                 "Industrial Engineering", 
                 "Mechanical Engineering"), 
         new = c("Race/ethnicity/sex", 
                 "Civil", 
                 "Electrical", 
                 "Industrial", 
                 "Mechanical"))
kable2html(temp, caption = "Table 2: Stickiness (row records)")
```

Depending on your software background, you may have encountered a block-record/row-record transformation using any of these function pairs---a list adapted from [@Mount+Zumel:2019:fluid-data].  

- pivot / anti-pivot or crosstab / shred (databases)
- pivot_to_rowrecs / unpivot_to_blocks (cdata)
- reshape "wide" / reshape "long" (R)
- pivot_wider / pivot_longer (tidyr)
- pivot / unpivot (Microsoft Excel)
- dcast / melt (data.table)
- cast / melt (reshape2)

In our example, we use `dcast()` from the data.table package to reshape the data for publication. We start by limiting the significant figures of floating-point numbers. 

```{r}
# create a new memory location  
block_form <- copy(DT)

# limit significant digits
block_form[, stick := round(stick, 2)]
```

If in the final result we want rows and columns ordered alphabetically, we convert factors to characters. 

```{r}
# create a new memory location 
row_form <- copy(block_form)

# convert factors to characters
row_form[, race_sex := as.character(race_sex)]
row_form[, program := as.character(program)]
```

`dcast()` is used to reshape the data frame from block-record form to row-record form. 

```{r}
# reshape
row_form <- dcast(row_form, race_sex ~ program, value.var = "stick")

# examine the result
row_form
```

For publication, we edit the new column names. 

```{r}
# edit column names
setnames(row_form, 
         old = c("race_sex", 
                 "Civil Engineering", 
                 "Electrical Engineering", 
                 "Industrial Engineering", 
                 "Mechanical Engineering"), 
         new = c("Race/ethnicity/sex", 
                 "Civil", 
                 "Electrical", 
                 "Industrial", 
                 "Mechanical"))

# examine the result
row_form
```

This result is the source for Table 2, repeated below. 

```{r echo = FALSE}
kable2html(temp, caption = "Table 2: Stickiness (row records)")
```


Other good tools for reshaping are provided in packages such as `reshape()` in base R, cdata  [@Mount+Zumel:2020:cdata], and tidyr [@Wickham+Henry:2020:tidyr]. 

## References

<div id="refs"></div>

## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# packages used
library("midfieldr")
library("data.table")
library("ggplot2")

# data preparation
DT <- copy(study_stickiness)
class(DT[, program])
class(DT[, race_sex])
DT <- condition_multiway(DT)
class(DT[, program])
levels(DT[, program])
class(DT[, race_sex])
levels(DT[, race_sex])
DT_med <- condition_multiway(DT, details = TRUE)
DT_med[order(program, race_sex)]
DT_med[order(race_sex, program)]

# creating a multiway graph 
ggplot(data = DT_med, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = med_program), 
             linetype = 2) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", 
       y = "", 
       title = "Practice data (not for research)", 
       caption = "Source: midfielddata")

# creating the dual multiway graph
ggplot(data = DT_med, aes(x = stick, y = program)) +
  facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = med_race_sex), 
             linetype = 2) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", 
       y = "", 
       title = "Practice data (not for research)", 
       caption = "Source: midfielddata")

# creating a table for publication
block_form <- copy(DT)
block_form[, stick := round(stick, 2)]
row_form <- copy(block_form)
row_form[, race_sex := as.character(race_sex)]
row_form[, program := as.character(program)]
row_form <- dcast(row_form, race_sex ~ program, value.var = "stick")
setnames(row_form, 
         old = c("race_sex", 
                 "Civil Engineering", 
                 "Electrical Engineering", 
                 "Industrial Engineering", 
                 "Mechanical Engineering"), 
         new = c("Race/ethnicity/sex", 
                 "Civil", 
                 "Electrical", 
                 "Industrial", 
                 "Mechanical"))

```


