---
title: "Timely completion"
author: "Richard Layton"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../inst/REFERENCES.bib
output: rmarkdown::html_vignette
csl: ../inst/ieee-with-url.csl
vignette: >
  %\VignetteIndexEntry{Timely completion}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
nocite: | 
resource_files: 
  - ../man/figures/art-05-timely-completion-fig1-1.png
  - ../man/figures/art-05-timely-completion-fig2-1.png
  - ../man/figures/art-05-timely-completion-fig3-1.png
---

```{r include = FALSE}
knitr::opts_chunk$set(fig.path = here::here(
  "man/figures",
  "art-05-timely-completion-"
))
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```






## Introduction

Program completion---a key component of most persistence measures---must be timely, that is, occur within a specific span of years. Our default set point is six years---150% of the typical "scheduled" span of four years.  

Students who complete their programs later than their *timely completion term* (introduced in the [Data sufficiency vignette](art-04-data-sufficiency.html)). are reclassified as non-graduates in most persistence metrics.  

In this vignette, we introduce the *timely completion criterion* and related midfieldr functions. 

timely completion (*TC*) term 
: The last term in which program completion would be considered timely for a given student. For example, the *TC* term in many cases is defined as the admission term plus 6 years.     

timely completion criterion
: Graduates completing their program later than their *TC* term are grouped with non-graduates for purposes of computing a persistence metric. 

In the figure we compare the history of two students: both enter in Fall 2010 and both graduate in Spring 2015. We assume a basis of 6 years for timely completion. 

- Student A is a first-time-in-college student. Their *TC* term is 6 academic years later, in Spring 2016. 
- Student B is a transfer student, entering as a junior. Having already satisfied 2 years of program requirements, their *TC* term is 4 years later, in Spring 2014.


<br>
```{r fig1, echo = FALSE, fig.asp = 5/10}
library("ggplot2")
library("data.table")
callout_color <- "gray60"
callout_line_size <- 0.3
anno_size <- 3.5 # 3.5 approx 10 point
vert_baseline <- 1.07

y1 <- 1.5
y2 <- 1.7
y_lim <- c(1.3, 1.8)

x1 <- 1990
x2 <- 1993
x3 <- 1994
x4 <- 1995
x_lim <- c(x1 - 1.5, x4 + 1.5)

# arrows
df_arrow <- wrapr::build_frame(
  "x", "x_end", "y" |
    x1, x4, y2 |
    x1, x2, y1
)
setDT(df_arrow)
df_arrow[, x_mid := (x + x_end) / 2]

# dimension lines
del <- 0.03
dim_lines <- wrapr::build_frame(
  "x", "y", "y_end" |
    x1, y1, y2 |
    x2, y1, y1 |
    x4, y2, y2 |
    x3, y1, y2
)
setDT(dim_lines)
dim_lines[, `:=`(y = y - 2 * del, y_end = y_end + del)]

# grad markers
grad <- wrapr::build_frame(
  "x", "y" |
    x3, y1 |
    x3, y2
)

ggplot(data = df_arrow) +
  scale_x_continuous(
    limits = x_lim,
    breaks = c(1990, 1993, 1994, 1995),
    labels = c("Fall 2010", "Spr 14", "Spr 15", "Spr 16")
  ) +
  scale_y_continuous(limits = y_lim) +
  labs(
    y = "",
    x = "Term",
    title = "Illustrating timely completion"
  ) +
  theme_light() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  # label start of student row
  annotate("text",
    x = df_arrow$x - 0.6,
    y = df_arrow$y,
    size = anno_size,
    label = c("student A", "student B"),
    vjust = 0.5,
    hjust = 1
  ) +
  # arrows
  geom_segment(aes(x = x, xend = x_end, y = y, yend = y),
    color = callout_color,
    size = callout_line_size,
    arrow = arrow(
      type = "closed",
      ends = "both",
      length = unit(2, "mm")
    ),
    arrow.fill = callout_color
  ) +
  # label arrows
  annotate("text",
    x = df_arrow$x_mid,
    y = df_arrow$y,
    size = anno_size,
    label = "timely span",
    vjust = -0.5
  ) +
  # dimension lines
  geom_segment(
    data = dim_lines,
    aes(x = x, xend = x, y = y, yend = y_end),
    color = callout_color,
    size = callout_line_size,
    linetype = c(1, 1, 1, 2)
  ) +
  # label lines
  annotate("text",
    x = dim_lines$x,
    y = dim_lines$y,
    size = anno_size,
    label = c(
      "enter",
      "TC term",
      "TC term",
      "degree term"
    ),
    vjust = 1,
    hjust = c(-0.1, 1.1, -0.1, -0.1)
  ) +
  # grad points
  geom_point(
    data = grad, aes(x = x, y = y),
    size = 2
  )
```
<br>

Student A's degree term is before their *TC* term. Their completion is timely and they would be counted among the program graduates. 

However, Student B's degree term is after their *TC* term. Their completion is not timely and their graduation classification (for purposes of computing the persistence metric only) would be converted from graduate to non-graduate. 


### Outline

- Start with the case study students (`study_students`)
- Determine whether completion is timely 
- Summarize the pre-metric criteria 
- Explore: Compare programs by completion status

### This vignette uses

midfieldr functions 

- `add_completion_timely()`  
- `add_timely_term()`
- `condition_multiway()`


packages

```{r}
# Packages used
library("midfieldr")
library("midfielddata")
library("data.table")
library("ggplot2")

# Optional code to control data.table printing
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5,
  datatable.print.class = TRUE
)
```

data

```{r}
# Load data tables from midfielddata
data(student, term, degree)
```








## Case study students

Having applied the data sufficiency criterion to our case study programs, we saved the result in the midfieldr package as the data set `study_students`. These students all have timely completion terms within the range of data provided by their institutions. 

Review its help page by running

```r
? study_students
```

Review the data frame, 

```{r}
study_students
```






## Determine whether completion is timely

We use `add_timely_term()` to estimate the timely completion term. See the [Data sufficiency vignette](art-04-data-sufficiency.html) for additional details about this function.  

```{r}
# Estimate the timely completion term
DT <- add_timely_term(study_students, midfield_term = term)
DT
```

Now that our working data frame has a `timely_term` column, we use `add_completion_timely()` to add a logical variable (TRUE/FALSE) that answers the question, "Is a program completion timely?" View its help page by running, 

```r
? add_completion_timely
```
This function has two required arguments and one optional `details` argument. The input `dframe` must have a `timely_term` column.  The details, in this case, add two extra columns to show whether a program was completed, and if so, in what term. 

```{r}
# Add the completion timely column  
DT <- add_completion_timely(
  dframe = DT, 
  midfield_degree = degree, 
  details = TRUE
  )

# Examine the result
DT
```

Interpreting the result: 

- `TRUE` in the `completion_timely` column indicates that a student completed their program (`completion = TRUE`) and their degree term is no later than their timely completion term (`term_degree <= timely_term`). 
- `FALSE` in the `completion_timely` column indicates non-completion (`completion = FALSE`) or a degree term later than the timely completion term (`term_degree > timely_term`). 

With details FALSE, only the `completion_timely` column is added.

```{r}
# With details FALSE 
DT <- add_completion_timely(
  dframe = DT, 
  midfield_degree = degree, 
  details = FALSE
  )

# Examine the result
DT
```

Like the other midfieldr `add_*` functions, existing columns are deleted or overwritten if they have the same name as the variables to be added. In this case, the `term_degree` and `completion` columns are deleted and the `completion_timely` column is overwritten.  





### Exercises

If we start with the result from above and add the details, 

```{r}
# Assign a new memory location 
x <- copy(DT)

# Add the details
x <- add_completion_timely(
  dframe = x, 
  midfield_degree = degree, 
  details = TRUE
  )
```

Then order the rows and examine the last 20 rows only, 

```{r}
# Examine the final 20 rows
options(datatable.print.topn = 20)
x <- x[order(cip6)]
x <- x[8180:8199]
```

1. Find a row in which `completion_timely` is TRUE. Confirm that the student completed their program and that their degree term precedes their *TC* term. 
2. Find a row in which `term_degree` is NA, that is, the student did not complete their program. Confirm that their `completion_timely` value is FALSE.  
3. Find a row with `completion` TRUE but `completion_timely` FALSE. Confirm that the student's degree term overshoots their *TC* term.




## Summarize the pre-metric criteria 

At this point in the case study example, we have illustrated the two criteria typically applied before developing any persistence metric, i.e., "pre-metric" criteria. Both depend on the estimate of the timely completion (*TC*) term: 

data sufficiency criterion 
: Excludes some students, limiting the study to those whose *TC* terms are within the range of data provided by their institution.  

timely completion criterion 
: Retains all students but reclassifies completers whose degree terms   overshoot their *TC* terms as non-graduates. 

Having applied both criteria, our data are ready for use in a persistence metric. 







## Explore: Compare programs by completion status

In a typical persistence metric, the untimely graduates and the non-completers are grouped together as non-graduates. However, we keep them separate in this example to explore the data. We construct a new `status` column with three possible values: 

- Timely grad
- Untimely grad
- Non-completer 

```{r}
# Need the completion column for this application 
DT <- add_completion_timely(
  dframe = DT, 
  midfield_degree = degree, 
  details = TRUE
  )
```
   
A timely graduate completed their program no later than their timely completion term. 

```{r}
# Timely grads
DT[completion == TRUE & completion_timely == TRUE, 
   status := "Timely grad"]
```

A untimely graduate completed their program after their timely completion term. 

```{r}
# Untimely grads
DT[completion == TRUE & completion_timely == FALSE, 
   status := "Untimely grad"]
```

A non-completer did not complete their program. 

```{r}
# Non-completers
DT[completion == FALSE, status := "Non-completer"]
```

Remove some unnecessary columns for less cluttered viewing.  

```{r}
# Improve viewing
cols_we_want <- c("cip6", "completion", "completion_timely", "status")
DT <- DT[order(cip6), ..cols_we_want]

# Examine the result
options(datatable.print.topn = 12)
DT
```

Recall that we have a data frame of case-study program CIP codes and names in the `study_programs` data set. To review its hep page, run

``` r
? study_programs
```

Join the program names to our working data frame and drop unnecessary columns, 

```{r}
# Join program names
DT <- merge(DT, study_programs, by = "cip6", all.x = TRUE)

# Keep columns we need for grouping and summarizing
cols_we_want <- c("status", "program")
DT <- DT[, ..cols_we_want]

# Examine the result
DT
```

Group and summarize, 

```{r}
# Count students by program and completion status 
DT <- DT[, .(N_group = .N), by = .(program, status)]
DT
```

In anticipation of the graph, we condition these multiway data with the categorical columns converted to factors with levels ordered by the median values of `N_group`. 

```{r}
# Create factors with ordered levels 
mw <- condition_multiway(DT)
mw
```

```{r echo = FALSE}
asp_ratio <- 1.2 * asp_ratio_mw(mw, categories = c("program", "status"))
```

Graph the numbers of students by completion status and program. 

```{r fig2, fig.asp = asp_ratio[1]}
# Graph the counts 
ggplot(mw, aes(x = N_group, y = status)) +
  geom_point() +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  scale_x_continuous(limits = c(0, 2000)) +
  labs(x = "Number of students", 
       y = "", 
       title = "Practice data (not for research)",
       caption = "Source: midfielddata"
  )
```
<br>

A more useful comparison might be as a percentage of the total number in each program (panel). 

```{r}
# Compute percentage of total in a program
mw[, N_program := sum(N_group), by = .(program)]
mw[, pct_of_program := round(100 * N_group / N_program, 1)]
mw
```

Repeat the graph, but with percentage of program instead of numbers per program. The rows and panels remain ordered by the factor levels. 

```{r fig3, fig.asp = asp_ratio[1]}
# Graph percentages
ggplot(mw, aes(x = pct_of_program, y = status)) +
  geom_point() +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(x = "Percentage of N students in the major", 
       y = "", 
       title = "Practice data (not for research)",
       caption = "Source: midfielddata"
  )
```
<br>

We can see that while the four programs have about the same percentage of timely graduates (just over 60%), Industrial Engineering has the lowest fraction of non-completers and a higher fraction of untimely graduates. 

The graph can't tell us why this is, but it does raise the question. While this particular question may not be significant, it does illustrate how exploratory data graphics can raise new research questions. 







## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval = FALSE}
# Packages used
library("midfieldr")
library("midfielddata")
library("data.table")
library("ggplot2")

# Timely completion 
data(student, term, degree)
DT <- add_timely_term(study_students, midfield_term = term)
DT <- add_completion_timely(
  dframe = DT, 
  midfield_degree = degree, 
  details = TRUE
  )

# Compare programs by completion status
DT[completion == TRUE & completion_timely == TRUE, 
   status := "Timely grad"]
DT[completion == TRUE & completion_timely == FALSE, 
   status := "Untimely grad"]
DT[completion == FALSE, status := "Non-completer"]
DT <- merge(DT, study_programs, by = "cip6", all.x = TRUE)
DT <- DT[, .(N_group = .N), by = .(program, status)]

# Graph counts
mw <- condition_multiway(DT)
ggplot(mw, aes(x = N_group, y = status)) +
  geom_point() +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  scale_x_continuous(limits = c(0, 2000)) +
  labs(x = "Number of students", 
       y = "", 
       title = "Practice data (not for research)",
       caption = "Source: midfielddata"
  )

# Graph percentages
mw[, N_program := sum(N_group), by = .(program)]
mw[, pct_of_program := round(100 * N_group / N_program, 1)]
ggplot(mw, aes(x = pct_of_program, y = status)) +
  geom_point() +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  scale_x_continuous(limits = c(0, 100), breaks = seq(0, 100, 20)) +
  labs(x = "Percentage of N students in the major", 
       y = "", 
       title = "Practice data (not for research)",
       caption = "Source: midfielddata"
  )
```


