---
title: "Data selection"
vignette: >
  %\VignetteIndexEntry{Data selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
    rmarkdown::html_vignette:
        highlight: "tango"
        css: "style.css"
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-141-", 
                      message = TRUE)
```

*Strategies* for data frame row/column operations.

- *base R.* Given a base R `data.frame`, we prefer the form `dframe[i, j, drop = FALSE]`, with row indices in `i`, column indices in `j`, and `drop = FALSE` to ensure a data frame is returned (when desired).

- *data.table.* Given a `data.table`-type data frame, the native form is `DT[i, j, by]`, with row indices in `i`, column operations (or any R expression) in `j` , and `by` to compute `j` expression by group.

- *dplyr.* Given a dplyr `tbl`-type data frame (tibble), we employ the pipe operator `%>%` preferred by many dplyr users. Thus, instead of `select(df, column)`, for example, we write `df %>% select(column)`. 

*Code chunks* 

- Are grouped by task and color-coded by dialect.

- Sometimes use a construction that can be used in any of the dialects, e.g., a character vector `columns_we_want` for column subsetting or a logical vector `rows_we_want` for row subsetting. 

*Results*

- Code chunks are generally followed by a display of some aspect of the result, e.g., `class()`, `str()`, `dim()`, `nrow()`, etc.

- If the three dialects return the same display, one representative result is shown; otherwise, a result is shown for each. 

## Getting started

Load packages. Because dplyr masks certain functions in data.table and base R, we have to be careful to select the right package if we use any of those functions. 

```{r}
library(midfielddata)
library(data.table)
library(dplyr)
```

For future reference, the package version numbers (as of this writing) are: 

```{r}
packageVersion("midfielddata")
packageVersion("data.table")
packageVersion("dplyr")
```

Load the `student` data set from the midfielddata package and  examine its contents.

```{r}
data(student)
student
```

We format the dataset three ways:

```{r class.source = "base"}
# base R
student_df <- data.frame(student)
```

```{r class.source = "dtbl"}
# data.table
student_dt <- copy(student)
```

```{r class.source = "dplyr"}
# dplyr
student_tbl <- tibble(student)
```

```{r}
# Results
class(student_df)
class(student_dt)
class(student_tbl)
```

## Subsetting columns

***One column, return a vector***

In all dialects, one may use the `$` or `[[` operators

```{r}
#| eval: false
# Equivalent forms
result <- dframe$column_name
result <- dframe[["column_name"]]
```

Alternatively, 

*base R.* If only one column is named, a vector is returned by default.

```{r class.source="base"}
# base R
result <- student_df[, "mcid"]
# or
result <- student_df[, "mcid", drop = TRUE]
```

*data.table.* Placing an *unquoted* column name in `j` extracts one column as a vector. 

```{r class.source="dtbl"}
# data.table
result <- student_dt[, mcid]
```

*dplyr.* To use `pull()` safely, use a *quoted* column name even though unquoted column names are conventional usage in dplyr. 

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    pull("mcid")
```

`dplyr::glimpse()` is similar to `str()` or a transposed version of `print()`

```{r}
# All dialects yield:
glimpse(result)
```

*For more information.* For a given task, whether a name is in plain code or in quotes depends on how a system evaluates the name: in standard evaluation (SE) like most of base R, or in nonstandard evaluation (NSE) like data.table and dplyr. For a brief but complete discussion of the difference, see [@Mount:2019].


<br>

***One column, return a data frame***

*base R.* Note the use of `drop = FALSE` to ensure a data frame is returned when a single column is selected. 

```{r class.source="base"}
# base R
result <- student_df[, "mcid", drop = FALSE]
```

*data.table.* Placing an *unquoted* column name in `.()` retains that column and omits all others. 

```{r class.source="dtbl"}
# data.table
result <- student_dt[, .(mcid)]
```

*dplyr.* Placing an *unquoted* column name in `select()` retains that column and omits all others.

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(mcid)
```

```{r}
# All dialects yield:
glimpse(result)
```

***Several columns***

```{r class.source="base"}
# base R
result <- student_df[, c("mcid", "institution", "race", "sex"), drop = FALSE]
```

```{r class.source="dtbl"}
# data.table
result <- student_dt[, .(mcid, institution, race, sex)]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(mcid, institution, race, sex)
```

```{r}
# All dialects yield:
glimpse(result)
```


***Character vector of column names***

```{r}
# Character vector of column names, all dialects 
columns_we_want <- c("mcid", "institution", "race", "sex")
```

```{r class.source="base"}
# base R
result <- student_df[, columns_we_want, drop = FALSE]
```

*data.table.* Refer to a vector of column names using `..` notation. 

```{r class.source="dtbl"}
# data.table
result <- student_dt[, ..columns_we_want]
```

*dplyr.* Refer to a vector of column names using `all_of()` or `any_of()` inside `select()`. `all_of()` returns an error if a requested column name does not exist; `any_of()` does not. 

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(all_of(columns_we_want))
```



```{r}
# All dialects yield:
glimpse(result)
```




***By partial-pattern matching to column names***

Example: Subset columns with "sat" in the column name. 

```{r}
# Character vector, subset of column names, all dialects 
all_columns <- colnames(student_df)
columns_we_want <- all_columns[grepl("sat", all_columns)]
```

Using a vector of column names, similar to the prvious example, 

```{r class.source="base"}
# base R
result <- student_df[, columns_we_want, drop = FALSE]
```

```{r class.source="dtbl"}
# data.table
result <- student_dt[, ..columns_we_want]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(all_of(columns_we_want))
```

Alternatively, using pattern matching provided by data.table or dplyr,  

```{r class.source="dtbl"}
# data.table
result <- student_dt[, .SD, .SDcols = patterns("sat")]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(matches("sat"))
```

*Comments*

- *base R.* Use `grepl()` to pattern-match column names. Accepts regular expressions if needed.

- *data.table.* Can use `patterns()` with `.SDcols`.

- *dplyr.* Similarly, can use `matches()` with `select()`. 


```{r}
# All dialects yield:
glimpse(result)
```
 

***By variable class***

Example: Subset numeric columns only. 

```{r class.source="base"}
# base R
result <- Filter(is.numeric, student_df)
```

```{r class.source="dtbl"}
# data.table
result <- student_dt[, .SD, .SDcols = is.numeric]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(where(is.numeric))
```

*Comments*

- *base R.* `Filter(f, x)`---when 
$\small x$ is a data frame---returns the columns for which function $\small f$ returns TRUE. 

- *data.table.* `.SD` in `j` is a subset of the data with columns specified by `.SDcols`. 

- *dplyr.* `where(f)` is a helper to choose the variables for which a function $\small f$---applied to columns---returns TRUE.


```{r}
# All dialects yield:
glimpse(result)
```


***Excluding columns***

In all dialects, one may use the `$` operator to exclude a single column, 

    dframe$column_name <- NULL

For excluding multiple columns, 

```{r class.source="base"}
# base R
result <- student_df
result[ , c("us_citizen", "high_school")] <- list(NULL)
```

```{r class.source="dtbl"}
# data.table
result <- copy(student_dt)
result[, c("us_citizen", "high_school") := NULL]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(-us_citizen, -high_school)
```

*Comments*

- *base R.* Assigned data frame to `result` before using `NULL` to preserve `student_df` unaltered.

- *data.table.*  Similarly, made a `copy()` because `:=` makes changes in place. 

- *dplyr.* Minus sign `-` to exclude a column. 

Alternatively, using a vector column names, 

```{r}
# Character vector of column names, all dialects 
columns_to_exclude <- c("us_citizen", "high_school")
```

```{r class.source="base"}
# base R
columns_we_want <- !names(student_df) %in% columns_to_exclude
result <- student_df[, columns_we_want, drop = FALSE]
```

```{r class.source="dtbl"}
# data.table
result <- student_dt[, !..columns_to_exclude]
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(-all_of(columns_to_exclude))
```

*Comments*

- *base R.* Use `!` to preserve names not in the `columns_to_exclude` vector. 

- *data.table.*  Similar.

- *dplyr.* Again, `all_of()` for a vector of names, but with a minus sign. 

```{r}
# All versions yield:
glimpse(result)
```
 




## Subsetting rows

***By one Boolean condition***

```{r class.source="base"}
# base R
rows_we_want <- student_df$institution == "Institution B"
result <- student_df[rows_we_want, , drop = FALSE]
```

```{r}
nrow(result)
```

```{r class.source="dtbl"}
# data.table
rows_we_want <- student_dt$institution == "Institution B"
result <- student_dt[rows_we_want, ]
```

```{r}
nrow(result)
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    filter(institution == "Institution B")
```

```{r}
nrow(result)
```

*Comments*

- *base R.* 

- *data.table.* 

- *dplyr.* 

<br>

***By multiple Boolean conditions***

```{r class.source="base"}
# base R
rows_we_want <- with(student_df, transfer == "First-Time Transfer"  & hours_transfer >= 15)
result <- student_df[rows_we_want, , drop = FALSE]
```

```{r}
nrow(result)
```

```{r class.source="dtbl"}
# data.table
rows_we_want <- with(student_dt, transfer == "First-Time Transfer"  & hours_transfer >= 3)
result <- student_dt[rows_we_want, ]
```

```{r}
nrow(result)
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    filter(transfer == "First-Time Transfer", hours_transfer >= 3)
```

```{r}
nrow(result)
```

*Comments*

- *base R.* 

- *data.table.* 

- *dplyr.* 

<br>

## Subsetting columns and rows

***Columns by name, rows by Boolean selection***

```{r class.source="base"}
# base R
columns_we_want <- c("mcid", "institution", "transfer", "hours_transfer")
rows_we_want <- student_df$transfer == "First-Time Transfer"
result <- student_df[rows_we_want, columns_we_want, drop = FALSE]
```

```{r}
dim(result)
```

```{r class.source="dtbl"}
# data.table
columns_we_want <- c("mcid", "institution", "transfer", "hours_transfer")
rows_we_want <- student_dt$transfer == "First-Time Transfer"
result <- student_dt[rows_we_want, ..columns_we_want]
```

```{r}
dim(result)
```

```{r class.source = "dplyr"}
# dplyr
result <- student_tbl %>%
    select(mcid, institution, transfer, hours_transfer) %>%
    filter(transfer == "First-Time Transfer")
```

```{r}
result
```

*Comments*

- *base R.* 

- *data.table.* 

- *dplyr.* 

<br>

## References

<div id="refs"></div>

## Appendices

### A subtle flaw

The only safe way to use `pull(y)` appears to be:

1) Ensure column `y` exists before using `pull(y)`, or
2) Use one of the argument forms `pull("y")` or `pull(.data$y)`. 

Why? Suppose we are trying to select the `student$mcid` column. 

```{r}
#| echo: true
student_tbl %>% 
    pull(mcid) %>%
    str()
```

Suppose during some operations, the ID column had been deleted and at some later point we tried to pull it. We would expect an error. 

```{r}
#| echo: true
#| error: true
case <- student_tbl %>% 
    select(-mcid)

# An error if we select a column that doesn't exist
case %>% 
    pull(mcid)
```

So far, so good. Now suppose a variable happens to exist in the workspace that has the same name as the column name. If the data frame contains the desired  column, `pull()` operates as expected. 

```{r}
#| echo: true
mcid <- "race"

student_tbl %>% 
    pull(mcid) %>%
    str()
```

However, if the column had been deleted for some reason, `pull()` does not necessarily return an error. Instead, it treats the argument as if it were a variable whose *value* may be a column name. Thus, in this example, `pull()` returns the `student_tbl$race` column. 

```{r}
#| echo: true
#| error: true
mcid <- "race"
case <- student_tbl %>% 
    select(-mcid)

# No error! Argument treated as if it refers to a column name
case %>% 
    pull(mcid) %>%
    str()
```

In other words, `pull(mcid)` has incorrectly returned the same result as the forms below that are correct for using the value associated with the variable `mcid` in nonstandard evaluation by adding the extra de-quoting symbols `!!`

In contrast, if we want to use the value associate with x in NSE we add the extra de-quoting symbols !!, as we show here.

```{r}
case %>% 
    pull(!!mcid) %>%
    str()

case %>% 
    pull(all_of(mcid)) %>%
    str()
```

<br>

***Safe usage***

The forms `pull("y")` or `pull(.data$y)` both yield expected results under normal circumstances and the error conditions described above. 

```{r}
#| echo: true
#| error: true
# Yields expected results
mcid <- "race"
student_tbl %>% 
    pull("mcid") %>%
    str()
student_tbl %>% 
    pull(.data$mcid) %>%
    str()
# Correctly yields errors when the column name doesn't exist
mcid <- "race"
case <- student_tbl %>% 
    select(-mcid)
case %>% 
    pull("mcid")
case %>% 
    pull(.data$mcid)
```




```{r child = "../man/rmd/common-closing.Rmd"}
```

