---
title: "How to in R"
vignette: >
  %\VignetteIndexEntry{How to in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
    rmarkdown::html_vignette:
        highlight: "tango"
        css: "style.css"
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-140-", 
                      message = TRUE)
```

Tutorials demonstrating how to use R to wrangle MIDFIELD data into a shape useful for analysis.  

Many of these methods appear in other midfieldr vignettes, but only as needed to meet the specific goals of the article. Here we take the time to systematically develop basic data shaping transforms in the context of MIDFIELD data tables. 

In addition, we add one feature not developed elsewhere in the midfieldr vignettes: showing solutions in base R, data.table, and dplyr. Our goal is to meet the needs of users who prefer to use base R or dplyr syntax in lieu of the data.table syntax that appears everywhere else in midfieldr. Sources for translating among the three dialects include [@atrebas:2019], [@Gimond:2019], [@Hajnala:2018], [@Mercer:2020], [@Vaughan:2023], and  [@Zumel+Mount:2020, Ch. 5]. As Zumel and Mount say, 

> We will show data wrangling solutions using base R, data.table, and dplyr. Each of these has its advantages, which is why we are presenting more than one solution. ... we are deliberately using a polyglot approach to data wrangling, mixing base R, data.table, and dplyr, as convenient. Each of these systems has its strengths ... 

<br>









## Tutorial outline

*Data selection*

- [Subsetting columns]()
- [Subsetting rows]()
- [Reordering rows]()
- [Reordering columns]()
- [Removing rows with incomplete data]()
- [Removing columns]()

*Basic transforms*

- [Creating new columns]()
- [Renaming columns]()
- [Recoding column values]()

*Aggregating transforms*

- [Counting]()
- [Summarizing]()

*Multitable transforms*

- [Combining two datasets by row]()
- [Combining two datasets by column]()
- [Joining two datasets]()

*Reshaping transforms*

- [Tall to wide: block records "pivot" to row records]()
- [Wide to tall: row records "unpivot" to block records]()











## References

<div id="refs"></div>





```{r child = "../man/rmd/common-closing.Rmd"}
```

