---
title: "Basic transforms"
vignette: >
  %\VignetteIndexEntry{Basic transforms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-142-")
```

Demonstrating how to use base R, data.table, and dplyr to wrangle MIDFIELD data into useful forms, adapted from [@Zumel+Mount:2020, Ch. 5.1]. Topics covered: 

- Adding new columns
- Renaming columns
- Removing columns
- Recoding column values


## Scenario

Continuing our investigation: If a student were ever enrolled in Engineering at Institution B, what was their level (first-year, second-year, etc.) when passing one of the core math courses (Calculus 1, 2, 3 and Ordinary Differential Equations)?

```{r}
#| cache: true
library(midfielddata)
data(course)
```

First we reproduce the results from the previous article. 

```{r}
library(data.table)
columns_we_want <- c("mcid", "institution", "term_course", 
                     "discipline_midfield", "course", "hours_course", 
                     "grade")
rows_we_want <- course$institution == "Institution B"
course_dt <- course[rows_we_want, ..columns_we_want]

rows_we_want <- complete.cases(course_dt)
course_dt <- course_dt[rows_we_want, ]

order_cols <- c("mcid", "term_course")
setorderv(course_dt, order_cols)

course_df <- as.data.frame(course_dt)

library(dplyr)
course_dplyr <- as_tibble(course_dt)
```





## Adding new columns

### Solution 1: base R

### Solution 2: data.table

### Solution 3: dplyr







## Renaming columns

### Solution 1: base R

### Solution 2: data.table

### Solution 3: dplyr









## Removing columns

The institution column is constant (all values "Institution B") so we can safely remove it. 

### Solution 1: base R

```{r}
course_df$institution <- NULL

str(course_df)
```


### Solution 2: data.table

Remove one column

```{r}
dt <- copy(course_dt)
dt$institution <- NULL
names(dt)

dt <- copy(course_dt)
dt[, institution := NULL]
names(dt)
```

Remove multiple columns

```{r}
dt <- copy(course_dt)
dt[, c("institution", "discipline_midfield") := NULL]
names(dt)

dt <- copy(course_dt)
columns_to_delete <- c("institution", "discipline_midfield")
dt[, (columns_to_delete) := NULL]
names(dt)
```






### Solution 3: dplyr

Again, the base R approach works fine, 

        course_dplyr$institution <- NULL


```{r}


course_dplyr
```







## Recoding column values

### Solution 1: base R

### Solution 2: data.table

### Solution 3: dplyr












## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
