---
title: "Selecting CIP codes"
output:
  rmarkdown::html_vignette: 
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Selecting CIP codes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, echo = FALSE, message = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(collapse = TRUE, comment = "#>")
opts_chunk$set(warning = FALSE, message = FALSE)
opts_chunk$set(error = TRUE, purl = FALSE)
opts_chunk$set(fig.width = 6)
options(tibble.print_max = 10L, tibble.print_min = 5L)
knit_hooks$set(inline = function(x) {
	if (!is.numeric(x)) { 
		x 
		} else if (x >= 10000) { 
		prettyNum(round(x, 2), big.mark = ",")
		} else {
		prettyNum(round(x, 2))
		}
	})
```

The Classification of Instructional Programs (CIP) is a taxonomy of academic  programs curated by an agency of the US Department of Education.[^1]

[^1]: The CIP is curated by the Integrated Postsecondary Education Data System (IPEDS), the the core postsecondary education data collection program of the National Center for Education Statistics (NCES), an agency of the US Department of Education (https://nces.ed.gov/ipeds/cipcode). 



```{r}
# packages used in the vignette
library(midfieldr)
library(dplyr) 
library(stringr)
library(knitr)
```

```{r echo = FALSE}
# For formatting the output seen by the reader
# but not used by the reader
library(kableExtra)
```

## Introduction

Instructional programs are encoded in `cip`, a data frame of character variables listing the 2, 4, and 6-digit CIP codes and program names in the MIDFILED database. See the help page (`?cip`) for more detail.  

```{r}
glimpse(cip)
```

`cip_filter()` filters `cip` for the programs we wish to study. The function has two arguments:   

- `series` <span style="display:inline-block; width: 1mm;"></span> Search terms. A character vector of codes or keywords.  
- `data` <span style="display:inline-block; width: 5.5mm;"></span>A searchable data frame (`cip` is the default). 

With no arguments, `cip_filter()` returns the unique 2-digit series and the first entry of the corresponding 4- and 6-digit series. 

```{r}
cip_filter()
```

`cip_filter()` searches all columns in `data` for the strings in `series` and returns a data frame whose rows contain the search terms.

Filtering is performed by `dplyr::filter()` with pattern detection by `stringr::str_detect()`.  

## Search by keywords

Because `cip_filter()` searches all columns in `data`, we can search by keyword to find the CIP codes that match the programs under study. The search is case-sensitive.  

For example, if we wanted all Civil Engineering programs, we might try keywords "civil" and "Civil". 

```{r results = "hide"}
# series with a keyword
df <- cip_filter(series = c("Civil", "civil"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

The search returns some programs with Civilization in their names as well as Engineering Technology. If we wanted Civil Engineering only, we should search for CIPs that start with the 4-digit "1408"  code. We use the regular expression notation `^` to match the start of the strings. 

```{r results = "hide"}
# series with a 4-digit code
df <- cip_filter(series = "^1408")
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

In the next example we find that the keyword "German" appears in three different 2-digit groups of programs. 

```{r results = "hide"}
# series with a keyword
df <- cip_filter(series = c("German", "german"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

 If we were interested in German Studies specifically, we might want to limit our CIP codes  to "050125" and "160501".
 
```{r results = "hide"}
# series with a 6-digit code
df <- cip_filter(series = c("050125", "160501"))
kable(df)
```
 
```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```





## Search by CIP codes 

As illustrated above, assigning CIP codes to `series` yields a data frame of all rows from `data` that contain those character strings. Invalid codes---and valid codes not in the midfieldr `cip` dataset---are silently ignored. 

Specifying 6-digit codes yields a data frame with a row for each 6-digit code.  

```{r results = "hide"}
# 6-digit example
df <- cip_filter(series = c("140802", "140803"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

Specifying 4-digit codes yields a data frame with as many rows as there are rows that include the search strings.   

```{r results = "hide"}
# 4-digit example
df <- cip_filter(series = c("1407", "1408"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

This search, in addition to the desired Engineering programs, returned two Foreign Language programs because their 6-digit codes contain "1407" and "1408". We use the regular expression notation `^` to match the start of the strings.

```{r results = "hide"}
# 4-digit example matching the start of the string 
df <- cip_filter(series = c("^1407", "^1408"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```


The 2-digit series represent the most general groupings of related programs. Here, the result includes all History codes and programs. 

```{r results = "hide"}
# 2-digit example
df <- cip_filter("^54")
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

The series argument can include any combination of 2, 4, and 6-digit codes. 

```{r results = "hide"}
# a series with 2, 4, and 6-digits specified
df <- cip_filter(c("^24", "^4102", "^450202"))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

A series assigned numerically is coerced to characters. 

```{r results = "hide"}
# series as a sequence
df <- cip_filter(series = seq(540102, 540110, 1))
kable(df)
```

```{r echo = FALSE}
kable(df, "html") %>% kable_styling(font_size = 12)
```

Invalid codes in the sequence, e.g., 540109 and 540110, are silently ignored. 



## Search by named series

midfieldr includes several datasets of *named series*. A named series is a character vector of 6-digit CIP codes that belong to a group of programs such as Engineering, Physical Sciences, STEM, etc. To see the full set of available named series, type `?named_series`. 

A named series is assigned to the series argument in `cip_filter()`. 

```{r}
# the engineering series
engr

df <- cip_filter(series = engr)
glimpse(df)
```










## Using a previous result as a data argument 

A data frame produced by `cip_filter()` can be used as the `data` argument in a subsequent use of `cip_filter()`. 

```{r}
df1 <- cip_filter("^49")
```

Suppose, after viewing the full "49" series above we decide we only need the "4903" subset. We can use the data frame `df1` from above as the `data` argument. 

```{r results = "hide"}
# use a previous result as a data argument
df2 <- cip_filter(series = "4903", data = df1)
kable(df2) %>% kable_styling(font_size = 12)
```


```{r echo = FALSE}
kable(df2, "html") %>% kable_styling(font_size = 12)
```




## Conclusion

Having used the `cip_filter()` function to identify the codes of the programs one wishes to study, the codes are used to filter the midfieldr   datasets to obtain the records of students who matriculate, ever enroll, or graduate in those programs. 
