---
title: "Computing the stickiness metric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing the stickiness metric}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/chicago-author-date.csl
link-citations: yes
resource_files:
  - ../man/figures/vignette-compute-stickiness-fig1-1.png
---

```{r setup, include = FALSE}
source("vignette-knitr-opts.R")
knitr::opts_chunk$set(fig.path = "../man/figures/vignette-compute-stickiness-")
```

In this vignette, we demonstrate computing the "stickiness" persistence metric in the ongoing case study in four engineering programs.

## Introduction 

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@Ohland+Orr+others:2012]. 

The stickiness metric accounts for all students in a program. Unlike the graduation rate metric, stickiness includes students who begin college part-time, who enroll mid-year, and who switch majors or transfer between institutions. 

Every student counted in a program's stickiness metric was enrolled in that program for at least one term. The program succeeded in attracting these students---stickiness tells us how well the program succeeded in retaining them. 

The workflow is: 

- choose the programs to study
- count the students ever enrolled
- count the students graduating  
- compute stickiness 
- graph the multiway


```{r echo = FALSE}
library(midfieldr)
```




## Students ever enrolled in a group of programs

In other vignettes, we describe in detail the process of choosing the programs to study, counting the students ever enrolled (??), and subsetting for completion feasibility. 

We have a built-in data set `exa_ever_fc` of the `r length(exa_ever_fc)` student IDs ever enrolled in our case study programs for whom program completion within 6 years is feasible. The vignette starts with these students. 

```{r}
# packages used
library(midfieldr)

# ever enrolled and completion feasible
enrollees <- exa_ever_fc

# examine the result
str(enrollees)
```




  gather ever in this vignette??








<br>
<a href="#top">&#9650; top of page</a>   




## References

<div id="refs"></div>





## Appendix

### Issue bin

- Add a note regarding the suggested 6-student minimum. 
- Account for "possible graduation in 6 years"
- Could use mice to impute unknown values of the race and sex variables 

### Complete script

The vignette code chunks are collected below in a single, condensed script.

```{r eval=FALSE}
# TBD
```

<br>
<a href="#top"         >&#9650;     top of page </a>     
<a href="../index.html">&#9665;       main page </a> 



