---
title: "MIDFIELD data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MIDFIELD data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo = FALSE, message = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(collapse = TRUE, comment = "#>")
opts_chunk$set(warning = FALSE, message = FALSE)
opts_chunk$set(error = TRUE, purl = FALSE)
opts_chunk$set(fig.width = 6)
options(tibble.print_min = 6L, tibble.print_max = 6L)
knit_hooks$set(inline = function(x) {
	if (!is.numeric(x)) { 
		x 
		} else if (x >= 10000) { 
		prettyNum(round(x, 2), big.mark = ",")
		} else {
		prettyNum(round(x, 2))
		}
	})
```

```{r echo = FALSE}
library(midfieldr)
library(stringr)
```



## CIP data 

The midfieldr package includes `cip`, a tidy data frame with `r data_bits[["obs_cip"]]` observations and `r data_bits[["var_cip"]]` CIP variables of program codes and names at the 2, 4, and 6-digit levels. Each observation is a unique program. This dataset occupies `r round(data_bits[["size_cip"]] / 1e+3, 0)` kb of memory.  

The dataset format and variable definitions are described on the help page:

```r
library(midfieldr)
?cip
```

midfieldr also includes collections of 6-digit CIP codes denoting groups of programs. To see the help page: 

```r
?named_series
```


## Data packages

The MIDFIELD datasets, because of their size, are provided in separate data packages. The  datasets are a stratified sample of the MIDFIELD database. 

- [midfieldstudents](https://github.com/MIDFIELDR/midfieldstudents) 
- [midfieldcourses](https://github.com/MIDFIELDR/midfieldcourses) 
- [midfieldterms](https://github.com/MIDFIELDR/midfieldterms) 
- [midfielddegrees](https://github.com/MIDFIELDR/midfielddegrees) 


To install the data packages using the devtools package:   

```r
# data packages
devtools::install_github("MIDFIELDR/midfieldstudents")
devtools::install_github("MIDFIELDR/midfieldcourses")
devtools::install_github("MIDFIELDR/midfieldterms")
devtools::install_github("MIDFIELDR/midfielddegrees")
```


## Student data

The [midfieldstudents](https://github.com/MIDFIELDR/midfieldstudents) package provides `midfieldstudents`, a tidy data frame with `r data_bits[["obs_student"]]` observations and `r data_bits[["var_student"]]` demographic variables. Each observation is a unique student. This dataset occupies `r round(data_bits[["size_student"]] / 1e+6, 1)` Mb of memory. 

The dataset format and variable definitions are described on the help page:

```r
library(midfieldstudents)
?midfieldstudents
```
   
## Course  data

The [midfieldcourses](https://github.com/MIDFIELDR/midfieldcourses) package provides `midfieldcourses`, a tidy data frame with `r round(data_bits[["obs_course"]] / 1e+6, 1)` M observations and `r data_bits[["var_course"]]` academic course variables. Each observation is one course in one term for one student. This dataset occupies `r round(data_bits[["size_course"]] / 1e+6, 1)` Mb of memory. 

The dataset format and variable definitions are described on the help page:

```r
library(midfieldcourses)
?midfieldcourses
```






## Term  data

The [midfieldterms](https://github.com/MIDFIELDR/midfieldterms) package provides `midfieldterms`, a tidy data frame with `r round(data_bits[["obs_term"]] / 1e+6, 1)` M observations and `r data_bits[["var_term"]]` academic term variables. Each observation is one term for one student. This dataset occupies `r round(data_bits[["size_term"]] / 1e+6, 1)` Mb of memory.  

The dataset format and variable definitions are described on the help page:

```r
library(midfieldterms)
?midfieldterms
```


`

## Degree  data

The [midfielddegrees](https://github.com/MIDFIELDR/midfielddegrees) package provides `midfielddegrees`, a tidy data frame with `r data_bits[["obs_degree"]]` observations and `r data_bits[["var_degree"]]` graduation variables. Each observation is a unique student. This dataset occupies `r round(data_bits[["size_degree"]] / 1e+6, 1)` Mb of memory. 

The dataset format and variable definitions are described on the help page:

```r
library(midfielddegrees)
?midfielddegrees
```
    
