---
title: "MIDFIELD data"
output:
  rmarkdown::html_vignette: 
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{MIDFIELD data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/REFERENCES.bib
---

```{r echo = FALSE, message = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(collapse = TRUE, comment = "#>")
opts_chunk$set(warning = FALSE, message = FALSE)
opts_chunk$set(error = TRUE, purl = FALSE)
opts_chunk$set(fig.width = 6)
options(tibble.print_min = 6L, tibble.print_max = 6L)
knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
```

```{r echo = FALSE}
library(midfieldr)
library(stringr)

# unpack data bits
data_bits <- midfieldr:::data_bits_cip
obs_cip <- round(data_bits$data[data_bits$name == "obs_cip"] / 1e+0, 1)
var_cip <- data_bits$data[data_bits$name == "var_cip"]
size_cip <- round(data_bits$data[data_bits$name == "size_cip"] / 1e+3, 0)

data_bits <- midfieldstudents:::data_bits_students
obs_student <- round(data_bits$data[data_bits$name == "obs_student"] / 1e+0, 1)
size_student <- round(data_bits$data[data_bits$name == "size_student"] / 1e+6, 0)
var_student <- data_bits$data[data_bits$name == "var_student"]
year_min <- data_bits$data[data_bits$name == "year_min"]

library(midfieldstudents)
values_sex  <- str_c(sort(unique(midfieldstudents$sex)), collapse = ", ")
values_race <- str_c(sort(unique(midfieldstudents$race)), collapse = ", ")

data_bits <- midfielddegrees:::data_bits_degrees
obs_degree <- round(data_bits$data[data_bits$name == "obs_degree"] / 1e+0, 1)
var_degree <- data_bits$data[data_bits$name == "var_degree"]
size_degree <- round(data_bits$data[data_bits$name == "size_degree"] / 1e+6, 0)
award_degree <- round(data_bits$data[data_bits$name == "award_degree"] / 1e+0, 1)
year_max <- data_bits$data[data_bits$name == "year_max"]

data_bits <- midfieldterms:::data_bits_terms
obs_term <- round(data_bits$data[data_bits$name == "obs_term"] / 1e+0, 1)
var_term <- data_bits$data[data_bits$name == "var_term"]
size_term <- round(data_bits$data[data_bits$name == "size_term"] / 1e+6, 0)

data_bits <- midfieldcourses:::data_bits_courses
obs_course <- round(data_bits$data[data_bits$name == "obs_course"] / 1e+6, 1)
var_course <- data_bits$data[data_bits$name == "var_course"]
size_course <- round(data_bits$data[data_bits$name == "size_course"] / 1e+6, 0)
```


The midfieldr package includes Classification of Instructional Programs (CIP) data from the US Department of Education [@cip2010].

Separate data packages provide a stratified sample of student record data from the MIDFIELD database. The process of designing, compiling, maintaining, protecting, and sharing the MIDFIELD data is described in [@aee2016]. 




## Installing the data packages

Four data packages provide the stratified data sample from the MIDFIELD database. To install the data packages, 

```r
# data packages
devtools::install_github("MIDFIELDR/midfieldstudents")
devtools::install_github("MIDFIELDR/midfieldcourses")
devtools::install_github("MIDFIELDR/midfieldterms")
devtools::install_github("MIDFIELDR/midfielddegrees")
```




## CIP data 

The midfieldr package includes `cip`, a tidy data frame with `r obs_cip` observations and `r var_cip` Classification of Instructional Programs (CIP)  variables. Each observation is a unique program; the variables are the program codes and names at the 2, 4, and 6-digit levels. This dataset occupies `r size_cip` kb of memory. Open the help page by running: 

```r
library(midfieldr)
?cip
```

midfieldr also includes collections of 6-digit CIP codes denoting groups of programs. To learn more, open the help page by running `?named_series`. 









## Student data

The midfieldstudents package provides `midfieldstudents`, a tidy data frame with `r obs_student` observations and `r var_student` demographic variables, occupying `r size_student` Mb of memory. Each observation is a unique student. Open the help page by running: 

```r
library(midfieldstudents)
?midfieldstudents
```





## Course  data

The midfieldcourses package provides `midfieldcourses`, a tidy data frame with `r obs_course` M observations and `r var_course` academic course variables, occupying `r size_course` Mb of memory. midfieldcourses is the largest of the midfieldr datasets. Each observation is one course in one term for one student. Open the help page by running: 

```r
library(midfieldcourses)
?midfieldcourses
```






## Term  data

The midfieldterms package provides `midfieldterms`, a tidy data frame with `r obs_term`  observations and `r var_term` academic term variables, occupying `r size_term` Mb of memory. Each observation is one term for one student. Open the help page by running: 

```r
library(midfieldterms)
?midfieldterms
```




## Degree  data

The midfielddegrees package provides `midfielddegrees`, a tidy data frame with `r obs_degree` observations and `r var_degree` graduation variables, occupying `r size_degree` Mb of memory. Each observation is a unique student. Open the help page by running: 

```r
library(midfielddegrees)
?midfielddegrees
```
    
## References
