---
title: "Explore multiway data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore multiway data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/chicago-author-date.csl
link-citations: yes
resource_files:
  - ../man/figures/vignette-explore-multiway-fig1-1.png
  - ../man/figures/vignette-explore-multiway-fig1-dual-1.png
---

```{r setup, include = FALSE}
source("vignette-knitr-opts.R")
knitr::opts_chunk$set(fig.path = "../man/figures/vignette-explore-multiway-")
```


`order_multiway()` is used to transform multiway data such that the two categorical variables are factors with levels ordered by the third, quantitative variable. In this vignette we explore its arguments and continue the ongoing case study in four engineering programs.


## Introduction

Many persistence metrics can be organized as *multiway data* in which there is is one quantitative variable and two categorical variables. To quote Bill Cleveland [-@Cleveland:1993], 

> ... the quantitative variable is a response, and the goal is to study how it depends on the categorical variables, which are factors. What distinguishes multiway data is the cross-classification of the categorical variables; there is a value of the response for each combination of levels of the two categorical variables. 





## Preparing the multiway data 

We'll use the "stickiness" metric to illustrate multiway graphs and data tables. Stickiness is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@Ohland+Orr+others:2012]. 

`exa_stickiness` is a sample dataset included with midfieldr. 

```{r}
# packages
library(midfieldr)
library(ggplot2)

# data.table printing options
options(datatable.print.nrows = 20, datatable.print.topn = 5)

# load stickiness data from case study
pre_mw <- exa_stickiness

# examine the result
pre_mw
```

To protect confidentiality and prevent identification of individuals or institutions with small sub-populations, we typically omit rows with a small population. Using the larger MIDFIELD database, we usually set this limit to 10. Here, using the midfielddata sample, we'll use 5. 

```{r}
# prepare rows
rows_we_want <- pre_mw$ever >= 5
pre_mw <- pre_mw[rows_we_want]

# examine the result
pre_mw
```

We typically delete ambiguous levels of race/ethnicity. And because of its sparse representation in the data, we often omit the Native American population as well. 

```{r}
# remove rows of ambiguous or underrepresented populations
rows_we_want <- !pre_mw$race %in% c(
  "Unknown",
  "International",
  "Other",
  "Native American"
)
pre_mw <- pre_mw[rows_we_want]

# examine the result
unique(pre_mw$race)
```

We combine the race/ethnicity and sex column to create one of the multiway categorical variables. 

```{r}
# complete the transformation to multiway form
data_mw <- pre_mw

# create a new combined framing variable
data_mw[, race_sex := paste(race, sex, sep = " ")]

# examine the result
data_mw
```

To complete the transformation to multiway form, we select the three multiway variables.  

```{r}
# select the multiway variables
columns_we_want <- c("program", "race_sex", "stick")
data_mw <- data_mw[, ..columns_we_want]

# examine the result
data_mw
```

Then we use `order_multiway()` to convert the character columns to factors ordered by the median stickiness.

```{r}
# transform characters to factors ordered by median stickiness
data_mw <- order_multiway(data_mw)

# examine the result
data_mw
```





## A multiway dot plot  

```{r echo = FALSE}
nlevel1 <- nlevels(data_mw$program)
nlevel2 <- nlevels(data_mw$race_sex)
r <- nlevel1 * nlevel2
q <- 32
asp_ratio1 <- (r + 2 * nlevel1) / q
asp_ratio2 <- (r + 2 * nlevel2) / q
```


We use conventional ggplot2 functions to create the multiway graphs. By previously ordering the levels of the factors, we have structured the data so that the rows and panels of the multiway graph are ordered by the appropriate medians. 

```{r fig1, fig.asp = asp_ratio1}
# create one multiway graph
ggplot(data = data_mw, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "")
```

<br>

This graph permits a direct visual comparison of how stickiness varies by race and sex within a particular major. 

Reading a multiway graph

- We can more effectively compare values within a panel than between panels.
- Because rows are ordered, one expects a generally increasing trend within a panel. A response greater or smaller than expected creates a visual asymmetry. The interesting stories are often in these visual anomalies.
- To explore the comparison between panels, we swap the row and panel categories, creating the dual multiway.




## The dual multiway dot plot 

The previous graph does not facilitate visual comparisons of members of the same race-sex group. As Cleveland says,

> Because of this asymmetry, it is often important to explore multiway data by as many multiway dot plots as there are categorical variables, with each variable assigned once to the levels.

To create the *dual multiway* graph, we swap the roles of the rows and panels. 

- `y = race_sex` becomes `y = program`
- `facet_wrap(vars(program) ...`  becomes `facet_wrap(vars(race_sex) ...` 

```{r fig1-dual, fig.asp = asp_ratio2}
# create the dual multiway graph
ggplot(data = data_mw, aes(x = stick, y = program)) +
  facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "")
```

<br>

This graph permits a direct visual comparison of how stickiness varies by major for one group of students. 

Both multiways should be created for discussion. Usually one or the other will convey a more compelling visual story that gets included in the published findings. 




## Creating a data table

Data tables are useful in discussions with one's research team and are often included in publications for editors or reviewers who want to see the values underlying the data graphics. The first thing to do before tabulating is to limit the significant figures of floating-point numbers. 

```{r results = "hide"}
# tabulate data in block record ("long") form
blockf <- data_mw

# limit significant digits
blockf$stick <- round(blockf$stick, 2)

# examine the result
blockf
```

These data are laid out in "block record" form [@Mount+Zumel:2019:fluid-data], where the `race_sex` key and the `program` key are both in columns and the finding is in the last column. This layout is also called "tidy data"  [@Wickham+Grolemund:2017]. 

The graphing package we use, ggplot2, is designed to work most effectively with block-record data, thus the block-record ("long") form is the default layout in midfieldr. 

<br>

```{r echo = FALSE}
kable2html(blockf, caption = "Table 1: Stickiness (block records)")
```

<br>

However, the visual convention preferred by publishers and readers is a row-record ("wide") form, as illustrated by Table 2. The `race_sex` key is in the left column and the `program key` is in the column names. The stickiness findings lie at the row-column intersections. 

<br>

```{r echo = FALSE}
temp <- reshape(
  data = blockf,
  idvar = "race_sex", # rowKeyColumns
  timevar = "program", # columnToTakeKeysFrom
  v.names = "stick", # columnToTakeValuesFrom
  direction = "wide"
)
row.names(temp) <- NULL
names(temp) <- c(
  "Group",
  "Civil",
  "Electrical",
  "Industrial",
  "Mechanical"
)
kable2html(temp, caption = "Table 2: Stickiness (row records)")
```

<br>

Depending on your software background, you may have encountered a block-record/row-record transformation using any of these function pairs---a list adapted from [@Mount+Zumel:2019:fluid-data]. 

- pivot / anti-pivot or crosstab / shred (databases)
- pivot_to_rowrecs / unpivot_to_blocks (cdata)
- reshape "wide" / reshape "long" (R)
- pivot_wider / pivot_longer (tidyr)
- pivot / unpivot (Microsoft Excel)
- cast / melt (reshape2)

In base R, using `reshape()`, the block-record form is transformed to row-record form to create Table 2 as follows. 

```{r}
# tabulate data in row-record ("wide") form
rowf <- reshape(
  blockf,
  idvar = "race_sex",  # remains a column
  timevar = "program", # becomes new column headings
  v.names = "stick",   # fills the table cells
  direction = "wide"
)

# reset the row numbers
row.names(rowf) <- NULL

# edit columns names
names(rowf) <- c("Group", "Civil", "Electrical", "Industrial", "Mechanical")

# examine the result
rowf
```

While we used the base R `reshape()` function in this example, there are good reshaping tools in other packages such as cdata  [@Mount+Zumel:2020:cdata] or tidyr [@Wickham+Henry:2020:tidyr]. 



## References

<div id="refs"></div>






## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

```{r eval=FALSE}
# packages
library(midfieldr)
library(ggplot2)

# load stickiness data from case study
pre_mw <- exa_stickiness

# prepare rows
rows_we_want <- pre_mw$ever >= 5 &
  !pre_mw$race %in% c(
    "Unknown",
    "International",
    "Other",
    "Native American"
  )
pre_mw <- pre_mw[rows_we_want, , drop = FALSE]

# prepare columns
pre_mw$race_sex <- paste(pre_mw$race, pre_mw$sex, sep = " ")
columns_we_want <- c("program", "race_sex", "stick")
pre_mw <- pre_mw[, columns_we_want, drop = FALSE]

# transform characters to factors ordered by median stickiness
data_mw <- order_multiway(pre_mw)

# create one multiway graph
ggplot(data = data_mw, aes(x = stick, y = race_sex)) +
  facet_wrap(vars(program), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "")

# create the dual multiway graph
ggplot(data = data_mw, aes(x = stick, y = program)) +
  facet_wrap(vars(race_sex), ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "")

# tabulate data in block record ("long") form
blockf <- data_mw
blockf$stick <- round(blockf$stick, 2)
columns_we_want <- c("race_sex", "program", "stick")
blockf <- blockf[, columns_we_want, drop = FALSE]
blockf

# tabulate data in row record ("wide") form
rowf <- reshape(
  data = blockf,
  idvar = "race_sex", # column of row keys to copy
  timevar = "program", # column to take keys from
  v.names = "stick", # column to take values from
  direction = "wide"
)
row.names(rowf) <- NULL
names(rowf) <- c("Group", "Civil", "Electrical", "Industrial", "Mechanical")
rowf
```

<br>
<a href="#top"         >&#9650;     top of page </a>     
<a href="../index.html">&#9665;       main page </a> 



