---
title: "Introduction to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
link-citations: yes
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
# knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = TRUE, # varies from one Rmd to another
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/intro-to-R-"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
options(tibble.print_min = 8L, tibble.print_max = 8L)

# normally, purl is FALSE. But in some code chunks, I want purl = TRUE so that the code in the vignette is executed to pass R CMD check. These next few lines set that up. In the code chunk header, add opts.label = 'dopurl'
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_template$set(dopurl = list(purl = TRUE, error = FALSE))
```

The midfieldr vignettes are written with the R novice in mind, but they do not constitute an R manual. Readers should know how to:  

- Install R and update R periodically 
- Install and update an IDE of their choosing, e.g., RStudio 
- Install and update R packages
- Write and run R scripts 

There are many good tutorials online to help the absolute beginner learn to do these things. The MIDFIELD Institute offers the following online materials for the absolute R beginner, each one running about 50 minutes. 


[**Getting started with R**](https://midfieldr.github.io/workshops/getting-started.html#getting-started)

How to install R and RStudio, install an R package, install midfielddata and midfieldr packages, and create a set of directories useful for organizing your work.


[**R basics**](https://midfieldr.github.io/workshops/r-basics.html) 

An introduction to an R style guide, objects in R, functions, packages, basic syntax, and scripts. 


[**Graph basics**](https://midfieldr.github.io/workshops/graph-basics.html) 

A brief introduction to the ggplot2 package, graph anatomy, layers, aesthetics, properties, facets, etc. 


[**Data basics**](https://midfieldr.github.io/workshops/data-basics.html)

An introduction to data structure and data carpentry with worked examples to illustrate typical operations. 


<br>
<a href="#top">&#9650; top of page</a> 










## References

<div id="refs"></div>


## Appendix

### Complete script

In response to requests from some of our workshop attendees, we collect the vignette code chunks in one script. We condense the script by omitting exploratory steps so we can focus on the steps that produce the results. 

```{r eval=FALSE}
# TBD
```

<br>
<a href="#top"         >&#9650;     top of page </a>     
<a href="../index.html">&#9665;       main page </a> 








