---
title: "If you are new to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{If you are new to R}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/body-and-society.csl
link-citations: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.path = "../man/figures/vignette-new-to-R-")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```

## Introduction

Researchers interested in exploring MIDFIELD data are often R novices,  learning to use midfieldr at the same time they are learning to use R. There are many good online resources for learning R, so we do not attempt to reproduce that work here. However, we would like to offer some suggestions to help address some of the obstacles our new users have encountered in the past.

## Keyboard shortcuts

If you are working in RStudio, you can see the menu of keyboard shortcuts using the menu  *Tools > Keyboard Shortcuts Help*. The (Windows) shortcuts we use regularly include 

```{r echo = FALSE}
library("data.table")
df <- wrapr::build_frame(
  "Keyboard shortcut", "Action" |
    "`ctrl L`"       , "Clear the RSDtudio Console" |
    "`ctrl Enter`"   , "Run selected line(s)" |
    "`ctrl alt B`"   , "Run from  begining to line" |
    "`ctrl alt E`"   , "Run from line to end" |
    "`alt` --"       , "Insert `<-`, the assign operator" |
    "`ctrl shift C`" , "Comment/uncomment line(s)" |
    "`ctrl X, C, V`" , "Cut, copy, paste" |
    "`ctrl Z`"       , "Undo" |
    "`ctrl F`"       , "Find in text" |
    "`ctrl I`"       , "Indent/re-indent lines"  |
    "`ctrl S`"       , "Save" |
    "`ctrl A`"       , "Select all text" |
    "`ctrl shift K`" , " Compile knitr or R Markdown document"
)
setDT(df)
setkey(df, Action)
knitr::kable(df)
```

## Work in an RStudio Project 

The first rule of working in an RStudio project, 

> Start your session by launching the `project-name.Rproj` file in the project's main directory. 

RStudio is an an integrated development environment (IDE) for R that includes a console, editor, and tools for plotting, history, debugging, and workspace management as well as access to GitHub for collaboration and version control [@Rstudio:2021]. If we provide IDE screenshots, they will be from the RStudio interface. As the folks at RStudio assert, 

> RStudio projects make it straightforward to divide your work into multiple contexts, each with their own working directory, workspace, history, and source documents.

The advantages of working in an RProject are: 

- the working directory is set to the project directory, enabling the use of relative file paths in your scripts to support portability, reproducibility, and  collaboration
- active tabs are restored to where they were the last time the project was closed 
- multiple projects can be open at the same time 

You can read more about RProjects [here](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects).

## Type once-only commands in the Console

The installation instructions for midfielddata and midfieldr start with 

    # install midfielddata first 
    install.packages("midfielddata", 
                     repos = "https://MIDFIELDR.github.io/drat/", 
                     type = "source")

These lines can be typed in the RStudio console rather than an R script, because they generally only have to be run once. If you write them in a working script, you unnecessarily re-install the package every time you run the script. 

## Stay current

Running old software can be considerably harder than running new
software. Get current at the start of a new project, but avoid updating if you are approaching a project deadline. Read more about it at [Maintaining R](https://whattheyforgot.org/maintaining-r.html), a chapter in Bryan and Hester [-@Bryan+Hester:2019]. 


**Update R.** On a Windows machine, update R using the R GUI running as an administrator.

  - Navigate to your most recent `Rgui.exe` file located in your
    Programs directory, e.g., `C:\Program
    Files\R\R-3.5.3\bin\x64\Rgui.exe`  
  - Right-click on `Rgui.exe` and run as administrator

A straightforward update procedure uses the pacman package and the following lines of code. The first line ensures that pacman is installed. Run these lines in the R GUI. Defaults are generally OK. 
    
```r
# ensure pacman is installed
if (!require("pacman")) install.packages("pacman")

# load installr
pacman::p_load("installr")

# update R 
installr::updateR()
```

**Update RStudio.** Check for updates from the menu *Help \> Check for Updates*. If *Check for Updates* does not appear in the menu,

  - Find the current version of RStudio from the menu *Help \> About
    RStudio*  
  - Navigate to the [RStudio
    website](https://www.rstudio.com/products/rstudio/#Desktop), find
    out what the current version is.

To update RStudio, close RStudio on your machine, download
the new version, and run the `RStudio-n.n.n.exe` as an administrator
(`n.n.n` is the current version number).

**Update R packages.** When updating packages, if a window pops up asking about compilation,

- NO saves time
- YES gets you the latest version but can be time-consuming. Don’t say yes if you are in a hurry to get some work done.

To update packages in RStudio,

- From the RStudio pane, Select *Packages \> Update*
- OR, from the menu, *Tools \> Check for Package updates …*



## Use relative paths 

Explicitly link files using relative file paths within an RStudio Project. Paths are are relative to the main directory of an RStudio Project.  For example, you might have an R script in the `scripts` directory that reads a raw data file, 

```r
DT <- fread("data-raw/2020-03-21-student-record-ver09.txt")
```
    
When the data have been cleaned up, you might save the results to the `data` directory, e.g., 

```r
fwrite(DT, "data/2020-03-21-clean-student-record.txt")
```

Another script in the `scripts` directory might read this data, 

```r
DT <- fread("data/2020-03-21-clean-student-record.txt")
```
    
and after constructing a graph, write the graph to file in the `figures` directory, 

```r
gggsave(filename = "fig-01-enrollees", 
        path = "figures/", 
        device = "png", 
        width = 4.5, 
        height = 3.5, 
        units = "in")
```

## Plan your directory structure 

At a minimum, a project might start with three directories, 

```
project-name/
  |- data/ 
  |- documents/
  |- scripts/
  |- project-name.Rproj
```
`data/`

: Data ready for analysis  
  File names carefully curated for reproducibility  
  
`documents/`

: Outlines, drafts, other  text  

`scripts/`

: R scripts for analysis and creating graphs  


Two additional directories are often useful:  


`data-raw/`  

: Data in their original form, never edited manually 

`figures/`

: Graphs created by scripts are written to this directory, making them easy to find and sort through later 


  
## Directories for larger projects 

At the start of a project, you should carefully consider the number and types of files you will create over the life of the project and lay out a directory structure that helps you consistently organize your work. 
  
For a longer-term project, the directory structure can become  more detailed. Each publication directory could have its own  sub-directories for `documents`, `figures`, `manage`, and `scripts`.  For example:  

```
project-name/
  |- 2018-conference-name/
      |- documents
      |- figures
      |- manage 
      |- scripts
  |- 2019-journal-name/
  |- 2020-presentation-name/
  |- admin/
  |- data/ 
  |- data-raw/
  |- resources/
  |- README.Rmd
  |- project-name.Rproj
```

`yyyy-publication-name/`  

: Everything that creates a specific publication, report, or presentation  
  Can include R scripts, reports, figures, and administrative materials relevant to the publication or report  
  
`admin/`

: Overall project management and correspondence  
  Proposals and RFPs  
  Contracts  

`manage/`

: Publication administrative materials  
  Conference registration  
  Correspondence with editors 
  Travel reservations, registration, etc. 

`resources/`

: Bibliography and CSL files accessed by any of the publications  
  Image downloads and screen-shots accessed by any of the publications  
  
`README.Rmd`

: Describes the overall project generally  
  Can facilitate collaboration if git and GitHub are being used  

For different opinions on directory structure schemes, see

- Karl Broman (n.d.) [Organize your data and code](https://kbroman.org/steps2rr/pages/organize.html)  
- The Carpentries (2018) [Project management with RStudio](https://swcarpentry.github.io/r-novice-gapminder/02-project-intro/)  
- Chris von Csefalvay (2018) [Structuring  projects](https://chrisvoncsefalvay.com/structuring-r-projects/)   
- Mira Céline Klein (2017) [A meaningful file structure for R projects](https://www.inwt-statistics.com/read-blog/a-meaningful-file-structure-for-r-projects.html)    










## References 

