---
title: "Course-level study"
vignette: >
  %\VignetteIndexEntry{Course-level study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-130-")
```

The course data table contains information about every course for every student in the database. As might be expected, values reported by an institution such as course name, abbreviation, number, and section have not been "cleaned" for uniformity. A course can be recorded at different times with different encodings at the same institution. For example, at one institution, we find Calculus 1 labeled two ways, 

        "Calculus For Engrs 1", 
        "Calculus 1 For Engineers".

At the same institution, Calculus 2 has two encodings, 

        "Analytic Geometry/Calc 2", 
        "Analytic/Geometry Calc 2".

And Calculus 3 as well, 

        "Calculus 3 For Engnrs", 
        "Calculus 3 For Engineers".

Even the abbreviations an institution uses can be inconsistent from term to term  ("Engrs" and "Engnrs"), illustrating that quite a bit of exploration and data carpentry are required to extract and recode the information one wants in a useful form. 

*Goal.* &nbsp; Our goal is to provide an example of working with course-level data by investigating the question: If a student were ever enrolled in Engineering, what was their level (first-year, second-year, etc.) when passing one of the core math courses (Calculus 1, 2, 3 and Ordinary Differential Equations)?

- We recognize that not all engineering programs require all four courses.   

- Even when required, some students will satisfy a requirement by means other than a course at their institution, e.g., AP, IB, or transfer credit. 

## Definitions

```{r child = "../man/rmd/define-program.Rmd"}
```

```{r child = "../man/rmd/define-ever-enrolled.Rmd"}
```

```{r child = "../man/rmd/define-level.Rmd"}
```

math core

: Calculus 1, 2, 3, and Ordinary Differential Equations (ODEs). We recognize that not all engineering programs require all four courses. 

satisfying a requirement

: A passing grade ($\small C-$ or better) or an administrative mark such as credit (CR) or satisfactory progress (S/P).  







## Method

Demonstrating the following elements of a MIDFIELD workflow.  

1. *Planning.*  &nbsp; 

1. *Initial processing.* &nbsp; Filter student IDs for for data sufficiency and degree-seeking. Correct errors in the course table.

1. *Blocs.* &nbsp;  

1. *Additional processing.* &nbsp;   

1. *Groupings.* &nbsp; 

1. *Metric.* &nbsp; 

1. *Displays* &nbsp; 



```{r child = "../man/rmd/note-for-practice-only-2.Rmd"}
```

## Local functions

Some functions (defined in this vignette only) for simplifying repeated tasks. May be incorporated in midfieldr in the future. 

`suv()`

: Sort unique values. Applies `sort()` and `unique()` to values in a vector, including any NAs. 

```{r}
suv <- function (x) {
    # na.last FALSE sorts NA (if any) to the top of the result
    x <- sort(unique(x), na.last = FALSE)
    return(x)
}
```

```{r}
#| echo: false

# For internal use only
luv <- function (x) {
    x <- length(unique(x))
    return(x)
}
```

`make_pattern()`

: Make a pattern for matching or partial matching. Given a character vector with elements $\small \{a_1, a_2,..., a_m\}$ constructs a regular expression of the form `".?(a1)|(a2)|...|(am).?"` for use as a pattern to match in subsequent row operations.  

```{r}
# Build a regular expression
make_pattern <- function(x){
        pattern <- paste0("(", x, ")")
        pattern <- paste0(pattern, collapse = "|")
        pattern <- paste0(".?", pattern, ".?")
}
```

## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
library(janitor)

# set data.table print options
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

*Load.* &nbsp; Practice datasets. View data dictionaries via `?student`, `?term`, `?course`, `?degree`.  

```{r}
#| cache: true
# Load practice data
data(student, term, course, degree)
```

*Copy* &nbsp; Preserve a copy of the original source data in case they are needed later. 

```{r}
#| cache: true
# Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)
source_course <- copy(course)
source_degree <- copy(degree)
```

*Select.* &nbsp; Reduce the number of columns. Code reproduced from [Getting started](art-000-getting-started.html#reusable-code). 

```{r}
# Select variables required by midfieldr functions
student <- select_required(student)
term <- select_required(term)
degree <- select_required(degree)

# We require additional variables from course
course <- select_required(course, select_add = c("course", "hours_course", "grade", "discipline_midfield"))
```

The `course` data table has `r round(nrow(course)/1e+6, 1)`M observations of `r luv(course$mcid)` unique students. 





## Cleaning the course data

```{r}
#| echo: false

# Save copy of DT before the data entry errors are corrected.
# Apply ever-enrolled and math discipline filters

ever <- term[cip6 %like% "^14"]
ever <- ever[, .(mcid)]
ever <- unique(ever)

x <- copy(course)
x <- ever[x, on = c("mcid"), nomatch = NULL]
x <- x[discipline_midfield == "Mathematics"]
before_correction <- unique(x)
```

*Filter.* &nbsp; Delete rows with only 1 or 2 credit hours. Zero hours are OK, because students may receive credit (a grade of "CR") without accumulating credit hours. 

```{r}
DT <- copy(course)
DT <- DT[hours_course == 0 | hours_course >= 3]
```

*Filter.* &nbsp; Delete rows with an NA for the course name. 

```{r}
# Deleting rows that cannot be used in this particular study
DT <- DT[!is.na(course)]
```

*Correcting an error.* &nbsp; The `discipline_midfield` column contains errors that would affect our results if not corrected. For example, consider the   student course record shown below.  The Calc 3 and ODE courses, labeled "Business: Management" should be labeled "Mathematics". 

```{r}
this_id <- "MCID3112196665"
DT[mcid == this_id][c(5, 7)]
```

Using `suv()`, we find that only institutions B and C are present. We review course names assigned to "Business: Management" and we find mislabeled math courses at Institution C only. 
 
```{r}
# Unique institutions
suv(DT$institution)
# Examine Inst C course names
mgmt_label <- DT[institution == "Institution C" & discipline_midfield == "Business: Management"]
# Un-comment to see full list
# mgmt_label
```

We reviewed the full list of course names, but to save space in this document we print out an example of 10 mislabeled math courses. 

```{r}
# Example of Mathematics courses mislabeled "Business: Management"
unique(mgmt_label)[1:10]
```

We search the `mgmt_label$course` column and keep rows containing "calc" (for calculus) and "diff" (for differential equations) and drop rows containing "adv", "geometry", "manage", etc. In these two lines of code, you see the result of several iterations, especially for adding terms to the `drop` vector. The result is a vector of course names to change from business to math.   

```{r}
keep <- make_pattern(c("calc", "diff"))
drop <- make_pattern(c("adv", "geometry", "manage", "part","precalc", "surv", "svy",  "topics"))
should_be_math <- mgmt_label[course %ilike% keep][!course %ilike% drop]
should_be_math <- suv(should_be_math$course)
should_be_math
```

Recode the `discipline_midfield` value for these courses. 

```{r}
DT <- DT[course %chin% should_be_math, discipline_midfield := "Mathematics"]
```

Compare the result, 

```{r}
# Recoded values
unique(DT[course %chin% should_be_math, .(course, discipline_midfield)])
```

to the original source data. We have successfully re-labeled the math courses germane to the study. 

```{r}
# From the original source data
unique(source_course[course %chin% should_be_math & 
                         discipline_midfield %ilike% "Business", .(course, discipline_midfield)])
```

*Filter.* &nbsp; Keep rows with "Mathematics" in the discipline column, then delete the (no longer needed) discipline column.   

```{r}
# Keep all math courses
DT <- DT[discipline_midfield == "Mathematics"]

# Delete the discipline column
DT$discipline_midfield <- NULL

# Examine current course data
DT
```

The result has `r nrow(DT)` observations of `r luv(DT$mcid)` unique students.
 
 
 









## Blocs

The bloc of students are those ever enrolled in Engineering. The `term` data table is our starting point.  

```{r}
# IDs of ever enrolled in engineering
ever_id <- term[cip6 %ilike% "^14", (mcid)]
ever_id <- unique(ever_id)
length(ever_id)
```

Filter course to keep ever enrolled in Engr. 

```{r}
# Course contains every math course by ID and term. 
DT <- DT[mcid %chin% ever_id]
DT
nrow(DT)
length(unique(DT$mcid))
```

```{r}
setnames(DT, old = "term_course", new = "term")
```

Join term ID, term, level left join keep all rows of DT

```{r}
term_levels <- term[, .(mcid, term, level)]

# Left join
DT <- term_levels[DT, on = c("mcid", "term")]
nrow(DT)
length(unique(DT$mcid))
setorderv(DT, c("mcid", "term"))
DT
sum(complete.cases(DT))
```

2236 obs with level = NA is OK

```{r}
DT[is.na(level)]
```


```{r}
#| echo: false
options(datatable.print.nrows = 20, datatable.print.topn = 10)
```


look at 

```{r}
this_id = "MCID3111169186"
term[mcid == this_id]
course[mcid == this_id]
DT[mcid == this_id]
```




## Transforming

Additional processing of the course data includes revising the course names, identifying course names in our defined math core, and deleting courses that have been grade-replaced.   

We start with cleaning the course names in the `DT$course` column with `janitor::make_clean_names()`. The function applies `snake_case` and replaces punctuation and spaces with underscores `_`. We waited until now to take this "cleaning" step to save computing time. Applying `make_clean_names()` to 3.2M rows (the original course table) takes significantly more time than applying it to the 21k rows we have now.

While users may decide for themselves whether to use the janitor package, the detailed examination of course names we undertake in later sections is made somewhat easier by re-coding the values using a consistent framework. 

```{r}
#| cache: true
DT$course <- janitor::make_clean_names(DT$course, allow_dupes = TRUE)
DT
```

Review all unique names in the `course` column (not displayed to save space in this document).

```{r}
# Commented-out to save space
# suv(DT$course)
```

After review, we'd like to make a few changes. Janitor converted the original "I", "II", and "III" to "i", "ii", and "iii", which we now convert to "1", "2", "3". 

```{r}
DT$course <- gsub("_iii$", "_3", DT$course)
DT$course <- gsub("_ii$" , "_2", DT$course)
DT$course <- gsub("_i$"  , "_1", DT$course)
DT$course <- gsub("iii_" , "3_", DT$course)
```

Some spelling variations are made more uniform.   

```{r}
DT$course <- gsub("_physicl_"  , "_physical_"  , DT$course)
DT$course <- gsub("_scientist_", "_scientists_", DT$course)
DT$course <- gsub("_scientst_" , "_scientists_", DT$course)
DT$course <- gsub("_engrs_"    , "_engineers_" , DT$course)
DT$course <- gsub("_engnrs"    , "_engineers"  , DT$course)
```

We change all "calc" abbreviations to "calculus". 

```{r}
#| echo: false
options(datatable.print.nrows = 10)
```

```{r}
DT$course <- gsub("calc_", "calculus_", DT$course)
DT <- unique(DT)
```

Shorten a long name.

```{r}
DT$course <- gsub(
    "introduction_to_differential_equations_with_linear_algebra",
    "intro_linear_alg_diff_eq", DT$course
    )
DT$course <- gsub("methods_in_applied_mathematics_", "", DT$course)
DT$course <- gsub("introduction_to_", "", DT$course)

DT <- unique(DT)
```










## Identify math core

```{r}
#| echo: false
options(datatable.print.nrows = 200)
```

After closely reviewing the full set of course names, we can iteratively filter and refine to obtain unique vectors of names for each course. 

```{r}
math_pool <- unique(DT[, .(course)])
```

```{r}
# First filter
keep = make_pattern(c("calc", "diff"))
math_pool <- math_pool[course %ilike% keep]
setorderv(math_pool, "course")

# Commented-out to save space
# suv(math_pool$course)
```

Review to identity terms to drop

```{r}
drop = make_pattern(c("game", "adv", "calculus/soc", "chaos",
                      "geometry", "lab", "manage", "part", "pre_calc",
                      "precalc", "soc_sci", "social_science"))
math_pool <- math_pool[!course %ilike% drop]
suv(math_pool$course)
```

A few courses are ambiguously named---I'm not sure to which of the math core they belong, if any. 

- `fund_integral_calculus`
- `fundament_diff_calculus`
- `hon_calculus_3_diff_equatns`
- `intensive_calculus`

To explore how these courses may or may not fit in a student's math sequence, we look at students' course histories that contain these ambiguous course names. The function below allows us to print a sample of student records that include a specific course. 

```{r}
sample_includes <- function(dframe, selected_course, max_sample = 5) {
  # Gather all IDs with which the course is listed
  all_ids <- unique(dframe[course %ilike% selected_course]$mcid)
  n <- min(length(all_ids), max_sample)
  
  # Sample the IDs
  # set.seed(2024007) # un-comment for reproducibility
  sample_ids <- sample(all_ids, n)
  
  # Print the records in dframe for the sample IDs
  for (jj in sample_ids) {
      cat("\n\n")
      print(paste0("mcid: ", jj))
      print(dframe[mcid == jj, .(term, level, course, hours_course, grade)])
  }
}
```

Usage: Applied to the "fund_integral_calculus" course. 

```{r}
#| eval: false
sample_includes(DT, "fund_integral_calculus", 5)
```

From the output (not shown), we select three IDs to highlight below. From these records, we assume that "fund_integral_calculus" and "fundament_diff_calculus" are not substitutes or equivalent to Calculus 1. We will drop these two courses from our math pool.

```{r}
math_record <- DT[, .(mcid, term, level, course, hours_course, grade)]
math_record[mcid == "MCID3111363671"][, !c("mcid")]
math_record[mcid == "MCID3111215921"][, !c("mcid")]
math_record[mcid == "MCID3111173937"][, !c("mcid")]
```

Exploring "intensive_calculus" with `sample_includes()` yields the following IDs. Their records suggest it can substitute for Calculus 1. 

```{r}
math_record[mcid == "MCID3111370748"][, !c("mcid")]
math_record[mcid == "MCID3111421519"][, !c("mcid")]
math_record[mcid == "MCID3111369869"][, !c("mcid")]
```

Exploring "hon_calculus_3_diff_equatns" with `sample_includes()` yields the following IDs. Their records suggest it can substitute for Calculus 3 but not ODEs. 

```{r}
math_record[mcid == "MCID3111217247"][, !c("mcid")]
math_record[mcid == "MCID3111214610"][, !c("mcid")]
math_record[mcid == "MCID3111220676"][, !c("mcid")]
```

Thus, of the ambiguous course names, only two get dropped at this point. 

```{r}
drop <- make_pattern(c("fund_integral", "fundament_diff"))
math_pool <- math_pool[!course %ilike% drop]
```










## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
