---
title: "Course-level study"
vignette: >
  %\VignetteIndexEntry{Course-level study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../man/rmd/common-setup.Rmd"}
```

```{r}
#| echo: false
knitr::opts_chunk$set(fig.path = "../man/figures/art-130-")
```

Course-level data has the finest level of granularity of all MIDFIELD data tables: one observation per student per term per course. With this level of detail, it also requires the greatest effort to extract the information one wants in a useful form.  

In this article, we use a question about math courses to illustrate issues one encounters with course-related data and how to address them. 

Our goal is to visualize how students progress through the engineering core-math sequence---Calculus 1, 2, 3 and Ordinary Differential Equations (ODEs)---inspired by a similar study at Texas A&M [@Froyd+Shryock+etal:2015]. We are interested in all students ever enrolled in Engineering, without regard to their completion status. 

## Definitions

```{r child = "../man/rmd/define-program.Rmd"}
```

```{r child = "../man/rmd/define-ever-enrolled.Rmd"}
```

math core

: The first four math courses required by most engineering programs---typically includes Calculus 1, 2, 3, and Ordinary Differential Equations (ODEs), or some variation thereof. The core is fairly uniform across institutions because of accreditation requirements [@ABET:2024].

satisfying a requirement

: A passing grade ($\small C-$ or better) in the course. Alternatively, if a missing course is a prerequisite for a course in the record, we assume the prerequisite requirement was satisfied in some other way, e.g., AP or IB credit, transfer credit, credit by exam, etc.  







## Method

Demonstrating the following elements of a MIDFIELD workflow.  

1. *Planning.*  &nbsp; The metric is progress through a math sequence, Calc 1, 2, 3 and ODEs or their equivalent. The bloc of students are those ever enrolled in an engineering program. 

1. *Initial processing.* &nbsp; Filter student IDs for for data sufficiency and degree-seeking. Correct errors in the course data. Filter course data for math courses only. 

1. *Blocs.* &nbsp; Ever-enrolled in Engineering (2-digit CIP).  

1. *Groupings.* &nbsp; The courses of the core math sequence students attempt; passing the course or otherwise satisfying the requirement; and sex. 

1. *Metric.* &nbsp; Summarize by grouping variables and determine the following percentages, starting with $\small N$ students attempting Calculus 1:   

    - ? % satisfy the Calc 1 requirement  
    - ? % attempt Calc 2
    - ? % satisfy the Calc 2 requirement  
    - ? % attempt Calc 3 
    - ? % satisfy the Calc 3 requirement  
    - ? % attempt ODEs
    - ? % satisfy the ODEs requirement  

1. *Displays* &nbsp; A pseudo-time series with the math sequence attempt/satisfy along the horizontal axis and percentage values disaggregated by sex on the vertical axis. 




```{r child = "../man/rmd/note-for-practice-only-2.Rmd"}
```



## Load data

*Start.* &nbsp; If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)

# set data.table print options
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 3,
  datatable.print.class = TRUE
)
```

*Load.* &nbsp; Practice datasets. View data dictionaries via `?student`, `?term`, `?course`, `?degree`.  

```{r}
#| cache: true
# Load practice data
data(student, term, course, degree)
```

*Loads with midfieldr.* &nbsp; Prepared data. View data dictionary via `?baseline_mcid`. 

- `baseline_mcid`  (derived in [Blocs](art-050-blocs.html#initial-processing)).







## Initial processing

*Select.* &nbsp; Reduce the number of columns. Code reproduced from [Getting started](art-000-getting-started.html#reusable-code). 

```{r}
# Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)
source_course <- copy(course)
source_degree <- copy(degree)

# Select variables required by midfieldr functions
student <- select_required(source_student)
term <- select_required(source_term)
degree <- select_required(source_degree)
```

*Select.* &nbsp; The `course` table has columns we need in addition to those in `select_required()`. 

```{r}
# Select variables required for this study.
course <- select_required(source_course, 
                          select_add = c("course", "hours_course", "grade", "discipline_midfield"))
```

The `course` practice data has `r nrow(course)` observations.

<br>

*Filter.* &nbsp; Delete rows with an NA for the course name to create our initial working data frame. 

```{r}
# Deleting rows that cannot be used in this particular study
DT <- copy(course)
DT <- DT[!is.na(course)]
DT
```

The result has `r nrow(DT)` observations.

<br>

*(Optional) Recode.* &nbsp; To compress the displays in this document, we recode some longer course names. As we work through this vignette and come across a long course name, we return to this step and add a line of code.  

```{r}
# Recode 
DT[course == "Introduction To Differential Equations With Linear Algebra", 
   course := "Intro Diff Eq W/Lin Alg"]
DT[course == "Differential Equations Computer Lab", 
   course := "Diff Eq Comp Lab"]
DT[course == "Calculus For Social Science And Business", 
   course := "Calculus/Soc Sci & Bus"]
DT[course == "Calculus BC Score of 3     ++", 
   course := "Calculus BC Score of 3"]
DT[course == "Calculus BC Score of 4     ++", 
   course := "Calculus BC Score of 4"]
DT[course == "Calculus BC Score of 5     ++", 
   course := "Calculus BC Score of 5"]
```


## Correct a data error

```{r}
#| echo: false

# Save copy of DT before the data entry errors are corrected.
# Apply ever-enrolled and math discipline filters

ever <- term[cip6 %like% "^14"]
ever <- ever[, .(mcid)]
ever <- unique(ever)

x <- copy(DT)
x <- ever[x, on = c("mcid"), nomatch = NULL]
x <- x[discipline_midfield == "Mathematics"]
before_correction <- unique(x)
```

The `course` data table, `discipline_midfield` column, Institution C, contains errors important to the goals of the study. In this section, we show the errors and how we corrected them. 

For example, consider the first few observation of a student course record shown below.  The Calc 3 and ODE courses (rows 5 and 7) are labeled as "Business: Management" courses. This error is repeated in about 6100 Institution C observations.  

```{r}
this_id <- "MCID3112196665"
DT[mcid == this_id][1:7]
```

Check which institutions are present (using the locally-defined `suv()` function, for "sort unique values"). 
 
```{r}
# Convenience function to sort unique values of a vector
suv <- function (x) {
    # na.last FALSE sorts NA (if any) to the top of the result
    x <- sort(unique(x), na.last = FALSE)
    return(x)
}

# Unique institutions in the working data frame
suv(DT$institution)
```

Review all Institution C course names assigned the "Business: Management" discipline. 

```{r}
# Examine Inst C course names
mgmt_label <- DT[institution == "Institution C" & 
            discipline_midfield == "Business: Management"]
mgmt_label <- unique(mgmt_label$course)
```

To save space in this document, I display only a subset of the full list to illustrate the mix of Mathematics and Business courses assigned the "Business: Management" discipline.  (A similar review was run for Institution B and no math courses were similarly mislabeled. )

```{r}
# Sample of Mathematics course labeled Business: Management
mgmt_label[1:20]

# Sample of Business courses labeled Business: Management
mgmt_label[c(53, 64, 137:141, 146, 148:150, 155:160, 166:168)]
```

To make the necessary corrections, we wrote a separate script in which the indices of math courses in the `mgmt_label` vector (like those in the code chunk above) were carefully collected. For the moment, I'll write that list of indices here. Later, we might incorporate it as a midfieldr prepared data file.

```{r}
# Inst C math courses that are mislabeled as Business
mislabeled <- mgmt_label[c(1:52, 54:63, 65:67, 69:79, 85:90, 
                           98:109, 111:117, 119:125, 127:130, 134:136, 142:145, 
                           147, 151, 154, 161, 169, 184)]
```

Recode the discipline column of the mislabeled math courses. 

```{r}
# Recode discipline
DT <- DT[institution == "Institution C" & course %chin% mislabeled, 
         discipline_midfield := "Mathematics"]
```

Redo the checks we ran earlier. First, review Institution C course names assigned the "Business: Management" discipline. Printing the first 20 course names, all are Business courses.  

```{r}
# Examine Inst C course names with corrected data entry error
mgmt_label <- DT[institution == "Institution C" & 
            discipline_midfield == "Business: Management"]
unique(mgmt_label$course)[1:20]
```

Second, redo the student example. Both math courses are correctly labeled in the discipline column. 

```{r}
this_id <- "MCID3112196665"
DT[mcid == this_id][1:7]
```

```{r}
#| echo: false

# Save copy of DT after the data entry errors are corrected.
# Apply ever-enrolled and math discipline filters
x <- copy(DT)
x <- ever[x, on = c("mcid"), nomatch = NULL]
x <- x[discipline_midfield == "Mathematics"]
after_correction <- unique(x)
```

Considering the bloc of students ever enrolled in Engineering, the number of math courses after and before the errors were corrected are: 

- `r nrow(after_correction)` observations after correction  

- `r nrow(before_correction)` observations before correction 

Thus our study was missing `r nrow(after_correction) - nrow(before_correction)` individual courses before correction---some (though not all) of which were in the core math sequence. 



## Filter by discipline

Reviewing the values in the `discipline_midfield` column, "Mathematics" is the only relevant discipline in this study. To save space in this document, we display here only a subset of the disciplines.  

```{r}
# All values
# suv(DT$discipline_midfield)

# Discipline subset
suv(DT$discipline_midfield)[181:190]
```

We conclude our initial data processing, then, by filtering for the "Mathematics" discipline, 

```{r}
DT <- DT[discipline_midfield == "Mathematics"]
```

then deleting the discipline column. 

```{r}
DT$discipline_midfield <- NULL
DT
```

The result has `r nrow(DT)` observations.

## Bloc

The original concept for the study was to investigate all Engineering students. However, after closely examining records of individual students (especially graduating students), we find that a number of programs (in the practice data) do not necessarily require all 4 courses of what we call the math core. 

Therefore, we restrict our student bloc to students ever enrolled in Aero (CIP 1402), Chemical (CIP 1407), Electrical (CIP 1410), and Mechanical (CIP 1419) Engineering---programs that are most likely to require the math core we have defined. 

```{r}
# CIP codes for the study
programs <- c("^1403", "^1407", "^1410", "^1419")
search_string <- paste(programs, collapse = "|")

ever_enrolled <- term[cip6 %ilike% search_string]

# Check result
suv(ever_enrolled$cip6)
```






*Filter.* &nbsp; Towards the end of the analysis, we check the number of math courses in each Engineering graduates' record, expecting at least the four core courses. During those checks, we found that not all Engineering majors require the four course we are expecting. And even those that generally do require all 4, e.g., Mechanical Engineering and Electrical Engineering, have graduates in this data set with less than 4 of the core courses. 

Of a total of 12243 Engineering graduates used in that exploration, four had Calc 1 only, nine had Calc 1 and another core course, and 110 had Calc 1 and two other core courses. Thus about 1% (123/12243) of Engineering graduates satisfied their program requirements with 1--3 core math courses, introducing a 1% uncertainty (at least) in the final counts. 

```{r}
# x <- degree[cip6 %like% "^14", .(cip6)]
# y <- x[, .N, by = cip6]
# setorder(y, N)
# y
# 
# # 1419 ME
# # 1408 Civil
# # 1410 EE
# # 1407 ChE
# # 1402 Aero
# # 1409 Computer Engr
# 
# filter_cip("^1402")
```






*Select.* &nbsp; Retain the ID column. 

```{r}
ever_enrolled <- ever_enrolled[, .(mcid)]
ever_enrolled <- unique(ever_enrolled)
ever_enrolled
```

*Join.*  &nbsp; Inner-join to filter courses by ever-enrolled IDs.

```{r}
# Filter for ever enrolled in Engr
DT <- ever_enrolled[DT, on = c("mcid"), nomatch = NULL]
DT <- unique(DT)
DT
```

The result has `r nrow(DT)` observations.

## Identify the math core

Our four math courses of interest are Calculus 1, 2, 3 and ODEs. These courses are recorded in the data under a variety of names. 

To be certain we don't miss any names, our approach is to collect terms from the course names that denote non-core courses. This vector is built up one term at a time; the remainder reviewed; adding to the list until the course names remaining are potential candidates for the math-core courses. 

Without knowledge of an institution's curriculum, this step is necessary though tedious.  

To drop (or keep) rows based on the value in the `course` column, we use the locally-defined `filter_column()` function.

```{r}
# Keep or drop rows by partial matches in one column
filter_column <- function(dframe, col_name = "course", keep = NULL, drop = NULL) {
    x <- copy(dframe)
    if (!is.null(keep)) {
        keep <- paste0(keep, collapse = "|")
        x <- x[get(col_name) %ilike% keep]
    }
    if (!is.null(drop)) {
        drop <- paste0(drop, collapse = "|")
        x <- x[!get(col_name) %ilike% drop]
    }
    return(x)
}
```

```{r}
#| echo: false
options(datatable.print.nrows = 200)
```

```{r}
noncore_terms <- c(
    "Abstract Algebra",
    "Adv Math 1 For Engineers", 
    "Advanced Engineering",
    "Analysis", 
    "Applied Algebra", 
    "Applied Mathematics",
    "Biometry", 
    "Calc Bound", 
    "Chaos", 
    "Col Algebra/Trigonometry", 
    "College", 
    "Combinatorial", 
    "Combinatoric", 
    "Communicating Math", 
    "Comp Variables",
    "Complex Var",
    "Computer",
    "Consumer Mathematics", 
    "Curves/Surfaces", 
    "Data And Models", 
    "Differential Geometry", 
    "Discrete", 
    "Ecology", 
    "Environment", 
    "Euclidean", 
    "Financial Mathematics", 
    "Finite", 
    "First Year Sem/Math Sciences", 
    "Foundations",
    "Fourier",
    "Functon/Limit", 
    "Fund/Tech Coll Algebra", 
    "Fundamentals And Technique",
    "Fundamental/Coll Algebra",
    "Games", 
    "Graphing",
    "Groups, Rings",
    "History", 
    "Independent Study",
    "Inequalities",
    "Info Security", 
    "Information and Coding Theory",
    "Informatn Integrity & Security",
    "Internship",
    "Intro-Differ Geometry 1",
    "Intro To Abstract", 
    "Java", 
    "Lab", 
    "Linear Algebra I",
    "Linear Algebra II",
    "Linear Equat/Matrices", 
    "Linear Programming", 
    "Linear Program & Network Flows", 
    "Logarith",
    "Maple", 
    "Markov", 
    "Master's Thesis",
    "Math in", 
    "Math Of Coding", 
    "Math-Social Sci/Business",
    "Mathematics for Economists" ,                               
    "Mathematics From The Visual Arts",                          
    "Matrices and Linear Equations", 
    "Matrix", 
    "Methds",
    "Methods",
    "Modeling", 
    "Modern Algebra", 
    "Multigrid Methods", 
    "Nonlinear", 
    "Non-Mathematics", 
    "Number Theory", 
    "Numer Analys",
    "Numeric", 
    "Operations", 
    "Optimal Control", 
    "Part Diff", 
    "Partial", 
    "Partial Differential",
    "Patterns of Phenomena",
    "PDE", 
    "Polyn", 
    "Pre-Calculus", 
    "Precalculus", 
    "Probability", 
    "Projects", 
    "Quant Reason",
    "Quantitative Reasoning", 
    "Rdng & Rsch", 
    "Research", 
    "Sem-", 
    "Seminar",
    "Set Theory", 
    "Stat", 
    "Statistic", 
    "Spirit/Uses", 
    "Supervised", 
    "Symbolic/Numer",
    "Techniques/Col",
    "Theory of Numbers", 
    "Time Series", 
    "Topics",
    "Topology", 
    "Tpc", 
    "Ud Mathematics", 
    "Undergraduate"
)

# Drop courses and print remaining course list
potential_core <- copy(DT)
potential_core <- filter_column(potential_core, drop = noncore_terms)
potential_core <- suv(potential_core$course)
potential_core
```

To check on any specific term, for example, "linear", we can use the following code. In this case, the result shows us that some linear algebra courses are stand-alone while others are combined with differential equations and could therefore be part of a core sequence. 

```{r}
# How to identify a term that could be added to the non-core terms
suv(DT[course %ilike% "linear", (course)])
```

```{r}
#| echo: false
options(datatable.print.nrows = 10)
```

Filter for potential core courses. 

```{r}
# Drop non-core observations
DT <- filter_column(DT, drop = noncore_terms)
DT
```

The result has `r nrow(DT)` observations.

<br>

From the list of courses in `potential_core`, we can start to compile separate lists for Calc 1, 2, 3 and ODEs. For example, 

```{r}
#| echo: false
options(datatable.print.nrows = 30)
```

```{r}
seq_incl_course <- function(dframe, course_to_find, n_mcid){
    id_to_find <- suv(dframe[course == course_to_find, (mcid)])
    n <- length(id_to_find)
    if (n > 0) {
        printn <- min(n_mcid, n)
        for (j in 1:printn) {
            print(dframe[mcid %chin% id_to_find[j]])
            cat("\n")
        } 
    } else {
        print(paste(course_to_find, "not found in data frame.")) 
    }
}

# seq_incl_course(DT, "Adv Calculus of One Variable", 5)
# seq_incl_course(DT, "Fundament/Diff Calculus", 5)
# seq_incl_course(DT, "Math for Scientists", 5)
# 
# this_id <- "MCID3111143103"
# source_student[mcid == this_id]
# source_course[mcid  == this_id]
# source_term[mcid    == this_id]
# source_degree[mcid  == this_id]

```


Here, we see that "Advanced Calculus of One Variable" generally comes after  Calc 3 or ODEs, so it is not a core course. On the other hand, we see that 
"Calculus-Physical Scientists I, II, and III" are clearly our Calc 1, 2, 3. 

Repeating this process a number of times, we conclude that the core math sequence for Institution B is: 

- Calculus 1 For Engineers or Analytic Geometry/Calc 1 or Calc 1a & 1b
- Calculus 2 For Engineers or Analytic Geometry/Calc 2
- Calculus 3 For Engineers or Analytic Geometry/Calc 3
- Intro Linear Alg-Diff Eq

And for Institution C we have: 

- Calculus-Physical Scientists I or Calc AB
- Calculus-Physical Scientists II or Calc BC
- Calculus-Physical Scientists III
- Intro-Ordinary Differen Equatn or Differential Equations

## Recoding the core

Create a vector for each of the 4 core courses. 

```{r}
# Constants (this code chunk)
dframe <- copy(DT)

calc1 <- filter_column(dframe, 
                          keep = c("Calculus 1", 
                                   "Calc 1", 
                                   "Calc I", 
                                   "Scientists I", 
                                   "Engineers I", 
                                   "Engrs 1", 
                                   "Intensive Calculus"), 
                          drop = c("1a", "Part A", 
                                   "1b", "Part B", 
                                   "AB", "BC", 
                                   "advanced"))[
                              , (course)] |> suv()

calc1a <- filter_column(dframe, 
                           keep = c("Calculus 1 With Algebra, Part A", 
                                    "Calculus 1a With Algebra"))[
                              , (course)] |> suv()

calc1b <- filter_column(dframe, 
                           keep = c("Calculus 1 With Algebra, Part B", 
                                    "Calculus 1b With Algebra"))[
                              , (course)] |> suv()

calcAB <- filter_column(dframe, 
                           keep = c("Calculus AB"))[
                              , (course)] |> suv()

calcBC <- filter_column(dframe, 
                           keep = c("Calculus BC"))[
                              , (course)] |> suv()






calc2 <- filter_column(dframe, 
                          keep = c("Calculus 2", 
                                   "Calc 2", 
                                   "Calc II",
                                   "Scientists II", 
                                   "Engineers 2"))[
                              , (course)] |> suv()







calc3 <- filter_column(dframe, 
                          keep = c("Calculus 3", 
                                   "Calc 3", 
                                   "Calc III",
                                   "Scientists III",
                                   "Scientist III",
                                   "Scientst III", 
                                   "Engineers 3", 
                                   "Engrs 3"))[
                              , (course)] |> suv()




# ODEs, Inst B, course names
ode <- filter_column(dframe, 
                       keep = c("Differential", 
                                "Differen", 
                                "Diff"), 
                       drop = c("Partial", 
                                "Part", 
                                "Geometry", 
                                "Fundament/Diff"))[
                              , (course)] |> suv()
```

Review the collection. 

```{r}
calc1
calc1a
calc1b
calcAB
calcBC
calc2
calc3
ode
```

*Recode.* &nbsp; Create the `math_core` variable with our uniform, abbreviated course names. 

```{r}
DT[, math_core := NA_character_]
DT[course %chin% calc1,  math_core := "Calc 1"]
DT[course %chin% calc1a, math_core := "Calc 1a"]
DT[course %chin% calc1b, math_core := "Calc 1b"]
DT[course %chin% calcAB, math_core := "Calc AB"]
DT[course %chin% calcBC, math_core := "Calc BC"]
DT[course %chin% calc2,  math_core := "Calc 2"]
DT[course %chin% calc3,  math_core := "Calc 3"]
DT[course %chin% ode,    math_core := "ODEs"]
```

## Check results

Review the course sequences of individuals: graduates again because we know they satisfied their program requirements. 

```{r}
# Engineering graduates CIP start with "14"
grad <- degree[cip6 %like% "^14"]
grad <- unique(grad[, .(mcid)])

# Inner join 
grad <- grad[DT, on = "mcid", nomatch = NULL]
grad
```


Delete all NA rows in the `math_core` column. 

```{r}
core_only <- grad[!is.na(math_core)]
core_only
```

Count the number of courses in a students core math sequence by ID

```{r}
course_count <- core_only[, .(N_seq = .N), by = c("mcid")]
course_count
```

Summarize the counts

```{r}
summary_count <- course_count[order(N_seq), .(N_students = .N), by = c("N_seq")]
summary_count
```

What math courses taken by students with only 1 core course?


```{r}
#| echo: false
options(datatable.print.nrows = 80, datatable.print.topn = 50)
```


```{r}
these_ids <- suv(course_count[N_seq == 3, (mcid)])
explore <- core_only[mcid %chin% these_ids, .(mcid, math_core)]
explore <- unique(explore)
suv(explore[math_core == "Calc 1", (mcid)])
```











Functions for reviewing one student record at a time

```{r}
#| echo: false
options(datatable.print.nrows = 80)
```


```{r}
# select a specific math course we are interested in
this_course <- "Calc BC"
j = 6;
   

this_id <- copy(grad)
this_id <- this_id[math_core %chin% this_course][, (mcid)][j]

source_student[mcid == this_id]
source_term[mcid == this_id]
DT[mcid == this_id]
source_degree[mcid  == this_id]
```

By student ID

```{r}

# this_id <- "MCID3111170434" # Civil full core

# Four grads with only Calc 1 in their core
# this_id <- "MCID3111262210" # Civil only two APPM courses
# this_id <- "MCID3111534273" # ME, one core math and three other math
# this_id <- "MCID3111675558" # Arch Engr one math
# this_id <- "MCID3112605117" #EE major, 1 math course

# Nine grads with two core
# these_ids <- c("MCID3111264265", "MCID3111479512", "MCID3111595138", "MCID3112036029", "MCID3112089901", "MCID3112135411", "MCID3112294710", "MCID3112295541", "MCID3112349622")

# 1, 2 140801 civil calc 1, 2
# 3, 140901 comp engr calc 1 and DEs
# 4, 141901 ME calc 1 calc 2 lin-alg (no calc 3 no DEs)
# 5, 140401 Architct engr, Calc 1, DEs
# 6, 140701 ChemE, calc 1, 2 stats
# 7, 140701 Chem E, calc 1, 2, adv calc
# 8, 140701 Chem E, calc 1, 2, algorithm
# 9  1419019, ME calc 1, 2 stats

# 110 with three core
these_ids <- c("MCID3111170543", "MCID3111171021")

# 1, 140901 Computer Engr no calc 3


j = 2
source_student[mcid == these_ids[j]]
source_term[mcid == these_ids[j]]
DT[mcid == these_ids[j]]
source_course[mcid == these_ids[j]]
source_degree[mcid == these_ids[j]]
```


Examples 

- MCID3111142988, EE grad, Calc 3 only, transfer student
- MCID3111144051, ME grad, ODEs only, transfer student
- MCID3111151027, Aero Engr, full seq
- MCID3111169577, Arch Engr, full seq 
- MCID3111169974, Computer Engr, full seq
- MCID3111170434, Civil Engr, full seq
- MCID3112243918, ME grad, Calc 1a and 1b and Calc 1

Inst C

- MCID3112501584, Civil Engr, full seq
- MCID3112501593, ME, full seq
- MCID3112291009, ChE, Calc BC satisfies Calc 1

Exceptions



## Sample sequences

Before continuing, we pause to review a sample of student records to gain insight into the variety of math sequences we might find. We will use the sub-bloc of Engineering graduates because we know they have satisfied their program requirements. 

```{r}
# Engineering graduates CIP start with "14"
grad <- degree[cip6 %like% "^14"]
grad <- unique(grad[, .(mcid)])

# Inner join to filter for grad only
grad <- grad[DT, on = "mcid", nomatch = NULL]
setorderv(grad, c("mcid", "term_course"))
grad
```

First, we count the number of math core courses taken per student.

```{r}
N_core <- grad[, .(repeats = .N), by = c("mcid", "math_core")]
N_core
```

Now the summary of the counts

```{r}
summ <- N_core[, .N, by = c("math_core", "repeats")]
setorderv(summ, c("math_core", "repeats"))
summ
```

Totals ever enrolled in the core course

```{r}
summ <- summ[, enroll := sum(N), by = c("math_core")]
summ[]
```

grad

```{r}
summ[, grad := length(unique(grad$mcid))]
summ[]
```

Estimate zero in course

```{r}
summ[, satisfied_0 := grad - enroll]
summ[]

x <- unique(summ[, .(math_core, N = satisfied_0, grad)])
x[, repeats := 0]
x[, enroll := 0]
x

summ[, satisfied_0 := NULL]
summ[]

setcolorder(x, names(summ))
x

summ <- rbindlist(list(x, summ))
setorderv(summ, c("math_core", "repeats"))
summ[]
```










```{r}
summ[, pct_grad := round(N / grad * 100, 1)]
summ[enroll > 0, pct_enroll := round(N / enroll * 100, 1)]
summ[]
```

```{r}
library(ggplot2)
ggplot(summ, aes(x = math_core, y = N)) +
    geom_point() +
    facet_wrap(vars(repeats), nrow = 1) +
    # scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 7)) +
    # scale_y_log10(breaks = c(1, 3.2, 10, 32, 100, 320, 1000, 3200)) +
    theme(panel.grid.major = element_line(linewidth = 0.5, color="gray80"))
    # labs(x = "Number of repeats of a course", 
    #      y = "Number of students", 
    #      title = "Engineering graduates satisfying the math core")

ggplot(summ, aes(x = repeats, y = pct_grad)) +
    geom_point() +
    facet_wrap(vars(math_core), nrow = 1) +
    scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 7)) +
    scale_y_continuous(breaks = seq(0, 100, 10))+
    theme(panel.grid.major = element_line(linewidth = 0.5, color="gray80")) +
    labs(x = "Number of repeats of a course", 
         y = "Engr graduates (%)", 
         title = "Graduates in Engineering")

ggplot(na.omit(summ), aes(x = math_core, y = pct_enroll)) +
    geom_point() +
    facet_wrap(vars(repeats), nrow = 1) +
    scale_x_continuous(breaks = seq(0, 10, 1), limits = c(0, 7)) +
    scale_y_continuous(breaks = seq(0, 100, 10))+
    theme(panel.grid.major = element_line(linewidth = 0.5, color="gray80")) +
    labs(x = "Number of repeats of a course", 
         y = "Percent of enrollees (%)", 
         title = "Graduates in Engineering")
```








## References

<div id="refs"></div>

```{r child = "../man/rmd/common-closing.Rmd"}
```
