---
title: "Select and label programs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Select and label programs}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = TRUE, # varies from one Rmd to another
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/selecting-programs-"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
options(tibble.print_min = 8L, tibble.print_max = 8L)

# normally, purl is FALSE. But in some code chunks, I want purl = TRUE so that the code in the vignette is executed to pass R CMD check. These next few lines set that up. In the code chunk header, add opts.label = 'dopurl'
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_template$set(dopurl = list(purl = TRUE, error = FALSE))
```

```{r echo = FALSE}
# for my use, load the following but conceal from novice user
library("kableExtra")

# function to print to html using kable, control font size
kable2html <- function(...) {
  kable(..., "html") %>%
    kable_styling(font_size = 12)
}
```



In the vignettes illustrating various persistence metrics, the first step is to select the programs to study. We will use the same set of programs in different vignettes, so here we show the details of selecting and labeling those programs.  




## Contents 

[Identify CIP codes]    
[Select and label programs]    
[Case study program group]    
[Program label options]  
[Non-midfieldr CIP variables]    





## Identify CIP codes 

The programs we are studying are Civil Engineering, Electrical Engineering, Industrial Engineering, and Mechanical Engineering. 

```{r message = FALSE}
# packages used
library("midfieldr")
library("dplyr")
```

We use `cip_filter()` to search the `cip` dataset first for Engineering and then filter the result for Civil, Electrical, Industrial, and Mechanical. 
 
```{r first-cip-search, echo = -1}
options(tibble.print_min = 15L)
cip_filter(cip, series = "engineering")%>%
  cip_filter(., series = c("civil", "electrical", "industrial", "mechanical")) %>%
  print()
```

The results show that the codes we want are at the 4-digit level:  1408, 1410, 1419, and 1435.  









## Select and label programs

In the next step, we extract these programs individually. 

```{r extract-cip}
cve <- cip_filter(cip, series = "^1408")
ele <- cip_filter(cip, series = "^1410")
mce <- cip_filter(cip, series = "^1419")
ise <- cip_filter(cip, series = "^1435")
```

We use `cip6_select()` to assign program names to these programs. One possibility is to use the 4-digit program names already available in the data. To view those names:    

```{r}
unique(cve[["cip4name"]])
unique(ele[["cip4name"]])
unique(mce[["cip4name"]])
unique(ise[["cip4name"]])
```

Except for Electrical Engineering, the default program names are concise. We use `cip6_select()` to assign default names to three of the programs and a shorter name for Electrical Engineering. 

```{r}
cve <- cip6_select(cve, program = "cip4name")
ele <- cip6_select(ele, program = "Electrical Engineering")
mce <- cip6_select(mce, program = "cip4name")
ise <- cip6_select(ise, program = "cip4name")
```

`cip6_select()` is pipe-compatible, so the two previous steps could have been combined, for example,  

```{r}
cve <- cip_filter(cip, "^1408") %>%
  cip6_select(., "cip4name")
```

We use `dplyr::bind_rows()` to bind the individual data frames into one data frame. 

```{r group-cip}
program_group <- bind_rows(cve, ele, mce, ise)
```

Examine the result: there are a total of `r length(unique(program_group[["cip6"]]))` CIP codes for these programs. 

```{r echo = -1}
options(tibble.print_min = 12L)
program_group
```

For additional information, try the [Identifying program CIP codes](cip_filter.html) vignette and the help pages by running `?cip_filter` and `?cip6_select`. 




<br>
<a href="#top">&#9650; top of page</a> 




## Case study program group

`case_program_group`, a data set identical to the `program_group` data frame above, is bundled in midfieldr for use in subsequent vignettes and example. 

The data are available if midfieldr is loaded. 

```{r}
library("midfieldr")
case_program_group %>%
  print()
```




<br>
<a href="#top">&#9650; top of page</a> 





## Program label options 

In this section, we explore the arguments of `cip6_select()`. There are four options for `program` argument 

<br>

**Option 1:** When a string of your own choosing is assigned to program.

```{r results = "hide"}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "Leisure Studies") %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "Leisure Studies") %>% 
  kable2html()
```




<br> 


**Option 2:** When `program = NULL` (default) and `data`  matches one of the named `cip_series`, then the 6-digit CIP program names are assigned to the new column.

```{r results = "hide"}
cip %>% 
  cip_filter(series = cip_math_stat) %>% 
  cip6_select() %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = cip_math_stat) %>% 
  cip6_select() %>% 
  kable2html()
```

<br>

**Option 3:** When `program` is assigned one of three strings ("cip2name", "cip4name", or "cip6name"), then the 2-digit, 4-digit, or 6-digit CIP program names are assigned to the new column.

```{r results = "hide"}
# 2-digit 
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip2name") %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip2name") %>% 
  kable2html()
```

```{r results = "hide"}
# 4-digit 
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip4name") %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip4name") %>% 
  kable2html()
```

```{r results = "hide"}
# 6-digit 
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip6name") %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "cip6name") %>% 
  kable2html()
```

<br>

**Option 4:** When `program = NULL` (default) and `data` does not match one of the  named `cip_series`, then the 6-digit CIP program names are assigned to the new column. This result is identical to `program = "cip6name"`. 

```{r results = "hide"}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select() %>% 
  print() # or View
```

```{r echo = FALSE}
cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select() %>% 
  kable2html()
```






<br>
<a href="#top">&#9650; top of page</a> 





## Non-midfieldr CIP variables

Optional arguments in `cip6_select()` accommodate the possibility that one uses a CIP data frame with different variable names than those in `midfieldr::cip`. The four columns accommodated are `cip6`, `cip6name`, `cip4name`, and `cip2name`. 

For illustration only, let's create a CIP data frame called `nonstandard_cip` with a different name for the two variables `cip6` and `cip6name`. 

```{r echo=-1}
options(tibble.print_min = 5L)
nonstandard_cip <- cip %>% 
  rename("code_6" = cip6, "name_6" = cip6name) %>% 
  print() # or View
```

**Option 1:** If the non-standard variable names must be kept intact, then we use the optional arguments `cip6` and `cip6name` so that `cip6_select()` can operate on the  nonstandard column names.  

```{r}
nonstandard_cip %>% 
  cip_filter(series = "^31") %>% 
  cip6_select(program = "Leisure Studies", cip6 = "code_6", cip6name = "name_6") %>% 
  print() # or View
```

**Option 2:** If the non-standard variable names do not have to be kept intact, the simple option is to `rename()` the nonstandard variable names to the standard names, e.g.,  

```{r}
standard_cip <- nonstandard_cip %>% 
  cip_filter(series = "^31") %>% 
  rename("cip6" = code_6, "cip6name" = name_6) %>% 
  print() # or View
```

Similar arguments are made when one has nonstandard column names for  `cip4name` and `cip2name`.






<br>
<a href="#top">&#9650; top of page</a> 




## References 


