---
title: "Stickiness metric"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stickiness metric}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
resource_files:
  - ../man/figures/stickiness-fig1-1.png
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = TRUE, # varies from one Rmd to another
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/stickiness-"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
options(tibble.print_min = 8L, tibble.print_max = 8L)

# normally, purl is FALSE. But in some code chunks, I want purl = TRUE so that the code in the vignette is executed to pass R CMD check. These next few lines set that up. In the code chunk header, add opts.label = 'dopurl'
knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_template$set(dopurl = list(purl = TRUE, error = FALSE))
```

Longitudinal "stickiness" is the ratio of the number of students graduating in a program to the number of students ever enrolled in that program [@stickiness2012]. 

The stickiness metric accounts for all students in a program. Unlike the graduation rate metric, stickiness includes students who begin college part-time, who enroll mid-year, and who transfer from one institution to another. 

Every student counted in a program's stickiness metric was enrolled in that program for at least one term. The program succeeded in attracting these students---stickiness tells us how well the program succeeded in retaining them.  

## Getting started 

Packages used: 

```{r message = FALSE}
library(tidyverse)
library(seplyr)
library(midfieldr)
```

In this article, we show you how to use midfieldr functions to compute program stickiness. Our workflow is: 

- select the programs 
- count the students ever enrolled
- count the students graduating  
- compute stickiness 
- create a multiway graph  


## Choose the programs to study

The [Selecting programs](selecting_programs.html) vignette describes the  process for obtaining the program codes and names for four engineering programs: Civil, Electrical, Industrial, and Mechanical. The output of that vignette is saved as a midfieldr case data set, `case_program_group`. You may 

- create the program group by running the code in the vignette 
- or load the saved case data 

To load the case data: 

```{r}
# information saved from an earlier vignette
data("case_program_group")
program_group <- case_program_group
```

Examine the result: there are a total of `r length(unique(program_group[["cip6"]]))` CIP codes for these programs. We have all six columns from the `cip` dataset and the `program` column we added using `cip_label()`. 

```{r echo = -1}
options(tibble.print_min = 12L)
# check the result
program_group
```




## gather_ever()  

We use `gather_ever()` to collect all students ever enrolled in the programs we are studying. `gather_ever()` has two arguments: 

- `.data` data frame of academic term data to be filtered. The default is `midfieldterms` 
- `series` atomic character vector of 6-digit CIP codes used to filter `.data` 

The `series` argument in this case is an atomic vector of the 6-digit CIP codes in our `program_group`. 

```{r}
program_cip6 <- program_group[["cip6"]] %>%
  unique() %>%
  sort()

program_cip6
```

Extract all students from midfieldterms ever enrolled in these programs. 

```{r opts.label = 'dopurl'}
students <- gather_ever(series = program_cip6)
```

Use `glimpse()` (a [tibble](http://tibble.tidyverse.org) function) to view the result to check the number of observations and to view the first few values of each variable. We see that we have `r nrow(students)` students ever enrolled in these programs. 

```{r}
# check the result
glimpse(students)
```

Open the help page by running `?gather_ever()`.

## race_sex_join() 

`race_sex_join()` accepts any data frame with the MIDFIELD student `id` variable and joins the student race and sex from the `midfieldstudents` data to these data. 

```{r}
students <- race_sex_join(students) %>%
  glimpse()
```

If the argument to `race_sex_join()` already includes a `sex` or `race` column, that column (or columns) is returned unchanged. Open the help page by running `?race_sex_join()`.

## Assign program names for grouping

Next we join our custom program names to the student data. We use `left_join()` (a dplyr function) to join `program_group` to `students` by the 6-digit CIP code. 

```{r}
students <- left_join(students, program_group, by = "cip6") %>%
  glimpse()
```

Optionally, we can select just the variables we want to work with next. 

```{r}
students <- students %>%
  select(program, race, sex) %>%
  glimpse()
```

This step is optional because when we group and summarize in the next step, all  variables not used for grouping are dropped anyway. 

## Count the numbers of students ever enrolled

We use `group_summarize()` (an [seplyr](https://winvector.github.io/seplyr/) function) to group and count the numbers of students ever enrolled in our programs.  All variables in `students` not used as grouping variables are quietly dropped. The assigned count variable is `ever`.  

```{r}
# assign the variables for grouping
grouping_variables <- c("program", "race", "sex")

# count the numbers of students ever in a program
ever_enrolled <- students %>%
  group_summarize(., grouping_variables, ever = n())

# check the result
ever_enrolled
```

## gather_grad() 

`gather_grad()` extracts a subset of the `midfielddegrees` data containing the  unique students ever graduating from the programs listed in `program_cip6 `. 

The procedure for gathering graduates is nearly identical to the procedure for gathering the ever-enrolled. We use `gather_grad()` to search the  `midfielddegrees` dataset using the same series argument; we join race, sex, and program in the same way; and when we count students, the totals are assigned to the new variable `grad`.

The functions can be chained in a code chunk of 4 lines. 

```{r}
# count students graduating from programs
graduated <- gather_grad(series = program_cip6) %>%
  race_sex_join(.) %>%
  left_join(., program_group, by = "cip6") %>%
  group_summarize(., grouping_variables, grad = n())

graduated
```

## Tally stickiness

To set up the stickiness calculation, we join `graduated` to `ever_enrolled` by program, race, and sex.  

```{r}
stickiness <- left_join(ever_enrolled, graduated, by = grouping_variables)

stickiness
```

We suggest omitting observations with 5 or fewer students ever enrolled. 

```{r}
stickiness <- stickiness %>%
  filter(., ever > 5)
```

Finally, we create `stick`, the ratio of `grad` to `ever`. Stickiness is reported as a fraction between 0 and 1. 

```{r echo = -1}
options(tibble.print_min = 10L)
# tally stickiness
stickiness <- stickiness %>%
  mutate(., stick = round(grad / ever, 3))

# check the result
stickiness
```

## multiway_order() 

Before structuring the data for graphing, we select the populations we are interested in. For example, we might omit students for whom the "race" designation is ambiguous. This step could have been taken earlier, but must be taken (if at all) before using `multiway_order()`. 

We use `filter()` (a [dplyr](http://dplyr.tidyverse.org) function) to keep all rows except those in which race is Unknown, International, or Other.  

```{r}
stickiness_mw <- stickiness %>%
  filter(., !race %in% c("Unknown", "International", "Other")) %>%
  filter(., !sex %in% "Unknown")
```

This example has no "Unknown" values in `sex`, but the larger data set does. For completeness, I've also filtered to remove those observations.   

We structure the stickiness results as *multiway* data by combining student race and sex into a single `race_sex` variable. The [Multiway](multiway.html) article explains multiway data and graphs in detail. 

We use: 

- `mutate()` (a [dplyr](http://dplyr.tidyverse.org) function) to create a new variable 
- `str_c()` (a [stringr](http://stringr.tidyverse.org) function) to combine the race and sex strings 
- `select()` (a [dplyr](http://dplyr.tidyverse.org) function) to keep three variables only 

```{r}
# create the race_sex variable
stickiness_mw <- stickiness_mw %>%
  mutate(., race_sex = str_c(race, sex, sep = " ")) %>%
  select(., program, race_sex, stick)

# check the result
stickiness_mw
```

`multiway_order()` converts the character variables `program` and `race_sex` into *factors* and orders the levels of the two factors by the relevant stickiness medians. `glimpse()` shows that the two categorical variables are now factors. Open the help page by running `?multiway_order()`.

```{r}
# convert the data to a multiway structure
stickiness_mw <- multiway_order(stickiness_mw)

# check the result
stickiness_mw
```

## Graphing the multiway

By using `multiway_order()` to condition the data, the rows and panels of the multiway graph will be ordered by the appropriate medians. We use conventional ggplot2 functions applied to the `stickiness` data frame: 

- `stick` is the quantitative x-variable 
- `race_sex` is the y-variable forming the rows of the panels 
- `program` in `facet_wrap()` conditions the panels  
- `as.table = FALSE` places the panels in ascending order from bottom to top 
- `geom_point()` creates the data markers; any NA values are removed by `na.rm = TRUE` 
- `labs()` assign axis labels 

```{r fig1, fig.asp = 2/1.6}
ggplot(stickiness_mw, aes(x = stick, y = race_sex)) +
  facet_wrap(~program, ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "") +
  theme_midfield()
```

We also apply our own `theme_midfield()` to edit the visual properties of the graph: axes, legends, panels, strips, etc. Open the help page by running

```r
?theme_midfield()
```

## Complete script

For reference, the complete script for this example is collected below.  

```{r eval = FALSE}
# packages used
library(tidyverse)
library(seplyr)
library(midfieldr)

# CIP information saved from an earlier vignette
program_group <- case_program_group

# extract the 6-digit CIPs as a search series
program_cip6 <- program_group[["cip6"]] %>%
  unique() %>%
  sort()

# assign the variables for grouping
grouping_variables <- c("program", "race", "sex")

# count students ever enrolled
ever_enrolled <- gather_ever(series = program_cip6) %>%
  race_sex_join(.) %>%
  left_join(., program_group, by = "cip6") %>%
  group_summarize(., grouping_variables, ever = n())

# count students graduating
graduated <- gather_grad(series = program_cip6) %>%
  race_sex_join(.) %>%
  left_join(., program_group, by = "cip6") %>%
  group_summarize(., grouping_variables, grad = n())

# tally stickiness, by group
stickiness <- left_join(ever_enrolled, graduated, by = grouping_variables) %>%
  filter(., ever > 5) %>%
  mutate(., stick = round(grad / ever, 3))

# prepare data for graph
stickiness_mw <- stickiness %>%
  filter(., !race %in% c("Unknown", "International", "Other")) %>%
  filter(., !sex %in% "Unknown") %>%
  mutate(., race_sex = str_c(race, sex, sep = " ")) %>%
  select(., program, race_sex, stick) %>%
  multiway_order(.)

# graph
ggplot(stickiness_mw, aes(x = stick, y = race_sex)) +
  facet_wrap(~program, ncol = 1, as.table = FALSE) +
  geom_point(na.rm = TRUE) +
  labs(x = "Stickiness", y = "") +
  theme_midfield()
```

## Issue bin

- Add a note regarding the suggested 6-student minimum. 
- Account for "possible graduation in 6 years" (posgrad6) 
- Could use mice to impute unknown values of the race and sex variables 

## References
