---
title: "Timely completion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Graduation rate}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: ../inst/REFERENCES.bib
csl: ../inst/body-and-society.csl
link-citations: yes
resource_files:
  - ../man/figures/vignette-graduation-rate-fig1-1.png
---

```{r include = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  fig.path = "../man/figures/vignette-graduation-rate-"
)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center"
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable2html <- function(x, font_size = NULL, caption = NULL) {
  font_size <- ifelse(is.null(font_size), 11, font_size)
  kable_in <- knitr::kable(x, format = "html", caption = caption)
  kableExtra::kable_styling(kable_input = kable_in, font_size = font_size)
}
asp_ratio_mw <- function(data, categories) {
  cat1 <- categories[1] # panels
  cat2 <- categories[2] # rows
  nlevel1 <- nlevels(data[, get(cat1)])
  nlevel2 <- nlevels(data[, get(cat2)])
  r <- nlevel1 * nlevel2
  q <- 32
  asp_ratio1 <- (r + 2 * nlevel1) / q
  asp_ratio2 <- (r + 2 * nlevel2) / q
  ratios <- c(asp_ratio1, asp_ratio2)
}
```




## Introduction

Some persistence metrics, e.g. graduation rate, require that students earn their degrees in a timely fashion, typically in no more than 6 years. When counting the number of graduates, only degrees earned within the timely completion span are included. All other degrees are re-coded for aggregation with the non-graduates.  

Timely completion terminology is illustrated in the figure. Definitions are given below.


<br>
```{r echo = FALSE, fig.asp = 5/10}
yyyy <- c(1990, 1992, 1994, 1995)
df <- data.frame(yyyyt = yyyy, y = 1)

library("ggplot2")
callout_color <- "gray60"
callout_line_size <- 0.3
anno_size <- 3.5 # 3.5 approx 10 point
vert_baseline <- 1.07

ggplot(df, aes(x = yyyyt, y = y)) +
  
  # adjusted start
  geom_segment(aes(x = yyyy[1], xend = yyyy[1], 
                   y = vert_baseline, yend = 1.75),
               color = callout_color, size = callout_line_size
  ) +
  annotate("text",
           x = yyyy[1], y = 0.89, size = anno_size,
           label = "offset start", hjust = 1, vjust = 0
  ) +
  # matriculation term
  geom_segment(aes(x = yyyy[2], xend = yyyy[2], 
                   y = vert_baseline, yend = 1.35),
               color = callout_color, size = callout_line_size
  ) +
  annotate("text",
           x = yyyy[2], y = 0.89, size = anno_size,
           label = "matriculation", hjust = 0.5, vjust = 0
  ) +
  # degree term
  geom_segment(aes(x = yyyy[3], xend = yyyy[3], 
                   y = vert_baseline, yend = 1.55),
               color = callout_color, size = callout_line_size
  ) +
  annotate("text",
           x = yyyy[3], y = 0.89, size = anno_size,
           label = "degree", hjust = 0.5, vjust = 0
  ) +
  # timely completion limit
  geom_segment(aes(x = yyyy[4], xend = yyyy[4], 
                   y = vert_baseline, yend = 1.75),
               color = callout_color, size = callout_line_size
  ) +
  annotate("text",
           x = yyyy[4], y = 0.89, size = anno_size,
           label = "limit", hjust = 0, vjust = 0
  ) +
  # timely completion arrow
  geom_segment(aes(x = yyyy[1], xend = yyyy[4], y = 1.7, yend = 1.7),
               color = callout_color,
               size = callout_line_size,
               arrow = arrow(
                 type = "closed",
                 ends = "both",
                 length = unit(2, "mm")
               ),
               arrow.fill = callout_color
  ) +
  annotate("text",
           x = 1/2 * (yyyy[1] + yyyy[4]) ,
           y = 1.75,
           size = anno_size,
           label = "timely completion span"
  ) +

  # offset span arrow
  geom_segment(aes(x = yyyy[1], xend = yyyy[3], y = 1.5, yend = 1.5),
               color = callout_color,
               size = callout_line_size,
               arrow = arrow(
                 type = "closed",
                 ends = "both",
                 length = unit(2, "mm")
               ),
               arrow.fill = callout_color
  ) +
  annotate("text",
           x = 1/2 * (yyyy[1] + yyyy[3]) ,
           y = 1.55,
           size = anno_size,
           label = "time to graduation"
  ) +
  # transfer arrow
  geom_segment(aes(x = yyyy[1], xend = yyyy[2], y = 1.3, yend = 1.3),
               color = callout_color,
               size = callout_line_size,
               arrow = arrow(
                 type = "closed",
                 ends = "both",
                 length = unit(2, "mm")
               ),
               arrow.fill = callout_color
  ) +
  annotate("text",
           x = 1/2 * (yyyy[1] + yyyy[2]) ,
           y = 1.3,
           size = anno_size,
           label = "transfer\noffset", 
           hjust = 0.5,
           vjust = 0.5
  ) +
  
  geom_point(size = 2) +
  scale_x_continuous(limits = c(1988, 1998), breaks = seq(1980, 2020, 2)) +
  scale_y_continuous(limits = c(0.8, 1.8)) +
  labs(
    y = "",
    x = "Start of academic year",
    title = "Illustrating timely completion"
  ) +
  
  theme_light() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```
<br>

matriculation term
: The term a student enters the institution from which they graduate. Applies to transfer students as well as first-time-in-college (FTIC) students. 

offset start term
: For transfer students, the effective start term as if their transfer credit hours been earned at the institution from which they graduate.  Typically, no more than 2 years of transfer credits are counted. For FTIC students, the offset is zero. 
  
degree term 
: The term in which a student earns their first degree. 

time to graduation
: The number of terms (or equivalent years) between the offset start term and the end of the degree term. 

timely completion span
: The maximum time to graduation (typically 6 years) for a degree to be included in the count of program graduates. 

When applying the timely completion constraint, students are counted as program graduates only if their time to graduation is no greater than the timely completion span; otherwise their they are aggregated with the program non-graduates. 



### This vignette uses

midfieldr functions 

- `subset_feasible()` 
- `timely_completion()`

packages

```{r}
# packages used
library(midfieldr)
library(midfielddata)
library(data.table)
library(ggplot2)

# print max 20 rows, otherwise 10 rows each head/tail
options(datatable.print.nrows = 20, datatable.print.topn = 10)
```




## How to use `timely_completion()`


change: use the `rep_ever` IDs to illustrate using the function

The program group for the engineering case study is included with midfieldr as the built-in data set `rep_group`. View its help page by running

```r
? rep_ever 
```

The data are loaded with midfieldr. 

```{r}
# examine the result
str(rep_ever)
```


Related vignettes 

- *Identify  programs* [(link)](identify_programs.html) illustrates program labeling options 






The argument of `timely_completion()` is a vector of student IDs. The function returns a data frame with two columns keyed by ID: a numerical variable  `years_to_grad`, giving the time to graduation; and a logical variable `is_timely`, stating if the graduate satisfies the timely completion criterion. 

Example output of `timely_completion(id)` 


    #>      id   years_to_grad   is_timely 
    #>  MID001               5        TRUE
    #>  MID002               7       FALSE        
    #>  MID003            <NA>       FALSE  
    #>  MID004               5        TRUE
    #>  MID005             6.5       FALSE
    #>   ---   

Next, merge these data frames with the starters.

    #>           id   cip6  is_timely  
    #>    1: MID001 141901      TRUE
    #>    2: MID002 140801      TRUE
    #>    3: MID003 141001     FALSE
    #>    4: MID004 141001     FALSE
    #>    5: MID005 140801     FALSE       
    #>   ---   


function arguments 

    timely_completion(
      id,                       # character vector of student IDs
      ...,
      span = NULL,              # numeric scalar, number of years to define 
                                  timely completion, default 6 years
                                  
      term_transfer_max = NULL, # numeric scalar, maximum number of terms of
                                  transfer credit to offset matriculation,  
                                  default 4 terms

      data_students = NULL,     # data frame of student attributes
      data_terms = NULL,        # data frame of term attributes
      data_degrees = NULL       # data frame of degree attributes
    )




## How `timely_completion()` works





## References

<div id="refs"></div>

## Appendix

### Complete script

The vignette code chunks are collected below in a single, condensed script. 

