---
title: "About MIDFIELD"
output: "github_document"
bibliography: "../../inst/REFERENCES.bib"
csl: "../../admin/chicago-author-date.csl"
always_allow_html: true
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = FALSE, # varies from one Rmd to another
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/about-"
)
```

<a href="https://engineering.purdue.edu/MIDFIELD" target="blank">MIDFIELD</a> is the acronym for the Multiple-Institution Database for Investigating Engineering Longitudinal Development. 

Longitudinal studies using MIDFIELD are not limited to Engineering---the database includes *all* student records from member institutions.


## Who we are

```{r}
df <- wrapr::build_frame(
  "name"             , "responsibility"                       , "contact"                  |
  "Matthew Ohland"   , "Principal Investigator/Director", "ohland@purdue.edu"      |
  "Marisa Orr"       , "Associate Director"             , "marisak@clemson.edu"    |
  "Catherine Brawner", "Director Policy Analysis"       , "brawnerc@bellsouth.net" |
  "Russell Long"     , "Managing Director/Data Steward" , "ralong@purdue.edu"      |
  "Susan Lord"       , "Director MIDFIELD Institute"    , "slord@sandiego.edu"     |
  "Richard Layton"   , "Director Data Display"          , "graphdoctor@gmail.com"
)
kable2html(df, font_size = 13)
```
 

 
## Funding

Current funding is provided by the US National Science Foundation, [Grant 1545667](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1545667) Expanding Access to and Participation in the Multiple-Institution Database for Investigating Engineering Longitudinal Development.    













