---
title: "About MIDFIELD"
output: "github_document"
bibliography: "../../inst/REFERENCES.bib"
csl: "../../admin/chicago-author-date.csl"
---

```{r setup, echo = FALSE, message = FALSE, purl = FALSE}
knitr::opts_knit$set(root.dir = "../")
knitr::opts_chunk$set(
  echo = FALSE, # varies from one Rmd to another
  message = TRUE,
  warning = TRUE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE,
  fig.width = 6,
  fig.asp = 1 / 1.6,
  out.width = "70%",
  fig.align = "center",
  fig.path = "../man/figures/about-"
)
```

<a href="https://engineering.purdue.edu/MIDFIELD" target="blank">MIDFIELD</a> is the acronym for the Multiple-Institution Database for Investigating Engineering Longitudinal Development. 

Longitudinal studies using MIDFIELD are not limited to Engineering---the database includes *all* student records from member institutions.


## Who we are

```{r}
library("tibble")
df <- tribble(
  ~name, ~position, ~email,
  "Matthew Ohland"    , "Principal Investigator/Director", "ohland@purdue.edu",
  "Marisa Orr"        , "Associate Director", "marisak@clemson.edu",
  "Catherine Brawner" , "Director Policy Analysis", "brawnerc@bellsouth.net",
  "Russell Long"      , "Managing Director/Data Steward", "ralong@purdue.edu",
  "Susan Lord"        , "Director MIDFIELD Institute", "slord@sandiego.edu",
  "Richard Layton"    , "Director Data Display", "layton@rose-hulman.edu"
)
knitr::kable(df, col.names = NULL)
```
 

 
## Funding

Current funding is provided by the US National Science Foundation, [Grant 1545667](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1545667) Expanding Access to and Participation in the Multiple-Institution Database for Investigating Engineering Longitudinal Development.    













