---
title: "About MIDFIELD"
output: rmarkdown::html_vignette
bibliography: "../../inst/REFERENCES.bib"
csl: "../../inst/body-and-society.csl"
always_allow_html: TRUE
---

<!-- generated from .Rmd Please edit that file -->

```{r setup, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  fig.path = "../../man/figures/pages-about-midfieldr-", 
  fig.align = "center",
  fig.show = "hold", 
  out.width = "70%",
  fig.width = 6,
  fig.asp = 0.7,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = FALSE
  # see note in README or error/purl settings
)
knitr::knit_hooks$set(inline = function(x) {
  if (!is.numeric(x)) {
    x
  } else if (x >= 10000) {
    prettyNum(round(x, 2), big.mark = ",")
  } else {
    prettyNum(round(x, 2))
  }
})
kable_no_header <- function(x, font_size) {
  kable_in <- knitr::kable(x, "html", col.names = NULL)
  kableExtra::kable_styling(kable_input = kable_in, 
                            font_size = font_size)
}
```

MIDFIELD is the acronym for the Multiple-Institution Database for Investigating Engineering Longitudinal Development <a href="https://engineering.purdue.edu/MIDFIELD" target="blank">(link)</a>. 

Longitudinal studies using MIDFIELD are not limited to Engineering---the database includes *all* student records from member institutions.


## Who we are

```{r}
df <- wrapr::build_frame(
  "name"             , "responsibility"                       , "contact"                  |
  "Matthew Ohland"   , "Principal Investigator/Director", "ohland@purdue.edu"      |
  "Marisa Orr"       , "Associate Director"             , "marisak@clemson.edu"    |
  "Catherine Brawner", "Director Policy Analysis"       , "brawnerc@bellsouth.net" |
  "Russell Long"     , "Managing Director/Data Steward" , "ralong@purdue.edu"      |
  "Susan Lord"       , "Director MIDFIELD Institute"    , "slord@sandiego.edu"     |
  "Richard Layton"   , "Director Data Display"          , "graphdoctor@gmail.com"
)
kable_no_header(df, font_size = 13)
```
 

 
## Funding

Current funding is provided by the US National Science Foundation, Grant 1545667, Expanding Access to and Participation in the Multiple-Institution Database for Investigating Engineering Longitudinal Development [(link)](https://www.nsf.gov/awardsearch/showAward?AWD_ID=1545667).    













