---
title: "Updating CIP data"
format:
  gfm:
    html-table-processing: none
csl: ../../inst/information-science-and-technology.csl
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-041-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))

# to read data file
path_to_raw_data <- find_root_file("data-raw",
  criterion = has_file("DESCRIPTION")
)
path_to_csv_file <- paste0(path_to_raw_data, "/CIPCode2010.csv")
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

Starting with a file of program data from the U.S. Integrated Postsecondary Education Data System (IPEDS), we construct a Classification of Instructional Programs (CIP) dataset in MIDFIELD format. The result, `cip2010,` is included with midfieldr.


##  Introduction

The IPEDS website [@cipcode:2026] provides downloadable CSV files for the 2010 and 2020 CIP data via their "Resources" tab under the file names: 

- `CIPCode2010.csv`
- `CIPCode2020.csv`

In this article, we document our procedure for converting `CIPCode2010.csv` data into MIDFIELD format. The result is `cip2010,` an update to midfieldr’s original `cip` dataset. Both datasets load with midfieldr.

```{r}
library("midfieldr")
look_at(cip)

look_at(cip2010)
```

Both datasets are based on the 2010 IPEDS data; `cip2010` has 267 more rows than `cip,` the majority of which (225 rows) are medical or dental residencies---programs that were not germane to the original MIDFIELD project. The new data set `cip2010` restores those missing rows. 

In documenting the conversion of IPEDS data to MIDFIELD format, this article  also provides a template for converting the 2020 IPEDS data to MIDFIELD format when needed. 



## Examine the raw data

```{r}
#| echo: false 
options(datatable.print.topn = 6)
```

Because we set some options as well as the random number seed, we save their existing settings to be restored at the end. 

```{r}
library("midfieldr")
library("data.table")
library("stringr")

backup_options <- options()
backup_seed <- NULL
if (exists(".Random.seed")) backup_seed <- .Random.seed
```

The CSV files are available for download from IPEDS [-@cipcode:2026]. Save the file to a convenient directory. Construct a path (in quotes) to that file, substitute your path for the placeholder `path_to_csv_file,` and import the data. 

```{r}
# import data
DT <- fread(path_to_csv_file, colClasses = "character")

# result
look_at(DT)
```

To construct a midfieldr-style data set, we need three columns: 

```{r}
cols_we_want <- c("CIPFamily", "CIPCode", "CIPTitle")
DT <- DT[, ..cols_we_want]

look_at(DT)
```

Closer look at a sample of our primary-source data. 

```{r}
# set the seed for reproducibility
set.seed(202612)

# sample 15 rows and order by the CIP code
x <- DT[sample(100:400, 15)]
setorderv(x, "CIPCode")

# result
x
```

Viewing the result, we observe:

- `CIPFamily` corresponds to `cip2` in midfieldr.
- `CIPCode` contains *all program codes* in a single column. In midfieldr, the 2-, 4-, and 6-digit codes are separated into columns `cip2`, `cip4`, and `cip6`. 
- `CIPTitle` contains *all program names* in a single column. In midfieldr, the the 2-, 4-, and 6-digit names are separated into columns `cip2name`, `cip4name`, and `cip6name`.

When constructing the midfieldr `cip` data set, some characters were changed from those in the source file, such as removing the decimal point separator in the 4- and 6-digit codes or replacing slashes ("/") with commas (","). For example, compare the `13.1209` entry above to its equivalent in the midfieldr `cip` data:

```{r}
# CIPCode2010
filter_programs(DT, "13.1209")

# midfieldr default
filter_programs(cip, "131209")[, .(cip2, cip6, cip6name)]
```

Additionally, source-file 2-digit program names are in all-caps; in midfieldr all program names are in title-case.

```{r}
# CIPCode2010
DT[CIPCode == "04", .(CIPCode, CIPTitle)]

# midfieldr default
cip[cip2 == "04", .(cip2, cip2name)][1]
```






## Transforming the raw data

### *Omit periods*

Omit the periods in the codes and titles. 

```{r}
DT <- DT[, CIPCode := gsub("[.]", "", CIPCode)]
DT <- DT[, CIPTitle := gsub("[.]", "", CIPTitle)]

# confirm result
filter_programs(DT, "131209")
```

### *Split the data frame*

Here we split the data frame into 2-, 4-, and 6-digit subsets by counting the number of digits in `CIPCode`, expected to be 2, 4, or 6. 

```{r}
# add a column of digit counts
DT[, N_digits := str_length(CIPCode)]

# confirm 2, 4, or 6 digits
unique(DT$N_digits)

# result
look_at(DT)
```

Create a separate data frame for 2-digit codes and names. We will use the new `cip2` variable to check our results---it should be a duplicate of  `CIPFamily.`

```{r}
DT2 <- DT[N_digits == 2, .(CIPFamily,
  cip2 = CIPCode,
  cip2name = CIPTitle
)]

look_at(DT2)
```

Repeat for the 4-digit codes and names. Extract the first two digits of `CIPCode` to compare to `CIPFamily.`

```{r}
DT4 <- DT[N_digits == 4, .(CIPFamily,
  cip2 = substr(CIPCode, 1, 2), 
  cip4 = CIPCode,
  cip4name = CIPTitle
)]

look_at(DT4)
```

Repeat for the 6-digit codes and names.

```{r}
DT6 <- DT[N_digits == 6, .(CIPFamily,
  cip2 = substr(CIPCode, 1, 2),
  cip4 = substr(CIPCode, 1, 4), 
  cip6 = CIPCode,
  cip6name = CIPTitle
)]

look_at(DT6)
```

Compare the 2-digit codes in each data frame. If the data were correctly entered, this check for equality should all be TRUE. 

```{r}
# This check for 2-digit equality should all be TRUE
all.equal(DT2$CIPFamily, DT2$cip2)
all.equal(DT4$CIPFamily, DT4$cip2)
all.equal(DT6$CIPFamily, DT6$cip2)
```


### *Editing*

Convert the 2-digit names from all-caps to title-case.

```{r}
DT2[, cip2name := str_to_title(cip2name)]
DT2[, cip2name := str_replace_all(cip2name, "And", "and")]

DT2[, .(cip2, cip2name)]
```


### *Rejoin data frames*

Select the columns we want for the joining operations.  

```{r}
cols_we_want <- c("cip2", "cip4", "cip6", "cip6name")
DT6 <- DT6[, ..cols_we_want]

cols_we_want <- c("cip2", "cip4", "cip4name")
DT4 <- DT4[, ..cols_we_want]

cols_we_want <- c("cip2", "cip2name")
DT2 <- DT2[, ..cols_we_want]
```

Left-join `cip4` to `cip6`, matching on the 2- and 4-digit code columns. Reuse the object name `DT` for our new working data frame. Left-join `cip2` to the result, matching on the 2-digit code columns. 

```{r}
DT <- DT4[DT6, on = c("cip2", "cip4")]

DT <- DT2[DT, on = c("cip2")]
```

Order rows and columns.

```{r}
setorderv(DT, c("cip6"))
setcolorder(DT, c("cip6name", "cip6", "cip4name", "cip4", "cip2name", "cip2"))

DT
```

If we've joined correctly, the following logical checks should yield `TRUE` in all cases. 

```{r}
# cip4(first 2 digits) should match cip2
unique(substr(DT$cip4, 1, 2) == DT$cip2) == TRUE

# cip6(first 2 digits) should match cip2
unique(substr(DT$cip6, 1, 2) == DT$cip2) == TRUE

# cip6(first 4 digits) should match cip4
unique(substr(DT$cip6, 1, 4) == DT$cip4) == TRUE
```



### Unspecified programs

Lastly, we add a row to the CIP data to includes one non-IPEDS code (999999) for Undecided or Unspecified, instances in which institutions reported no program information or that students were not enrolled in a program.

Create a data frame with one row.

```{r}
txt_99 <- "Undecided/Unspecified (non-IPEDS)"
row_99 <- data.table(
  cip2 = "99",
  cip4 = "9999",
  cip6 = "999999",
  cip2name = txt_99,
  cip4name = txt_99,
  cip6name = txt_99
)
```

Row-bind to our working data frame. 

```{r}
DT <- rbindlist(list(DT, row_99), use.names = TRUE)
DT
```

Here, we verify that the CIP dataset just completed is identical to the `cip2010` dataset that loads with midfieldr. 

```{r}
check_equiv_frames(DT, cip2010)
```




## Usage

To use the alternative CIP data set included with midfieldr, use `cip2010` as the value of the first argument in `filter_programs().` Compare, for example, 

```{r}
filter_programs(cip, "^1408")[, .(cip6name, cip6)]

filter_programs(cip2010, "^1408")[, .(cip6name, cip6)]
```

Here we see an instance of a minor difference in program names between the two datasets.

- `cip:` 140802 `r filter_programs(cip, "140802")[, (cip6name)]`
- `cip2010:` 140802  `r filter_programs(cip2010, "140802")[, (cip6name)]`

Alternatively, one can simply rename the alternate dataset for the convenience of the shorter name and overwrite `cip.`

```{r}
cip <- copy(cip2010)
check_equiv_frames(cip, cip2010)
```

The originals can be recovered as follows, 

```{r}
data(list = c("cip", "cip2010"), 
     package = "midfieldr", 
     overwrite = TRUE)
```

And the two datasets are no longer equivalent. Using the example from above, 

```{r}
check_equiv_frames(cip, cip2010)

filter_programs(cip, "^1408")[, .(cip6name, cip6)]

filter_programs(cip2010, "^1408")[, .(cip6name, cip6)]
```



## Wrap up

Restore options and random number generator to their original settings. 

```{r}
if (!is.null(backup_seed)) .Random.seed <- backup_seed
options(backup_options)
```


```{r child = paste0(path_to_rmd, "/common-wrap.Rmd")}
```
