---
title: "Programs"
format:
  gfm:
    html-table-processing: none
csl: ../../inst/information-science-and-technology.csl
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-040-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))

library("midfieldr")
library("data.table")
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

In the US, instructional programs are encoded by 6-digit numbers curated by the US Department of Education. The standard encoding format is a two-digit number followed by a period, followed by a four-digit number, for example, 14.0102. 

MIDFIELD encodes programs using the same 6 digits without the period, e.g., 140102, recorded as character strings under the  `cip6` variable in the relevant data tables. As strings, any leading zeros are preserved, e.g., 010101, 030101, etc. 

## Introduction

Academic programs have three levels of codes and names:

- 6-digit code, a specific program
- 4-digit code, a group of 6-digit programs of comparable content 
- 2-digit code, a grouping of 4-digit groups of related content 

Specialties within a discipline are encoded at the 6-digit level, the discipline itself is represented by one or more 4-digit codes (roughly  corresponding to an academic department), and a collection of disciplines  are represented by one or more 2-digit codes (roughly corresponding to an academic college).  

For example, Geotechnical Engineering (140802) is a specialty of Civil Engineering (1408) which is a department in a College of Engineering (14). 


```{r}
#| echo: false
df <- filter_programs(cip, c("^51", "^14", "^24", "^41"))

# N 6-digit codes
df <- unique(df[, .(cip2name, cip2, cip4, cip6)])
ncip6_51 <- nrow(df[cip2 == "51"])
ncip6_14 <- nrow(df[cip2 == "14"])
ncip6_24 <- nrow(df[cip2 == "24"])
ncip6_41 <- nrow(df[cip2 == "41"])

# N 4-digit codes
df <- unique(df[, .(cip2name, cip2, cip4)])
ncip4_51 <- nrow(df[cip2 == "51"])
ncip4_14 <- nrow(df[cip2 == "14"])
ncip4_24 <- nrow(df[cip2 == "24"])
ncip4_41 <- nrow(df[cip2 == "41"])

# 2-digit names
df <- unique(df[, .(cip2name, cip2)])
name51 <- df[cip2 == "51", cip2name]
name14 <- df[cip2 == "14", cip2name]
name24 <- df[cip2 == "24", cip2name]
name41 <- df[cip2 == "41", cip2name]
```

To illustrate the taxonomy in a little more detail, the table shows all programs under CIP 41 *`r name41`*, subdivided into (`r ncip4_41`) programs at the 4-digit level and (`r ncip6_41`) programs at the 6-digit level. Some 4-digit codes include only (1) 6-digit code, e.g., 4100 and 4101, while others include more than one, e.g., 4102 and 4103. 

```{r}
#| echo: false
library(gt)
x <- filter_programs(cip, "^41")
x[2:9, cip2name := "\U2003\U02193"]
x[c(4, 5, 7, 8), cip4name := "\U2003\U02193"]
setcolorder(x, c("cip2", "cip2name", "cip4", "cip4name", "cip6", "cip6name"))
x |>
  gt() |>
  tab_caption("Table 1. Example of CIP taxonomy") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  )
```


The number of programs represented by 2-digit codes vary over a wide range, for example, 

- CIP 14 *`r name14`* comprises (`r ncip4_14`) 4-digit codes and (`r ncip6_14`) 6-digit codes
- CIP 24 *`r name24`* comprise (`r ncip4_24`) 4-digit code and (`r ncip6_24`) 6-digit codes
- CIP 51 *`r name51`* comprise (`r ncip4_51`) 4-digit codes and (`r ncip6_51`) 6-digit codes











## Data

The dataset `cip` that loads with midfieldr contains program names and codes at the 6-digit, 4-digit, and 2-digit level.

```{r}
library("midfieldr")
library("data.table")

# Loads with midfieldr
cip
```

All variables in `cip` are character strings, which protects the leading zeros of CIP codes when present. 

```{r}
# 2-digit codes with leading zeros
cip[cip2 %like% "^0", .(cip2, cip2name)] |> unique()
```

The number of unique programs.

```{r}
# 2-digit level
length(unique(cip$cip2))

# 4-digit level
length(unique(cip$cip4))

# 6-digit level
length(unique(cip$cip6))
```

A sample of program names uses a random number generator, so your result will differ from that shown. 

```{r}
#| echo: false
# seed is reset in common closing
set.seed(20210613)
```

```{r}
# 2-digit name sample
sample(cip[, cip2name], 10)

# 4-digit name sample
sample(cip[, cip4name], 10)

# 6-digit name sample
sample(cip[, cip6name], 10)
```




## How to search

### `filter_programs()`

*`r filter_programs_one_line`*

```{r}
#| eval: false
# usage
filter_programs(dframe, # cip or equivalent
  pattern,              # search pattern
  ...,                  # subsequent arguments referable only by name
  negate = NULL         # default FALSE
)
```

The first argument is usually `cip` or a subset of `cip`. The output is a data frame with rows that contain matches or partial matches to the search pattern.  The forward pipe operator `|>` can be used if desired. Here, we use `check_equiv_frames()` to compare the results of equivalent statements.  

```{r}
# equivalent statements
x <- filter_programs(dframe = cip, pattern = c("engineering"))
y <- filter_programs(cip, "engineering")
z <- cip |> filter_programs("engineering")

# equivalent results
check_equiv_frames(x, y)
check_equiv_frames(x, z)
```

The `negate` argument, if true, drops rows that contain the search terms.

```{r}
x <- filter_programs(cip, "engineering")
x

filter_programs(x, c("^15", "^29", "51"), negate = TRUE)
```

## Examples

**Example 1**

```{r}
#| echo: false
options(datatable.print.nrows = 12)
```

Suppose we want to determine the 6-digit codes for literature programs. We could start with a keyword.

```{r}
pass_1 <- filter_programs(cip, "literature")
pass_1
```

To refine the search further, we might first examine the highest level, 2-digit categories. 

```{r}
unique(pass_1[, .(cip2name, cip2)])
```

If our search is for English-language literature, we can restrict the search for codes that start with 23 (regular expression `"^23"`) and drop the 2-digit values from the working data frame. 

```{r}
pass_2 <- pass_1[, .(cip6name, cip6, cip4name, cip4)]
pass_2 <- filter_programs(pass_2, "^23")
pass_2
```

Searching the result on "literature."

```{r}
pass_3 <- filter_programs(pass_2, "literature")
pass_3
```

If we wanted Canadian, US, or UK literature specifically, we can search for those terms and retain the 6-digit names and codes only. 

```{r}
pass_4 <- pass_3[, .(cip6name, cip6)]
filter_programs(pass_4, c("united", "canadian", "british"))
```

Alternatively, we could select the codes themselves, 

```{r}
filter_programs(pass_4, c("^2307", "^2308", "231402", "231403", "231404"))
```




**Example 2**

Suppose we are searching for history programs. We can start, as we did above, with a keyword search across all 2-, 4-, and 6-digit names then examine the resulting top-level programs 

```{r}
pass_1 <- filter_programs(cip, "history")
pass_1

unique(pass_1[, .(cip2name, cip2)])
```

It appears that the 2-digit code we want is 54. In the second pass, we focus on the 6-digit names and codes. 

```{r}
pass_2 <- pass_1[, .(cip6name, cip6)]
pass_2 <- filter_programs(pass_2, "^54")
pass_2
```

Assuming the programs we want are a subset of those shown, we can use the `negate` argument to drop selected programs by their ending string (e.g., regular expression `01$`).

```{r}
pass_3 <- filter_programs(pass_2, 
                          c("01$", "04$", "05$", "08$", "99$"), 
                          negate = TRUE)
pass_3
```


**Example 3**

Illustrating details. `catch_error()` is a midfieldr utility.

1. Search expressions must be strings.

```{r}
# incorrect
catch_error(
  filter_programs(cip, 050125)
)

# correct
filter_programs(cip, "050125")
```

2. The first two arguments do not have to be named.

```{r}
# equivalent statements
x <- filter_programs(dframe = cip, pattern = "^14")
y <- filter_programs(cip, "^14")

# equivalent results
check_equiv_frames(x, y)
```

3. The `negate` argument, if used, must be named.

```{r}
# incorrect
catch_error(
  filter_programs(pass_2, 
                  c("01$", "04$", "05$", "08$", "99$"), 
                  TRUE)
  
)

# correct
filter_programs(pass_2, 
                c("01$", "04$", "05$", "08$", "99$"), 
                negate = TRUE)
```


























```{r child = paste0(path_to_rmd, "/common-wrap-no-ref.Rmd")}
```
