---
title: "Case study"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-004-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

In this study we present a complete case, from metric to records to charts, with an emphasis on how midfieldr supports the *process* of working with longitudinal data, leaving the details of the code itself to later detailed articles. 

## Introduction 

*Metric.* &nbsp;  *Stickiness* is the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$. This metric includes part-time students, transfers, students admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

$$
  \small S = \frac{N_\textrm{grad}}{N_\textrm{ever}} = \frac{\mathrm{number\ of\ graduates\ of\ a\ program}}{\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

*Records.* &nbsp; Records up to and including a student's first degree term, if any. 

*Population.* &nbsp; Degree-seeking students for whom the data sufficiency constraint is satisfied. No exclusions due to part-time status, transfer status, admission term, or starting program. 

*Programs.* &nbsp; Civil, Electrical, Industrial, and Mechanical Engineering. 

*Groupings.* &nbsp; The metric requires counts of timely graduates and of students ever enrolled in a program. We will group by program, race/ethnicity and sex. 

We use the following packages:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```



## Choose the baseline records

*Load the data.* &nbsp; We use the practice data provided in the midfielddata package. Data are given in four tables: `student`, `course`, `term`, and `degree`, three of which we load for this case.  
 
```{r}
data(student, term, degree)
```

*Keep a copy of the original source data.* &nbsp; We copy the original data to preserve them should we need them later. This step allows us to continue using the simple names `student`, `term`, etc. as we work without affecting the original source data due to data.table's "change by reference" behavior. 

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

<!-- Having `student`, `term`, etc. in our computational workspace is convenient because they also appear as default argument values in some midfieldr functions. For example, if the object `degree`  exists in the environment, then we can write `add_post_bacc(dframe)` instead of `add_post_bacc(dframe, midfield_degree = degree)`. -->

*Reduce the number of columns.* &nbsp; Optional, but convenient for viewing data frames at intermediate stages. We subset our data to select the minimum set of columns required by midfieldr functions. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

look_at(student)

look_at(term)

look_at(degree)
```

*Exclude post-baccalaureate terms.* &nbsp; Identify rows of post-baccalaureate terms to exclude. This step does not apply to the `student` table because it contains no term information. This step adds two temporary columns to each data frame:

```{r}
term <- add_post_bacc(term, midfield_degree = degree)
degree <- add_post_bacc(degree, midfield_degree = degree)

look_at(term)

look_at(degree)
```

```{r}
#| echo: false
N_post_term <- nrow(term[term_status == "post-first-degree"])
N_post_degree <- nrow(degree[term_status == "post-first-degree"])
```

Summarizing by term status, we have a total of `r N_post_term` post-baccalaureate terms in `term` and `r N_post_degree` in degree. 

```{r}
term[, .N, by = c("term_status")]

degree[, .N, by = c("term_status")]
```

Excluding the post-baccalaureate terms, then dropping the two temporary columns, yields the final shape of the baseline records. 

```{r}
term <- term[term_status != "post-first-degree"]
degree <- degree[term_status != "post-first-degree"]

cols_to_drop <- c("first_degree_term", "term_status")
term[, (cols_to_drop) := NULL]
degree[, (cols_to_drop) := NULL]

look_at(term)

look_at(degree)
```

*Result.* &nbsp;

- baseline `student`  
- baseline `term`  
- baseline `degree`  


## Identifying the baseline population

The goal here is to develop a data frame with one column of IDs that satisfies our constraints of data sufficiency and degree-seeking. We start with the unique IDs from the `term` data table we just derived. 

```{r}
DT <- copy(term)
DT <- DT[, .(mcid)]
DT <- unique(DT)

DT
```

*Filter rows for data sufficiency.* &nbsp; Data sufficiency requires the timely completion term estimate. 

```{r}
DT <- add_timely_term(DT, midfield_term = term)

DT
```

We can drop the columns of supporting variables and add the data sufficiency columns.

```{r}
cols_to_drop <- c("level_i", "adj_span")
DT[, (cols_to_drop) := NULL]

DT <- add_data_sufficiency(DT, midfield_term = term)

DT
```

```{r}
#| echo: false
N_exclude <- nrow(DT[data_sufficiency != "include"])
N_include <- nrow(DT[data_sufficiency == "include"])
```

Summarizing by data sufficiency, we have a total of `r N_exclude` rows to exclude due to insufficient data, leaving a population of `r N_include`.  

```{r}
DT[, .N, by = c("data_sufficiency")]
```

Filtering for the "include" label, and dropping the extra columns, 

```{r}
DT <- DT[data_sufficiency == "include", .(mcid)]

DT
```

*Filter rows for degree seeking.* &nbsp; Using this result, we apply an inner join with the student data table, which contains degree-seeking students only. The result  contains rows that are common to both by ID---in effect, removing rows from `DT` of any non-degree-seeking students. We also select the ID column alone to return. 

```{r}
population <- student[DT, .(mcid), on = "mcid", nomatch = NULL]

population
```

It happens that all students in the practice data are degree-seeking, so this step did not reduce the size of our population. We include the step to illustrate our complete process. The result is our final, baseline population. 

*Result.* &nbsp; 

- baseline `population`




## Collecting and labeling study program codes

In MIDFIELD data tables, the `cip6` variable identifies the 6-digit code for the program in which a student is enrolled in a given term. 

Finding your program codes can be a bit of a trial and error process (although the NCES website does have a search feature as well). Let's start with Civil Engineering. 

```{r}
filter_cip("civil engineering")
```

We see that the 4-digit code 1408 represents the  Civil Engineering programs we want. We also note from the 2-digit code and name that Engineering in general has a 2-digit code of 14. So we'll take a first pass on "14" and use that result as the `cip` *argument* in the next pass. We select a subset of the columns to display.

```{r}
engr_cip <- filter_cip("^14")
civil_engr <- filter_cip("civil", cip = engr_cip)

civil_engr[, .(cip6, cip6name, cip4name)]
```

Moving on to Electrical Engineering, we find a 4-digit code of 1410. 

```{r}
elec_engr <- filter_cip("electrical", cip = engr_cip)
elec_engr

elec_engr[, .(cip6, cip6name, cip4name)]
```


Continuing in a similar fashion, we establish that the 4-digit codes we need for study are: 

- Civil Engineering 1408 
- Electrical Engineering 1410 
- Mechanical Engineering 1419  
- Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 

```{r}
#| echo: false
op <- options()
options(datatable.print.topn = 15)

x <- filter_cip(c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437"))

N_codes <- nrow(x)
```

From `cip`, we obtain `r N_codes` codes that start with any of the selected 4-digit codes. We can drop all columns except the 6-digit code and 4-digit name, leaving

```{r}
programs <- filter_cip(c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437"))
programs <- programs[, .(cip6, cip4name)]

programs
```

User-defined program names are nearly always required. We add a `program`  variable to label each of these `r N_codes` programs with abbreviations we will use in comparing metrics, i.e., Civil (CE), Electrical (EE), Mechanical (ME), and Industrial/Systems Engineering (ISE). 

```{r}
programs[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %chin% c("142701", "143501", "143601", "143701"), "ISE"
)]
```

Now we can drop all but the 6-digit code and the program labels we created. 

```{r}
programs <- programs[, .(cip6, program)]

programs
```

 The result is our programs data frame for the study. 

*Result.* &nbsp; 

- study `programs`


```{r}
#| echo: false
options(op)
```




## Blocs and groupings

Before computing our metric, we must group and summarize our final population. We have two blocs (from the definition of the metric):

- ever-enrolled  
- graduates  


and three grouping variables

- program
- race/ethnicity
- sex

We have a lot of flexibility in the order in which we construct our blocs and groupings, so what follows is only one of several effective solutions. 


## Timely-graduates bloc

We start with the baseline population. Like we did with the original source data files, we copy it to protect `population` from "by reference" changes.

```{r}
graduates <- copy(population)
```

Because we plan on this becoming a bloc of timely graduates, we add a `bloc` variable with the value "grad".

```{r}
graduates[, bloc := "grad"]

graduates
```

Inner join with `degree` excludes rows of non-completers. 

```{r}
graduates <- degree[graduates, .(mcid, bloc, cip6, term_degree),
  on = "mcid", nomatch = NULL
]

graduates
```

*Filter rows for timely completion.* &nbsp; Determining a student's completion status requires the timely term variable. 

```{r}
graduates <- add_timely_term(graduates, midfield_term = term)

graduates
```

We can drop the columns of supporting variables and add the completion status columns.

```{r}
cols_to_drop <- c("term_i", "level_i", "adj_span")
graduates[, (cols_to_drop) := NULL]

graduates <- add_completion_status(graduates, midfield_degree = degree)

graduates
```


```{r}
#| echo: false
N_include <- nrow(graduates[completion_status == "timely"])
```

Summarizing by completion status, we find `r N_include` rows with timely completion. 

```{r}
graduates[, .N, by = "completion_status"]
```

Filtering for the “timely” label and keeping the ID, bloc, and CIP columns, 

```{r}
graduates <- graduates[completion_status == "timely", .(mcid, bloc, cip6)]

graduates
```

We can complete this stage by filtering for programs.

```{r}
graduates <- programs[graduates, .(mcid, bloc, program, cip6), on = "cip6"]

graduates
```

Then drop all rows with an NA program value

```{r}
graduates <- graduates[!is.na(program)]

graduates
```

Drop the CIP code

```{r}
graduates[, cip6 := NULL]
graduates <- unique(graduates)

graduates
```

Summarize to check results so far

```{r}
graduates[, .N, by = "program"]
```







## Ever-enrolled group

Again, we start with the baseline set of IDs and we add a `bloc` variable with the value "ever". 

```{r}
ever_enrolled <- copy(population)
ever_enrolled[, bloc := "ever"]

ever_enrolled
```


Left-join adds all terms

```{r}
ever_enrolled <- term[ever_enrolled, .(mcid, bloc, cip6), on = "mcid"]

ever_enrolled
```


Filter by programs

```{r}
ever_enrolled <- programs[ever_enrolled, .(mcid, bloc, program, cip6), on = "cip6"]
ever_enrolled <- ever_enrolled[!is.na(program)]
ever_enrolled <- unique(ever_enrolled)
ever_enrolled
```

















```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
