---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---
  
```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))


# path_to_data_raw <- find_root_file("data-raw", criterion = has_file("DESCRIPTION"))
# source(paste0(path_to_data_raw, "/new_functions.R"))
```
  
```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

When working with student-level records to develop quantitative metrics, we recommend that you:
  
- Know the structure of your data.
- Define your metrics and the blocs of records they require.
- Plan your process of refining and shaping the data.

midfieldr helps you refine and shape your data in these areas:

- *Programs.* &nbsp; Collect 6-digit program codes. 
- *Records.* &nbsp; Credibly subset source data and refine the population to suit your study. 
- *Blocs.* &nbsp; Construct group of records for computing metrics.  
  
This document introduces you to midfieldr's basic set of tools.

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

```{r}
#| echo: false

options(
  datatable.print.topn = 3,
  datatable.print.nrows = 6
)
```


Please note:

1. *On syntax:* &nbsp; All data manipulation in our sample scripts is coded using  data.table syntax, e.g., choosing rows and columns, merging, grouping, summarizing, and reshaping data frames. Users preferring a different syntax would have to translate from data.table to their preferred system. However, midfieldr does provide modest support for tibbles in that functions attempt to return a data frame of the same class as the input data frame. 

2. *On functions:* &nbsp; Here, we provide a brief introduction only. Details are discussed at length in subsequent articles. You can always access function documentation, e.g., `?select_basic_cols`, `?add_term_cluster`, etc. for more  information.  







## midfieldr functions

The major midfieldr functions can be organized into three categories based on their contribution to a typical workflow:

Programs

- `filter_cip()` chooses rows of CIP data based on search terms. 

Records

- `select_basic_cols()` chooses columns required by midfieldr functions.
- `add_term_cluster()` identifies rows of post-baccalaureate terms to exclude. 
- `add_timely_term()` estimates a student's timely completion term. 
- `add_data_sufficiency()` identifies rows to exclude due to insufficient data.
- `add_completion_status()` determines if program completion is timely or late.

Special conditioning

- `prep_fye_mice()` conditions data for imputing starting majors of FYE students. 
- `order_multiway()` conditions data for Cleveland multiway charts. 










## Program data

*Collecting and labeling 6-digit program codes.*

The *Classification of Instructional Programs (CIP)* is a taxonomy of academic programs, encoded by 6-digit numeric codes curated by the US Department of Education [@NCES:2010]. 

The `cip` data set, loaded with midfieldr, is a subset of the NCES CIP2010 data that contains codes and names for `r nrow(cip)` instructional programs organized on three levels---a 6-digit series, a 4-digit series, and a 2-digit series---keyed by the `cip6` variable. 

```{r}
#| echo: false
options(
  datatable.print.topn = 5,
  datatable.print.nrows = 5
)
```

```{r}
cip
```





## `filter_cip()`

*Chooses rows of CIP data based on search terms.*

`filter_cip()` acts on the data frame assigned to its `cip` argument (default is `cip = cip`) to select rows that match or partially match search strings. Search strings are case-independent and can include regular expressions. 

```{r}
filter_cip("music", cip = cip)
```

To refine our results, we can assign the results of a first pass to the `cip` argument of a second pass. For example, our first pass below searches the default `cip` dataset for "music". Our second pass searches the results of the  first pass for any line that starts with "50". 

```{r}
first_pass <- filter_cip("music")
second_pass <- filter_cip("^50", cip = first_pass)
second_pass
```

Assuming we are looking for programs in a School of Music, the 4-digit code "5009" appears to be the correct finding. Our third pass searches the results of the second pass for any line that starts with "5009". 

```{r} 
third_pass <- filter_cip("^5009", cip = second_pass)
third_pass
```

```{r}
#| echo: false
options(
  datatable.print.topn = 20,
  datatable.print.nrows = 20
)
```

If these were part of our study programs, we would save the 6-digit codes and names and add a `program` variable to label each row as needed to suit our goals. Here, we insert a placeholder value "label_TBD". We might save the data frame as our `programs` data frame. 

```{r}
programs <- third_pass[, .(cip6name, cip6, program = "label_TBD")]
programs
```

Which specific programs we select depend on the study, of course, but in all cases the end result is a data frame like the one shown above with the 6-digit CIP codes, possibly the NCES 6-digit program name, and our own program labels.  







## Student-level data

*Credibly subset source data and refine the population to suit your study.*

```{r}
#| echo: false
options(
  datatable.print.topn = 5,
  datatable.print.nrows = 5
)
```

For this article we load the `student, term,` and `degree` tables from midfielddata. 

```{r}
data(student, term, degree)
```

We usually copy the source data, giving them new names (and new locations in memory), to keep them intact while we use the original names  --- `student`, `term`, and `degree` --- to do our work. 

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

The data tables are linked by `mcid`, the anonymized student ID. 

```{r}
look_at(student)

look_at(term)

look_at(degree)
```

`look_at()` is a midfieldr convenience function that wraps `base::str()`. 











## `select_basic_cols()` 

*Choose columns required by midfieldr functions.* 

`select_basic_cols()` operates on student records to reduce the number of columns to those required by other midfieldr functions plus the key or composite key variables of the four data tables. Shown below, the records have been reduced to no more than 5 columns required by other midfieldr functions. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

look_at(student)

look_at(term)

look_at(degree)
```

With a smaller number of columns, the printout of the data frame is more readable, a benefit when working with the data interactively. 

```{r}
term
```







## `add_term_cluster()`

*Identify rows of post-baccalaureate terms to exclude.*

`add_term_cluster()` identifies terms later than the first baccalaureate, if any. We are not generally interested in terms beyond the first degree term, so we use the results of this function to exclude post-first-degree terms. 

```{r}
term <- add_term_cluster(term)
degree <- add_term_cluster(degree)
```

`add_term_cluster()` adds a column indicating the cluster a term belongs to with respect to the first degree term.

```{r}
term
```

Using `sort_uniq()`, a midfieldr convenience function to sort unique values of a vector, we find the possible term cluster values to be:

```{r}
sort_uniq(term$term_cluster)
```

We filter to exclude all terms labeled "post-first-degree" and remove the extra columns. 

```{r}
term <- term[!"post-first-degree", on = "term_cluster"]
degree <- degree[!"post-first-degree", on = "term_cluster"]

term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

term
```




## `add_timely_term()`

*Determine the term by which degree completion would be considered timely.*

In this section, we begin refining the population, so we start with a unique set of IDs from the `term` record obtained above. 

```{r}
DT <- term[, .(mcid)]
DT <- unique(DT)

DT
```

`add_timely_term()` adds a column indicating the last term by which a student’s degree completion would be considered timely, generally 6 years after admission.  The timely completion term is required to determine data sufficiency as well as assessing timely completion.  

```{r}
DT <- add_timely_term(DT)

DT
```

## `add_data_sufficiency()`

*Identify members of the population to exclude due to insufficient data.*

The data sufficiency criterion limits student records to those for which available data are sufficient to assess timely completion.

`add_data_sufficiency()` adds a column that labels each row for inclusion or exclusion based on data sufficiency near the upper and lower bounds of an institution’s data range.

```{r}
DT <- add_data_sufficiency(DT)

DT
```

The possible values for data sufficiency are:

```{r}
sort_uniq(DT$data_sufficiency)
```

We filter to retain rows labeled "include". The resulting IDs define our baseline population. 

```{r}
population <- DT["include", on = "data_sufficiency", .(mcid)]

population
```

This population is then used to further refine the student records. An inner-join filters `student, term,` and `degree` to retain these IDs only, finalizing our records for the remainder of the study. 

```{r}
student <- population[student, on = "mcid", nomatch = NULL]
term <- population[term, on = "mcid", nomatch = NULL]
degree <- population[degree, on = "mcid", nomatch = NULL]
```




## `add_completion_status()`

*Determines if program completion is timely or late.*

This section often pertains to constructing a bloc of graduates, starting with the baseline population we obtained above. Completion status indicates whether or not a student completes a degree, and if they do, labeling it timely or late compared to their timely completion term. Thus we run `add_timely_term()` again. 

```{r} 
DT <- copy(population)
DT <- add_timely_term(DT)
```

`add_completion_status()` adds a column indicating completion status for every student.

```{r}
DT <- add_completion_status(DT)

DT
```

The possible values for completion status are:

```{r}
sort_uniq(DT$completion_status)
```

If we were constructing a bloc of timely graduates, we would filter to retain rows labeled "timely". The resulting IDs would define our graduates bloc.

```{r}
graduates <- DT["timely", on = "completion_status", .(mcid)]
graduates[, bloc := "grad"]
graduates
```






## Other functions

`prep_fye_mice()` &nbsp; Conditions data for imputing the starting majors of First-Year Engineering (FYE) students. Used when blocs of starters in Engineering are needed and an  institution has a required FYE program. For details see [FYE proxies](art-060-fye-proxies.html).   

`order_multiway()` &nbsp; Conditions data for Cleveland multiway charts. The ordering of its rows and panels is crucial to the perception of effects. Used when data have a multiway structure. For details see [Multiway data and charts](art-120-multiway.html). 

Utilities 

- `look_at()` wraps base `str()`.  
- `sort_uniq()` wraps base `sort()` and `unique()`.  
- `catch_error()` wraps base `tryCatch()` for errors.  
- `check_equiv_frames()` re-exported from the wrapr package  



```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```



