---
title: "Planning"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-011-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

The description section of the  [Case study](articles/art-004-case-study.html) article provides a good initial outline of how to plan your analysis using MIDFIELD student records. Under each heading, we'll pose some prompts to help you assemble the resources you'll need. 

*Data.* &nbsp; 

- What programs are to be included? 
- Do you know their CIP codes? 
- If you don't plan to use the `cip` dataset in midfieldr, do you have another source available? Do you know how it is structured?
- Are you planning to use the MIDFIELD research database? Have you requested access from ASEE?
- If not, do you have an alternative set of student-level data?  To use midfieldr, your data must be modeled on the MIDFIELD database.




*Metric.* &nbsp;  Program *stickiness:* the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$, including part-time students, migrators, transfers, and students admitted in any term  [@Ohland+Orr+others:2012]. 

$$
  \small S = \frac{\small N_\textrm{grad}}{\small N_\textrm{ever}} = \frac{\small\mathrm{number\ of\ graduates\ of\ a\ program}}{\small\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

*Programs.* &nbsp; Civil, Electrical, Industrial/Systems, and Mechanical Engineering. 

*Records.* &nbsp; Exclude records later than a student's first degree term; filter for data sufficiency and degree seeking; no exclusions due to part-time status, transfer status, admission term, or starting program.

*Population.* &nbsp; The set of unique IDs from the above records. 

*Blocs.* &nbsp; Students ever enrolled in the programs and timely graduates of the programs are required by the metric. 

*Groupings.* &nbsp; We select program, race/ethnicity, and sex for grouping and summarizing.

*Outcome.* &nbsp; To calculate the metric, we construct a data frame with columns for each grouping variable (program, race/ethnicity, and sex) and the counts by group $\small N_\textrm{grad}$ and $\small N_\textrm{ever}$. 

*Dissemination.* &nbsp; Exclude groupings too small to preserve anonymity. Edit column names to suit the audience. Condition/transform data as needed for tables or charts. 






```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
