---
title: "Alternative CIPs"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-041-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r}
#| echo: false
options(datatable.print.topn = 10)
```


Transforming primary-source program data from the US Integrated Postsecondary Education Data System (IPEDS) to the CIP data structure used by midfieldr. 

As of this writing (2026--05), the 2020 and 2010 (and earlier) versions of the US Classification of Instructional Programs (CIP) codes are downloadable from the IPEDS website [@cipcode:2026] under file names: 

- `CIPCode2020.csv`
- `CIPCode2010.csv`

In this article, we demonstrate transforming the data in the 2010 source file into the form required by midfieldr, resulting in the new data set `cip2010` included with the latest version of midfieldr. The procedure could be used to perform a similar transformation of `CIPCode2020.csv` as well. 



## Goals

The data set `cip` included with the earliest version of midfieldr is based on the IPEDS 2010 CIP data. However, `cip` content is not identical to the `cip2010` content. There are some typographical differences and the original `cip` data has fewer rows, thus fewer 6-digit codes. 

If for any reason the legacy `cip` data set fails to meet a user's needs, an alternative is to use the 2010 or 2020 primary-source data files from IPEDS. In this article, we demonstrate how to transform such data to the CIP format usable by the midfieldr tools. 


If you are writing your own script to follow along, we use these packages in this article:

```{r}
# Load packages
library(midfieldr)
library(data.table)
```


*Usage.* &nbsp; To use the alternative CIP data set included with midfieldr, use `cip2010` as the value of the `cip` argument. For example, 

```{r}
# Using the default cip
filter_cip("^1408", cip = cip)


# Using the alternate cip
filter_cip("^1408", cip = cip2010)
```




## Get started

Assign a new name to the legacy `cip` data set included with midfieldr, allowing us retain the original data set while using the name `cip` for the new work. 

```{r}
# Allows us to use "cip" as an object name  
legacy_cip <- copy(midfieldr::cip)
```

```{r}
#| echo: false
# Use package rprojrot to help with relative file paths
path_to_raw_data <- find_root_file("data-raw", 
                                   criterion = has_file("DESCRIPTION"))
path_to_csv_file <- paste0(path_to_raw_data, "/CIPCode2010.csv")
```

The first step is to download the CSV file from the IPEDs website. After saving it to a convenient directory, construct a path to that file (`path_to_csv_file`) and import the data with: 

```{r}
# Import the 2010 CIP file in data.table format
cip <- fread(path_to_csv_file, colClasses = "character")
```

The `dplyr::glimpse()` function is quite useful for a first impression of the data. (One could also use base R `str()`.)

```{r}
# View the result
dplyr::glimpse(cip)
```

To construct our own CIP data set, we need three columns: 

```{r}
# Select columns
cols_we_want <- c("CIPFamily", "CIPCode", "CIPTitle")
cip <- cip[, ..cols_we_want]

# View the result
dplyr::glimpse(cip)
```

Closer look at a sample of our primary-source data. 

```{r}
# Set the seed for reproducibility 
set.seed(202612)

# Extract a sample of 15 rows and reset the seed
x <- cip[sample(100:400, 15)]
set.seed(NULL)

# Order the rows by the CIP code
setorderv(x, "CIPCode")

# View the result
x
```

Viewing the result, we observe:

- `CIPFamily` corresponds to `cip2` in midfieldr.
- `CIPCode` contains *all program codes* in a single column. In midfieldr, the 2-, 4-, and 6-digit codes are separated into columns `cip2`, `cip4`, and `cip6`. 
- `CIPTitle` contains *all program names* in a single column. In midfieldr, the the 2-, 4-, and 6-digit names are separated into columns `cip2name`, `cip4name`, and `cip6name`.

When constructing the midfieldr legacy CIP data set, some values were changed from those in the source file, such as removing the decimal point separator in the 4- and 6-digit codes or replacing slashes ("/") with commas (","). For example, compare the `13.1209` entry above to its equivalent in the midfieldr legacy CIP data:

```{r}
# CIPCode2010
filter_cip("13.1209", cip = cip)

# midfieldr
filter_cip("131209", cip = legacy_cip)[, .(cip2, cip6, cip6name)]
```

Additionally, source-file 2-digit program names are in all-caps; in midfieldr all program names are in title-case.

```{r}
# CIPCode2010
unique(cip[CIPCode == "04"])[, .(CIPCode, CIPTitle)]

# midfieldr
unique(filter_cip("^04", cip = legacy_cip)[, .(cip2, cip2name)])
```






## Transforming the source data

```{r}
#| echo: false
options(datatable.print.topn = 5)
```

*Editing.* &nbsp; Omit the periods in the codes and titles. 

```{r}
# Omit the period separator on the CIP code
cip <- cip[, CIPCode := gsub("[.]", "", CIPCode)]

# Omit the period at the end of each title
cip <- cip[, CIPTitle := gsub("[.]", "", CIPTitle)]

# Verify the result
filter_cip("131209", cip = cip)
```

Count the number of digits in `CIPCode`, expected to be 2, 4, or 6. 

```{r}
# Add a column
cip[, N_digits := nchar(CIPCode)]

# View result
dplyr::glimpse(cip)

# Confirm 2, 4, or 6 digits
unique(cip$N_digits)
```

*Split* &nbsp; Create a separate data frame for 2-digit codes and names, keyed by `CIPFamily`.

```{r}
# Unique 2-digit codes and names
cip2 <- cip[N_digits == 2, .(CIPFamily, 
                             cip2 = CIPCode, 
                             cip2name = CIPTitle)]
# View the result
dplyr::glimpse(cip2)
```

Repeat for the 4-digit codes and names and add a 2-digit code extracted from the `CIPCode` column we can compare to `CIPFamily` to ensure they agree. 

```{r}
# Unique 4-digit codes and names, with cip2 codes derived from cip4
cip4 <- cip[N_digits == 4, .(CIPFamily, 
                             cip2 = substr(CIPCode, 1, 2), 
                             cip4 = CIPCode, 
                             cip4name = CIPTitle)]
# View the result
dplyr::glimpse(cip4)
```

Repeat for the 6-digit codes and names.

```{r}
# Unique 6-digit codes and names, with cip2 and cip4 codes derived from cip6
cip6 <- cip[N_digits == 6, .(CIPFamily, 
                             cip2 = substr(CIPCode, 1, 2), 
                             cip4 = substr(CIPCode, 1, 4), 
                             cip6 = CIPCode, 
                             cip6name = CIPTitle)]
# View the result
dplyr::glimpse(cip6)
```

Compare the 2-digit codes in each data frame. If the data were correctly entered, this check for equality should all be TRUE. 

```{r}
# This check for 2-digit equality should all be TRUE
all.equal(cip2$CIPFamily, cip2$cip2)
all.equal(cip4$CIPFamily, cip4$cip2)
all.equal(cip6$CIPFamily, cip6$cip2)
```

*Editing.* &nbsp;  Convert the 2-digit names from all-caps to title-case. Here we use two functions from the stringr package. 

```{r}
library(stringr)

# Convert to title case
cip2[, cip2name := str_to_title(cip2name)]

# Use lowercase "and"
cip2[, cip2name := str_replace_all(cip2name, "And", "and")]

# View result
cip2[, .(cip2, cip2name)]
```

*Combine.* &nbsp; Next, select the columns we want for the joining operations.  

```{r}
# Select 6-digit columns
cols_we_want <- c("cip2", "cip4", "cip6", "cip6name")
cip6 <- cip6[, ..cols_we_want]

# Select 4-digit columns
cols_we_want <- c("cip2", "cip4", "cip4name")
cip4 <- cip4[, ..cols_we_want]

# Select 2-digit columns
cols_we_want <- c("cip2", "cip2name")
cip2 <- cip2[, ..cols_we_want]
```

Now, left-join `cip4` to `cip6`, matching on the 2- and 4-digit code columns. 

```{r}
# Join `cip4` to `cip6` 
cip <- cip4[cip6, on = c("cip2", "cip4")]
```

Now, left-join `cip2` to the result, matching on the 2-digit code columns. 

```{r}
# Left join `cip2` to `cip` from above
cip <- cip2[cip, on = c("cip2")]

# Ensure row order and examine the result
setorderv(cip, c("cip6"))
cip
```


## Program undecided or unspecified

Lastly, we add a row to the CIP data to includes one non-IPEDS code (999999) for Undecided or Unspecified, instances in which institutions reported no program information or that students were not enrolled in a program.


```{r}
name_99 <- "Undecided/Unspecified (non-IPEDS)"
row_99  <- data.table(cip2 = "99", 
                            cip4 = "9999", 
                            cip6 = "999999", 
                            cip2name = name_99, 
                            cip4name = name_99, 
                            cip6name = name_99)


cip <- rbindlist(list(cip, row_99), use.names = TRUE)
cip
```


*Verify prepared data.* &nbsp; `cip2010`, included with midfieldr, contains the CIP data frame developed above. Here we verify that the two data frames have the same content.

```{r}
# Demonstrate equivalence
check_equiv_frames(cip, cip2010)
```





```{r}
#| echo: false
#| eval: false

# Execute manually to store data to package

# cip2010 <- copy(cip)
# 
# usethis::use_data(
#   cip2010,
#   internal  = FALSE,
#   overwrite = TRUE
# )
```



















***

[&#9665; Programs](art-040-programs.html) &nbsp;&nbsp; <a href="#top">&#9650; top of page</a> &nbsp;&nbsp; [Blocs &#9655;](art-050-blocs.html)

***






```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
