---
title: "Degree seeking"
output: github_document 
date: "`r Sys.Date()`"
csl: ../../inst/information-science-and-technology.csl
bibliography: ../../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = "../../man/rmd/common-setup.Rmd"}
```


Most analyses of Student Unit Records (SURs) omit records of students not seeking degrees. We use an *inner join* merging operation as a degree-seeking filter. This vignette treats degree seeking and inner joins generally.

This vignette in the MIDFIELD workflow.  

1. Planning  
1. `r accent("Initial processing")`  
    - Data sufficiency  
    - `r accent("Degree seeking")` 
    - Identify programs  
1. Blocs  
1. Groupings  
1. Metrics  
1. Displays   



## Method

By design, the `student` data table contains records of degree-seeking students only. We use an inner join with `student` to filter our working data frame for degree-seeking students. 

1. Filter source SURs for data sufficiency. 

1. Filter for degree seeking. 

```{r child = "../../man/rmd/note-practice-not-research.Rmd"}
```





## Load data

*Open.* &nbsp; If you are writing your own script to follow along, we use these packages in this vignette:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

*Load.* &nbsp; MIDFIELD practice data tables. Described in [Getting started](art-000-getting-started.html#load-data).

```{r}
# Load data
data(student, term)
```





## Initial processing

*(Optional) Select.* &nbsp; Columns required by midfieldr functions. Code reproduced from [Getting started](art-000-getting-started.html#reusable-code). 

```{r}
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_term <- copy(term)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
term <- select_required(source_term)
```

*Work.* &nbsp; Using `copy()` avoids *by-reference* effects  [@data.table-reference-semantics].

```{r}
# Working data frame
DT <- copy(term)
```

*Data sufficiency.*  &nbsp; Filter to satisfy the data sufficiency criterion. Code reproduced from [Data sufficiency](art-020-data-sufficiency.html#reusable-code).

```{r}
# Filter for data sufficiency, output unique IDs
DT <- add_timely_term(DT, term)
DT <- add_data_sufficiency(DT, term)
DT <- DT[data_sufficiency == "include", .(mcid)]
DT <- unique(DT)
```

We preserve this data frame as a baseline for examples in the vignette. 

```{r}
baseline <- copy(DT)
```









## Inner joins

An *inner join* is a merge operation that returns all observations (rows) from two data frames that match specified conditions in both. Using data.table syntax, we have two approaches: `merge(X, Y, by)` (similar to base R) and `Y[X, j, on]` (native to data.table).



### Using `merge(X, Y, by)`

The general form for an inner join is 

        merge(X, Y, by, all = FALSE)

where

- `X` is a data frame, matching rows returned
- `Y` is a data frame, matching rows returned
- by is the vector of shared column names to merge by 
- `all = FALSE` ensures the inner join 

In this example, the `Y` data frame is `student`, from which we extract the ID column before joining. Otherwise, all columns from both data frames would be returned. 

```{r}
# Select columns in Y
cols_we_want <- student[, .(mcid)]

# merge(X, Y) inner join
merge(DT, cols_we_want, by = c("mcid"), all = FALSE)
```





### Using `Y[X, j, on]`

The second approach---native to data.table and computationally more efficient---has the form

        Y[X, j, on, nomatch = NULL]

where

- `X` is a data frame, matching rows returned
- `Y` is a data frame, matching rows returned
- `j` selects columns from the joined data frame to retain (default all columns)
- `on` is the vector of shared column names to merge by 
- `nomatch = NULL` ensures the inner join 

```{r}
# Y[X] inner join
DT[student, .(mcid), on = c("mcid"), nomatch = NULL]
```

*Demonstrate equivalence.* &nbsp; Showing that the two approaches produce the same result and that, for inner joins, `X` and `Y` are interchangeable. 

```{r}
#| collapse: true

# merge(X, Y)
w <- merge(DT, cols_we_want, by = c("mcid"), all = FALSE)
# merge(Y, X)
x <- merge(cols_we_want, DT, by = c("mcid"), all = FALSE)
# X[Y]
y <- DT[student, .(mcid), on = c("mcid"), nomatch = NULL]
# Y[X]
z <- student[DT, .(mcid), on = c("mcid"), nomatch = NULL]

# Demonstrate equivalence
same_content(w, x)
same_content(w, y)
same_content(w, z)
```


### Selecting columns

In either method, we can select columns from both data frames. Using `merge()` we select the columns by explicitly subsetting the two data frames.  

```{r}
#| collapse: true

# Selecting columns from both data frames, merge() inner join
x <- merge(DT[, .(mcid)], student[, .(mcid, institution)], by = c("mcid"), all = FALSE)
setkey(x, NULL)
x[]
```

In the `X[Y, j]` syntax, however, we can list the columns to be returned from both data frames in the `j` list, that is, `.(var1, var2, etc.)`, without subsetting the original two data frames.  

```{r}
#| collapse: true

# Selecting columns from both data frames, X[Y] inner join
y <- DT[student, .(mcid, institution), on = c("mcid"), nomatch = NULL]
y[]
```

*Demonstrate equivalence.* &nbsp; Showing that the two approaches produce the same result. 

```{r}
#| collapse: true
# Demonstrate equivalence
same_content(x, y)
```






## Degree seeking

*Work.* &nbsp; The baseline data frame we preserved earlier is the intake for this section. 

```{r}
# Reusable starting state
DT <- copy(baseline)
DT[]
```

*Filter.* Use an inner join with `student` to filter `DT` to retain degree-seeking students. 

```{r}
# Inner join for degree seeking
DT <- student[DT, .(mcid), on = c("mcid"), nomatch = NULL]
```

*Filter.* &nbsp;  Filter to ensure IDs are unique.

```{r}
# One observation per ID
DT <- unique(DT)
DT[]
```






## Reusable code: Degree seeking

*Preparation.* &nbsp; The baseline data frame we preserved earlier is the intake for this section.

```{r}
DT <- copy(baseline)
```

*Degree seeking.* &nbsp; A summary code chunk for ready reference.   

```{r}
# Filter for degree seeking, output unique IDs
DT <- student[DT, .(mcid), on = c("mcid"), nomatch = NULL]
DT <- unique(DT)
```













## References

<div id="refs"></div>










```{r}
#| echo: false

# to change the CSS file
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding:     10px 20px;
    margin:      0 0 20px;
    border-left: 0px
}
caption {
    color:       #525252;
    text-align:  left;
    font-weight: normal;
    font-size:   medium;
    line-height: 1.5;
}
```

