---
title: "Case study"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-004-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

In this study we present a complete case, from initial description to graphing the results, in as concise a fashion as we can, emphasizing *process* over details. (Terminology and functions are described in detail in subsequent articles.) Here we emphasize how we work with longitudinal data and how midfieldr supports that process.

Like most midfieldr articles, this case is worked using data.table syntax. An alternate version using dplyr syntax is available [here](articles/art-005-case-dplyr.html). 



## Description

We define the parameters of our case study as follows:

*Data.* &nbsp; Program CIP codes from midfieldr `cip`. Student records from  midfielddata `student, term,` and `degree`. 

*Metric.* &nbsp;  Program *stickiness:* the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$, including part-time students, migrators, transfers, and students admitted in any term  [@Ohland+Orr+others:2012]. 

$$
  \small S = \frac{\small N_\textrm{grad}}{\small N_\textrm{ever}} = \frac{\small\mathrm{number\ of\ graduates\ of\ a\ program}}{\small\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

*Programs.* &nbsp; Civil, Electrical, Industrial/Systems, and Mechanical Engineering. 

*Records.* &nbsp; Exclude records later than a student's first degree term; filter for data sufficiency and degree seeking; no exclusions due to part-time status, transfer status, admission term, or starting program.

*Population.* &nbsp; The set of unique IDs from the above records. 

*Blocs.* &nbsp; The metric requires two blocs: students ever enrolled in the programs; and timely graduates of the programs. 

*Groupings.* &nbsp; The metric will be grouped by program, race/ethnicity, and sex.

*Outcome.* &nbsp; To calculate the metric, we construct a data frame with columns for each grouping variable (program, race/ethnicity, and sex) and bloc summary counts $\small N_\textrm{grad}$ and $\small N_\textrm{ever}$ by group.

*Dissemination.* &nbsp; Exclude groupings too small to preserve anonymity. Edit column names to suit the audience. Condition/transform data as needed for tables or charts. 

If you are writing your own script to follow along, we use these packages in this article:

```{r}
library("midfieldr")
library("midfielddata")
library("data.table")
library("cdata")
library("gt")
library("ggplot2")
```







## Programs

```{r}
#| echo: false
options(
  datatable.print.nrows = 8,
  datatable.print.topn = 4
)
```

One can start an analysis with program data or with student record data---the choice is arbitrary. We start with programs and set the results aside until needed when constructing our blocs. Our goal in this section is to search the CIP data table for the 6-digit codes for our programs. The `cip` dataset loads with midfieldr. 

### *Search for program codes*

Unless you already know your program CIP codes, finding them entails some trial and error.

`filter_programs()` searches `dframe` for string patterns. Searching for "civil engineering" yields programs in Engineering that we want and some in Engineering Technology that we do not. 

```{r}
filter_programs(cip, "civil engineering")
```

These results suggest that Engineering has the 2-digit code "14" and that Civil Engineering has the 4-digit code "1408". We can extract Civil Engineering alone by searching `cip` for lines that start with "1408", yielding six 6-digit codes. Regular expressions such as "^1408" are accepted. 

```{r}
filter_programs(cip, "^1408")
```

```{r}
#| echo: false
N_engr_programs <- nrow(filter_programs(cip, "^14"))
```

Knowing the 2-digit code for Engineering programs, our next search is for lines that start with "14". Note that the `cip` *argument* takes the `cip` dataset as its default *value.* The result is an Engineering subset of `cip` with `r N_engr_programs` rows.

```{r}
engr_cip <- filter_programs(cip, "^14")
engr_cip
```

Next, to search this result for Electrical Engineering, we assign `engr_cip` to the `cip` argument, yielding four 6-digit codes. 

```{r}
filter_programs(engr_cip, "electrical")
```


Continuing in a similar fashion, we find that our programs have the following 4-digit codes: 

- Civil Engineering 1408 
- Electrical Engineering 1410 
- Mechanical Engineering 1419  
- Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 


```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

### *Construct the programs table*

To collect all our 6-digit codes, we create a search string of the desired 4-digit codes. We drop all columns except the 6-digit names and 6-digit codes.

```{r}
codes_we_want <- c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437")
programs <- filter_programs(cip, codes_we_want)
programs <- programs[, .(cip6name, cip6)]

programs
```

The program names in `cip` are usually too long for effective use---user-defined names are nearly always required. So we add a `program`  variable with values "CE" (Civil Engineering), "EE" (electrical), "ME" (Mechanical), and "ISE" (Industrial/Systems Engineering). We also abbreviate a couple of terms for a slightly more compact display. 

```{r}
programs[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %like% c("^1427|^1435|^1436|^1437"), "ISE",
  default = NA_character_
)]
programs[, cip6name := gsub("Engineering", "Engng", cip6name)]
programs[, cip6name := gsub("Communications", "Commn", cip6name)]
programs[, cip6name := gsub("Electrical, Electronics", "Elec, Electr", cip6name)]

programs
```

Our programs data frame is complete: `r nrow(programs)` six-digit codes are encoded using 4 program labels. This data frame can sit in memory (or written to file) until we're ready to filter the blocs by program, joining data frames by matching on the `cip6` variable. 

```{r}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)
```








## Records

For this study we load three of the midfielddata data tables. 

```{r}
data(student, term, degree)
```

We usually copy the source data, giving them new names (and new locations in memory), to keep them intact while we use the original names  --- `student`, `term`, and `degree` --- to do our work, preventing the source data from being updated "by reference" as we work. *Reference semantics* in data.table is discussed in [@reference-semantics:2026].

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

For reference, these data frames have the following number  of rows.  

- student: `r nrow(student_source)` rows
- term: `r nrow(term_source)` rows
- degree: `r nrow(degree_source)` rows

```{r}
#| echo: false
orig_N_students <- length(unique(student$mcid))
```

Our approach is to apply the data sufficiency and degree-seeking criteria to refine the population, then exclude any post-baccalaureate terms. The resulting set of "source" data tables is a suitable foundation for most studies. 

### *Data sufficiency*

*Data sufficiency* assesses whether an institution's data range is sufficient to determine a student's completion status---timely, late, or NA---without ambiguity. Students for whom the data range is insufficient must be excluded from the population to avoid biased counts of both completers and non-completers. For details, see the discussion in [Data sufficiency](articles/art-020-data-sufficiency.html). 

We start with the full set of unique student IDs in the source data.

```{r}
DT <- term[, .(mcid)]
DT <- unique(DT)
DT
```

Our usual definition of *timely* program completion is 6 years after admission (4 or 8 years are also commonly encountered). The term at the upper limit of that span is the *timely completion term,* obtained using `timely_term().` Output variables are documented in `?timely_term.`

```{r}
DT <- timely_term(DT, midfield_table = term)
DT
```

Operating on this output, `data_sufficiency()` identifies records at the upper and lower bounds of an institution's data range that must be excluded. Output variables are documented in `?data_sufficiency.`

```{r}
DT <- data_sufficiency(DT, midfield_table = term)
DT
```

To assess the relative number of records to include or exclude, we count observations by the `data_sufficiency` variable. 

```{r}
DT[, .N, by = c("data_sufficiency")][order(-N)]
```

We filter to retain rows labeled "include" and drop all but the ID column.

```{r}
DT <- DT[data_sufficiency == "include", .(mcid)]
DT
```


### *Degree seeking*

We require all students in our study to be degree-seeking. By design, the `student` table contains only degree-seeking students. We inner-join the ID column  from the `student` table, matching on `mcid`. 

In effect, the inner join filters our population to remove any non-degree-seeking students. 

```{r}
student_cols <- student[, .(mcid)]
DT <- student_cols[DT, on = "mcid", nomatch = NULL]
DT
```

It happens that all students in this case are degree-seeking, so this step did not reduce the size of our population. Still, we include the step to illustrate our complete process.

### *Population*

Filtering for data sufficiency and degree-seeking gives us the starting population for most of our studies. 

```{r}
population <- copy(DT)
```

We use this population to filter the source records using an inner join, matching on ID.

```{r}
student_source <- population[student_source, on = "mcid", nomatch = NULL]
term_source <- population[term_source, on = "mcid", nomatch = NULL]
degree_source <- population[degree_source, on = "mcid", nomatch = NULL]
```

The number of rows is now 

- student: `r nrow(student_source)` rows
- term: `r nrow(term_source)` rows
- degree: `r nrow(degree_source)` rows







### *Post-baccalaureate terms*

We are not generally interested in terms beyond the first degree term, so we identify and exclude terms later than the first degree term in all the source data frames. 

```{r}
term <- copy(term_source)
degree <- copy(degree_source)
```

For each student and term in a data frame, `post_bacc_terms()` assigns every row a label indicating that a term belongs to one of three clusters: terms that are prior to the first degree term ("pre-degree"), equal to it ("first-degree"), or subsequent to it ("post-first-degree"). Output variables are documented in `?post_bacc_terms.` We don't apply the function to `student` because it has no term column. 

```{r}
term <- post_bacc_terms(term, midfield_table = degree)
degree <- post_bacc_terms(degree, midfield_table = degree)

term
```

To assess the relative size of the three clusters, we count observations by the `term_cluster` variable. 

```{r}
term[, .N, by = c("term_cluster")][order(-N)]

degree[, .N, by = c("term_cluster")][order(-N)]
```

We exclude the rows labeled "post-first-degree." Note that we are dropping terms but not reducing the population of student IDs. 

```{r}
term <- term[term_cluster != "post-first-degree"]
degree <- degree[term_cluster != "post-first-degree"]
```

We drop the temporary columns. 

```{r}
term[, c("term_cluster", "first_degree_term") := NULL]
degree[, c("term_cluster", "first_degree_term") := NULL]
```

We redefine our source material to incorporate the exclusion of post-baccalaureate terms. 

```{r}
term_source <- copy(term)
degree_source <- copy(degree)
```

The number of rows is now 

- student: `r nrow(student_source)` rows
- term: `r nrow(term_source)` rows
- degree: `r nrow(degree_source)` rows

We've reduced the number of unique students from `r orig_N_students` in the original source data to `r nrow(student_source)` that have satisfied our several constraints. 

Review the results.

```{r}
look_at(student_source)

look_at(term_source)

look_at(degree_source)

look_at(population)
```

From this point forward, anytime we need a fresh copy of any of the data tables, we copy the "source" version. Anytime we need a starting population, we copy  `population` or the unique IDs from `student_source.` 












## Blocs and groupings

The work up to this point is applicable to most studies. In summary, we have configured our:

- `programs` 6-digit program codes, names, and custom labels
- `student, term,` and `degree` records with post-baccalaureate terms removed and filtered for data sufficiency and degree seeking
- `population` the unique IDs in these records

The next steps depend on the metric and the groupings we assigned at the beginning. The stickiness metric requires these blocs:

- students with timely completion from the study programs
- students ever enrolled in these programs

And we selected these groupings:

- program
- race/ethnicity
- sex

We have a lot of flexibility in the order in which we construct our blocs and groupings, so what follows is only one of several effective solutions. Our approach here is to construct a bloc, filter by program, join the demographics, and repeat for the next bloc. 

First, we copy the source data tables so our work will not affect the source material by reference.

```{r}
student <- copy(student_source)
term <- copy(term_source)
degree <- copy(degree_source)
```




### *Select basic columns*

Convenient for viewing data frames at intermediate stages. We reduce the number of columns to those required by other midfieldr functions plus the key or composite key variables of the data tables. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

student

term

degree
```





## Timely graduates


We start with the baseline population. Like we did with the original source data files, we copy it to protect `population` from changes by reference.

```{r}
DT <- copy(population)
DT
```


### *Filter by program*

We left-join the CIP column from the `degree` table, matching on `mcid`. That we increase the number of rows indicates that some students have more than one degree in their first degree term. 

```{r}
degree_cols <- degree[, .(mcid, cip6)]
DT <- degree_cols[DT, on = "mcid"]
DT
```

Now we use an inner-join with our `programs` data frame, matching on `cip6`, to retain only those students who complete one of our study programs. We retain the `program`  column and drop the `cip6` column. 

```{r}
programs_cols <- programs[, .(cip6, program)]
DT <- programs_cols[DT, on = "cip6", nomatch = NULL]
DT[, cip6 := NULL]
DT <- unique(DT)
DT
```



### *Filter for timely completion*

We want to retain timely graduates. 

`completion_status()` builds on the output from `timely_term()` to label rows to indicate whether a student completes a degree timely or late compared to their timely completion term (or NA for no completion). Output variables are documented in `?completion_status.`

```{r}
DT <- timely_term(DT)
DT <- completion_status(DT)
DT
```

Another brief assessment. Here we compare the relative size of the three possible status labels. 

```{r}
DT[, .N, by = c("completion_status")][order(-N)]
```

We retain the rows labeled "timely" and the drop all the columns except the ID and program columns. 

```{r}
cols_we_want <- c("mcid", "program")
DT <- DT[completion_status == "timely", ..cols_we_want]
DT
```





### *Join demographics*

To add columns for student demographics, we left-join selected columns from the `student` table, matching on `mcid`. 

```{r}
DT <- student[DT, on = "mcid"]
DT
```







### *Bloc of timely graduates*

This is the bloc of timely graduates required by our metric. We add a `bloc` variable with the value "grad" and ensure we have unique rows.

```{r}
graduates <- copy(DT)
graduates[, bloc := "grad"]
graduates <- unique(graduates)
graduates
```





## Ever enrolled



Again we start with the baseline population. 

```{r}
DT <- copy(population)
DT
```


### *Filter by program*

We left-join the CIP column from the `term` table, matching on `mcid`. 

```{r}
term_cols <- term[, .(mcid, cip6)]
term_cols <- unique(term_cols)
DT <- term_cols[DT, on = "mcid"]
DT
```

CIP codes are also present in the `degree` table. Students working in a  multidisciplinary program may have CIP codes at graduation that do not appear in the `term` data, where only their primary major is recorded. We assume that if a student earns a degree in such a program we can consider them "ever enrolled" in the program.

From `degree,` we extract the CIP codes by ID and join them by rows to the previous data frame. 

```{r}
extra_cip <- copy(population)
degree_cols <- unique(degree[, .(mcid, cip6)])
extra_cip <- degree_cols[extra_cip, on = "mcid", nomatch = NULL]
DT <- unique(rbindlist(list(DT, extra_cip)))
DT
```

We repeat the process we used earlier to inner-join our `programs` data frame, matching on `cip6`. 

```{r}
programs_cols <- programs[, .(cip6, program)]
DT <- programs_cols[DT, on = "cip6", nomatch = NULL]
DT[, cip6 := NULL]
```

With the CIP code removed, we filter for unique rows. A student may switch CIP codes yet stay within a program as defined by our custom labels. We want to avoid counting that student as ever-enrolled in the same program more than once. 

```{r}
DT <- unique(DT)
DT
```

Another brief assessment. Here we compare the relative numbers of students ever enrolled in our programs. 

```{r}
DT[, .N, by = c("program")][order(-N)]
```



### *Join demographics*

Again, we left-join selected columns from the `student` table, matching on `mcid`.

```{r}
DT <- student[DT, on = "mcid"]
DT
```

### *Bloc of ever-enrolled*

This is the bloc of students ever enrolled in our programs required by our metric. We add a `bloc` variable with the value "ever" and ensure we have unique rows.

```{r}
ever_enrolled <- copy(DT)
ever_enrolled[, bloc := "ever"]
ever_enrolled <- unique(ever_enrolled)
ever_enrolled
```


## Outcomes

Combining the two data frames (blocs) by rows, we obtain the data structure we need for grouping and summarizing. 

```{r}
DT <- rbindlist(list(graduates, ever_enrolled), use.names = TRUE)
DT
```

### *Group and summarize*

Count the numbers of observations for each combination of the grouping variables. This data frame is our initial block-records form with four keys and one measurement *N*. 

```{r}
DT <- DT[, .N, by = c("bloc", "program", "race", "sex")]
DT
```

### *Reshape*

*Reshaping the data frame to calculate the metric.*

Transform from block-record form to row-record form. This operation is  known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data]. 

The data.table package uses `dcast()` for this operation. The key columns `{program, race, sex}` remain in place, the `bloc` column yields the new key columns `{grad, ever}`, and the values in the new columns are taken from the `{N}` column. 

```{r}
DT <- dcast(DT, program + sex + race ~ bloc, value.var = "N", fill = 0)
setkey(DT, NULL)
DT
```

Any missing values in `grad` can be set to zero because it is the numerator in the stickiness metric. Missing values in `ever` (the denominator) are removed. 

```{r}
DT[is.na(grad), grad := 0]
DT <- na.omit(DT)
setorderv(DT, c("program", "sex", "race"))
DT
```

The result has the data structure we called out in our project description for calculating the metric.  






 

### *Calculate the metric* 

*Completes the initial analysis.*

Stickiness is the ratio of the number of graduates to the number ever enrolled, expressed as a percentage. Stickiness is calculated for each combination of program, race/ethnicity, and sex. 

```{r}
DT[, stickiness := round(100 * grad / ever, 1)]
setkey(DT, NULL)
DT[order(-grad, -ever)]
```


## Dissemination

We take several additional steps before disseminating these results.  

To preserve the anonymity of the people involved, we remove observations with   $\small N$ or fewer observations. When dealing with the full MIDFIELD research data, we typically use $\small N = 10$, but for these practice data we illustrate the procedure using  $\small N = 3$. 

```{r}
DT <- DT[grad > 3]
DT
```

We have found it useful to report such data with a variable that combines race/ethnicity and sex. 

```{r}
DT[, people := paste(race, sex)]
setcolorder(DT)
DT
```

Readers can more readily interpret our charts and tables if the programs are unabbreviated. 

```{r}
DT[, program := fcase(
  program %like% "CE", "Civil",
  program %like% "EE", "Electrical",
  program %like% "ME", "Mechanical",
  program %like% "ISE", "Industrial/Systems"
)]
DT
```

### *Table*

```{r}
#| echo: false
options(datatable.print.nrows = 12)
```

Omit columns that won't appear in the table. 

```{r}
DT_table <- copy(DT)
DT_table[, c("race", "sex", "ever", "grad") := NULL]
DT_table
```

Transform the data from block-records to row-records with one row per "people" category (race/ethnicity/sex grouping).  

```{r}
DT_table <- dcast(DT_table, people ~ program, value.var = "stickiness")
setnames(DT_table, old = "people", new = "People", skip_absent = TRUE)
setkey(DT_table, NULL)
DT_table
```

Format the table for publication. 

```{r}
DT_table |>
  gt() |>
  tab_caption("Table 1. Engineering program stickiness (%)") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  )
```

```{r}
#| echo: false
options(datatable.print.nrows = 5)
```


### *Chart*

To use `ggplot()`, we want the data in its block-record form. 

```{r}
DT_chart <- copy(DT)
DT_chart
```

With one quantitative variable (stickiness) for every combination of the levels of two categorical variables (program and people), these are *multiway data* [@Cleveland:1993]. How one orders the categorical variables is critical for visualizing effects. 

`order_multiway()` converts the two categorical variables to ordered factors to support the ordering of rows and panels in the chart. The calculated stickiness values by group---which determine the ordering---are added in new columns. 

```{r}
DT_chart <- order_multiway(DT_chart,
  quantity = "stickiness",
  categories = c("program", "people"),
  method = "percent",
  ratio_of = c("grad", "ever")
)
DT_chart
```

Format the chart for publication. 

```{r}
#| label: fig01
#| fig-asp: 1.2
#| fig-cap: "Figure 1: Program stickiness."
ggplot(DT_chart, aes(x = stickiness, y = people)) +
  facet_wrap(vars(program),
    ncol = 1,
    as.table = FALSE
  ) +
  geom_vline(aes(xintercept = program_stickiness),
    linetype = 2,
    color = "gray60"
  ) +
  geom_point(size = 1.8) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```




```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
