---
title: "Stickiness"
format:
  gfm:
    html-table-processing: none
csl: ../../inst/information-science-and-technology.csl
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-110-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r}
#| include: false
library("ggplot2")
```

Stickiness is a more-inclusive alternative to graduation rate as a measure of a program's success in attracting, keeping, and graduating their undergraduates. Students excluded by a conventional graduation rate metric--including migrators---are included in the stickiness metric [@Ohland+Orr+others:2012]. 

## An inclusive metric 

Program *stickiness* $\small(S)$ is the ratio of the number of graduates of a program $\small(N_g)$ to the number ever enrolled in the program $\small(N_e)$.

$$
S = \frac{N_g}{N_e}
$$

Stickiness, in comparison to graduation rate, has these characteristics:

- Includes migrators, where graduation rate does not. 

- Is based on the bloc of ever enrolled rather than starters, so there is no need for FYE proxies.

- Counts all graduates (timely completers) in a program, eliminating the need to filter graduates based on their starting program. 

- Like the MIDFIELD definition of graduation rate (in contrast to the IPEDS definition), includes students who attend college part-time, who transfer between institutions, and who start in any term. 

As they pertain to the stickiness metric, relationships among starters, migrators, and graduates (timely completers) of a given program *P* are illustrated in Figure 1. 

- The overall rectangle represents the stickiness denominator $(N_e)$, the number of students ever enrolled in program *P*, including starters and migrators.

- The interior rectangle represents the stickiness numerator $(N_g)$, the number of graduates (timely completers) of program *P*.   


```{r}
#| echo: false
#| label: fig01
#| fig-width: 12
#| fig-asp: 0.7
#| fig-cap: "Figure 1. Stickiness metric. Starters, migrators, and timely completers."

df_tile <- data.frame(
  x = rep(c(2, 4), 2), # centerline of rectangle
  y = rep(c(1), each = 2), # centerline
  z = factor(rep(1:2))
)

delta <- 0.02

# x-position, center of circled numbers
c1 <- 2.57
c2 <- 1.75 # 1.55
c3 <- 4 # 3.43
c4 <- 4.25

df_box1 <- data.frame(
  x = c(1, 5, 5) + delta * c(-1, 1, 1),
  y = c(0.5, 0.5, 1.5) + delta / 2 * c(-1, -1, 1)
)
df_box2 <- data.frame(
  x = c(1, 1, 5) + delta * c(-1, -1, 1),
  y = c(0.5, 1.5, 1.5) + delta / 2 * c(-1, 1, 1)
)
df_circ <- data.frame(x = c(c1, c2, c3, c4), y = c(1, 1, 1, 1.35))
df_hash <- data.frame(
  x = c(3, 3), xend = c(5, 5),
  y = c(0.5, 1.5), yend = c(1.5, 0.5)
)
df_circ2 <- data.frame(x = 4, y = 0.99)

ggplot(df_tile, aes(x, y)) +
  geom_tile(aes(fill = z)) +
  scale_x_continuous(breaks = seq(0, 16, 2)) +
  scale_fill_manual(
    values = c("#80cdc1", "#80cdc1"), # "#dfc27d"  "#80cdc1"
    aesthetics = c("colour", "fill")
  ) +
  geom_vline(aes(xintercept = 3), color = "white") +
  geom_point(
    data = data.frame(x = 3, y = 0.95),
    aes(x = x, y = y),
    shape = 22,
    size = 190,
    color = "white",
    fill = "white",
    alpha = 0.4
  ) +
  theme_void() +
  theme(legend.position = "none") +
  geom_line(data = df_box1, aes(x = x, y = y), linewidth = 1, linetype = 2) +
  geom_line(data = df_box2, aes(x = x, y = y), linewidth = 1, linetype = 2) +
  scale_y_continuous(limits = c(0.4, 1.6)) +
  annotate("text",
    x = c(c1, c2, c3, c4, 3, 3),
    y = c(0.925, 1.45, 0.925, 1.45, 1.27, 1.55), # 1.37
    label = c(
      "", # starter-completers
      "starters in program P",
      "", #  migrator-completers
      "migrators into program P",
      "timely completers of program P",
      "ever enrolled in program P"
    ),
    hjust = 0.5,
    vjust = 0.5,
    size = 6
  )
```




## Application

In our [Case study](art-004-case-study.html), we construct the two major blocs required to calculate stickiness---the numbers of graduates $\small(N_g)$ and ever-enrolled $\small(N_e)$.

Rather than duplicate the complete case, we pick it up at an intermediate step just before grouping and summarizing. Those data load with midfieldr as `case_blocs.`

```{r}
library("midfieldr")
library("data.table")

DT <- copy(case_blocs)
DT
```

### *Group and summarize*

Count the numbers of observations for each combination of the grouping variables. Convert the count from integer to double format.

```{r}
DT <- DT[, .(N = as.double(.N)), by = c("program", "people", "bloc")]
DT
```



### *Reshape*

```{r}
#| echo: false
options(
  datatable.print.nrows = 28,
  datatable.print.topn = 14
)
```

We want to separate the $\small N$ column into two columns---one for the number of graduates and the other for the number of ever enrolled. This operation is  known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data]. 

The data.table package uses `dcast()` for this operation. The key columns `program` and `people` remain in place. The `bloc` column yields the new key columns `ever` and `grad` with values taken from the `N` column.

```{r}
DT <- dcast(DT,
  program + people ~ bloc,
  value.var = "N",
  drop = FALSE, # keep all combinations
  fill = NA_real_ # NA if no value
)
setkey(DT, NULL)
DT
```



### *Calculate the metric* 

Before calculating the metric, we address possible "divide by zero" errors by converting any zero values of `ever` to NA. Not required in this case, but included for completeness. 
  
```{r}
DT[ever == 0, ever := NA_real_]
```

Stickiness is calculated for each combination of program and people.

```{r}
DT[, stick := round(100 * grad / ever, 1)]
DT
```

We plot a subset of the results below for a quick overview of its range and distribution. For charts better designed for making comparisons, see the [Case study](art-004-case-study.html).

```{r}
#| label: fig02
#| fig-asp: 0.7
#| fig-cap: "Figure 2: Stickiness overview"

library("ggplot2")
dframe <- DT[grad > 10, group := paste(people, program)]
dframe <- na.omit(dframe)
ggplot(dframe, aes(x = stick, y = reorder(group, stick))) +
  geom_point(size = 1.8, na.rm = TRUE) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```





```{r child = paste0(path_to_rmd, "/common-wrap.Rmd")}
```
