---
title: "Case study"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-004-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
op <- options
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

In this study we present a complete case, from metric to charts, emphasizing the *process* of manipulating longitudinal data rather than the code, functions, or arguments. Subsequent articles deal with those details.






## Outline

In any study, the data handling process depends on how you structure your data  and how you define your metrics, population, blocs, programs, and groupings. In this study, we select the following:

*Data.* &nbsp; The midfieldr program codes in `cip` and the midfielddata practice data tables `student`, `term`, and `degree`. 

*Programs.* &nbsp; Civil, Electrical, Industrial, and Mechanical Engineering.  

*Metric.* &nbsp;  Program *stickiness:* the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$. This metric includes part-time students, transfers, students admitted in any term, and migrators [@Ohland+Orr+others:2012]. 

$$
  \small S = \frac{\small N_\textrm{grad}}{\small N_\textrm{ever}} = \frac{\small\mathrm{number\ of\ graduates\ of\ a\ program}}{\small\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

*Records.* &nbsp; Exclude records later than a student's first degree term. 

*Population.* &nbsp; Degree-seeking students for whom the data sufficiency constraint is satisfied. No exclusions due to part-time status, transfer status, admission term, or starting program. 

*Blocs.* &nbsp; The stickiness metric requires two blocs: students ever enrolled in the programs; and timely graduates of the programs without regard to their starting programs.

*Groupings.* &nbsp; Program, race/ethnicity, and sex will be used for grouping and summarizing. Groups too small to preserve anonymity will be excluded. 

*Results.* &nbsp; To group and summarize, we require a data frame with columns for bloc, program, race/ethnicity, and sex, keyed by student ID. After summarizing and computing the metric we should have columns for the metric (stickiness) and for each grouping variable (program, race/ethnicity, sex). 

If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```




## Programs

```{r}
#| echo: false
options(
  datatable.print.nrows = 15,
  datatable.print.topn = 15
)
```

One can start an analysis with program data or with student record data---the choice is arbitrary. We start with programs and set the results aside until needed. Our goal is to search the CIP data table for the 6-digit codes for our programs. The `cip` dataset loads with midfieldr, and is documented in `?cip`. 

Finding the program codes entails some trial and error. Let's start with Civil Engineering. 

```{r}
filter_cip("civil engineering", cip = cip)
```

We note two things:

- The 2-digit code for Engineering is "14"
- Civil Engineering (not "technology") has 4-digit code of "1408"

We take a first pass searching for "^14" (regular expression for "line that starts with 14") and use that result as the cip-argument in the next pass.

```{r}
engr_cip <- filter_cip("^14", cip = cip)
filter_cip("civil", cip = engr_cip)
```

In Engineering (at least), the 4-digit code seems to correspond to department-level, degree-granting programs. Again using `engr_cip`, we search for "electrical" and find the 4-digit code of "1410." 

```{r}
filter_cip("electrical", cip = engr_cip)
```

Continuing in a similar fashion, we find that our programs have the following 4-digit codes: 

- Civil Engineering 1408 
- Electrical Engineering 1410 
- Mechanical Engineering 1419  
- Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 

We create a search string of 4-digit codes and search. We drop all columns except the 6-digit names and 6-digit codes: `cip6` is the link from these data to the student records. 

```{r}
codes_we_want <- c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437")
programs <- filter_cip(codes_we_want, cip = engr_cip)
programs <- programs[, .(cip6name, cip6)]

programs
```

The program names in `cip` are usually too long for effective use---user-defined names are nearly always required. So we add a `program`  variable with values "CE" (Civil Engineering), "EE" (electrical), "ME" (Mechanical), and "ISE" (Industrial/Systems Engineering). We also abbreviate a couple of terms for a slightly more compact display. 

```{r}
programs[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %like% c("^1427|^1435|^1436|^1437"), "ISE",
  default = NA_character_
)]
programs[, cip6name := gsub("Engineering", "Engng", cip6name)]
programs[, cip6name := gsub("Communications", "Commn", cip6name)]

programs
```

Our programs data frame is complete: `r nrow(programs)` six-digit codes are encoded using 4 program names The original 6-digit names act as a check on the abbreviations we assigned. Later, as we prepare the blocs required for the metric, we will join the `program` values to the blocs by matching on `cip6`. 

This data frame can sit in memory (or written to file) until we're ready to filter the blocs by program. 




## Records

*Load the data.* &nbsp; We load three of the midfielddata data tables. 

```{r}
data(student, term, degree)
```

*Copy the source data.* &nbsp; We usually copy the source data, giving them new names (and new locations in memory), to keep them intact while we use the original names  --- `student`, `term`, and `degree` --- to do our work. Otherwise, the source data would be altered "by reference" as we work. 

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

Having `student`, `term`, and `degree` in our computational environment is convenient because these names appear as the default values for arguments  in some midfieldr functions. For example, if the object `degree`  exists in the environment, then we can write `add_post_bacc(dframe)` instead of `add_post_bacc(dframe, midfield_degree = degree)`.

*Choose columns required by midfieldr functions.* &nbsp; Optional, but convenient for viewing data frames at intermediate stages. We subset our data to select the minimum set of columns required by midfieldr functions. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

look_at(student)

look_at(term)

look_at(degree)
```

*Identify rows of post-baccalaureate terms to exclude.* &nbsp; This step identifies terms later than the first baccalaureate, if any. We are not generally interested in terms beyond the first degree term, so we use the results to exclude post-first-degree terms. This step does not apply to the `student` table because it contains no term information. 

This step adds two columns (`first_degree_term` and `term_status`) to a data frame:

```{r}
term <- add_post_bacc(term, midfield_degree = degree)
degree <- add_post_bacc(degree, midfield_degree = degree)

look_at(term)

look_at(degree)
```

```{r}
#| echo: false
N_post_term <- nrow(term[term_status == "post-first-degree"])
N_post_degree <- nrow(degree[term_status == "post-first-degree"])
```

Summarizing by term status, we have a total of `r N_post_term` post-baccalaureate terms in `term` and `r N_post_degree` in degree. 

```{r}
term[, .N, by = c("term_status")]

degree[, .N, by = c("term_status")]
```

We exclude rows with post-baccalaureate terms. 

```{r}
term <- term[term_status != "post-first-degree"]
degree <- degree[term_status != "post-first-degree"]
```

Dropping the two temporary columns and ensuring rows are unique yields the baseline records in their final configuration.

```{r}
cols_to_drop <- c("first_degree_term", "term_status")
term[, (cols_to_drop) := NULL]
degree[, (cols_to_drop) := NULL]

student <- unique(student)
term <- unique(term)
degree <- unique(degree)
```

These three data frames are the basis of all further analysis. 

```{r}
look_at(student)

look_at(term)

look_at(degree)
```




## Population

```{r}
#| echo: false
options(
  datatable.print.nrows = 5,
  datatable.print.topn = 5
)
```

Our goal is developing a data frame with one column of IDs that satisfies our constraints of data sufficiency and degree-seeking. We start with the unique IDs from the `term` data table we just derived. 

```{r}
DT <- copy(term)
DT <- DT[, .(mcid)]
DT <- unique(DT)
DT
```

*Filter rows for data sufficiency.* &nbsp; Data sufficiency requires the timely completion term estimate. 

```{r}
DT <- add_timely_term(DT, midfield_term = term)
DT
```

We can drop the columns of supporting variables and add the data sufficiency columns.

```{r}
cols_to_drop <- c("level_i", "adj_span")
DT[, (cols_to_drop) := NULL]

DT <- add_data_sufficiency(DT, midfield_term = term)
DT
```

```{r}
#| echo: false
N_exclude <- nrow(DT[data_sufficiency != "include"])
N_include <- nrow(DT[data_sufficiency == "include"])
```

Summarizing by data sufficiency, we have a total of `r N_exclude` rows to exclude due to insufficient data, leaving a population of `r N_include`.  

```{r}
DT[, .N, by = c("data_sufficiency")]
```

Filtering for the "include" label and dropping all but the ID column, 

```{r}
DT <- DT[data_sufficiency == "include", .(mcid)]
DT
```

*Filter rows for degree seeking.* &nbsp; Using this result, we apply an inner join with the student data table, which contains degree-seeking students only. The result  contains rows that are common to both by ID---in effect, removing rows from `DT` of any non-degree-seeking students.

```{r}
DT <- student[DT, on = "mcid", nomatch = NULL]
DT
```

It happens that all students in the practice data are degree-seeking, so this step did not reduce the size of our population. (We include the step to illustrate our complete process.) 

Extracting the column of IDs yields our baseline population.

```{r}
population <- DT[, .(mcid)]
population
```









## Blocs and groupings

Before computing our metric, we must group and summarize our final population. We have two blocs (from the definition of the metric):

- ever-enrolled  
- graduates  


and three grouping variables

- program
- race/ethnicity
- sex

We have a lot of flexibility in the order in which we construct our blocs and groupings, so what follows is only one of several effective solutions. 


## Timely-graduates bloc

We start with the baseline population. Like we did with the original source data files, we copy it to protect `population` from "by reference" changes.

```{r}
graduates <- copy(population)
```

Because we plan on this becoming a bloc of timely graduates, we add a `bloc` variable with the value "grad".

```{r}
graduates[, bloc := "grad"]

graduates
```

Inner join with `degree` excludes rows of non-completers. 

```{r}
graduates <- degree[graduates, .(mcid, bloc, cip6, term_degree),
  on = "mcid", nomatch = NULL
]

graduates
```

*Filter rows for timely completion.* &nbsp; Determining a student's completion status requires the timely term variable. 

```{r}
graduates <- add_timely_term(graduates, midfield_term = term)

graduates
```

We can drop the columns of supporting variables and add the completion status columns.

```{r}
cols_to_drop <- c("term_i", "level_i", "adj_span")
graduates[, (cols_to_drop) := NULL]

graduates <- add_completion_status(graduates, midfield_degree = degree)

graduates
```


```{r}
#| echo: false
N_include <- nrow(graduates[completion_status == "timely"])
```

Summarizing by completion status, we find `r N_include` rows with timely completion. 

```{r}
graduates[, .N, by = "completion_status"]
```

Filtering for the “timely” label and keeping the ID, bloc, and CIP columns, 

```{r}
graduates <- graduates[completion_status == "timely", .(mcid, bloc, cip6)]

graduates
```

We can complete this stage by filtering for programs.

```{r}
graduates <- programs[graduates, .(mcid, bloc, program, cip6), on = "cip6"]

graduates
```

Then drop all rows with an NA program value

```{r}
graduates <- graduates[!is.na(program)]

graduates
```

Drop the CIP code

```{r}
graduates[, cip6 := NULL]
graduates <- unique(graduates)

graduates
```

Summarize to check results so far

```{r}
graduates[, .N, by = "program"]
```







## Ever-enrolled group

Again, we start with the baseline set of IDs and we add a `bloc` variable with the value "ever". 

```{r}
ever_enrolled <- copy(population)
ever_enrolled[, bloc := "ever"]

ever_enrolled
```


Left-join adds all terms

```{r}
ever_enrolled <- term[ever_enrolled, .(mcid, bloc, cip6), on = "mcid"]

ever_enrolled
```


Filter by programs

```{r}
ever_enrolled <- programs[ever_enrolled, .(mcid, bloc, program, cip6), on = "cip6"]
ever_enrolled <- ever_enrolled[!is.na(program)]
ever_enrolled <- unique(ever_enrolled)
ever_enrolled
```

















```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
