---
title: "Case study"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-004-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

## Introduction

We illustrate how to work with longitudinal student records using midfieldr, focusing on the overall process and leaving detailed explanations of the coding to subsequent articles. The work is organized in three major topics: 

1. Obtaining a credible population and filtering the records to match.
2. Manipulating the data to produce the desired metric and groupings. 
3. Conditioning the results for dissemination in tables and charts. 


### *Scope*    

*Data.* &nbsp; Student records `student, term,` and `degree` from midfielddata and program codes `cip` from midfieldr.

*Population.* &nbsp; The IDs of degree-seeking students whose institutional data-ranges satisfy the data sufficiency requirement. 

*Records.* &nbsp; Data filtered to match the population and focused on undergraduate terms.

*Programs.* &nbsp; Four Engineering majors: Civil, Electrical, Industrial/Systems, and Mechanical.   

*Quantitative metric.* &nbsp;  Program *stickiness,* the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$.  

$$
\small S = \frac{\small N_\textrm{grad}}{\small N_\textrm{ever}} = \frac{\small\mathrm{number\ of\ graduates\ of\ a\ program}}{\small\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

*Blocs.* &nbsp; The metric requires two blocs: students ever enrolled in the programs; and timely graduates of the programs. 

*Groupings.* &nbsp; Group the findings by program, race/ethnicity, and sex.

*Dissemination.* &nbsp; Exclude groupings too small to preserve anonymity;   condition or transform data as needed for tables or charts. 

### *Essential terminology*

- The students comprising the population are expected to be undergraduates and  *degree-seeking,* i.e., attempting to complete a program. 

- *Program completion* means satisfying the requirements for a first baccalaureate degree. 

- The *timely-completion term* is the term by which we would consider their completion "timely", default 6 years after admission. 

- *Completion status* is "timely" for students graduating no later than their timely-completion term; "late" for graduating afterwards; and "NA" for non-completion. Completion status can be established only for records consistent with data sufficiency. 

- The *data sufficiency* test identifies students whose admission term and projected timely completion term lie within the range of data available from their institution---a necessary and sufficient condition for determining completion status. 




### *Getting started*

Packages used in this article:

```{r}
library("midfieldr") # working with student records
library("midfielddata") # practice data
library("data.table") # data manipulation
library("gt") # tables
library("ggplot2") # charts
```

We load three of the midfielddata data tables.

```{r}
data(student, term, degree)
```

We copy the tables, giving them new names (suffix `_source`) and new locations in memory. We can then use the original names without inadvertently updating the source tables "by reference" [@reference-semantics:2026].  

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

For comparing results as we subset the data, we start with the following number of rows in the original data frames. 

```{r}
#| echo: false
make_table <- function(x, cap) {
  gt(x) |>
    sub_missing() |>
    tab_caption(cap) |>
    tab_options(table.font.size = "small") |>
    opt_stylize(style = 1, color = "gray") |>
    tab_style(
      style = list(cell_fill(color = "#c7eae5")),
      locations = cells_column_labels(columns = everything())
    ) |>
    fmt_number(use_seps = TRUE, decimals = 0)
}
Ncompare <- wrapr::build_frame(
  "Table", "Source" |
    "student", eval(nrow(student_source)) |
    "term", eval(nrow(term_source)) |
    "degree", eval(nrow(degree_source))
)
setDT(Ncompare)
setnames(Ncompare, old = "Source", new = "Original tables")
make_table(Ncompare, "Table 1(a). Number of observations")
```


## Population

The baseline population used for most studies consists of degree-seeking students whose records satisfy the data sufficiency requirement.

### *Degree-seeking*

By design, the `student` data table contains all (and only) degree-seeking students. Thus we start with those IDs. 

```{r}
DT <- student[, .(mcid)]
DT
```

### *Data sufficiency*

We use `timely_term()` to determine the timely completion term for the degree-seeking students and add columns to the data frame to support those findings. 

```{r}
DT <- timely_term(DT, midf_table = term)
DT[order(entry_level)]
```

When displaying an intermediate result like this one, we often order the rows to illustrate some feature of the result. Here, for example, we order rows by  entry level which determines the adjusted span that leads to the timely completion term. 

We can reduce the number of columns to just those we need and use  `data_sufficiency()` to identify records that pass or fail the data sufficiency test. The added columns support the findings.

```{r}
DT <- DT[, .(mcid, entry_term, timely_term)]
DT <- data_sufficiency(DT, midf_table = term)
DT[order(-sufficiency)]
```

*Summary check.* &nbsp; A brief credibility check by summarizing the numbers of students in each category.

```{r}
DT[, .N, by = c("sufficiency")][order(-N)]
```

We retain rows labeled "include" and drop all but the ID column. This set of IDs is our baseline population.

```{r}
population <- DT[sufficiency == "satisfied", .(mcid)]
population
```


## Records

To obtain our baseline records, we exclude students not in our population and we exclude terms that are post-baccalaureate.

### *Population filter*

An inner join of the population data frame and each data table, matching on `mcid,` excludes IDs not in our population. 

```{r}
student <- population[student, on = "mcid", nomatch = NULL]
term <- population[term, on = "mcid", nomatch = NULL]
degree <- population[degree, on = "mcid", nomatch = NULL]
```

```{r}
#| echo: false
Ncompare[, x := c(
  nrow(student),
  nrow(term),
  nrow(degree)
)]
setnames(Ncompare, old = "x", new = "Population filter")
make_table(Ncompare, "Table 1(b). Number of observations")
```



### *Qualification level*

We are interested in *undergraduate* records: academic terms before a student's  first degree. We use `qualification_level()` to categorize terms as "undergrad" for terms before the first degree or "post-bacc" (post-baccalaureate) for terms after the first degree.

```{r}
term <- qualification_level(term, midf_table = degree)
degree <- qualification_level(degree, midf_table = degree)

term[order(-qual_level), .(mcid, term, bacc, qual_level)]

degree[order(-qual_level), .(mcid, term_degree, bacc, qual_level)]
```

*Summary check.* &nbsp; Summarize the numbers of students in each category.

```{r}
term[, .N, by = c("qual_level")][order(-N)]

degree[, .N, by = c("qual_level")][order(-N)]
```

We keep the terms at the undergraduate level.  

```{r}
term <- term[qual_level == "undergrad"]
degree <- degree[qual_level == "undergrad"]

# drop temporary columns
term[, c("bacc", "qual_level") := NULL]
degree[, c("bacc", "qual_level") := NULL]
```

We copy the current data tables to reserve them as our "baseline" records. From this point forward, anytime we need a fresh copy of any of the data tables, we copy the `*_baseline` version. Anytime we need a starting population, we copy  `population.`

```{r}
student_baseline <- copy(student)
term_baseline <- copy(term)
degree_baseline <- copy(degree)
```

```{r}
#| echo: false
Ncompare[, x := c(
  nrow(student),
  nrow(term),
  nrow(degree)
)]
setnames(Ncompare, old = "x", new = "Undergrad filter")
make_table(Ncompare, "Table 1(c). Number of observations")
```

Review the baseline records. `look_at()` is a midfieldr convenience function that wraps `base::str()` using our preferred arguments. 

```{r}
look_at(student_baseline)

look_at(term_baseline)

look_at(degree_baseline)

look_at(population)
```






## Blocs and groupings

With blocs and groupings we start narrowing our data from the general to the case-specific. We would usually start by copying the baseline tables, e.g.,

        student <- copy(student_baseline)

but in this particular script the baseline tables were created in the code chunk just above, so we know our working data tables `{student, term, degree}` already have the desired content. 

From our scope of work, the stickiness metric requires two blocs:

- students (timely) graduating from the programs
- students ever enrolled in the programs

And we listed three groupings:

- program
- race/ethnicity
- sex

This is a convenient point to work on the programs, given their importance to both blocs and groupings. 

## Programs

```{r}
#| echo: false
options(
  datatable.print.nrows = 16,
  datatable.print.topn = 8
)
```

In this section we search the `cip` dataset for the 6-digit codes for our case-study programs.

### *Search for program codes*

`filter_programs()` searches `dframe` for string patterns. Searching for "engineering" yields 2-digit CIP codes 14, 15, 29, and 51. From the program names, the 2-digit code we want is 14. 

```{r}
filter_programs(dframe = cip, pattern = "engineering")
```

Subset `cip` for the Engineering code that starts with 14. To unclutter our printouts, we can drop the 2-digit name and code. 

```{r}
engr_cip <- filter_programs(cip, "^14")
engr_cip <- engr_cip[, !c("cip2name", "cip2")]
```

Search that result for "civil". We find the Civil Engineering 4-digit code is 1408. 

```{r}
filter_programs(engr_cip, "civil")
```

Repeat for "electrical" and obtain the 4-digit code 1410.

```{r}
filter_programs(engr_cip, "electrical")
```

Continuing in a similar fashion, we find that our programs have the following 4-digit codes. 

- Civil Engineering 1408 
- Electrical Engineering 1410 
- Mechanical Engineering 1419  
- Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 








### *Construct the programs table*

We create a character vector of all the desired 4-digit codes.

```{r}
codes_we_want <- c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437")
programs <- filter_programs(engr_cip, codes_we_want)
programs
```

The `cip4name` values are not quite what we want for program identifiers. The Electrical Engineering name is inconveniently long and the four programs that comprise Industrial/Systems Engineering are not linked by name. We assign our own abbreviations to a user-defined `program` variable. 

```{r}
programs[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %like% c("^1427|^1435|^1436|^1437"), "ISE",
  default = NA_character_
)]
```

```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

In the student records, all program data is encoded using 6-digit codes, so we retain the 6-digit codes, the 6-digit names (for reference), and our program labels.  We shorten the longer names for a less cluttered display.

```{r}
programs <- programs[, .(cip6name, cip6, program)]
programs[, cip6name := gsub("Engineering", "Engng", cip6name)]
programs[, cip6name := gsub("Communications", "Commns", cip6name)]
programs[, cip6name := gsub("Electrical, Electronics", "Elec, Electr,", cip6name)]

programs
```

Our `programs` data frame is complete: `r nrow(programs)` six-digit codes labeled using four user-defined program abbreviations. This data frame can sit in memory (or written to file) until we're ready to join the programs to our blocs, matching on  `cip6.`  

```{r}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)
```











## Timely graduates

We start with the baseline population. Like we did with the original data files, we copy it to protect `population` from changes by reference.

```{r}
DT <- copy(population)
DT
```

### *Filter for timely completion*

We want to retain timely graduates only. We start by obtaining the timely completion term then selecting the variables we need to go forward. The `term` table here is identical to `term_baseline.`

```{r}
DT <- timely_term(DT, midf_table = term)
DT <- DT[, .(mcid, timely_term)]
DT
```

`completion_status()` builds on the output from `timely_term()` to label rows to indicate whether a student completes a degree timely or late compared to their timely completion term (or NA for non-completion). The `degree` table here is identical to `degree_baseline.`

```{r}
DT <- completion_status(DT, midf_table = degree)
DT
```

*Summary check.* &nbsp; Numbers of students in each category.

```{r}
DT[, .N, by = c("completion")][order(-completion)]
```

We retain the rows labeled "timely", drop all the columns except ID, and ensure unique rows. 

```{r}
DT <- DT[completion == "timely", .(mcid)]
DT <- unique(DT)
DT
```




### *Filter by program*

We left-join the CIP column from the baseline `degree` table, matching on `mcid.`

```{r}
#| echo: false
prev_nrows <- nrow(DT)
```

```{r}
degree_cip <- degree[, .(mcid, cip6)]
DT <- degree_cip[DT, on = "mcid"]
DT
```

```{r}
#| echo: false
new_nrows <- nrow(DT)
```

*Summary check.* &nbsp; That the number of rows above has increased from `r prev_nrows` to `r new_nrows` indicates that some students have more than one degree in their first degree term. To check, we count how many students completed $\small N$ degrees.  

```{r}
x <- DT[, .(N_degrees = .N), by = "mcid"]
x[, .(N_students = .N), by = "N_degrees"][order(N_degrees)]
```

We use an inner-join with our `programs` data frame to retain only the rows that have matching CIP codes in both tables. 

```{r}
program_abbrev <- programs[, .(cip6, program)]
DT <- DT[program_abbrev, on = "cip6", nomatch = NULL]
DT
```

We drop the `cip6` column, leaving the `program` column with our user-defined program labels. These students all have a degree in one or more of our four engineering programs. 

```{r}
DT[, cip6 := NULL]
DT <- unique(DT)
DT
```




### *Bloc of timely graduates*

We now have the bloc of timely graduates required by our metric. We add a `bloc` variable with the value "grad".

```{r}
graduates <- copy(DT)
graduates[, bloc := "grad"]
graduates
```

*Summary check.* &nbsp; Numbers of timely graduates by program. 

```{r}
graduates[, .N, by = c("program")][order(-N)]
```





## Ever-enrolled

Again we start with the baseline population. 

```{r}
DT <- copy(population)
DT
```


### *Filter by program*

We left-join the CIP column from the `term` table, matching on `mcid`. 

```{r}
term_cip <- term[, .(mcid, cip6)]
term_cip <- unique(term_cip)
DT <- term_cip[DT, on = "mcid"]
DT
```

CIP codes are also present in the `degree` table. Students working in a  multidisciplinary program may have CIP codes at graduation that do not appear in the `term` data, where only their primary major is recorded. We assume that if a student earns a degree in such a program we can consider them "ever enrolled" in the program.

From `degree,` we extract the CIP codes by IDs in our population and join them by rows to the previous data frame. We remove duplicate rows. 

```{r}
degree_cip <- unique(degree[, .(mcid, cip6)])
degree_cip <- population[degree_cip, on = "mcid", nomatch = NULL]
DT <- rbindlist(list(DT, degree_cip))
DT <- unique(DT)
DT
```

We repeat the process we used earlier to inner-join our `programs` data frame, matching on `cip6`. With the CIP code removed, we ensure unique rows. 

```{r}
program_abbrev <- programs[, .(cip6, program)]
DT <- DT[program_abbrev, on = "cip6", nomatch = NULL]
DT[, cip6 := NULL]
DT <- unique(DT)
DT
```

### *Bloc of ever-enrolled*

We now have the bloc of students ever enrolled in our programs required by our metric. We add a `bloc` variable with the value "ever."

```{r}
ever_enrolled <- copy(DT)
ever_enrolled[, bloc := "ever"]
ever_enrolled
```

*Summary check.* &nbsp; Numbers of students ever enrolled by program.

```{r}
ever_enrolled[, .N, by = c("program")][order(-N)]
```





## Outcomes

Combine the two data frames (blocs) by rows.

```{r}
DT <- rbindlist(list(graduates, ever_enrolled), use.names = TRUE)
DT
```

### *Join demographics*

Race/ethnicity and sex are recorded in the `student` table. We have found it useful to combine the two demographic columns into one column. 

```{r}
demographics <- student[, .(mcid, people = paste(race, sex))]
demographics
```

We left-join the demographics data frame, matching on `mcid`. 

```{r}
DT <- demographics[DT, on = "mcid"]
DT
```

We now have the data structure we need for grouping and summarizing. 

### *Group and summarize*

Count the numbers of observations for each combination of the grouping variables. These data are in block-record form with three keys, one value column `N`, and one row per value. We convert the count from integer to double. 

```{r}
DT <- DT[, .N, by = c("bloc", "program", "people")]
DT[, N := as.double(N)]
DT
```



### *Reshape*

```{r}
#| echo: false
options(
  datatable.print.nrows = 28,
  datatable.print.topn = 14
)
```

*Reshaping the data frame to calculate the metric.*

We want to separate the $\small N$ column into two columns---one for the number of graduates and the other for the number of ever enrolled. This operation is  known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data]. 

The data.table package uses `dcast()` for this operation. The key columns `program` and `people` remain in place. The existing `bloc` column yields the new key columns `ever` and `grad` with values taken from the `N` column.

```{r}
DT <- dcast(DT,
  program + people ~ bloc,
  value.var = "N",
  drop = FALSE, # keep all combinations
  fill = NA_real_ # NA if no value
)
setkey(DT, NULL)
DT
```

These data are in block-record form with two keys and two value columns. This is the data structure we require for calculating the metric.  






 

### *Calculate the metric* 

*Completes the analysis.*

Before calculating the metric, we address possible "divide by zero" errors by converting any zero values of `ever` to NA. Not required in this case, but included for completeness. 
  
```{r}
DT[ever == 0, ever := NA_real_]
```

Stickiness is calculated for each combination of program, race/ethnicity, and sex. 

```{r}
DT[, stick := round(100 * grad / ever, 1)]
DT
```

These data are in block-record form with two key columns (the grouping variables), one value column (stick), and thus one row per value. 




## Dissemination

We take several additional steps before disseminating these results.  

First, we remove rows with summary values that are small enough that student anonymity can no longer be assured. In this case we have a total of 11 rows with only 1, 2, or 3 graduates.  

```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

```{r}
DT[grad <= 3][order(grad)]
```

When dealing with the full MIDFIELD research data, we typically omit data from  rows in which $\small N_\mathrm{grad}\leq 10$, but for these practice data we illustrate the process using $\small N_\mathrm{grad}\leq 1$. We convert the values to NA. 

```{r}
DT[grad <= 1, c("ever", "grad", "stick") := NA_real_]
DT
```

We can optionally remove all NA rows for tables and charts. 

```{r}
DT <- na.omit(DT)
```

Readers can more readily interpret our charts and tables if the programs are unabbreviated. 

```{r}
DT[, program := fcase(
  program %like% "CE", "Civil",
  program %like% "EE", "Electrical",
  program %like% "ME", "Mechanical",
  program %like% "ISE", "Industrial/Systems"
)]
DT
```

### *Table*

```{r}
#| echo: false
options(
  datatable.print.nrows = 14,
  datatable.print.topn = 6
)
```

Retain the columns that appear in the table. The result is in block-record with two key column and one value column. 

```{r}
DT_table <- DT[, .(people, program, stick)]
setorderv(DT_table, c("people", "program"))
DT_table
```

Transform the data to row-record form. The key column `people` remains in place.  The existing `program` column yields the new key columns `Civil, Electrical,` etc., with values taken from the `stick` column. 

```{r}
DT_table <- dcast(DT_table,
  people ~ program,
  value.var = "stick",
  drop = FALSE,
  fill = NA_real_
)
setnames(DT_table, old = "people", new = "People")
setkey(DT_table, NULL)
DT_table
```

Format the table for publication. An em-dash indicates the number of graduates is too small to ensure confidentiality.

```{r}
DT_table |>
  gt() |>
  tab_caption("Table 1. Engineering program stickiness (%)") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  sub_missing() |>
  tab_style(
    locations = cells_column_labels(columns = everything()),
    style = list(cell_fill(color = "#c7eae5"))
  )
```

```{r}
#| echo: false
options(datatable.print.nrows = 5)
```


### *Chart*

To use `ggplot()`, we want the data in its original block-record form with one value column (stickiness). With one quantitative variable for every combination of the levels of two categorical variables (program and people), these are *multiway data* [@Cleveland:1993]. How one orders the categorical variables is critical for visualizing effects. 

`order_multiway()` converts the two categorical variables to ordered factors to support the ordering of rows and panels in the chart. Ordering is based on a calculation of aggregate stickiness values reported in two columns added to the data frame, one column per category. 

```{r}
DT_chart <- copy(DT)
DT_chart <- order_multiway(DT_chart,
  quantity = "stick",
  categories = c("people", "program"),
  method = "percent",
  ratio_of = c("grad", "ever")
)
DT_chart[, c("ever", "grad") := NULL]
DT_chart
```

Format the chart for publication. No arguments for ordering the data are required because the two categorical variables are ordered factors. 

```{r}
#| label: fig01
#| fig-asp: 1.2
#| fig-cap: "Figure 1: Program stickiness."
ggplot(DT_chart, aes(x = stick, y = people)) +
  facet_wrap(vars(program),
    ncol = 1,
    as.table = FALSE
  ) +
  geom_vline(aes(xintercept = program_metric),
    linetype = 2,
    color = "gray60"
  ) +
  geom_point(size = 1.8, na.rm = TRUE) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```

The vertical dashed line in each panel represents the overall stickiness of the program, calculated without regard to race/ethnicity and sex. The panels are ordered by those values, increasing from bottom to top. The stickiness of each "people" group, calculated without regard to program, determines the row order,  also increasing from bottom to top. 

In the next chart, we swap the roles of panels and rows. 

```{r}
#| label: fig02
#| fig-asp: 0.75
#| fig-cap: "Figure 2: Program stickiness."
ggplot(DT_chart, aes(x = stick, y = program)) +
  facet_wrap(vars(people),
    ncol = 2,
    as.table = FALSE
  ) +
  geom_vline(aes(xintercept = people_metric),
    linetype = 2,
    color = "gray60"
  ) +
  geom_point(size = 1.8, na.rm = TRUE) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```

In this version, the vertical dashed line in each panel represents the overall stickiness of the people group, calculated without regard to program. Panels are ordered by increasing group stickiness from left to right and from bottom to top. 


## Summary

We have presented a complete study, from a sample of registrar's data to charts comparing a quantitative metric, illustrating how we use midfieldr and other R packages to work with longitudinal student records.

Please note that the data in midfielddata are for *practice*, not *research*. These results cannot be used for drawing inferences about people or programs. 

```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
