---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
nocite: |
  @Layton+Long+Ohland:2026:midfielddata, 
  @Layton+Long+Ohland:2026:midfieldr
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

When working with student-level records, you should:

- Know the structure of your data
- Identify the programs to include
- Define your metrics and the relevant student blocs
- Develop the data processing steps that transform data to metrics

_midfieldr_ provides functions that contribute to this workflow in three ways:

1. Subsetting data frames in a MIDFIELD-tailored fashion. 
    - `filter_cip()`
    - `select_required()`
2. Adding decision variables to data frames for retaining records that satisfy  constraints.  
    - `add_timely_term()`
    - `add_data_sufficiency()`
    - `add_completion_status()`
3. Conditioning data frames for specific tasks.
    - `prep_fye_mice()`
    - `order_multiway()`
    
_midfielddata_ contributes to this workflow by providing data sets suitable for learning and practicing basic data skills in manipulating student-level records. midfielddata also serves as a model database for registrar's data in general. midfielddata are not, however, suitable for drawing inferences about program attributes or student experiences---midfielddata is for *practice*, not *research*.






If you are writing your own script to follow along, we use these packages in this article:

```{r}
library(midfieldr)
library(midfielddata)
library(data.table)
```

We use a couple of other packages as well for incidental tasks. We will load them at the point of use. 




## Data structure

*CIP data.* &nbsp; Classification of Instructional Programs (CIP) is a taxonomy of academic programs, encoded by 6-digit numeric codes curated by the US Department of Education [@NCES:2010]. 

The 2010 codes are loaded with midfieldr. `cip` contains codes and names for `r nrow(cip)` instructional programs organized on three levels: a 2-digit series, a 4-digit series, and a 6-digit series. 

CIP data are documented in `?cip`.

```{r}
# View data set
cip
```

`cip` is a "data.table" class of data frame (a format we use throughout midfieldr and midfielddata). The data.table "enhanced data.frame" is designed for fast manipulation of large data sets.   

```{r}
# Reveal the data.table class
class(cip)
```

Users can transform a data.table to a base R data.frame or a tidyverse tibble easily enough if they prefer. 

```{r}
# Base R
cip_DF <- as.data.frame(cip)

class(cip_DF)

# Tidyverse tibble
library(tibble)
cip_tbl <- as_tibble(cip)

class(cip_tbl)
```





<br>

*Student-records data.*  &nbsp; midfielddata is an R data package containing a sample of the larger MIDFIELD database [@ohland:midfield:2023]. midfieldr is designed to work with either database, as both are organized in four data tables (student, course, term, and degree) and have variable names in common. 


```{r}
#| echo: false

library(gt)
wrapr::build_frame(
  "Table", "Each row is", "N students", "N rows", "N columns", "Mb memory" |
    "course" , "one student per course per term", 97555, 3289532, 12L, 324.3 |
    "term"   , "one student per term"           , 97555,  639915, 13L,  72.8 |
    "student", "one student"                    , 97555,   97555, 13L,  17.3 |
    "degree" , "one student per degree"         , 49543,   49665,  5L,   5.2
) |>
  gt() |>
  tab_caption("Table 1. Student-level records in midfielddata") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  ) |>
  fmt_number(columns = 3:4, decimals = 0, sep_mark = ",", dec_mark = ".") |>
  fmt_number(columns = 6, decimals = 1, sep_mark = ",", dec_mark = ".")
```

<br>

Data tables from midfielddata can be loaded individually or collectively as needed.

```{r}
#| echo: false
#| eval: true

# Speed up the process by NOT loading course
data(student, term, degree)
```


```{r}
#| echo: true
#| eval: false
# Load one table as needed
data(student)

# Or load multiple tables
data(course, term, degree)
```


*Resources.*  &nbsp; If you are new to these data, the best place to start is the following articles at the midfielddata website:

- Data dictionaries are documented in `?student`, `?course`, `?term`, and `?degree`.

 - [MIDFIELD data structure.](https://midfieldr.github.io/midfielddata/articles/data-structure.html) We examine the structure of the four data tables in midfielddata: number of observations, number and class of variables, representative values, and database keys.  
 
 - [Data linked by student ID.](https://midfieldr.github.io/midfielddata/articles/individual-students.html)  To examine the variables and some representative values in midfielddata, we take a closer look at the records of individual students across the four data tables. 

 
<br>











## Subset CIP data with `filter_cip()`

`filter_cip()` acts on the `cip` data frame to select rows that match case-independent search terms, including regular expressions.

```{r}
# Sociology program codes
filter_cip("sociology")

# All music-related programs
filter_cip("music")
```

Having started a search, multiple passes are often need to narrow the search results. 

```{r}
# data.table printout option
options(datatable.print.topn = 20)

# First pass
first_pass <- filter_cip("music")

# Use the first pass as input to the second pass, select columns
filter_cip("^50", 
           cip = first_pass, 
           select = c("cip6", "cip6name"))

# restore original data.table printout option
options(datatable.print.topn = 3)
```

We develop these concepts further in [Programs](articles/art-040-programs.html).

<br>


## Subset student-level data with `select_required()`

`select_required()` selects key columns as well as all columns required by other midfieldr functions. Operates on a data frame to retain columns having names that match or partially match default search terms. Rows are unaffected.

Term records are significantly more compact if we select this minimum set of columns.

```{r}
# Select variables required by midfieldr functions
select_required(term)
```

We can include additional columns if we need them.

```{r}
# Select additional columns
select_required(term, select_add = c("gpa_term"))
```

Regular expressions can be used. 

```{r}
# Add columns using a regular expression
select_required(term, select_add = c("^gpa"))
```










## Add decision variables with `add_timely_term()`

`add_timely_term()` adds a new variable (plus supporting variables) for the term in which a student is expected to complete their program in a timely manner. This variable is used to make decisions about including or excluding records due to data sufficiency and whether or not a graduate is counted as a timely or late completer. Timely term is required for both `add_data_sufficiency()` and `add_completion_status()`. 

```{r timely_term}
# Initialize the working data frame using term
DT <- term[, .(mcid)]

# Retain a unique set of IDs
DT <- unique(DT)

# Add variables relating to the timely term
DT <- add_timely_term(DT, term)

# View the result
DT
```

We develop this concept further in the Data sufficiency article, [Timely term section](art-020-data-sufficiency.html#add_timely_term). 




## Add decision variables with `add_data_sufficiency()`

`add_data_sufficiency()` adds a new variable (plus supporting variables) for data sufficiency status.

```{r data_sufficiency}
# Retain a minimum number of variables
DT <- DT[, .(mcid, timely_term)]

# Add variables relating to data sufficiency
DT <- add_data_sufficiency(DT, term)

# View the result
DT[order(data_sufficiency)]
```


We develop these concepts further in [Data sufficiency article](art-020-data-sufficiency.html). 

<br>



## Add decision variables with `add_completion_status()`

`add_completion_status()` adds a new variable (plus supporting variables) for program completion status.

```{r completion_status}
# Retain a minimum number of variables
DT <- DT[, .(mcid, timely_term, data_sufficiency)]

# Add variables relating to completion status
DT <- add_completion_status(DT, degree)

# View the result
DT[order(completion_status)]
```

Note we have not yet excluded any student records. We still have the `r nrow(DT)` unique students from midfielddata. 

We develop these concepts further in [Graduates](articles/art-080-graduates.html). 







## Decisions

The data sufficiency criterion is nearly always imposed. We retain the records labeled "include". 

```{r}
# Retain a minimum number of variables
DT <- DT[, .(mcid, data_sufficiency, completion_status)]
DT <- unique(DT)

# Filter for data sufficiency
DT <- DT[data_sufficiency == "include"]

# View the result
DT
```

The number of students has dropped to `r nrow(DT)`.  When the timely completion criterion is imposed, we retain the records labeled "timely". 

```{r}
# Filter for timely completion
DT <- DT[completion_status == "timely"]
DT <- unique(DT)

# View the result
DT
```

The number of students has dropped to `r nrow(DT)`. 






## Next steps

Having filtered our set of student records to satisfy the constraints of data sufficiency and timely completion, we typically omit all variables except ID and use this data frame of IDs as the baseline for the next step of the analysis. 

```{r}
# Retain ID only
DT <- DT[, .(mcid)]

# Ensure unique IDs
DT <- unique(DT)

# View the result
DT
```


## Condition data for imputation with `prep_fye_mice()`

One of the most common (though flawed) program metrics is "graduation rate", roughly defined as the ratio of the number of timely completers to the number of  starters in a major. However, some engineering programs require that students enroll in a First-Year Engineering (FYE) program (CIP "140102") before they can enroll in their preferred, degree-granting, engineering major. 

To credibly assign FYE students to a "starting" major, we introduce the idea of “FYE proxies”---estimates of the 6-digit CIP codes of the degree-granting engineering programs that FYE students would have declared had they not been required to enroll in FYE. `prep_fye_mice()` estimates some of the FYE proxies, treats the remainder as missing values, and conditions the data frame for imputation using the mice R package. 

```{r}
# Condition data for imputation by mice
prep_fye_mice(student, term)
```

The function returns one row per FYE student keyed by student ID. All variables except ID are returned as factors to meet the requirements of `mice()`. The NA values in the proxy column are the unknowns CIP 6-digit codes to be imputed. 

We develop these concepts further in [FYE proxies](articles/art-060-fye-proxies.html).






## Condition data for multiway charts with `order_multiway()`

In working with longitudinal student-level records, we regularly encounter data structured as *multiway data*. The basic structure of multiway data is:

- a categorical variable with $\small m$ levels
- a categorical variable (independent of the first variable) with $\small n$ levels 
- a quantitative variable (the response) of length $\small m \times n$ with a value of the response for each combination of categorical levels

```{r}
#| echo: false
# count values for the inline chunks below
x <- student[, .(institution, race_ethn = race)]
x <- x[, .(N_matric = .N), by = c("institution", "race_ethn")]
cat1 <- length(unique(x$institution))
cat2 <- length(unique(x$race_ethn))
resp <- nrow(x)
```

For example, from the `student` data table, if we count the number of students at matriculation by race/ethnicity and institution, the resulting data structure is: 

- category 1: institution with `r cat1` levels
- category 2: race/ethnicity with `r cat2` levels
- quantitative response: N students with `r resp` observations

```{r}
# Extract two columns from student
DT <- student[, .(inst = institution, race_ethn = race)]

# Count by race/ethnicity and institution
DT <- DT[, .N, by = c("inst", "race_ethn")]

# Convert integer N to numeric N
DT[, N := as.double(N)]

# Order the rows for viewing
setorderv(DT, c("inst", "race_ethn"))

# View the result
DT
```

`order_multiway()` conditions such data for plotting in a *multiway chart*, a set of faceted dot charts. The categorical variables are converted to factors with levels ordered by the response variable. This ordering creates the row and panel ordering that is so crucial to the visual perception of effects in a multiway chart. 

```{r}
# Order the categorical variables by the median of the response
DT_multiway <- order_multiway(DT, 
                              quantity = "N", 
                              categories = c("inst", "race_ethn"), 
                              method = "median") 
# View result
DT_multiway
```

`order_multiway()` adds variables documenting the median values used to order each of the categories. This ordering contributes to the row and panel order in the faceted chart we plot below.

```{r}
#| echo: true
#| label: fig01
#| fig-width: 8.4

library(ggplot2)
ggplot(DT_multiway, aes(x = N, y = race_ethn)) +
  facet_wrap(vars(inst), ncol = 1, as.table = FALSE) +
  geom_vline(aes(xintercept = inst_median), 
             linetype = 2, 
             color = "darkgray") +
  scale_x_log10() +
  labs(x = "Number of students (log base 10)", 
       y = "", 
       title = "Population distribution at matriculation", 
       subtitle = "1998-2018", 
       caption = "Source: midfielddata") +
  geom_point()
```

The vertical references lines are the median value by race/ethnicity at the institution. We use a log-base-10 scale because of the orders of magnitude difference in values. Observations:

- Inst. B: International students in greater numbers than expected 
- Inst. J: Hispanic students in fewer numbers than expected
- Inst. B and C: Asian students in fewer numbers than expected

By greater or fewer "than expected", we mean in comparison to their overall numbers across all institutions. 

We develop these concepts further in [Multiway data and charts](articles/art-120-multiway.html).






## Reusable code

*Preparation.* &nbsp; The immediate prerequisites or "intake" required by the reusable code chunk are the source data tables.  

```{r}
#| eval: false
# Load source data (as needed)
data(student, course, term, degree)
```

*Initial data processing.* &nbsp; A summary code chunk for ready reference. 

```{r}
#| eval: false
# Optional. Copy of source files with all variables
source_student <- copy(student)
source_course  <- copy(course)
source_term    <- copy(term)
source_degree  <- copy(degree)

# Optional. Select variables required by midfieldr functions
student <- select_required(source_student)
course  <- select_required(source_course)
term    <- select_required(source_term)
degree  <- select_required(source_degree)
```

The `copy()` function ensures that "by-reference" changes to the working copies have no effect on the source copies we set aside. For example, editing `student` will have no effect on `source_student`. Thus the variables in the source material are always available if we need them. Concepts of data.table "reference semantics" are discussed further in [@data.table-reference-semantics]. 


***

[&#9665; midfieldr](../index.html) &nbsp; <a href="#top">&#9650; top of page</a> &nbsp;[Case study goals &#9655;](art-001-case-goals.html) 

***


```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```


 
