---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r, setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r, chunk01,  child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r, chunk02}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)

# display variables added by midfieldr functions
library(gt)
added_variables_table <- function(df) {
  gt(df) |>
    tab_options(table.font.size = pct(80), 
                container.height = pct(80)) |>
    fmt_markdown(columns = 1) |>
    cols_label(
      var = "variable",
      val = "description"
    )
}
```

In this document, we introduce midfieldr's basic tools and show how to apply them to data frames of student-level records (registrar's data). We organize the topics following a typical workflow:

- data
- population
- records
- blocs
- dissemination

*Packages.* We manipulate data using data.table syntax. 

```{r, chunk04}
library("midfieldr")
library("midfielddata")
library("data.table")
```



## Data

### *Student records*

midfieldr is designed to work with the MIDFIELD research database or any  database with a similar structure such as the practice data in the [midfielddata](https://midfieldr.github.io/midfielddata/index.html) package. If you are new to MIDFIELD, the best place to start is the article on [Data structure](https://midfieldr.github.io/midfielddata/articles/data-structure.html).

For this article we load all four midfielddata tables. 

```{r, chunk06}
# data(student, term, course, degree)
data(student, term, degree)
```

`select_basic_cols()` returns a subset of each table with the variables most often encountered in the early stages of a project. The tables are linked by the anonymized student ID variable `mcid.`

```{r, chunk07}
select_basic_cols(student)

select_basic_cols(term)

# select_basic_cols(course)

select_basic_cols(degree)
```

*Notes.*

1.  Like the data frames above, all data frames in midfieldr are `data.table` enhanced. However, if you convert these tables to another class such as `tbl_df` (tibbles) or `data.frame` (base R), midfieldr functions attempt to return data frames of the same class to support your preferred syntax. 

2.  Term variables `{term, term_course, term_degree}` are encoded as character strings `YYYYT`, where `YYYY` is the year at the start of the academic year and `T` encodes the semester or quarter within an academic year as Fall (1), Winter (2), Spring (3), or Summer (4, 5, and 6).

3.  The `cip6` columns contain 6-digit CIP program codes.

### *CIP data*

The US Classification of Instructional Programs (CIP) is a taxonomy of fields of study [@NCES:2010]. The dataset `cip` that loads with midfieldr contains program names and codes at the 2-digit, 4-digit, and 6-digit level.

```{r, chunk08}
cip
```

## Population

Relevant functions:

- `timely_term()`
- `data_sufficiency()`


### timely_term()

The *timely-completion term* is the latest term by which a student's program completion would be considered timely (default 6 academic years after admission). *Program completion* means satisfying the requirements for a degree. 

`timely_term()` determines the timely completion term for each student. The principal data frame must include the variable `{mcid}.` Entry term and level are pulled from the `term` table. The data frame is returned with the following variables added:

```{r, chunk09}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`entry_term`", "term when admitted" |
    "`entry_level`", "level when admitted" |
    "`adj_span`", "default span adjusted to account for entry level" |
    "`timely_term`", "timely completion term"
)
added_variables_table(x)
```

```{r, chunk10}
# setup
DT <- student[, .(mcid)]

# apply
DT <- timely_term(DT)

# result
DT[order(-adj_span)]
```

The optional `span` argument allows you to change the time span you consider "timely."

```{r, chunk11}
timely_term(DT, span = 8)[order(-adj_span)]
```



### data_sufficiency()

*Data sufficiency* is a necessary condition for including a student in a population if a metric depends on program completion. To meet the condition, an institution's data range must bracket a student's entry and timely completion terms.

`data_sufficiency()` evaluates whether the condition is met for each student. The principal data frame must include the variables `{mcid, entry_term, timely_term}.`  Institutions' data ranges are pulled from the `term` table. The data frame is returned with the following variables added:

```{r}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`data_range`", "an institution's min and max terms in the database" |
    "`sufficiency`", "indicates whether a record satisfies the condition"
)
added_variables_table(x)
```

```{r}
# setup
DT <- student[, .(mcid)]
DT <- timely_term(DT)
DT <- DT[, .(mcid, entry_term, timely_term)]

# apply
DT <- data_sufficiency(DT)

# result
DT[order(-sufficiency)]
```

For a sufficiency result of "satisfied," the entry term must be at least one term later than the lower limit of the data range and the timely completion term must be no later than the upper limit. We filter to retain rows with sufficiency labeled "satisfied."




## Records

Relevant functions:

- `is_undergrad()`

### is_undergrad()


*Undergraduate terms* are those prior to (and including) the first degree term.

`is_undergrad()` evaluates whether a term is before or after a student's first degree. The principal data frame is one of `term, course,` or `degree.` In all cases, the first degree term is pulled from the `degree` table. The data table is returned with the following variables added:

```{r}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`bacc_term`", "term of a student's first baccalaureate or NA" |
    "`term_focus`", 'indicates whether a term is before or after a first degree'
)
added_variables_table(x)
```

```{r}
is_undergrad(term)
```

We usually retain all columns, but the results are easier to examine if we view a selection of columns,

```{r}
term <- is_undergrad(term)
term[order(-term_focus), .(mcid, term, bacc_term, term_focus)]
```

We apply `is_undergrad()` to the student records having term-value variables `{term, term_course, term_degree}.` In all cases, we filter to retain terms with an "undergrad" focus.

```{r}
#| eval: false
term <- is_undergrad(term)
course <- is_undergrad(course)
degree <- is_undergrad(degree)
```



## Blocs

A *bloc* is a grouping of student-level data dealt with as a unit, for example, administrative groupings such as transfer students, traditional students, and non-traditional students, or program-based groupings such as program starters, ever-enrolled, graduates, or migrators.

Relevant functions:

- `filter_programs()`
- `completion_status()`
- `prep_fye_mice()`



### filter_programs()

Contributes to assembling a bloc of programs.

*Programs* are academic fields of study---specialties within a field or a collection of fields within a Department, College, or University---encoded in the `cip` dataset. 

`filter_programs()` acts on a CIP data frame to choose rows that match or partially match search strings. Search strings are case-independent. For example, to search for music programs, we might start with, 

```{r}
#| echo: false
options(
  datatable.print.nrows = 16,
  datatable.print.topn = 8
)
```

```{r}
filter_programs(cip, "music")
```

For music as a component of the Visual and Performing Arts, we filter `cip` again using a regular expression. We can also select specific columns for a more compact display. 

```{r}
filter_programs(cip, "^5009")[, .(cip6name, cip6, cip4name)]
```

We would continue in a similar fashion until we had determined all 6-digit codes needed for our study, combining them into one `programs` data frame. 


```{r}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)
```



### completion_status()

Contributes to assembling a bloc of graduates.

 *Completion status* is "timely" for students graduating no later than their timely-completion term; "late" or "NA" otherwise. Only records satisfying data sufficiency can be processed for completion status.

`completion_status()` yields a status label for each student. The principal data frame must include the variables `{mcid, timely_term}.` The first degree term is pulled from the `degree` table. The data frame is returned with the following variables added:

```{r}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`bacc_term`", "term of a student's first baccalaureate" |
    "`completion`", "indicates whether status is timely, late, or NA"
)
added_variables_table(x)
```

```{r}
# setup
DT <- student[, .(mcid)]
DT <- timely_term(DT)
DT <- data_sufficiency(DT)
DT <- DT[sufficiency == "satisfied"]
DT <- DT[, .(mcid, timely_term)]

# apply
DT <- completion_status(DT)

# result
DT[order(-completion)]
```



### prep_fye_mice()

Contributes to assembling a bloc of starters.

```{r}
#| echo: false
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5
)
```

At some U.S. institutions, completing a *First-Year Engineering (FYE)* program is a prerequisite for admission to specific engineering majors. FYE programs complicate the identification of *starters* because the student's preferred starting major is unknown. 

`prep_fye_mice()` helps us develop an *FYE proxy,* a likely CIP code of the  student's starting program had FYE not been required. Results are a CIP code or NA, treated as missing data to be imputed. 

```{r}
# seed set for reproducibility only
set.seed(202060909)

# apply
prep_fye_mice(student, term)
```

If FYE programs are involved in your study, 

- [FYE proxies](art-060-fye-proxies.html) describes the inner workings of the function and how to process the results using the R mice package for multiple imputation.
- [Starters](art-070-starters.html) describes how FYE proxies are incorporated in a starter bloc.




## Dissemination

Relevant functions:

- `order_multiway()`



### order_multiway()

Conditions data for Cleveland multiway charts.

*Multiway data* comprise three variables: a category with $\small m$ levels; a second independent category with $\small n$ levels; and a quantitative variable (the response) of length $\small m \times n$ with a value of the response for each combination of levels of the two categorical variables.

`order_multiway()` converts the categorical variables to factors ordered by the quantitative variable. The ordering of the rows and panels is crucial to the perception of effects.

Complete information on using `order_multiway()` can be found in  [Multiway data and charts](art-120-multiway.html).





## Utilities

See the relevant help page for more information, e.g. `?look_at.`

- `look_at()` for data frames, wraps base `str()` with preset arguments.
- `sort_uniq()` for vectors, wraps base `sort(unique())` with preset arguments.
- `catch_error()` wraps base `tryCatch()` for errors with preset arguments.
- `check_equiv_frames()` re-exported from the wrapr package.









```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
