---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---
  
```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))


# path_to_data_raw <- find_root_file("data-raw", criterion = has_file("DESCRIPTION"))
# source(paste0(path_to_data_raw, "/new_functions.R"))
```
  
```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

When working with student-level records to develop quantitative metrics, we recommend that you:
  
- Know the structure of your data
- Define your metrics and the relevant blocs of students
- Develop a process to refine and shape your data to produce the blocs you need
- Use chart designs suited to the types of results

midfieldr helps you with this process in these areas:

- *Records.* &nbsp; Credibly choosing rows and columns of the source data.  
- *Population.* &nbsp; Selectively subsetting a population to avoid miscounts.
- *Programs.* &nbsp; Collecting and labeling 6-digit program codes. 
- *Blocs.* &nbsp; Grouping and summarizing before computing metrics.  
- *Charts.* &nbsp; Conditioning data for multiway charts.  
  
This document introduces you to midfieldr's basic set of tools, and shows you how to apply them to data frames of student-level records. 

```{r}
library(midfieldr)
library(data.table)
```

```{r}
#| echo: false
opt_print_rows <- 3
options(
  datatable.print.topn = opt_print_rows,
  datatable.print.nrows = opt_print_rows
)
```




## Data: midfielddata

To explore the basic functionality of midfieldr, we use the datasets in  *midfielddata*: an R data package containing four tables --- `student`, `course`, `term`, and `degree` --- providing anonymized student-level records for 98,000 undergraduates at three US institutions from 1988 through 2018. The data tables are documented at the [midfielddata](https://midfieldr.github.io/midfielddata/reference/index.html) website. 

For this article we load the student, term, and degree tables. 

```{r}
library(midfielddata)
data(student, term, degree)
```

We usually copy the source data, giving them new names (and new locations in memory), to keep them intact while we use the original names  --- `student`, `term`, and `degree` --- to do our work. Also, these names are the default values when the term or degree tables, for example, are required arguments in a midfieldr function. 

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

The data tables are linked by `mcid`, the anonymized student ID. `look_at()` is a midfieldr convenience function that wraps `base::str()`. 

```{r}
look_at(student)

look_at(term)

look_at(degree)
```






## midfieldr functions

For this article, we organize midfieldr functions into three main categories. 

Working with student records

- `select_basic_cols()` chooses columns required by midfieldr functions.
- `add_post_bacc()` identifies rows of post-baccalaureate terms to exclude. 
- `add_timely_term()` estimates a student's timely graduation term. 
- `add_data_sufficiency()` identifies rows to exclude due to insufficient data.  
- `add_completion_status()` determines if program completion is timely or late.

Working with programs

- `filter_cip()` chooses rows of CIP data based on search terms. 

Special conditioning

- `prep_fye_mice()` conditions data for imputing starting majors of FYE students. 
- `order_multiway()` conditions data for Cleveland multiway charts. 







***midfieldr and data.table.*** &nbsp; midfieldr functions (internally) use data.table syntax; our scripts in these articles use it as well; and data provided in midfieldr and midfielddata are of the `data.table` class. For example,  

```{r}
degree_in <- copy(degree)
class(degree_in)
```

In default usage, midfieldr functions return data frames in data.table format, e.g., 

```{r}
degree_out <- select_basic_cols(degree_in)

class(degree_out)

degree_out
```


 

 
***midfieldr and tibbles.*** &nbsp; Knowing that some of our users prefer tibbles  (from the tidyverse package) to data.tables, midfieldr functions do attempt to return data frames of the same class as the input. So if we convert `degree` to a tibble, 

```{r}
library(tibble)
degree_tbl <- as_tibble(degree)

degree_tbl
```

and use it as an input value, a midfieldr function (in most cases) yields a tibble output. 

```{r}
select_basic_cols(degree_tbl)
```

Outside of the functions, however, all of our sample scripts employ data.table syntax. If you prefer dplyr (and friends) for data manipulation, you would have to translate our sample scripts into your preferred dialect. Good translation resources are available, e.g., the MIDFIELD Institute tutorials [-@midfieldinstitute:processing-data:2024] or Atrebas [-@atrebas:2019] 












## `select_basic_cols()` 

*Chooses columns required by midfieldr functions.* 

`select_basic_cols()` operates on source data frames to reduce the number of columns to those required by other midfieldr functions plus the key or composite key variables of the four data tables. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)
```

For example, compare `term` before and after applying `select_basic_cols()`. 

```{r}
look_at(term_source)

look_at(term)
```

With a smaller number of columns, the printout of the data frame is more readable, a benefit when working with the data interactively. 

```{r}
term
```







## `add_post_bacc()`

*Identifies rows of post-baccalaureate terms to exclude.*

`add_post_bacc()` identifies terms later than the first baccalaureate, if any. Because we are not generally interested in terms beyond the first degree term, we use the results of this function to exclude post-first-degree terms. 

```{r}
term <- add_post_bacc(dframe = term, midfield_degree = degree)
degree <- add_post_bacc(dframe = degree, midfield_degree = degree)
```

As shown below, two new columns have been added: the term of a student's first degree and a term-status variable that indicates the term is earlier than, equal to, or later than the first degree term. 

```{r}
term
```

Using `sort_uniq()`, a midfieldr convenience function to sort unique values of a vector, we find the possible values for term status to be:

```{r}
sort_uniq(term$term_status)
```

We filter to exclude all terms labeled "post-first-degree".

```{r}
term <- term[term_status != "post-first-degree"]
degree <- degree[term_status != "post-first-degree"]

term
```




## `add_timely_term()`

*Estimates a student's timely graduation term.*

The timely completion term is the last term in which a student’s degree completion would be considered timely. In many cases the timely completion (TC) term is 6 years after admission. 

We regularly use this step to begin defining our population for a study, so we begin with the unique IDs from the term record obtained above. 

```{r}
DT <- term[, .(mcid)]
DT <- unique(DT)

DT
```


`add_timely_term()` adds a column indicating the TC term for every student in the record. New columns of supporting information are also added: the student's initial term; their academic level upon entry; and the span in years, adjusted by their starting level, for timely completion.

```{r}
DT <- add_timely_term(DT, midfield_term = term)

DT
```








## `add_data_sufficiency()`

*Identifies rows to exclude due to insufficient data.*

The data sufficiency criterion limits student records to those for which available data are sufficient to assess timely completion without biased counts of completers or non-completers.

`add_data_sufficiency()` adds a column that labels each row for inclusion or exclusion based on data sufficiency near the upper and lower bounds of an institution’s data range. New columns of supporting information: the lower and upper limits of the institution's data range.

```{r}
DT <- add_data_sufficiency(DT, midfield_term = term)

DT
```

The possible values for data sufficiency are:

```{r}
sort_uniq(DT$data_sufficiency)
```

We filter to retain rows labeled "include".

```{r}
DT <- DT[data_sufficiency == "include"]

DT
```


## `add_completion_status()`

*Determines if program completion is timely or late.*

`add_completion_status()` adds a column of labels indicating whether or not a student completes a degree, and if they do, labeling it timely or late compared to their timely completion term. 

```{r} 
DT <- add_completion_status(DT, midfield_degree = degree)

DT
```

The possible values for completion status are:

```{r}
sort_uniq(DT$completion_status)
```

If we were constructing a bloc of timely graduates, we would filter to retain rows labeled "timely".

```{r}
DT[completion_status == "timely"]
```





## Data: cip

*Collecting and labeling 6-digit program codes.*

The *Classification of Instructional Programs (CIP)* is a taxonomy of academic programs, encoded by 6-digit numeric codes curated by the US Department of Education [@NCES:2010]. 

The `cip` data set, loaded with midfieldr, is a subset of the NCES CIP2010 data that contains codes and names for `r nrow(cip)` instructional programs organized on three levels---a 2-digit series, a 4-digit series, and a 6-digit series---keyed by the `cip6` variable. 

```{r}
#| echo: false
options(
  datatable.print.topn = 5,
  datatable.print.nrows = 5
)
```

```{r}
look_at(cip)

# 6-digit codes
cip[, .(cip6, cip6name)]

# 4-digit codes
cip[, .(cip4, cip4name)]

# 2-digit codes
cip[, .(cip2, cip2name)]
```





## `filter_cip()`

*Chooses rows of CIP data based on search terms.*

`filter_cip()` acts on the `cip` data frame to select rows that match or partially match search strings. Search strings are case-independent and can include regular expressions. 

```{r}
filter_cip("music")
```

To refine the results of one pass, assign those results to the `cip` argument in a subsequent pass. To illustrate the process, we collect programs related to "music" within the Visual and Performing Arts (CIP code 50). 

```{r}
first_pass <- filter_cip("music")

second_pass <- filter_cip("^50", cip = first_pass)

second_pass
```

In this case, the 4-digit CIP code 5009 identifies "Music" programs, distinguishing them from other programs in the Visual and Performing Arts. We assign the second pass to the `cip` argument and restrict the search to codes starting with "5009". 

```{r} 
third_pass <- filter_cip("^5009", cip = second_pass)

third_pass
```

```{r}
#| echo: false
options(
  datatable.print.topn = 20,
  datatable.print.nrows = 20
)
```

If these programs were needed for our study, we would save the 6-digit codes and probably edit the program names for later use.  

```{r}
third_pass[, .(cip6, cip6name)]
```

```{r}
#| echo: false
options(
  datatable.print.topn = opt_print_rows,
  datatable.print.nrows = opt_print_rows
)
```







## Other functions

`prep_fye_mice()` &nbsp; Conditions data for imputing the starting majors of First-Year Engineering (FYE) students. Used when blocs of starters in Engineering are needed and an  institution has a required FYE program. For details see [FYE proxies](art-060-fye-proxies.html).   

`order_multiway()` &nbsp; Conditions data for Cleveland multiway charts. The ordering of its rows and panels is crucial to the perception of effects. Used when data have a multiway structure. For details see [Multiway data and charts](art-120-multiway.html). 

Utilities 

- `look_at()` wraps base `str()`.  
- `sort_uniq()` wraps base `sort()` and `unique()`.  
- `catch_error()` wraps base `tryCatch()` for errors.  
- `check_equiv_frames()` re-exported from the wrapr package  



```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```



