---
title: "Planning"
output: github_document 
date: "`r Sys.Date()`"
csl: ../../inst/information-science-and-technology.csl
bibliography: ../../inst/REFERENCES.bib
link-citations: yes
always_allow_html: true
---

```{r child = here::here("man/rmd/common-setup.Rmd")}
```

Analysis of MIDFIELD data begins by identifying the groups of students, programs, and metrics with which we intend to work. 

## Workflow

Working with MIDFIELD data is iterative---intermediate results often cause us to revisit an earlier assumption or select a different bloc or student attributes to work with. Nevertheless, a completed analysis usually comprises the following steps in roughly the sequence given below. Accented entries indicate topics in the open vignette. 

1. `r accent("Planning")`  
    - `r accent("Records")`  
    - `r accent("Programs")`  
    - `r accent("Metrics, blocs, and groupings")`  
1. Initial processing 
    - Data sufficiency 
    - Degree seeking  
    - Identify programs  
1. Blocs
    - Ever-enrolled  
    - FYE proxies  
    - Starters  
    - Graduates  
1. Groupings    
    - Program labels  
    - Demographics    
    - Other variables 
1. Metrics  
    - Graduation rate  
    - Stickiness  
1. Displays 
    - Multiway charts  
    - Tables  



## Definitions

```{r child = "../../man/rmd/define-student-level-data.Rmd"}
```

```{r child = "../../man/rmd/define-program.Rmd"}
```

```{r child = "../../man/rmd/define-metric.Rmd"}
```

```{r child = "../../man/rmd/define-bloc.Rmd"}
```

```{r child = "../../man/rmd/define-grouping-variables.Rmd"}
``` 





## Which records?

There are currently two points of access to MIDFIELD data: 

```{r child = "../../man/rmd/note-midfielddata-and-midfield.Rmd"}
``` 

Records in both collections are organized in four tables---student, course, term, and degree---keyed by student ID.  Details in [Getting started](art-000-getting-started.html). 

To maintain confidentiality, the practice data include fewer variables than the research data, but the variable names and values are generally a subset of those in the research data. Thus a researcher transitioning from working with the practice data to the research data will find that their practice scripts will work with the research data with only minor modifications (if any). 

```{r child = "../../man/rmd/note-practice-not-research.Rmd"}
```

 



### Load data 

*Practice data.* &nbsp; Load from the midfielddata package.  

```{r}
#| eval: false

# Load practice data
data(student, course, term, degree)
```

*Research data.* &nbsp; Assuming one has MIDFIELD access, import data via the user's preferred file read function, e.g., 

```{r}
#| eval: false

# Not run
student <- fread("path_to_student_research_data")
course  <- fread("path_to_course_research_data")
term    <- fread("path_to_term_research_data")
degree  <- fread("path_to_degree_research_data")
```






## Which programs? 

Identify programs in *general* terms, for example, 

- All Engineering  
- Engineering, Business, Social Sciences, and Arts and Humanities  
- Electrical Engineering and Computer Engineering  

As shown in [Programs](art-040-programs.html#filter_cip), we use `filter_cip()` acting on the `cip` data set to determine the set of CIP codes relevant to a study. 

A majority of our vignette examples involve engineering programs. However, 

> *MIDFIELD research is not limited to Engineering*. 
> 
> *MIDFIELD research data contain SURs of all undergraduates in all programs at their institution over the time span given.* 

The practice data, as a sample of the research data, have similar attributes.








## Which metrics, blocs, and groupings?

Before the data processing starts, we have to decide the metrics we want to compare among which blocs of students grouped by what variables. Metrics can include bloc counts or comparative ratios, for example:

- [Blocs](art-050-blocs.html), e.g., counts of [starters](art-070-starters.html), [graduates](art-080-graduates.html), migrators, etc. 
- Four-year persistence 
- Six-year [graduation rate](art-100-grad-rate.html)  
- Six-year stickiness  

The metric determines the blocs to gather, for example:

- [Graduation rate](art-100-grad-rate.html) requires a bloc of starters and the subset who graduate in their starting program. 
- Stickiness requires the bloc [ever enrolled](art-050-blocs.html#ever-enrolled) in a program and the subset who graduate in that program. 

The research study design determines the grouping variables, for example,  

- [Completion status](art-080-graduates.html#add_completion_status) is a critical variable whenever graduation (program completion) is involved.  
- [Programs](art-090-groupings.html#program-labels), [race/ethnicity and sex](art-090-groupings.html#demographics) are important grouping variables in many studies. 
- Other student-level variables such as institution, GPA, grade level, etc., can also be used for grouping and summarizing.  






## References

<div id="refs"></div>


```{r}
#| echo: false

# to change the CSS file
# per https://github.com/rstudio/rmarkdown/issues/732
knitr::opts_chunk$set(echo = FALSE)
```

```{css}
blockquote {
    padding:     10px 20px;
    margin:      0 0 20px;
    border-left: 0px
}
caption {
    color:       #525252;
    text-align:  left;
    font-weight: normal;
    font-size:   medium;
    line-height: 1.5;
}
```
