---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r, setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r, chunk01,  child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r, chunk02}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)

# display variables added by midfieldr functions
library(gt)
added_variables_table <- function(df) {
  gt(df) |>
    tab_options(
      table.font.size = pct(80),
      container.height = pct(80)
    ) |>
    fmt_markdown(columns = 1) |>
    cols_label(
      var = "variable",
      val = "description"
    )
}
```

In this document, we introduce midfieldr's basic tools and show how to apply them to data frames of student-level records (registrar's data). We organize the topics around a typical workflow:

- data
- population
- records
- blocs
- special conditioning

*Packages.* We load midfielddata for its practice data and data.table for its data manipulation syntax. 

```{r, chunk04}
library("midfieldr")
library("midfielddata")
library("data.table")
```





## Data


### *Student records*

midfieldr is designed to work with the MIDFIELD research database or any  database with a similar structure such as the practice data in the [midfielddata](https://midfieldr.github.io/midfielddata/index.html) package. If you are new to MIDFIELD, the best place to start is the article on [Data structure](https://midfieldr.github.io/midfielddata/articles/data-structure.html).

For this article we load all four midfielddata tables. 

```{r, chunk06}
# data(student, term, course, degree)
data(student, term, degree)
```

`select_basic_cols()` returns a subset of each table with the variables most often encountered in the early stages of a project. The tables are linked by the anonymized student ID variable `mcid.`

```{r, chunk07}
select_basic_cols(student)

select_basic_cols(term)

# select_basic_cols(course)

select_basic_cols(degree)
```

*Notes.*

1.  Like the data frames above, all data frames in midfieldr are `data.table` enhanced. However, if you convert these tables to another class such as `tbl_df` (tibbles) or `data.frame` (base R), midfieldr functions attempt to return data frames of the same class to support your preferred syntax. 

2.  Term variables `{term, term_course, term_degree}` are encoded as character strings `YYYYT`, where `YYYY` is the year at the start of the academic year and `T` encodes the semester or quarter within an academic year as Fall (1), Winter (2), Spring (3), or Summer (4, 5, and 6).

3.  The `cip6` columns contain 6-digit CIP program codes.



### *CIP data*

The US Classification of Instructional Programs (CIP) is a taxonomy of fields of study [@NCES:2010]. The dataset `cip` that loads with midfieldr contains program names and codes at the 2-digit, 4-digit, and 6-digit level.

```{r, chunk08}
cip
```

## Population

Relevant functions:

- `timely_term()`
- `data_sufficiency()`



### timely_term()

The *timely-completion term* is the latest term by which a student's program completion would be considered timely (default 6 academic years after admission). *Program completion* means satisfying the requirements for a degree. 

`timely_term()` determines the timely completion term for each student. The principal data frame must include the variable `{mcid}.` Entry term and level are pulled from the `term` table. The data frame is returned with the following variables added:

```{r, chunk09}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`entry_term`", "term when admitted" |
    "`entry_level`", "level when admitted" |
    "`adj_span`", "default span adjusted to account for entry level" |
    "`timely_term`", "timely completion term"
)
added_variables_table(x)
```

```{r, chunk10}
# setup
DT <- student[, .(mcid)]

# apply
DT <- timely_term(DT)

# result
DT[order(-adj_span)]
```

The optional `span` argument allows you to change the time span you consider "timely."

```{r, chunk11}
DT <- student[, .(mcid)]
timely_term(DT, span = 8)[order(-adj_span)]
```



### data_sufficiency()

*Data sufficiency* is a necessary condition for including a student in a population if a metric depends on program completion. To meet the condition, an institution's data range must bracket a student's entry and timely completion terms.

`data_sufficiency()` evaluates whether the condition is met for each student. The principal data frame must include the variables `{mcid, entry_term, timely_term}.`  Institutions' data ranges are pulled from the `term` table. The data frame is returned with the following variables added:

```{r}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`data_range`", "an institution's min and max terms in the database" |
    "`sufficiency`", "indicates whether a record satisfies the condition"
)
added_variables_table(x)
```

```{r}
# setup
DT <- student[, .(mcid)]
DT <- timely_term(DT)
DT <- DT[, .(mcid, entry_term, timely_term)]

# apply
DT <- data_sufficiency(DT)

# result
DT[order(-sufficiency)]
```

We usually filter to retain rows with sufficiency labeled "satisfied," indicating that the entry term is at least one term later than the lower limit of the data range and the timely completion term is no later than the upper limit.



## Records

Relevant functions:

- `filter_undergrad()`



### filter_undergrad()

*Undergraduate terms* are those prior to (and including) the first degree term.

`filter_undergrad()` distinguishes post-baccalaureate terms from undergraduate terms and retains the undergraduate terms. Here we filter the `term` table and count the rows before and after the operation. 

```{r}
nrow(term)
x <- filter_undergrad(term, midf_table = degree)
nrow(x)
```

The optional `add_bacc_term` argument adds the baccalaureate term to the data frame if you want verify the result; test for yourself that only terms predating the degree are retained. The results match those above. 

```{r}
y <- filter_undergrad(term, add_bacc_term = TRUE)
y <- y[, .(mcid, term, bacc_term)][order(bacc_term)]
y

# Filter and compare to previous result
y <- y[bacc_term >= term | is.na(bacc_term)]
check_equiv_frames(x[, .(mcid, term)], y[, .(mcid, term)])
```

The principal data frame is any of the data tables having a term-value variable. In all cases, the term of the student's first degree is pulled from the `degree` table.

```{r}
#| eval: false
term <- filter_undergrad(term)
course <- filter_undergrad(course)
degree <- filter_undergrad(degree)
```





## Blocs

A *bloc* is a grouping of student-level data dealt with as a unit, for example, administrative groupings such as transfer students and  traditional or non-traditional students, as well as program-based groupings such as students starting in, ever-enrolling in, migrating into or out of, or graduating from a program.

Relevant functions:

- `filter_programs()`
- `completion_status()`



### filter_programs()

Contributes to assembling a bloc of programs.

*Programs* are academic fields of study---specialties within a field or a collection of fields within a Department, College, or University---encoded in the `cip` dataset. 

`filter_programs()` acts on a CIP data frame to choose rows that match or partially match search strings. Search strings are case-independent. For example, to search for music programs, we might start with, 

```{r}
#| echo: false
options(
  datatable.print.nrows = 16,
  datatable.print.topn = 8
)
```

```{r}
filter_programs(cip, "music")
```

For music as a component of the Visual and Performing Arts, we filter `cip` again using a regular expression. We can also select specific columns for a more compact display. 

```{r}
filter_programs(cip, "^5009")[, .(cip6name, cip6, cip4name)]
```

The `negate` argument allows you to drop the rows that match the search patterns.

```{r}
x <- filter_programs(cip, "^5009")[, .(cip6name, cip6, cip4name)]
filter_programs(x, c("Other", "General"), negate = TRUE)
```

In a study, we would continue in a similar fashion until we had merged all the 6-digit codes we needed into one data frame.

```{r}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)
```



### completion_status()

Contributes to assembling a bloc of graduates.

 *Completion status* is "timely" for students graduating no later than their timely-completion term; "late" or "NA" otherwise. Only records satisfying data sufficiency can be processed for completion status.

`completion_status()` yields a status label for each student. The principal data frame must include the variables `{mcid, timely_term}.` The first degree term is pulled from the `degree` table. The data frame is returned with the following variables added:

```{r}
#| echo: false
x <- wrapr::build_frame(
  "var", "val" |
    "`bacc_term`", "term of a student's first baccalaureate" |
    "`completion`", "indicates whether status is timely, late, or NA"
)
added_variables_table(x)
```

```{r}
# setup
DT <- student[, .(mcid)]
DT <- timely_term(DT)
DT <- data_sufficiency(DT)
DT <- DT[sufficiency == "satisfied"]
DT <- DT[, .(mcid, timely_term)]

# apply
DT <- completion_status(DT)

# result
DT[order(-completion)]
```

When we want of bloc of timely graduates, we filter to retain rows with completion "timely."



## Commonalities

You may have noticed similarities in several of the midfieldr functions. For example, in these functions,  

- `timely_term(dframe, midf_table = term)`
- `data_sufficiency(dframe, midf_table = term)`
- `filter_undergrad(dframe, midf_table = degree)`
- `completion status(dframe, midf_table = degree)`

the similarities include: 

- The first argument is a data frame.
- The second argument is one of the MIDFIELD data tables.
- The result is a new data frame with columns added that support the main finding.
- You can use the main finding to subset the result by rows.
 
Moreover, because in each case the `midf_table` argument has a default value, it can be assigned explicitly or not. For example, the three formulations below all yield the same result.

```{r}
# setup
DT <- student[, .(mcid)]

# equivalent statements
x <- timely_term(dframe = DT, midf_table = term)
y <- timely_term(DT, term)
z <- timely_term(DT)

# equivalent results
check_equiv_frames(x, y)
check_equiv_frames(x, z)
```

Of other functions used earlier, 

- `filter_programs()`
- `select_basic_cols()`

have in common that: 

- The first argument is a data frame. 
- The result is a new data frame, a subset of the input: "filter" chooses rows; "select" chooses columns. 



## Special conditioning

Relevant functions:

- `initialize_fye_proxies()`
- `order_multiway()`



### initialize_fye_proxies()

Contributes to assembling a bloc of starters.

```{r}
#| echo: false
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5
)
```

At some U.S. institutions, completing a *First-Year Engineering (FYE)* program is a prerequisite for admission to specific engineering majors. FYE programs complicate the identification of *starters* because these students' preferred starting majors are unknown. 

`initialize_fye_proxies()` constructs a data frame with the IDs of all FYE students with an initial *FYE proxy,* the 6-digit CIP code of their first specific engineering major, if any, or NA if not. 

```{r}
initialize_fye_proxies(student, term)
```

The NAs are ultimately treated as missing data that are replaced by imputed CIP codes. If FYE programs are involved in your study, 

- [FYE proxies](art-060-fye-proxies.html) describes the inner workings of the function and how to process the results using the R mice package for multiple imputation.
- [Starters](art-070-starters.html) describes how FYE proxies are incorporated in a starter bloc.









### order_multiway()

Conditions data for Cleveland multiway charts.

*Multiway data* comprise two independent categorical variables and one quantitative variable with a value for each combination of levels of the two categories.

`order_multiway()` converts the categorical variables to factors ordered by the quantitative variable. This ordering determines the order of the panels and rows in a multiway chart, crucial to our ability to perceive effects.

In this example, we use the `case_results` multiway data that loads with midfieldr. The two categories are `program` and `people`. The quantity is `stick` (program stickiness) which is the ratio of `grad` (number of program graduates) to `ever` (number ever enrolled in the program). 

```{r}
case_results
```

In the data frame returned, the factor levels are ordered by the aggregate stickiness calculated separately by category and reported in the new columns (suffix `_metric`). A multiway chart can be constructed from the data in this form. 

```{r}
DT <- order_multiway(case_results,
  quantity = "stick",
  categories = c("program", "people"),
  method = "percent",
  ratio_of = c("grad", "ever")
)
DT[, c("grad", "ever") := NULL]
DT
```

Complete information on using `order_multiway()` and constructing multiway charts can be found in  [Multiway data and charts](art-120-multiway.html).





## Utilities

See the relevant help page for more information, e.g. `?catch_error`. 

- `catch_error()` wraps base `tryCatch()` for errors with preset arguments.
- `check_equiv_frames()` re-exported from the wrapr package.
- `look_at()` for data frames, wraps base `str()` with preset arguments.
- `rm_redundant_cols()` primarily used internally to drop duplicate columns.
- `sort_uniq()` for vectors, wraps base `sort(unique())` with preset arguments.






```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
