---
title: "Case study"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-004-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

How to work with longitudinal student data using midfieldr, focusing less on software details and more on the overall  *process*. The work is organized in three major topics: 

1. Obtaining a credible population and filtering the records to match.
2. Manipulating the data to produce the desired metric and groupings. 
3. Conditioning the results for dissemination in tables and charts. 


### *Scope*    

*Records.* &nbsp; Data: `student, term,` and `degree` from midfielddata, excluding records later than a student's first degree term (if any), restricted to degree-seeking students, and filtered for data sufficiency. 

*Population.* &nbsp; The set of unique student IDs from the above records. 

*Programs.* &nbsp; Data: `cip` from midfieldr. We study four Engineering majors: Civil, Electrical, Industrial/Systems, and Mechanical.   

*Metric.* &nbsp;  Program *stickiness:* the ratio $\small (S)$ of the number of graduates of a program $\small (N_\textrm{grad})$ to the number ever enrolled in the program $\small (N_\textrm{ever})$.  

$$
  \small S = \frac{\small N_\textrm{grad}}{\small N_\textrm{ever}} = \frac{\small\mathrm{number\ of\ graduates\ of\ a\ program}}{\small\mathrm{number\ ever\ enrolled\ in\ the\ program}}
$$

However a student finds their way to a program (just entering college, transferring, or switching majors), the program has an opportunity to see the student through to graduation. The likelihood that a student will "stick to" and complete that program is the "stickiness" metric [@Ohland+Orr+others:2012]. 

*Blocs.* &nbsp; The metric requires two blocs: students ever enrolled in the programs; and timely graduates of the programs. 

*Groupings.* &nbsp; Group the findings by program, race/ethnicity, and sex.

*Dissemination.* &nbsp; Exclude groupings too small to preserve anonymity. Edit column names to suit the audience. Condition/transform data as needed for tables or charts. 

### *Definitions*

For understanding our data manipulation process. 

- The population is usually expected to be *degree-seeking,* i.e., attempting to complete a program. 

- *Program completion* means satisfying the requirements for a first baccalaureate degree. 

- The *timely-completion term* is the term by which we would consider their completion "timely", default 6 years after admission. 

- The *data sufficiency* test identifies students whose admission term and projected timely completion term lie within the range of data available from their institution---a necessary and sufficient condition for determining completion status. 

- *Completion status* is "timely" for students graduating no later than their timely-completion term; "late" if they graduate after that term; and "NA" for non-completion.

### *R packages*

```{r}
library("midfieldr") # working with student records
library("midfielddata") # practice data
library("data.table") # data manipulation
library("gt") # tables
library("ggplot2") # charts
```


## Refining the records

We load three of the midfielddata data tables. This study does not require the `course` data table. If it had been required, it would be included here and in similar steps throughout the article. 

```{r}
data(student, term, degree)
```

We copy the original data sets, giving them new names `{student_source, term_source, degree_source}` and new locations in memory. This step allows us to use the more convenient names `{student, term, degree}` for our "working" data tables without updating the source tables "by reference." Reference semantics in data.table is documented in [@reference-semantics:2026].  

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)
```

For comparing results as we refine the population, we start with the following number of rows in the original data frames. 

```{r}
#| echo: false
make_table <- function(x, cap) {
  gt(x) |>
    sub_missing() |>
    tab_caption(cap) |>
    tab_options(table.font.size = "small") |>
    opt_stylize(style = 1, color = "gray") |>
    tab_style(
      style = list(cell_fill(color = "#c7eae5")),
      locations = cells_column_labels(columns = everything())
    ) |>
    fmt_number(use_seps = TRUE, decimals = 0)
}


Ncompare <- wrapr::build_frame(
  "Table", "Source" |
    "student", eval(nrow(student_source)) |
    "term", eval(nrow(term_source)) |
    "degree", eval(nrow(degree_source))
)
setDT(Ncompare)
setnames(Ncompare, old = "Source", new = "Original tables")
make_table(Ncompare, "Table 1(a). Number of rows.")
```





### *Required variables*

For this first section of the work, we use `select_basic_cols()` to subset the data tables, retaining only those variables needed by  midfieldr functions. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

student

term

degree
```




### *Data sufficiency*

Only those records passing the data sufficiency test are included in a population study. We start with the full set of unique student IDs. 

```{r}
DT <- term[, .(mcid)]
DT <- unique(DT)
DT
```

We use `timely_term()` to determine the timely completion term for every student and add columns to the data frame to support those findings. 

```{r}
DT <- timely_term(DT, midf_table = term)
DT
```

Next, we can (optionally) reduce the number of columns to just those needed in the next step. 

```{r}
DT <- DT[, .(mcid, term_i, timely_term)]
DT
```

We operate on this output with `data_sufficiency()` to identify records that pass the data sufficiency test and those that do not and add columns to the data frame to support those findings.

```{r}
DT <- data_sufficiency(DT, midf_table = term)
DT
```

*Summary check.* &nbsp; A brief aside to summarize the numbers of students identified to include and exclude. We run a similar summary at a number of points in the analysis as a  credibility check. 

```{r}
DT[, .N, by = c("data_sufficiency")][order(-N)]
```

We filter to retain rows labeled "include" and drop all but the ID column.

```{r}
DT <- DT[data_sufficiency == "include", .(mcid)]
DT
```


### *Degree seeking*

We require all students in our study to be degree-seeking. By design, the `student` table contains only degree-seeking students. We inner-join the ID column  from the `student` table with our working data frame `DT`, matching on `mcid, to remove any non-degree-seeking students. 

```{r}
student_cols <- student[, .(mcid)]
DT <- student_cols[DT, on = "mcid", nomatch = NULL]
DT
```

In this instance, no students were removed because the midfielddata practice data contains only degree-seeking students. However, our process would not be complete without this step. 

### *Population*

The steps above---filtering for data sufficiency and degree-seeking---yield the baseline  population for this study. 

```{r}
population <- copy(DT)
```

We use an inner join, matching on `mcid,` to yield records for the IDs in the population only.  

```{r}
student_refined <- population[student_source, on = "mcid", nomatch = NULL]
term_refined <- population[term_source, on = "mcid", nomatch = NULL]
degree_refined <- population[degree_source, on = "mcid", nomatch = NULL]
```

```{r}
#| echo: false
Ncompare[, Population := c(
  nrow(student_refined),
  nrow(term_refined),
  nrow(degree_refined)
)]
setnames(Ncompare, old = "Population", new = "Refined population")
make_table(Ncompare, "Table 1(b). Number of rows")
```

### *Post-completion terms*

We are not generally interested in terms beyond the first degree term, so we identify and exclude terms later than the first degree term in the source data frames. The procedure applies to all all but the `student` data table because it  contains no columns of encoded-term data. 

```{r}
term <- copy(term_refined)
degree <- copy(degree_refined)
```

For every term per student ID, `post_completion_terms()` identifies terms that are before, equal to, or after the student's first degree term (if any) and adds columns to the data frame to support those findings. Note that here we retain the full set of columns in each data table. 

```{r}
term <- post_completion_terms(term, midf_table = degree)
degree <- post_completion_terms(degree, midf_table = degree)

term
```

*Summary check.* &nbsp; Numbers of students in each term cluster.

```{r}
term[, .N, by = c("term_cluster")][order(-N)]

degree[, .N, by = c("term_cluster")][order(-N)]
```

We exclude the rows labeled "post-first-degree." Note that we are dropping rows by values in the *term* variable without affecting the number of students in the *population.*  

```{r}
term <- term[term_cluster != "post-first-degree"]
degree <- degree[term_cluster != "post-first-degree"]
```

We drop the temporary columns. 

```{r}
term[, c("term_cluster", "first_degree_term") := NULL]
degree[, c("term_cluster", "first_degree_term") := NULL]
```

What we're calling our `_baseline` tables have been filtered for data sufficiency, degree-seeking, and post-completion terms. 

```{r}
student_baseline <- copy(student_refined)
term_baseline <- copy(term)
degree_baseline <- copy(degree)
```

```{r}
#| echo: false
Ncompare[, Post_bacc := c(
  nrow(student_baseline),
  nrow(term_baseline),
  nrow(degree_baseline)
)]
setnames(Ncompare, old = "Post_bacc", new = "Baseline tables")
make_table(Ncompare, "Table 1(c). Number of rows.")
```

Review the results.

```{r}
look_at(student_baseline)

look_at(term_baseline)

look_at(degree_baseline)

look_at(population)
```

From this point forward, anytime we need a fresh copy of any of the data tables, we copy the `_baseline` version. Anytime we need a starting population, we copy  `population.`





## Programs

```{r}
#| echo: false
options(
  datatable.print.nrows = 12,
  datatable.print.topn = 6
)
```

In this section, we begin the procedures that transform our baseline records and population into case-specific programs, blocs, metrics, and groupings. 

Our goal in this section is to search the CIP data table for the 6-digit codes for our programs. The `cip` dataset loads with midfieldr. 

### *Search for program codes*

`filter_programs()` searches `dframe` for string patterns. Searching for "engineering" yields 2-digit CIP codes 14, 15, 29, and 51. From the program names, the 2-digit code we want is 14. 

```{r}
filter_programs(dframe = cip, pattern = "engineering")
```

Subset the full `cip` for the Engineering code 14 and search that result for "civil". We find that Civil Engineering is encoded by the 4-digit code 1408. 

```{r}
engr_cip <- filter_programs(cip, "^14")
filter_programs(engr_cip, "civil")
```

Repeat for "electrical" and obtain the 4-digit code 1410.

```{r}
filter_programs(engr_cip, "electrical")
```

Continuing in a similar fashion, we find that our programs have the following 4-digit codes. 

- Civil Engineering 1408 
- Electrical Engineering 1410 
- Mechanical Engineering 1419  
- Industrial/Systems Engineering 1427, 1435, 1436, and 1437. 

Note that in general an accredited engineering degree-granting program is encoded by a single 4-digit CIP code, but not always. In addition, the code or codes used by a degree-granting program can vary over time and by institution.  






### *Construct the programs table*

We create a search string of the desired 4-digit codes.

```{r}
codes_we_want <- c("^1408", "^1410", "^1419", "^1427", "^1435", "^1436", "^1437")
programs <- filter_programs(engr_cip, codes_we_want)
programs
```

While we searched `cip` using 4-digit codes, it is the 6-digit codes that we want for matching to the student records later. Here we drop all but the 6-digit information. 

```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

```{r}
programs <- programs[, .(cip6name, cip6)]
programs
```

The program names in `cip` are usually too long for effective use---user-defined names are nearly always required. So we add a `program`  variable with values "CE" (Civil Engineering), "EE" (electrical), "ME" (Mechanical), and "ISE" (Industrial/Systems Engineering). 

```{r}
programs[, program := fcase(
  cip6 %like% "^1408", "CE",
  cip6 %like% "^1410", "EE",
  cip6 %like% "^1419", "ME",
  cip6 %like% c("^1427|^1435|^1436|^1437"), "ISE",
  default = NA_character_
)]
```

While we don't use the original 6-digit program names, we can preserve them for reference. Here we abbreviate some of them for a more compact display. 

```{r}
programs[, cip6name := gsub("Engineering", "Engng", cip6name)]
programs[, cip6name := gsub("Communications", "Commns", cip6name)]
programs[, cip6name := gsub("Electrical, Electronics", "Elec, Electr,", cip6name)]

programs
```

Our programs data frame is complete: `r nrow(programs)` six-digit codes are encoded using 4 program labels. This data frame can sit in memory (or written to file) until we're ready to filter the blocs by program, joining data frames by matching on the `cip6` variable. 

```{r}
#| echo: false
options(
  datatable.print.nrows = 6,
  datatable.print.topn = 3
)
```









## Blocs and groupings

The stickiness metric requires these blocs:

- students with timely completion from the study programs
- students ever enrolled in these programs

And we selected these groupings:

- program
- race/ethnicity
- sex

We have a lot of flexibility in the order in which we construct our blocs and groupings, so what follows is only one of several effective solutions. Our approach here is to construct a bloc, filter by program, join the demographics, and repeat for the next bloc. 

First, we copy so our work will not affect the baseline material by reference.

```{r}
student <- copy(student_baseline)
term <- copy(term_baseline)
degree <- copy(degree_baseline)
```






## Timely graduates

We start with the baseline population. Like we did with the original data files, we copy it to protect `population` from changes by reference.

```{r}
DT <- copy(population)
DT
```





### *Filter by program*

We left-join the CIP column from the `degree` table, matching on `mcid`. That the number of rows has increased indicates that some students have more than one degree in their first degree term.

```{r}
degree_cols <- degree[, .(mcid, cip6)]
DT <- degree_cols[DT, on = "mcid"]
DT
```

*Summary check.* &nbsp; Numbers of students completing $\small N$ degrees.  

```{r}
x <- DT[, .(N_degrees = .N), by = "mcid"]
x[, .(N_students = .N), by = "N_degrees"]
```

We use an inner-join with our `programs` data frame to retain only the rows that have matching CIP codes in both tables. 

```{r}
programs_cols <- programs[, .(cip6, program)]
DT <- programs_cols[DT, on = "cip6", nomatch = NULL]
DT
```

We drop the `cip6` column, leaving the `program` column with our user-defined program labels. These students all have a degree in one of our four engineering programs. 

```{r}
DT[, cip6 := NULL]
DT <- unique(DT)
DT
```



### *Filter for timely completion*

We want to retain timely graduates only. We start by obtaining the timely completion term then selecting the variables we need to go forward. 

```{r}
DT <- timely_term(DT)
DT <- DT[, .(mcid, program, timely_term)]
DT
```

`completion_status()` builds on the output from `timely_term()` to label rows to indicate whether a student completes a degree timely or late compared to their timely completion term (or NA for no completion).

```{r}
DT <- completion_status(DT)
DT
```

*Summary check.* &nbsp; Numbers of students by completion status.

```{r}
DT[, .N, by = c("completion_status")][order(-N)]
```

We retain the rows labeled "timely" and the drop all the columns except the ID and program columns. 

```{r}
DT <- DT[completion_status == "timely", .(mcid, program)]
DT
```




### *Join demographics*

To add columns for student demographics, we left-join selected columns from the `student` table, matching on `mcid`. 

```{r}
student <- select_basic_cols(student)
DT <- student[DT, on = "mcid"]
DT
```







### *Bloc of timely graduates*

This is the bloc of timely graduates required by our metric. We add a `bloc` variable with the value "grad" and ensure we have unique rows.

```{r}
DT[, bloc := "grad"]
graduates <- unique(DT)
graduates
```

*Summary check.* &nbsp; Numbers of timely graduates by program. 

```{r}
graduates[, .N, by = c("program")][order(-N)]
```




## Ever enrolled

Again we start with the baseline population. 

```{r}
DT <- copy(population)
DT
```


### *Filter by program*

We left-join the CIP column from the `term` table, matching on `mcid`. 

```{r}
term_cols <- term[, .(mcid, cip6)]
term_cols <- unique(term_cols)
DT <- term_cols[DT, on = "mcid"]
DT
```

CIP codes are also present in the `degree` table. Students working in a  multidisciplinary program may have CIP codes at graduation that do not appear in the `term` data, where only their primary major is recorded. We assume that if a student earns a degree in such a program we can consider them "ever enrolled" in the program.

From `degree,` we extract the CIP codes by ID and join them by rows to the previous data frame. 

```{r}
extra_cip <- copy(population)
degree_cols <- unique(degree[, .(mcid, cip6)])
extra_cip <- degree_cols[extra_cip, on = "mcid", nomatch = NULL]
DT <- unique(rbindlist(list(DT, extra_cip)))
DT
```

We repeat the process we used earlier to inner-join our `programs` data frame, matching on `cip6`. 

```{r}
programs_cols <- programs[, .(cip6, program)]
DT <- programs_cols[DT, on = "cip6", nomatch = NULL]
DT[, cip6 := NULL]
```

With the CIP code removed, we filter for unique rows. A student may switch CIP codes yet stay within a program as defined by our custom labels. We want to avoid counting that student as ever-enrolled in the same program more than once. 

```{r}
DT <- unique(DT)
DT
```


### *Join demographics*

Again, we left-join selected columns from the `student` table, matching on `mcid`.

```{r}
DT <- student[DT, on = "mcid"]
DT
```

### *Bloc of ever-enrolled*

This is the bloc of students ever enrolled in our programs required by our metric. We add a `bloc` variable with the value "ever" and ensure we have unique rows.

```{r}
DT[, bloc := "ever"]
ever_enrolled <- unique(DT)
ever_enrolled
```

*Summary check.* &nbsp; Numbers of students ever enrolled by program.

```{r}
ever_enrolled[, .N, by = c("program")][order(-N)]
```



## Outcomes

Combining the two data frames (blocs) by rows, we obtain the data structure we need for grouping and summarizing. 

```{r}
DT <- rbindlist(list(graduates, ever_enrolled))
DT
```

### *Group and summarize*

Count the numbers of observations for each combination of the grouping variables. These data are in block-record form with four keys, one value column, and one row per value.

```{r}
DT <- DT[, .N, by = c("bloc", "program", "race", "sex")]
DT
```




### *Reshape*

*Reshaping the data frame to calculate the metric.*

We want to separate the $\small N$ column into two columns---one for the number of graduates and the other for the number of ever enrolled. This operation is  known by a number of different names, e.g., pivot, crosstab, unstack, spread, or widen [@Mount+Zumel:2019:fluid-data]. 

The data.table package uses `dcast()` for this operation. The key columns `program, race,` and `sex` remain in place, the `bloc` column yields the new key columns `ever` and `grad,` and the values in the new columns are taken from the `N` column. The `fill` argument replaces missing values with zero. 

```{r}
DT <- dcast(DT,
  program + sex + race ~ bloc,
  value.var = "N",
  fill = 0
)
setkey(DT, NULL)
DT
```

These data are in block-record form with three keys and two value columns. This is the data structure we called out in our project description for calculating the metric.  






 

### *Calculate the metric* 

*Completes the analysis.*

Before calculating the metric, we have to omit rows in which `ever` (the denominator of the metric) is zero. In this example, no rows are removed. 
  
```{r}
DT <- DT[ever != 0]
setorderv(DT, c("program", "sex", "race"))
DT
```


Stickiness is the ratio of the number of graduates to the number ever enrolled, expressed as a percentage. Stickiness is calculated for each combination of program, race/ethnicity, and sex. 

```{r}
DT[, stick := round(100 * grad / ever, 1)]
setkey(DT, NULL)
DT
```

These data are in block-record form with three key columns (the grouping variables), one value column (stick), and thus one row per value. 




## Dissemination

We take several additional steps before disseminating these results.  

First, we remove rows with summary values that are small enough that student anonymity can no longer be assured. Here, for example, we have 13 rows with three or fewer graduates. 

```{r}
#| echo: false
options(datatable.print.nrows = 15)
```

```{r}
head(DT[order(grad, ever)], 15L)
```

When dealing with the full MIDFIELD research data, we typically use $\small N > 10$, but for these practice data we illustrate the procedure using  $\small N > 3$ in the graduate column. 

```{r}
DT <- DT[grad > 3]
DT
```

We have found it useful to report such data with a variable that combines race/ethnicity and sex. 

```{r}
DT[, people := paste(race, sex)]
setcolorder(DT)
DT
```

Readers can more readily interpret our charts and tables if the programs are unabbreviated. 

```{r}
DT[, program := fcase(
  program %like% "CE", "Civil",
  program %like% "EE", "Electrical",
  program %like% "ME", "Mechanical",
  program %like% "ISE", "Industrial/Systems"
)]
DT
```

### *Table*

```{r}
#| echo: false
options(
  datatable.print.nrows = 14,
  datatable.print.topn = 6
)
```

Retain the columns that appear in the table. The result is in block-record with two key column and one value column. 

```{r}
DT_table <- DT[, .(people, program, stick)]
setorderv(DT_table, c("people", "program"))
DT_table
```

Transform the data to row-record form. The key column `{people}` remains in place, the `{program}` column yields the new key columns `{Civil, Electrical, etc.}`, and the values in the new columns are taken from the `{stick}` column. The `fill` argument replaces missing values with NA.  

```{r}
DT_table <- dcast(DT_table,
  people ~ program,
  value.var = "stick",
  fill = NA
)
setnames(DT_table, old = "people", new = "People")
setkey(DT_table, NULL)
DT_table
```

Format the table for publication. The `sub_missing()` argument replaces NAs with an em-dash to improve readability. 

```{r}
DT_table |>
  gt() |>
  sub_missing() |>
  tab_caption("Table 1. Engineering program stickiness (%)") |>
  tab_options(table.font.size = "small") |>
  opt_stylize(style = 1, color = "gray") |>
  tab_style(
    style = list(cell_fill(color = "#c7eae5")),
    locations = cells_column_labels(columns = everything())
  )
```

```{r}
#| echo: false
options(datatable.print.nrows = 5)
```


### *Chart*

To use `ggplot()`, we want the data in its original block-record form with one value column (stickiness). We can drop two columns as well.

```{r}
DT_chart <- copy(DT)
DT_chart[, c("race", "sex") := NULL]
```

We (optionally) rearrange the order of columns and rows.

```{r}
setcolorder(DT_chart, c("people", "program"))
setkeyv(DT_chart, c("people", "program"))
setkey(DT_chart, NULL)
DT_chart
```

With one quantitative variable (stickiness) for every combination of the levels of two categorical variables (program and people), these are *multiway data* [@Cleveland:1993]. How one orders the categorical variables is critical for visualizing effects. 

`order_multiway()` converts the two categorical variables to ordered factors to support the ordering of rows and panels in the chart. Ordering is based on a calculation of aggregate stickiness values reported in two columns added to the data frame, one column per category. 

```{r}
DT_chart <- order_multiway(DT_chart,
  quantity = "stick",
  categories = c("people", "program"),
  method = "percent",
  ratio_of = c("grad", "ever")
)
DT_chart[, c("ever", "grad") := NULL]
DT_chart
```

Format the chart for publication. No arguments for ordering the data are required because the two categorical variables are ordered factors. 

```{r}
#| label: fig01
#| fig-asp: 1.2
#| fig-cap: "Figure 1: Program stickiness."
ggplot(DT_chart, aes(x = stick, y = people)) +
  facet_wrap(vars(program),
    ncol = 1,
    as.table = FALSE
  ) +
  geom_vline(aes(xintercept = program_metric),
    linetype = 2,
    color = "gray60"
  ) +
  geom_point(size = 1.8) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```

The vertical dashed line in each panel represents the overall stickiness of the program, calculated without regard to race/ethnicity and sex. The panels are ordered by those values, increasing from bottom to top. The stickiness of each "people" group, calculated without regard to program, determines the row order,  also increasing from bottom to top. 

In the next chart, we swap the roles of panels and rows. 

```{r}
#| label: fig02
#| fig-asp: 0.7
#| fig-cap: "Figure 2: Program stickiness."
ggplot(DT_chart, aes(x = stick, y = program)) +
  facet_wrap(vars(people),
    ncol = 2,
    as.table = FALSE
  ) +
  geom_vline(aes(xintercept = people_metric),
    linetype = 2,
    color = "gray60"
  ) +
  geom_point(size = 1.8) +
  labs(x = "Stickiness (%)", y = "") +
  theme_light(base_size = 10)
```

In this version, the vertical dashed line in each panel represents the overall stickiness of the people group, calculated without regard to program. Panels are ordered by increasing group stickiness from left to right and from bottom to top. 



```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
