---
title: "Alternative CIPs"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---

```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-041-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

```{r}
#| echo: false
options(datatable.print.topn = 10)
```

In this article, we demonstrate constructing a Classification of Instructional Programs (CIP) dataset using primary source data from the US Integrated Postsecondary Education Data System (IPEDS). The result is a dataset (`cip2010`) that can be used as an alternative to midfieldr's original CIP data (`cip`). The latest version of midfieldr includes both datasets. 

This article in the MIDFIELD workflow:

1. Planning  
1. Initial processing
- Data sufficiency  
- Degree seeking  
- [Identify programs]{.accent}
1. Blocs  
1. Groupings  
1. Metrics  
1. Displays 



##  Overview

The original midfieldr dataset `cip` and the new dataset `cip2010` derived here are both based on the IPEDS 2010 CIP data. However, the content of the two datasets is not identical. Aside from some typographical differences, the main difference is that `cip2010` has 267 more rows than `cip`, about a 15% increase in program codes (225 of these 267 rows are medical or dental *residency programs*).  Essentially, `cip` is a subset of `cip2010`. 

This is not to say that `cip` is deficient---it was designed for compatibility with the MIDFIELD database and thus the midfielddata practice data as well. However, `cip2010` now makes the full set of 2010 CIP codes available should they be needed. 

As of this writing (2026--05--25), the IPEDS website [@cipcode:2026] provides downloadable CSV files for the 2010 and 2020 CIP data via their "Resources" tab under the file names: 

- `CIPCode2010.csv`
- `CIPCode2020.csv`

In this article, we document a procedure for converting `CIPCode2010.csv` data into the midfieldr-friendly `cip2010` dataset. In future, should a researcher require the 2020 CIP codes in a midfieldr-friendly format, they can start by applying the  procedure documented here to `CIPCode2020.csv`. 



## Get started

If you are writing your own script to follow along, we use these packages in this article:

```{r}
# Load packages
library(midfieldr)
library(data.table)

# From stringr package we use
# str_to_title(), str_replace(), and str_length()
library(stringr)
```

```{r}
#| echo: false
# Use package rprojrot to help with relative file paths
path_to_raw_data <- find_root_file("data-raw", 
                                   criterion = has_file("DESCRIPTION"))
path_to_csv_file <- paste0(path_to_raw_data, "/CIPCode2010.csv")
```

The first step is to download the CSV file from the IPEDs website. After saving it to a convenient directory, label a path to that file (`path_to_csv_file`) and import the data, using `DT` as our working data table. 

```{r}
# Import the 2010 CIP file in data.table format
DT <- fread(path_to_csv_file, colClasses = "character")

# View the result
str(DT, strict.width = "cut", give.attr = FALSE, vec.len = 12)
```

To construct our own CIP data set, we need three columns: 

```{r}
# Select columns
cols_we_want <- c("CIPFamily", "CIPCode", "CIPTitle")
DT <- DT[, ..cols_we_want]

# View the result
str(DT, strict.width = "cut", give.attr = FALSE, vec.len = 12)
```

Closer look at a sample of our primary-source data. 

```{r}
# Set the seed for reproducibility 
set.seed(202612)

# Extract a sample of 15 rows and reset the seed
x <- DT[sample(100:400, 15)]
set.seed(NULL)

# Order the rows by the CIP code
setorderv(x, "CIPCode")

# View the result
x
```

Viewing the result, we observe:

- `CIPFamily` corresponds to `cip2` in midfieldr.
- `CIPCode` contains *all program codes* in a single column. In midfieldr, the 2-, 4-, and 6-digit codes are separated into columns `cip2`, `cip4`, and `cip6`. 
- `CIPTitle` contains *all program names* in a single column. In midfieldr, the the 2-, 4-, and 6-digit names are separated into columns `cip2name`, `cip4name`, and `cip6name`.

When constructing the midfieldr `cip` data set, some characters were changed from those in the source file, such as removing the decimal point separator in the 4- and 6-digit codes or replacing slashes ("/") with commas (","). For example, compare the `13.1209` entry above to its equivalent in the midfieldr `cip` data:

```{r}
# CIPCode2010
filter_cip("13.1209", cip = DT)

# midfieldr default
filter_cip("131209", cip = cip)[, .(cip2, cip6, cip6name)]
```

Additionally, source-file 2-digit program names are in all-caps; in midfieldr all program names are in title-case.

```{r}
# CIPCode2010
DT[CIPCode == "04", .(CIPCode, CIPTitle)]

# midfieldr default
cip[cip2 == "04", .(cip2, cip2name)][1]
```






## Transforming the source data

```{r}
#| echo: false
options(datatable.print.topn = 5)
```

*Editing values.* &nbsp; Omit the periods in the codes and titles. 

```{r}
# Omit the period separator on the CIP code
DT <- DT[, CIPCode := gsub("[.]", "", CIPCode)]

# Omit the period at the end of each title
DT <- DT[, CIPTitle := gsub("[.]", "", CIPTitle)]

# Verify the result
filter_cip("131209", cip = DT)
```

*Split the data frame.* &nbsp;  Count the number of digits in `CIPCode`, expected to be 2, 4, or 6. 

```{r}
# Add a column
DT[, N_digits := str_length(CIPCode)]

# View result
str(DT, strict.width = "cut", give.attr = FALSE, vec.len = 12)

# Confirm 2, 4, or 6 digits
unique(DT$N_digits)
```

Create a separate data frame for 2-digit codes and names, keyed by `CIPFamily`.

```{r}
# Unique 2-digit codes and names
DT2 <- DT[N_digits == 2, .(CIPFamily, 
                           cip2 = CIPCode, 
                           cip2name = CIPTitle)]
# View the result
str(DT2, strict.width = "cut", give.attr = FALSE, vec.len = 12)
```

Repeat for the 4-digit codes and names. Extract the first two digits of the 4-digit code to compare to `CIPFamily` (also 2-digits)

```{r}
# Unique 4-digit codes and names, with cip2 codes derived from cip4
DT4 <- DT[N_digits == 4, .(CIPFamily, 
                           cip2 = substr(CIPCode, 1, 2), 
                           cip4 = CIPCode, 
                           cip4name = CIPTitle)]
# View the result
str(DT4, strict.width = "cut", give.attr = FALSE)
```

Repeat for the 6-digit codes and names, extracting both 2- and 4-digit codes for comparison and joining later. 

```{r}
# Unique 6-digit codes and names, with cip2 and cip4 codes derived from cip6
DT6 <- DT[N_digits == 6, .(CIPFamily, 
                           cip2 = substr(CIPCode, 1, 2), 
                           cip4 = substr(CIPCode, 1, 4), 
                           cip6 = CIPCode, 
                           cip6name = CIPTitle)]
# View the result
str(DT6, strict.width = "cut", give.attr = FALSE, vec.len = 12)
```

Compare the 2-digit codes in each data frame. If the data were correctly entered, this check for equality should all be TRUE. 

```{r}
# This check for 2-digit equality should all be TRUE
all.equal(DT2$CIPFamily, DT2$cip2)
all.equal(DT4$CIPFamily, DT4$cip2)
all.equal(DT6$CIPFamily, DT6$cip2)
```


*Editing.* &nbsp;  Convert the 2-digit names from all-caps to title-case. Here we use two functions from the stringr package. 

```{r}
# Convert to title case
DT2[, cip2name := str_to_title(cip2name)]

# Use lowercase "and"
DT2[, cip2name := str_replace_all(cip2name, "And", "and")]

# View result
DT2[, .(cip2, cip2name)]
```

*Combine.* &nbsp; Next, select the columns we want for the joining operations.  

```{r}
# Select 6-digit columns
cols_we_want <- c("cip2", "cip4", "cip6", "cip6name")
DT6 <- DT6[, ..cols_we_want]

# Select 4-digit columns
cols_we_want <- c("cip2", "cip4", "cip4name")
DT4 <- DT4[, ..cols_we_want]

# Select 2-digit columns
cols_we_want <- c("cip2", "cip2name")
DT2 <- DT2[, ..cols_we_want]
```

Now, left-join `cip4` to `cip6`, matching on the 2- and 4-digit code columns. Reuse the object name `DT` for our new working data frame. 

```{r}
# Join `cip4` to `cip6` , overwrite earlier DT
DT <- DT4[DT6, on = c("cip2", "cip4")]
```

Now, left-join `cip2` to the result, matching on the 2-digit code columns. 

```{r}
# Left join `cip2` to `DT` from above
DT <- DT2[DT, on = c("cip2")]

# Ensure row order
setorderv(DT, c("cip6"))

# View the result
DT
```

*Reality check.* &nbsp; If we've joined correctly, the following logical checks should yield all TRUE (no FALSE) results.

```{r}
# The first 4 digits of every cip6 should match cip4
unique(substr(DT$cip6, 1, 4) == DT$cip4)

# The first 2 digits of every cip6 should match cip2
unique(substr(DT$cip6, 1, 2) == DT$cip2)

# The first 2 digits of every cip4 should match cip2
unique(substr(DT$cip4, 1, 2) == DT$cip2)
```


## Undecided or unspecified programs

Lastly, we add a row to the CIP data to includes one non-IPEDS code (999999) for Undecided or Unspecified, instances in which institutions reported no program information or that students were not enrolled in a program.


```{r}
# Create a data table with one row
txt_99 <- "Undecided/Unspecified (non-IPEDS)"
row_99 <- data.table(cip2 = "99", 
                     cip4 = "9999", 
                     cip6 = "999999", 
                     cip2name = txt_99, 
                     cip4name = txt_99, 
                     cip6name = txt_99)

# Bind the new row to the working data table
DT <- rbindlist(list(DT, row_99), use.names = TRUE)

# View the result
DT
```


*Verify prepared data.* &nbsp; `cip2010`, included with midfieldr, contains the CIP data frame developed above. Here we verify that the two data frames have the same content.

```{r}
# Demonstrate equivalence
check_equiv_frames(DT, cip2010)
```




## Usage

To use the alternative CIP data set included with midfieldr, use `cip2010` as the value of the `cip` argument. For example, 

```{r}
# Using the midfieldr default
filter_cip("^1408", cip = cip)


# Using the alternate cip
filter_cip("^1408", cip = cip2010)
```

Here we see an instance of a minor program title difference between the two datasets. The other program names in this major are identical. 

- 140802 `r filter_cip("140802")[, (cip6name)]` (`cip`)
- 140802  `r filter_cip("140802", cip = cip2010)[, (cip6name)]` (`cip2010`)

```{r}
#| echo: false
#| eval: false

# Execute manually to store data to package

# cip2010 <- copy(DT)
# 
# usethis::use_data(
#   cip2010,
#   internal  = FALSE,
#   overwrite = TRUE
# )
```



















***

[&#9665; Programs](art-040-programs.html) &nbsp;&nbsp; <a href="#top">&#9650; top of page</a> &nbsp;&nbsp; [Blocs &#9655;](art-050-blocs.html)

***






```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
