---
title: "Planning"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---
```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-011-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```

```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

In any study, your data manipulation procedure depends on your answers to following prompts. Because data exploration is iterative, your answers may change over time. Nevertheless, your work will proceed far more smoothly if you answer these at the beginning of a project and keep them in mind as you proceed. 

Data

: What data sources is your work based on? How is the data organized? What variables are available? 

Programs 

: You are limiting your study to which programs?

Metric  

: How do you define your quantitative metric, expressed in words and mathematically. 

Records

: What exclusions apply to the source data?

Population

: How do you define your population? What exclusions apply?

Blocs

: The metric requires what blocs of records? 

Groupings

: What groupings will you apply when you group and summarize?

Results

: To group and summarize, what variables do you need? After grouping, summarizing, and computing the metric, what variables do you expect to have? 


```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```
