---
title: "Introduction to midfieldr"
format:
  gfm:
    html-table-processing: none
bibliography: ../../inst/REFERENCES.bib
link-citations: true
---
  
```{r setup}
#| include: false

# for any figures generated in the article
knitr::opts_chunk$set(fig.path = "figures/art-000-")

# for any "child" arguments in r-code-chunks
library("rprojroot")
path_to_rmd <- find_root_file("man", "rmd", criterion = has_file("DESCRIPTION"))
```
  
```{r child = paste0(path_to_rmd, "/common-setup.Rmd")}
```

When working with student-level records to develop quantitative metrics, midfieldr helps you refine and shape your data in these areas:

- *Programs.* &nbsp; Collect 6-digit program codes. 
- *Records.* &nbsp; Credibly subset source data and refine the population. 
- *Blocs.* &nbsp; Construct group of records for computing metrics.  
  
This document introduces you to midfieldr's basic set of tools.

```{r}
library("midfieldr")
library("midfielddata")
library("data.table")
```

Before you start:

1. *On syntax:* &nbsp; We use data.table syntax for data manipulation throughout midfieldr and all data frames are of the `data.table` class. However, if you happen to prefer tidyverse syntax, midfieldr functions do attempt to preserve data frame attributes such as the `tbl_df` class. 

2. *On functions:* &nbsp; In getting started, we provide a brief introduction only. Details are discussed at length in subsequent articles. You can always access the  documentation, e.g., `?function_name`, for more information.  



## midfieldr functions

The major functions for treating student records can be categorized based on their contribution to a typical workflow:

Programs  

- `filter_programs()`  &nbsp; helps you find program names and CIP codes.  

Records and population  

- `timely_term()`  &nbsp; estimates timely completion terms.  
- `data_sufficiency()`  &nbsp; identifies IDs to exclude due to insufficient data.  
- `undergrad_terms()`  &nbsp; identifies rows with post-baccalaureate terms to exclude.  

Blocs  

- `completion_status()`  &nbsp; labels program completion as timely, late, or NA.  

Special data conditioning

- `prep_fye_mice()` &nbsp; for imputing the starting majors of First-Year Engineering (FYE) students.  

- `order_multiway()` &nbsp; for ordering Cleveland multiway data. 








## Program data


*Collecting and labeling 6-digit program codes.*

```{r}
#| echo: false
options(
  datatable.print.nrows = 10,
  datatable.print.topn = 5
)
```


The *Classification of Instructional Programs (CIP)* is a taxonomy of academic programs, encoded by 6-digit numeric codes curated by the US Department of Education [@NCES:2010]. 

The `cip` data set, loaded with midfieldr, is a subset of the NCES CIP2010 data that contains codes and names for `r nrow(cip)` instructional programs organized on three levels---a 6-digit series, a 4-digit series, and a 2-digit series---keyed by the `cip6` variable. 

```{r}
cip
```





## `filter_programs()`


*Chooses rows of CIP data based on search terms.*

`filter_programs()` acts on the data frame assigned to its `dframe` argument to select rows that match or partially match search strings. Search strings are case-independent and can include regular expressions. 

```{r}
filter_programs(cip, "music")
```

To refine our results, we can assign the results of a first pass to the `dframe` argument of a second pass. For example, our first pass below searches the default `cip` dataset for "music". Our second pass searches the results of the  first pass for any line that starts with "50". We can also drop the 2-digit level codes and names to reduce the visual clutter. Because these programs have the same 2-digit code and name, we can drop two columns to reduce the visual clutter. 

```{r}
first_pass <- filter_programs(cip, "music")
second_pass <- filter_programs(first_pass, "^50")
second_pass[, c("cip2", "cip2name") := NULL]
second_pass
```

```{r}
#| echo: false
options(
  datatable.print.topn = 10,
  datatable.print.nrows = 20
)
```

Assuming we are looking for programs in a School of Music, the 4-digit code "5009" appears to be the correct finding. Our third pass searches the results of the second pass for any line that starts with "5009". Again (in this case) we can reduce the visual clutter and remove two more columns. 

```{r} 
third_pass <- filter_programs(second_pass, "^5009")
third_pass[, c("cip4", "cip4name") := NULL]
third_pass
```

Assuming these were our study programs, we save the 6-digit codes and names and optionally abbreviate some of the longer names. More importantly, we add a `program` column with our own program abbreviations (here I use the placeholder `label_TBD`). These custom program labels aggregate the 6-digit codes into groups that are relevant to the study goals. 

```{r}
programs <- third_pass[, .(cip6name, cip6, program = "to be determined")]
programs[, cip6name := gsub("Violin, Viola, Guitar", "Vn, Va, Gtr", cip6name)]
programs[, cip6name := gsub("Instruments", "Instr", cip6name)]
programs
```

For example, here I label all programs I might consider part of an "orchestra" grouping. 
```{r}
orch_instr <- c("brass|wood|perc|conduct|vn")
programs[cip6name %ilike% orch_instr, program := "Orchestra"]
programs
```

The structure of this data frame is representative of the `programs` data frame you would find in any study: 6-digit CIP codes, possibly the NCES 6-digit program name, and our own program labels.  







## Student-level data

*Credibly subset the source data and refine the population.*

```{r}
#| echo: false
options(
  datatable.print.topn = 3,
  datatable.print.nrows = 6
)
```

For this article we load the `student,` `term,` and `degree` tables from midfielddata. 

```{r}
data(student, term, degree)
```

The data tables are linked by `mcid`, the anonymized student ID. `look_at()` is a midfieldr convenience function that wraps base  `str()` with preset arguments. 

```{r}
look_at(student)

look_at(term)

look_at(degree)
```


We copy our "source" material under separate names (and locations in memory).   

```{r}
student_source <- copy(student)
term_source <- copy(term)
degree_source <- copy(degree)

# demonstrate that memory addresses are different
address(student)
address(student_source)
```

Then we can use the shorter names such as `term` and `degree` as we work. The shorter names are also the default values for arguments in several midfieldr functions. 

### *Terminology*

The following sequence of definitions provides the context for the next few functions. 

- Program *completion* means satisfying the requirements for a baccalaureate degree. 
- Program completion is *timely* if accomplished within a set time span, typically 4, 6, or 8 years after admission depending on the definition one adopts. (The midfieldr default is 6 academic years.) The *timely-completion term* is the term at the end of that span. 
- A student's *completion status* is "timely" if they graduate no later than their timely-completion term, "late" if they graduate after their timely-completion term, and NA for non-completers.
- To avoid biased results, completion status can be assessed only for students for whom the data tables include a sufficient number of terms before and after their admission term. The test for *data sufficiency* identifies such students. Only those records passing the data sufficiency test are included in a population study. 

In midfieldr:

- `timely_term()` yields the timely-completion term for every student. 
- `data_sufficiency()` tests for data sufficiency for every student .
- `undergrad_terms()` identifies post-baccalaureate terms to exclude.
- `completion_status()` yields the completion status for every student passing the data sufficiency test. 










## `timely_term()`

*Determine the term by which degree completion would be considered timely.*

We start with a unique set of IDs from the `term` table. 

```{r}
DT <- term[, .(mcid)]
DT <- unique(DT)

DT
```

`timely_term()` builds a data frame with one row per student, a column for the timely completion term, and columns of supporting information. This data frame contains the required inputs for both `data_sufficiency()` and `timely_completion()`. 

```{r}
DT <- timely_term(DT, midf_table = term)

DT
```

## `data_sufficiency()`

*Identify members of the population to exclude due to insufficient data.*








`data_sufficiency()` builds on the output from `timely_term()`, labels rows to be included or excluded based on the data sufficiency finding, and generates additional columns of supporting information. 

```{r}
DT <- data_sufficiency(DT, midf_table = term)

DT
```

The possible values for data sufficiency are:

```{r}
DT[, sort(unique(data_sufficiency), na.last = FALSE)]
```

We filter to retain rows labeled "include". The resulting IDs define our baseline population. 

```{r}
population <- DT["include", on = "data_sufficiency", .(mcid)]
population <- unique(population)

population
```

We use this population to filter our source material one last time. We use an inner join to return records for this population only.

```{r}
student_source <- population[student_source, on = "mcid", nomatch = NULL]
term_source <- population[term_source, on = "mcid", nomatch = NULL]
degree_source <- population[degree_source, on = "mcid", nomatch = NULL]

look_at(term_source)
```

In subsequent analysis, any variables we need from the source data has already been filtered to satisfy the data sufficiency constraint and to exclude post-baccalaureate terms. 

From this point forward, anytime we need a fresh copy of any of the data tables, we copy the "source" version. Anytime we need a starting population, we use `population` or unique IDs from `student_source` or `term_source.`

```{r}
student <- copy(student_source)
term <- copy(term_source)
degree <- copy(degree_source)

all.equal(population$mcid, student_source$mcid)
all.equal(population$mcid, unique(term_source$mcid))
```


## `record_bracket()`

*Identify rows of post-baccalaureate terms to exclude.*

In most cases, we are not generally interested in academic terms beyond the first degree term, so we use the results of this function to exclude post-first-degree terms from the source data. 

`record_bracket()` identifies terms later than the first baccalaureate, if any.  

```{r}
term <- record_bracket(term_source, midf_table = degree)
degree <- record_bracket(degree_source, midf_table = degree)
```

`record_bracket()` adds a column indicating the bracket a term belongs to with respect to the first degree term.

```{r}
look_at(term)
```

The possible bracket values are given by, 

```{r}
term[, sort(unique(bracket), na.last = FALSE)]
```

We filter to exclude all terms labeled "post-bacc" and drop the temporary columns. 

```{r}
term <- term["undergrad", on = "bracket"]
degree <- degree["undergrad", on = "bracket"]

term[, c("term_1st_degree", "bracket") := NULL]
degree[, c("term_1st_degree", "bracket") := NULL]

look_at(term)
```

We redefine our source material to incorporate the exclusion of post-baccalaureate terms. 

```{r}
term_source <- copy(term)
degree_source <- copy(degree)
```










## `select_basic_cols()` 

*Choose columns required by midfieldr functions.* 

`select_basic_cols()` operates on student records to reduce the number of columns to those required by other midfieldr functions plus the key or composite key variables of the four data tables. With a smaller number of columns, the printout of the data frame is more readable, a benefit when working with the data interactively. 

```{r}
student <- select_basic_cols(student)
term <- select_basic_cols(term)
degree <- select_basic_cols(degree)

student

term

degree
```

Any variables you might need that have been dropped can always be recovered from the source tables we saved earlier. For example, if we needed GPA in our working `term` table, we can use a left join, knowing that student ID and term are the composite keys in this case.

```{r}
x <- copy(term)
source_cols <- term_source[, .(mcid, term, gpa_term, gpa_cumul)]
x <- source_cols[x, on = c("mcid", "term")]
x
```

Keys and composite keys to the four data tables are described in the midfielddata [Data structure](https://midfieldr.github.io/midfielddata/articles/data-structure.html) article. 




## `completion_status()`

*Determines if program completion is timely or late.*

This section often pertains to constructing a bloc of graduates, starting with the baseline population we obtained above. 

```{r}
DT <- copy(population)
```

`completion_status()` builds on the output from `timely_term()`, labels rows to indicate whether a student completes a degree timely or late compared to their timely completion term (or NA for no completion), and includes columns for the timely term and degree term as supporting information. 

```{r} 
DT <- timely_term(DT, midf_table = term)
DT <- completion_status(DT, midf_table = degree)

DT
```

The possible values for completion status are:

```{r}
DT[, sort(unique(completion_status), na.last = FALSE)]
```

If we were constructing a bloc of timely graduates, we would filter to retain rows labeled "timely". The resulting IDs would define our graduates bloc.

```{r}
graduates <- DT["timely", on = "completion_status", .(mcid)]
graduates[, bloc := "grad"]
graduates
```






## Other functions

`prep_fye_mice()` &nbsp; Conditions data for imputing the starting majors of First-Year Engineering (FYE) students. Used when blocs of starters in Engineering are needed and an  institution has a required FYE program. For details see [FYE proxies](art-060-fye-proxies.html).   

`order_multiway()` &nbsp; Conditions data for Cleveland multiway charts. The ordering of its rows and panels is crucial to the perception of effects. Used when data have a multiway structure. For details see [Multiway data and charts](art-120-multiway.html). 

Utilities 

- `look_at()` for data frames, wraps base `str()` with preset arguments.
- `sort_uniq()` for vectors, wraps base `sort(unique())` with preset arguments.
- `catch_error()` wraps base `tryCatch()` for errors with preset arguments. 
- `check_equiv_frames()` re-exported from the wrapr package  



```{r child = paste0(path_to_rmd, "/common-closing.Rmd")}
```



