---
title: "Try a QMD script"
format: html
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig-dpi: 300
    fig-asp: 0.618
    fig-width: 6
    out-width: "80%"
---

```{=html}
<style>
.content h3 {
margin-top: -30px !important;
}

details {
margin-bottom: 40px;
}
</style>
```
```{r}
#| include: false
library(midfieldr)
library(midfielddata)
library(data.table)
library(dplyr)
library(tibble)
```

## Hello

```{r}
data(student, term, degree)
```


::: {.panel-tabset}

### data.table

```{r}
student_DT <- copy(student)
term_DT <- copy(term)
degree_DT <- copy(degree)
```

### dplyr

```{r}
student_TB <- as_tibble(student)
term_TB <- as_tibble(term)
degree_TB <- as_tibble(degree)
```

:::

```{r}
class(student)
class(student_DT)
class(student_TB)

check_equiv_frames(student_DT, student_TB)
```

# Population

The baseline population used for most studies consists of degree-seeking students whose records satisfy the data sufficiency requirement.

## Degree-seeking

By design, the `student` data table contains all (and only) degree-seeking students. Thus we start with those IDs. 

::: {.panel-tabset}

### data.table

```{r}
DT <- student_DT[, .(mcid)]
```

### dplyr

```{r}
TB <- student_DT %>%
    select(mcid)
```

:::

```{r}
DT

check_equiv_frames(DT, TB)
```

## Data sufficiency

We use `timely_term()` to determine the timely completion term for the degree-seeking students and add columns to the data frame to support those findings. 

::: {.panel-tabset}

### data.table

```{r}
DT <- timely_term(DT, midf_table = term_DT)
DT
```

### dplyr

```{r}
TB <- timely_term(TB, midf_table = term_TB)
TB
```

:::



```{r}
DT[order(entry_level)]
```
